# Module 2 — Test Bank {.unnumbered}

```{r}
#| include: false
#| eval: true
# console() renders a code listing plus the console session it produces.
source("../_console.R")
```

::: {.callout-note}
## How this test bank works

This bank holds **120 questions**, thirty in each section. Section 1 gives you the data inside the question; Sections 2--4 each hold three blocks of ten, one block per dataset (the datasets are described below). On the real test you receive **four questions -- one from each section** -- and you submit **a single, well-commented R script**: a header block at the top, a clearly labelled section for each question, and your sentence answers written as comments. The script should run from top to bottom in a project folder that has the csvs in `data/`. Each question is designed to take about **8--10 minutes**.

Expand each **Answer** block to check your work. Every value has been computed from the real data; if the source file is refreshed the exact numbers may shift slightly, but the method stays the same.
:::

**The data.** Three datasets, six files, all in your project's `data/` folder:

*Saskatchewan RM crop yields* -- average yields in every Saskatchewan Rural Municipality, one row per RM-year(-crop), columns `Year`, `RM`, `Crop`, `Yield`, `Unit`:

- [`rm_canola_yields_1990_2025.csv`](data/rm_canola_yields_1990_2025.csv) -- the **canola rows only**, for Section 2: one crop, one unit, so its summary statistics mean something.
- [`rm_yields_1990_2025.csv`](data/rm_yields_1990_2025.csv) -- all **eight crops**, for Sections 3--4. You worked with the wide version in Module 1; this is the long form, with `Crop` as a variable. **Lentils are recorded in pounds per acre; every other crop is in bushels per acre** -- the `Unit` column says which, and several questions turn on it.

*Manitoba wheat variety yields* -- MASC's variety-level spring wheat yields, one row per municipality-variety-year, 2020--2025, columns `Year`, `Municipality`, `Variety`, `Farms`, `Acres`, `Yield_bu_ac`, `Reported`. Where too few farms grew a variety, the row is **suppressed** for privacy: `Reported` is FALSE and the numeric columns are missing.

- [`mb_wheat_reported_2020_2025.csv`](data/mb_wheat_reported_2020_2025.csv) -- the **reported rows only**, for Section 2.
- [`mb_wheat_varieties.csv`](data/mb_wheat_varieties.csv) -- the **full file**, suppressed rows included, for Sections 3--4.

*Canada field crops* -- Statistics Canada's provincial estimates, one row per province-crop-year, 2015--2025, columns `Year`, `Province`, `Crop`, `Seeded_acres`, `Yield_bu_ac`:

- [`statcan_spring_wheat_2015_2025.csv`](data/statcan_spring_wheat_2015_2025.csv) -- the **spring wheat rows only**, for Section 2.
- [`statcan_field_crops.csv`](data/statcan_field_crops.csv) -- all **thirteen crops**, for Sections 3--4.

**Grading model.** **20%** for script presentation -- a header block, comments in the style of the module, clearly labelled question sections, and a script that runs from top to bottom in a fresh session. The remaining **80%** is split evenly across the four questions, and evenly across the parts within each question. Parts are marked on **follow-through**: if an early part is wrong but the later parts are correctly worked *from your own earlier answer*, the later parts still earn full marks. Where a part asks for a sentence, write it as a comment in your script.

> **Vocabulary note.** Everything here uses only Module 2 tools: `<-`, arithmetic on vectors, `c()`, `data.frame()`, `$`, `mean`, `median`, `sd`, `var`, `min`, `max`, `range`, `sum`, `quantile` (with `probs`), `IQR`, `nrow`, `ncol`, `names`, `head`, `print`, `glimpse`, `summary`, `library()`, `install.packages()`, `read_csv` with relative paths, `filter` (with `==`, `>`, `<`, `>=`, `&`, `|`, `!`, `%in%`), `select` (including `-` and `:`), `rename`, `mutate`, `arrange` and `desc`, the pipe `|>`, `summarise`, `n()`, `group_by`, and `write_csv`. Nothing beyond that is needed. The chapter's `summarise()` examples write `na.rm = TRUE` inside `mean()` and `sd()`; none of the data these questions filter to has missing yields, so answers with or without it get the same numbers and the same marks.

::: {.callout-note collapse="true"}
## A sample test, with a full-marks answer script

A real test looks like this -- one question from each section (here Questions 1, 36, 65, and 96):

1. *Five canola fields yielded 52.3, 47.8, 55.1, 44.6, and 50.9 bu/ac, on 160, 320, 240, 130, and 200 acres. Create the two vectors; convert the yields to t/ha; compute the mean and sd of the yields; compute total production and the acreage-weighted average yield.*
2. *Compute the mean, median, and standard deviation of canola `Yield`, and its 90th percentile. In a comment: the mean sits a little above the median -- what does that direction of gap suggest about the shape of canola yields?*
3. *Filter to the lentil rows, saving as `lentils` (how many rows?); add a `yield_kg_ha` column (multiply by 1.12), saving back to `lentils`; compute the mean of the new column and of the original, and check their ratio.*
4. *In one pipeline, compute each RM's mean Spring Wheat yield across all years, sorted best-first. Report the top three RMs. In a comment: what did each row of the input represent, and what does each row of the result represent?*

And here is a script that would earn full marks, including all of the presentation component -- with the console session it produces shown beneath it:

```{r}
#| eval: true
#| echo: false
console('
# ---
# Title: Module 2 test
# Author: Jordan Field
# Date: 2026-10-15
# Description:
#   Answers to the four test questions, one section per block.
#   Sentence answers are written as comments below each result.
# ---

# Load packages
library(tidyverse)

# ============================================================
# Section 1 (Question 1): five canola fields
# ============================================================

# Yields (bu/ac) and field sizes (acres)
yields <- c(52.3, 47.8, 55.1, 44.6, 50.9)
acres  <- c(160, 320, 240, 130, 200)

# (a/b) Convert every yield to tonnes per hectare in one step
yields_t_ha <- yields * 0.0560
yields_t_ha

# (c) Centre and spread of the yields
mean(yields)   # 50.14 bu/ac
sd(yields)     # 4.06 bu/ac

# (d) Total production, then the average weighted by acres
total_bu <- sum(yields * acres)
total_bu                 # 52,866 bu
total_bu / sum(acres)    # 50.35 bu/ac
# The weighted average differs from the plain mean because the
# fields differ in size, so each yield should count by its acres.

# ============================================================
# Section 2 (Question 36): summary statistics for canola
# ============================================================

# Read the canola file using a relative path from the project folder
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")

# Centre and spread of the canola yields
mean(rm_canola$Yield)      # 28.29 bu/ac
median(rm_canola$Yield)    # 26.9 bu/ac
sd(rm_canola$Yield)        # 10.06 bu/ac

# 90th percentile
quantile(rm_canola$Yield, 0.9)   # 42.5 bu/ac

# The mean sits a little above the median, which suggests mild
# right skew: the best RM-years pull the mean up more than the
# worst years pull it down.

# ============================================================
# Section 3 (Question 65): converting the lentil yields
# ============================================================

# Read the full eight-crop file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# Keep only the lentil rows
lentils <- filter(rm_yields, Crop == "Lentils")
nrow(lentils)    # 6,338 rows

# Add the converted column (1 lb/ac is about 1.12 kg/ha) and
# overwrite lentils with the version that has the extra column
lentils <- mutate(lentils, yield_kg_ha = Yield * 1.12)

# Means of the new and original columns
mean(lentils$yield_kg_ha)   # 1,353.4 kg/ha
mean(lentils$Yield)         # 1,208.39 lb/ac
# Their ratio is 1.12, the conversion factor -- multiplying every
# value by a constant multiplies the mean by the same constant.

# ============================================================
# Section 4 (Question 96): best wheat-growing RMs
# ============================================================

rm_yields |>                              # take the data, THEN
  filter(Crop == "Spring Wheat") |>       # keep spring wheat, THEN
  group_by(RM) |>                         # split the rows by RM, THEN
  summarise(mean_yield = mean(Yield)) |>  # one row per RM: its mean, THEN
  arrange(desc(mean_yield))               # best RMs on top

# Top three: RM 369 (47.6), RM 333 (47.2), RM 368 (47.2) bu/ac.
# Each input row was one RM-year observation of spring wheat; each
# result row is one RM, its years collapsed into a single mean.
')
```
:::

---

# Section 1 — R Basics

*These questions are about the building blocks: objects, vectors, data frames, and functions. No data file is needed -- the numbers are in the question. Answer in the Section 1 block of your script, with sentence answers written as comments.*

### Question 1

Five canola fields yielded 52.3, 47.8, 55.1, 44.6, and 50.9 bu/ac, on 160, 320, 240, 130, and 200 acres.

**(a)** Create a vector `yields` and a vector `acres` holding these values.

**(b)** Create a vector `yields_t_ha` converting the yields to tonnes per hectare (multiply by 0.0560). No loop, no repetition -- one line.

**(c)** Compute the mean and standard deviation of `yields`.

**(d)** Compute total production in bushels (each field's yield times its acres, summed), and the farm's average yield weighted by acres (total bushels divided by total acres).

<details><summary>Answer</summary>

```r
yields <- c(52.3, 47.8, 55.1, 44.6, 50.9)
acres  <- c(160, 320, 240, 130, 200)
yields_t_ha <- yields * 0.0560
mean(yields)
sd(yields)
total_bu <- sum(yields * acres)
total_bu / sum(acres)
```

- (b) 2.93, 2.68, 3.09, 2.50, 2.85 t/ha.
- (c) Mean **50.14**, sd **4.06**.
- (d) Total **52,866 bu**; weighted average **50.35 bu/ac**.
</details>

### Question 2

**(a)** Build a data frame called `fields` with three columns: `field_id` containing `"F1"` to `"F5"`, `crop` containing Canola, Wheat, Canola, Peas, Wheat, and `yield` containing 44.1, 52.7, 39.8, 41.5, 49.3.

**(b)** Report the number of rows, the number of columns, and the column names using functions (not by counting).

**(c)** Pull out the `yield` column with `$` and compute its mean and median.

**(d)** In a comment: each column of a data frame is itself a familiar R object -- which one?

<details><summary>Answer</summary>

```r
fields <- data.frame(
  field_id = c("F1", "F2", "F3", "F4", "F5"),
  crop = c("Canola", "Wheat", "Canola", "Peas", "Wheat"),
  yield = c(44.1, 52.7, 39.8, 41.5, 49.3)
)
nrow(fields); ncol(fields); names(fields)
mean(fields$yield); median(fields$yield)
```

- (b) 5 rows, 3 columns; `field_id`, `crop`, `yield`.
- (c) Mean **45.48**, median **44.1**.
- (d) A vector.
</details>

### Question 3

Ten pea fields yielded 31, 44, 27, 50, 38, 42, 35, 47, 29, and 40 bu/ac.

**(a)** Save the values as a vector and compute the mean.

**(b)** Compute the 25th and 75th percentiles using `quantile()` with the arguments **in order, unnamed**.

**(c)** Compute the same two percentiles again with **named arguments in the reverse order**, and confirm you get the same answers.

**(d)** In a comment: what happens if you write `quantile(0.25, peas)` -- unnamed arguments in the wrong order -- and why?

<details><summary>Answer</summary>

```r
peas <- c(31, 44, 27, 50, 38, 42, 35, 47, 29, 40)
mean(peas)
quantile(peas, c(0.25, 0.75))
quantile(probs = c(0.25, 0.75), x = peas)
```

- (a) Mean **38.3**.
- (b, c) 25th percentile **32**, 75th percentile **43.5** -- identical both ways, because naming the arguments frees them from their positions.
- (d) R reads 0.25 as the data `x` and the yields as `probs`; since `probs` must lie between 0 and 1, it stops with `Error: 'probs' outside [0,1]`. Unnamed arguments only work in the order the function expects.
</details>

### Question 4

A student types the following four lines into the console, in order:

```r
x <- 8
x * 3
y <- x - 3
x * y
```

**(a)** In a comment: which lines print something to the console, and what exactly do they print (including the `[1]`)?

**(b)** In a comment: after all four lines, what objects does the Variables pane show, with what values? Is the result of `x * 3` among them?

**(c)** A fifth line, `y <- y * 2`, is now run. In a comment: what does the Variables pane show afterwards, and did the line print anything?

**(d)** Predict what a sixth line, `x * y`, would print now. Then type all six lines into the console yourself and check every prediction from (a)--(d) against what actually happens.

<details><summary>Answer</summary>

- (a) Only the second and fourth lines print: `[1] 24` and `[1] 40`. Assignments print nothing.
- (b) `x = 8` and `y = 5`. The result of `x * 3` was displayed but never assigned, so it is not saved anywhere -- displaying a result and saving one are different things.
- (c) `y` is now 10 (the old value was used to compute the new one, then overwritten); `x` is still 8. Nothing printed -- it is an assignment.
- (d) `[1] 80` -- x is 8 and y is now 10.
</details>

### Question 5

A student opens a new file in Positron, types a script that creates three objects, and saves the file -- but the Variables pane still says "No variables have been created."

**(a)** In a comment: why does nothing exist yet, in one sentence?

**(b)** In a comment: give the two ways to run the script from Module 2 (one uses a button, one a keyboard shortcut).

**(c)** In a comment: after running it, where do results appear, and where do the created objects appear?

**(d)** Reproduce the situation yourself: write a script that creates three objects, save it, and confirm the Variables pane is still empty. Then run it and, in a comment, report what the console showed and what the Variables pane now lists.

<details><summary>Answer</summary>

- (a) Writing code in a script does not run it -- typing puts text in a file, and nothing happens until the code is executed.
- (b) The **Run** button (source the whole file, or Execute code for selected lines), or **Cmd+Enter** (Mac) / **Ctrl+Enter** (Windows) to run the selected lines or the line under the cursor.
- (c) Results print in the **console**; created objects appear in the **Variables** pane.
- (d) Three assignments print nothing to the console; the Variables pane lists the three objects with their values.
</details>

### Question 6

Consider these three lines:

```r
n <- 12
n == 10
n = 15
```

**(a)** In a comment: say what each line does. Which ones assign, which one asks a question, and what does the question print?

**(b)** In a comment: all three symbols (`<-`, `==`, `=`) are legal R. Which one does this book use for assignment, and why does keeping the three visually distinct matter?

**(c)** `mean(x = c(4, 9, 11))` also contains an `=`. In a comment: what is `=` doing there?

<details><summary>Answer</summary>

- (a) Line 1 assigns 12 to `n`. Line 2 asks whether `n` equals 10 and prints `[1] FALSE`. Line 3 also assigns -- `n` is now 15.
- (b) `<-` for assignment. It keeps assignment visually distinct from `=` naming arguments inside function calls and from `==` testing equality -- three different jobs that are easy to confuse when they all look alike.
- (c) Naming an argument: it tells `mean()` that the vector is its `x` argument. It does not create an object called `x` in the Variables pane.
</details>

### Question 7

Here is a working script:

```r
a <- c(38.2, 45.6, 41.9, 36.8)
b <- c(210, 180, 260, 240)
d <- sum(a * b)
d / sum(b)
```

**(a)** Rewrite it with a proper header block (Title, Author, Date, Description) and a comment before each step, following the module's conventions.

**(b)** Replace the object names with descriptive ones.

**(c)** In a comment: state the module's test for whether a script is reproducible.

<details><summary>Answer</summary>

```r
# ---
# Title: Farm average yield
# Author: Your Name
# Date: 2026-10-08
# Description:
#   Computes total production and the acreage-weighted
#   average yield for four fields.
# ---

# Field yields (bu/ac) and sizes (acres)
yields <- c(38.2, 45.6, 41.9, 36.8)
acres  <- c(210, 180, 260, 240)

# Total production in bushels
total_bu <- sum(yields * acres)

# Average yield weighted by acres
total_bu / sum(acres)
```

- (c) Restart R and run the script from top to bottom: it should reproduce the analysis without any commands typed into the console. (The weighted average, for reference, is **40.4 bu/ac**.)
</details>

### Question 8

Five barley fields yielded 68.2, 71.5, 64.9, 74.1, and 66.3 bu/ac. Five lentil fields yielded 1450, 1720, 1280, 1610, and 1390 lb/ac.

**(a)** Compute the mean, the standard deviation, the variance, and the range (maximum minus minimum) of each crop's yields.

**(b)** The lentil standard deviation is far larger. In a comment: why can the two standard deviations not be compared directly?

**(c)** Compute the coefficient of variation (sd divided by mean) for each crop. Which crop's yields are more variable *relative to their own average*?

<details><summary>Answer</summary>

```r
barley  <- c(68.2, 71.5, 64.9, 74.1, 66.3)
lentils <- c(1450, 1720, 1280, 1610, 1390)
var(barley)
var(lentils)
max(barley) - min(barley)
max(lentils) - min(lentils)
sd(barley) / mean(barley)
sd(lentils) / mean(lentils)
```

- (a) Barley mean **69.0**, sd **3.77**, variance **14.25**, range **9.2**; lentils mean **1490**, sd **175.36**, variance **30,750**, range **440**.
- (b) They are in different units (bushels vs pounds per acre) and on very different scales, so the raw spreads are not comparable.
- (c) CV barley **0.055**, lentils **0.118** -- lentil yields are more variable relative to their own average.
</details>

### Question 9

**(a)** Using the pea yields from Question 3, compute the 10th and 90th percentiles in one `quantile()` call with **named** arguments.

**(b)** In a comment: `quantile()` has two main arguments. Name them and say what each is.

**(c)** In a comment: you cannot remember what the `probs` argument of `quantile()` expects. What do you type to get the official documentation?

**(d)** A classmate writes `Quantile(peas, 0.5)` and gets `could not find function "Quantile"`. In a comment: what went wrong?

<details><summary>Answer</summary>

```r
peas <- c(31, 44, 27, 50, 38, 42, 35, 47, 29, 40)
quantile(x = peas, probs = c(0.1, 0.9))
```

- (a) 10th percentile **28.8**, 90th percentile **47.3**.
- (b) `x`, the data, and `probs`, the percentile you want expressed in decimal form.
- (c) `?quantile`.
- (d) R is exact about names: `Quantile` with a capital Q is not the same as `quantile`, and no function by that name exists.
</details>

### Question 10

A trial records, for each field: the crop grown, the seeded acres, and whether the field was irrigated.

**(a)** Create three vectors of length four holding made-up values for these three variables, choosing the appropriate type for each (character, numeric, logical).

**(b)** In a comment: which values needed quotation marks and which must not have them?

**(c)** Combine your three vectors into a data frame called `trial` and print it.

<details><summary>Answer</summary>

```r
crop      <- c("Canola", "Wheat", "Peas", "Canola")
acres     <- c(160, 320, 240, 130)
irrigated <- c(TRUE, FALSE, FALSE, TRUE)
trial <- data.frame(crop = crop,
                    acres = acres,
                    irrigated = irrigated)
trial
```

- (b) The character values (crop names) need quotes. Numbers must not have them (quoted numbers become text), and `TRUE`/`FALSE` must not have them -- quoted, they would be text rather than logical values.
</details>

---

### Question 11

Four barley fields yielded 66.4, 72.1, 58.9, and 69.5 bu/ac, on 280, 150, 320, and 210 acres. A bushel of barley is 48 lb, so one bu/ac is about 53.8 kg/ha.

**(a)** Create a vector `yields` and a vector `acres` holding these values.

**(b)** Create a vector `yields_kg_ha` converting the yields to kg/ha in one line.

**(c)** Compute the mean and standard deviation of `yields`.

**(d)** Compute total production in bushels, and the farm's average yield weighted by acres.

<details><summary>Answer</summary>

```r
yields <- c(66.4, 72.1, 58.9, 69.5)
acres  <- c(280, 150, 320, 210)
yields_kg_ha <- yields * 53.8
mean(yields)
sd(yields)
total_bu <- sum(yields * acres)
total_bu / sum(acres)
```

- (b) 3572.32, 3878.98, 3168.82, 3739.10 kg/ha.
- (c) Mean **66.72**, sd **5.71**.
- (d) Total **62,850 bu**; weighted average **65.47 bu/ac**.
</details>

### Question 12

**(a)** Build a data frame called `pea_fields` with three columns: `field_id` containing `"P1"` to `"P5"`, `seeded_may` containing TRUE, TRUE, FALSE, TRUE, FALSE, and `yield` containing 38.4, 45.2, 33.9, 41.7, 36.8.

**(b)** Report its number of rows, number of columns, and column names using functions.

**(c)** Compute the mean and median of the `yield` column using `$`.

**(d)** In a comment: which of the three columns is character, which is logical, and which is numeric?

<details><summary>Answer</summary>

```r
pea_fields <- data.frame(
  field_id = c("P1", "P2", "P3", "P4", "P5"),
  seeded_may = c(TRUE, TRUE, FALSE, TRUE, FALSE),
  yield = c(38.4, 45.2, 33.9, 41.7, 36.8)
)
nrow(pea_fields); ncol(pea_fields); names(pea_fields)
mean(pea_fields$yield); median(pea_fields$yield)
```

- (b) 5 rows, 3 columns; `field_id`, `seeded_may`, `yield`.
- (c) Mean **39.2**, median **38.4**.
- (d) `field_id` is character, `seeded_may` is logical, `yield` is numeric.
</details>

### Question 13

Twelve canola fields yielded 24.1, 31.5, 28.8, 35.2, 22.7, 30.4, 27.9, 33.6, 25.5, 29.3, 32.8, and 26.4 bu/ac.

**(a)** Save the values as a vector and compute the mean.

**(b)** Compute the 10th, 50th, and 90th percentiles in a single `quantile()` call.

**(c)** In a comment: one of your three percentiles should equal the median. Which one, and does it?

<details><summary>Answer</summary>

```r
canola <- c(24.1, 31.5, 28.8, 35.2, 22.7, 30.4, 27.9, 33.6, 25.5, 29.3, 32.8, 26.4)
mean(canola)
quantile(canola, c(0.1, 0.5, 0.9))
```

- (a) Mean **29.02**.
- (b) 10th **24.24**, 50th **29.05**, 90th **33.52**.
- (c) The 50th percentile is the median by definition; `median(canola)` confirms **29.05**.
</details>

### Question 14

A student types these four lines into the console, in order:

```r
p <- 40
p + 10
p <- p + 10
p * 2
```

**(a)** In a comment: which lines print something, and what exactly?

**(b)** In a comment: what is the value of `p` after all four lines, and how do the second and third lines differ even though both contain `p + 10`?

**(c)** Type the four lines in and check your predictions.

<details><summary>Answer</summary>

- (a) The second line prints `[1] 50` and the fourth prints `[1] 100`. The two assignments print nothing.
- (b) `p` is 50. The second line only *displays* `p + 10` -- `p` stays 40 -- while the third *assigns* the result back to `p`, overwriting it.
- (c) The console shows exactly the two printed values, and the Variables pane shows `p = 50`.
</details>

### Question 15

A student writes a script that reads data and computes three statistics. They run it once, and it works. The next day they open Positron, open the script, and press **Cmd+Enter** on the *last* line only -- it fails with `object not found`.

**(a)** In a comment: why did the last line fail today when the whole script worked yesterday?

**(b)** In a comment: what should they do instead of running the last line alone?

**(c)** In a comment: which of the module's script-writing habits is this scenario about?

<details><summary>Answer</summary>

- (a) A fresh R session starts empty. Yesterday's objects are gone, and the last line depends on objects created by the earlier lines, which have not been run today.
- (b) Run the script from the top, so every object the later lines need gets created first.
- (c) Make the script runnable from the beginning -- a script should reproduce the analysis from a clean start.
</details>

### Question 16

Consider these three lines:

```r
k <- 25
k == 30
mean(x = c(10, 20, 60))
```

**(a)** In a comment: say what each line does and what, if anything, it prints.

**(b)** In a comment: after the third line runs, does an object called `x` appear in the Variables pane? Why or why not?

**(c)** In a comment: which single symbol would you change to make line 2 *assign* 30 to `k`, and which does this book prefer for assignment?

<details><summary>Answer</summary>

- (a) Line 1 assigns 25 to `k` (prints nothing). Line 2 asks whether `k` equals 30 and prints `[1] FALSE`. Line 3 computes a mean and prints `[1] 30`.
- (b) No -- the `x =` inside the call names an argument for `mean()`; it does not create an object.
- (c) `==` would become `<-` (or `=`, which also assigns) -- and the book uses `<-` for assignment.
</details>

### Question 17

Here is a working script:

```r
q <- c(21.4, 25.8, 19.6, 23.2)
z <- c(140, 190, 160, 220)
t <- sum(q * z)
t / sum(z)
```

**(a)** Rewrite it with a proper header block and a comment before each step, following the module's conventions.

**(b)** Replace the object names with descriptive ones.

**(c)** Run your version. What is the final number, and what does it mean in words?

<details><summary>Answer</summary>

```r
# ---
# Title: Flax farm average yield
# Author: Your Name
# Date: 2026-10-08
# Description:
#   Computes total production and the acreage-weighted
#   average yield for four flax fields.
# ---

# Flax yields (bu/ac) and field sizes (acres)
yields <- c(21.4, 25.8, 19.6, 23.2)
acres  <- c(140, 190, 160, 220)

# Total production in bushels
total_bu <- sum(yields * acres)

# Average yield weighted by acres
total_bu / sum(acres)
```

- (c) **22.73 bu/ac** (from 16,138 total bushels over 710 acres) -- the farm's average flax yield with each field counted by its size.
</details>

### Question 18

Five oat fields yielded 88.2, 95.1, 79.8, 91.4, and 84.6 bu/ac. Five pea fields yielded 39.5, 42.1, 36.8, 44.0, and 38.2 bu/ac.

**(a)** Compute the mean and standard deviation of each crop's yields.

**(b)** In a comment: the oat standard deviation is about twice the pea one. Does that make oats the more variable crop *relative to its own average*?

**(c)** Compute the coefficient of variation for each crop and answer (b) with numbers.

<details><summary>Answer</summary>

```r
oats <- c(88.2, 95.1, 79.8, 91.4, 84.6)
peas <- c(39.5, 42.1, 36.8, 44.0, 38.2)
sd(oats) / mean(oats)
sd(peas) / mean(peas)
```

- (a) Oats mean **87.82**, sd **5.93**; peas mean **40.12**, sd **2.92**.
- (b, c) No -- CV oats **0.068**, peas **0.073**. Relative to their own averages, pea yields are slightly the more variable; the oat sd is bigger mainly because oat yields are bigger.
</details>

### Question 19

**(a)** In a comment: `mean()` takes its data as an argument named `x`. Write two calls that compute the mean of 12, 18, and 30 -- one with the argument unnamed, one with it named -- and confirm they agree.

**(b)** In a comment: what do you type to open the documentation for `sd()`?

**(c)** A classmate writes `Mean(c(12, 18, 30))` and R replies `could not find function "Mean"`. In a comment: what went wrong?

**(d)** Compute the standard deviation of the same three values, with the argument named.

<details><summary>Answer</summary>

```r
mean(c(12, 18, 30))
mean(x = c(12, 18, 30))
sd(x = c(12, 18, 30))
```

- (a) Both print `[1] 20`.
- (b) `?sd`.
- (c) R is exact about names: `Mean` with a capital M is not the same function as `mean`.
- (d) `[1] 9.165151`.
</details>

### Question 20

A survey records, for each farm: the nearest town, the number of quarters farmed, and whether the farm has irrigation.

**(a)** Create three vectors of length four with made-up values, choosing the appropriate type for each.

**(b)** In a comment: a classmate wrote their quarters vector as `c("6", "9", "4", "11")`. What type did they create, and what will happen when they try `mean()` on it?

**(c)** Combine your three vectors into a data frame called `farms` and print it.

<details><summary>Answer</summary>

```r
town      <- c("Rosetown", "Davidson", "Outlook", "Elrose")
quarters  <- c(6, 9, 4, 11)
irrigated <- c(FALSE, FALSE, TRUE, FALSE)
farms <- data.frame(town = town,
                    quarters = quarters,
                    irrigated = irrigated)
farms
```

- (b) A character vector -- the quotes make the numbers text, and `mean()` on it stops with an error because text has no mean.
</details>

### Question 21

Four oat fields yielded 84.3, 91.7, 77.2, and 88.5 bu/ac, and oats are selling at $4.15 per bushel.

**(a)** Save the yields as a vector and the price as a single object.

**(b)** Compute a revenue-per-acre vector (yield times price) in one line, using the price object rather than retyping the number.

**(c)** Compute the mean revenue per acre.

**(d)** In a comment: the price rises to $4.60. What two lines would you rerun to update every answer, and why is that easier than if you had typed the price into each formula?

<details><summary>Answer</summary>

```r
yields <- c(84.3, 91.7, 77.2, 88.5)
price  <- 4.15
revenue <- yields * price
revenue
mean(revenue)
```

- (b) 349.85, 380.56, 320.38, 367.28 $/ac, rounded to the cent (the console shows 349.845 and so on).
- (c) Mean **$354.51** per acre.
- (d) Change `price <- 4.60`, then rerun the lines below it. The price lives in one object, so one edit updates everything -- the same reason a spreadsheet keeps a price in one cell.
</details>

### Question 22

**(a)** Build a data frame called `durum_fields` with columns `field_id` (`"D1"` to `"D4"`), `region` (South, South, Central, North), and `yield` (28.3, 41.6, 35.2, 44.8).

**(b)** Report its shape (rows, columns, names) using functions.

**(c)** Using `$`, compute the mean yield and the range (maximum minus minimum).

<details><summary>Answer</summary>

```r
durum_fields <- data.frame(
  field_id = c("D1", "D2", "D3", "D4"),
  region = c("South", "South", "Central", "North"),
  yield = c(28.3, 41.6, 35.2, 44.8)
)
nrow(durum_fields); ncol(durum_fields); names(durum_fields)
mean(durum_fields$yield)
max(durum_fields$yield) - min(durum_fields$yield)
```

- (b) 4 rows, 3 columns; `field_id`, `region`, `yield`.
- (c) Mean **37.48**; range **16.5**.
</details>

### Question 23

Eight durum fields yielded 28.3, 41.6, 35.2, 44.8, 31.9, 38.7, 26.5, and 42.3 bu/ac.

**(a)** Compute the median with `median()`, and again as the 50th percentile with `quantile()`. Confirm they agree.

**(b)** Compute the same 50th percentile with the arguments named and in reverse order.

**(c)** In a comment: the mean is 36.16 and the median is 36.95. In Module 1's language, what does a mean slightly *below* the median suggest?

<details><summary>Answer</summary>

```r
durum <- c(28.3, 41.6, 35.2, 44.8, 31.9, 38.7, 26.5, 42.3)
median(durum)
quantile(durum, 0.5)
quantile(probs = 0.5, x = durum)
```

- (a, b) All three give **36.95**.
- (c) Mild left skew -- a few low values pull the mean down more than the high values pull it up.
</details>

### Question 24

A student runs these lines in the console:

```r
bins <- c(1450, 2200, 1875)
bins * 2
total <- sum(bins)
```

**(a)** In a comment: what exactly does the second line print, `[1]` included, and is that doubled vector saved anywhere?

**(b)** In a comment: what does the third line print, and what does the Variables pane show afterwards?

**(c)** The student now wants the doubled vector kept for later. Write the line that does it.

<details><summary>Answer</summary>

- (a) `[1] 2900 4400 3750` -- displayed only, saved nowhere.
- (b) Nothing prints (it is an assignment); the pane shows `bins` (the three values) and `total` = 5525.
- (c) `bins_doubled <- bins * 2`.
</details>

### Question 25

A classmate's script begins with the analysis and creates its objects in a scattered order; it only runs correctly because they ran some lines in the console first.

**(a)** In a comment: state the module's test for whether a script is reproducible.

**(b)** In a comment: name the two things the module says belong at the *top* of every script.

**(c)** In a comment: why does "it works on my machine, right now" not count as a working script for this course?

**(d)** Write the header block you would put at the top of your own fixed version of their script.

<details><summary>Answer</summary>

```r
# ---
# Title: Canola field analysis
# Author: Your Name
# Date: 2026-10-08
# Description: Reads the field data and computes summary statistics.
# ---
```

- (a) Restart R and run the script from top to bottom in a fresh session; it should reproduce the analysis with nothing typed in the console.
- (b) A header block (title, author, date, description) and the `library()` calls for any packages the script uses.
- (c) The grader runs the script from scratch in a clean session. Anything that depended on console leftovers or run order will fail there, whatever it did on the author's machine.
</details>

### Question 26

Consider:

```r
n_fields <- 12
n_fields = 8
n_fields == 8
```

**(a)** In a comment: what is `n_fields` after each of the three lines, and what does each line print?

**(b)** In a comment: the second line worked -- so why does this book still tell you to write `<-` for assignment?

**(c)** In a comment: in `mean(x = c(12, 18, 30))`, is the `=` assigning anything? What job is it doing?

<details><summary>Answer</summary>

- (a) After line 1 it is 12, after line 2 it is 8 (both assignments, printing nothing); line 3 leaves it at 8 and prints `[1] TRUE`.
- (b) Because `=` already has another job -- naming arguments inside function calls -- and `==` a third. Keeping `<-` for assignment keeps the three jobs visually distinct, which prevents exactly the confusion this question is exploiting.
- (c) No object is created -- inside a function call, `=` names an argument. It hands the vector to `mean()`'s `x` and leaves the workspace untouched.
</details>

### Question 27

Here is a working script:

```r
v1 <- c(47.2, 49.8, 46.5, 50.1, 48.4)
v2 <- c(38.9, 55.6, 44.2, 58.3, 41.0)
mean(v1)
mean(v2)
```

**(a)** Rewrite it with a header and comments. The two vectors are canola yields from Farm A and Farm B.

**(b)** Add lines computing each farm's standard deviation and range.

**(c)** In a comment: the two farms have nearly the same mean. Using your numbers from (b), say in one sentence how they differ.

<details><summary>Answer</summary>

```r
# ---
# Title: Comparing two canola farms
# Author: Your Name
# Date: 2026-10-08
# Description:
#   Compares the centre and spread of canola yields
#   on Farm A and Farm B.
# ---

# Canola yields by field (bu/ac)
farm_a <- c(47.2, 49.8, 46.5, 50.1, 48.4)
farm_b <- c(38.9, 55.6, 44.2, 58.3, 41.0)

# Centre
mean(farm_a)
mean(farm_b)

# Spread
sd(farm_a)
sd(farm_b)
max(farm_a) - min(farm_a)
max(farm_b) - min(farm_b)
```

- (b) Farm A sd **1.57**, range **3.6**; Farm B sd **8.79**, range **19.4**.
- (c) The means are close (48.4 vs 47.6), but Farm B's yields swing far more from field to field -- its spread is five times Farm A's.
</details>

### Question 28

**(a)** In a comment: a data frame's columns can hold different types, but a single vector cannot. Say what each column of this data frame would need to be: a farm name, a seeded date written as text, an acre count, and a certified-seed yes/no.

**(b)** Create the four vectors (length three, made-up values) and combine them into a data frame called `seeding`.

**(c)** Report the column names and the number of rows using functions.

<details><summary>Answer</summary>

```r
farm      <- c("Sunrise", "Coulee", "Big Sky")
seeded    <- c("2026-05-04", "2026-05-09", "2026-05-12")
acres     <- c(1120, 940, 1560)
certified <- c(TRUE, FALSE, TRUE)
seeding <- data.frame(farm = farm,
                      seeded = seeded,
                      acres = acres,
                      certified = certified)
names(seeding)
nrow(seeding)
```

- (a) Character, character (a date stored as text), numeric, logical.
- (c) `farm`, `seeded`, `acres`, `certified`; 3 rows.
</details>

### Question 29

Eight wheat fields yielded 52.1, 58.4, 49.7, 61.2, 55.8, 47.3, 59.6, and 53.9 bu/ac.

**(a)** Compute the 25th and 75th percentiles in one `quantile()` call.

**(b)** Compute the interquartile range from your two percentiles.

**(c)** In a comment: say what the number from (b) means in words for these eight fields.

<details><summary>Answer</summary>

```r
wheat <- c(52.1, 58.4, 49.7, 61.2, 55.8, 47.3, 59.6, 53.9)
quantile(wheat, c(0.25, 0.75))
```

- (a) 25th **51.5**, 75th **58.7**.
- (b) IQR **7.2**.
- (c) The middle half of these fields' yields falls within a 7.2 bu/ac window, from about 51.5 to 58.7.
</details>

### Question 30

**(a)** In a comment: three of the following literals are written incorrectly for their intended type. Which three, and what is wrong? Intended: the number twelve, the text Canola, the logical true, the number 8.5, the text 2026.

```r
"12"    Canola    TRUE    8.5    2026
```

**(b)** Write all five correctly as a `c()` call each would belong in (one numeric vector, one character vector, one logical vector).

**(c)** In a comment: a classmate writes `c(12, "Canola", TRUE)`. A vector holds one type -- so what does R do with this, and what happens if they then try `mean()` on it?

<details><summary>Answer</summary>

- (a) `"12"` is text but twelve was meant as a number (drop the quotes); `Canola` without quotes would be read as an object name (add quotes); `2026` is a number but the text 2026 was intended (add quotes). `TRUE` and `8.5` are correct as written.
- (b)

```r
values  <- c(12, 8.5)
text    <- c("Canola", "2026")
seeded  <- c(TRUE)
```

- (c) R converts everything to text -- `"12"`, `"Canola"`, `"TRUE"` -- because that is the only type all three values can share. `mean()` on it stops with an error, since text has no mean.
</details>

# Section 2 — Reading and Inspecting Data

*Three blocks of ten, one per dataset. Each block starts by naming its file; on a real test your Section 2 question comes from one of the three. Answer in the Section 2 block of your script.*

## Questions 31--40 · Saskatchewan RM Canola {.unnumbered}

*These questions use the canola-only file, [`rm_canola_yields_1990_2025.csv`](data/rm_canola_yields_1990_2025.csv), in your project's `data/` folder.*

### Question 31

**(a)** In a comment at the top of your script, sketch your project's folder layout (the subfolders and where the csv sits).

**(b)** Load the tidyverse and read the canola file into an object called `rm_canola`, using a relative path.

**(c)** Report the number of rows, the number of columns, and the column names using functions.

**(d)** In a comment: why will your `read_csv()` line work on the grader's computer, when a full path starting `C:/Users/...` would not?

<details><summary>Answer</summary>

```r
library(tidyverse)
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")
nrow(rm_canola); ncol(rm_canola); names(rm_canola)
```

- (c) **10,039 rows**, **5 columns**: `Year`, `RM`, `Crop`, `Yield`, `Unit`.
- (d) The path is relative to the project folder, and the grader opens the same self-contained folder -- so `data/...` exists on their machine too. An absolute path names one particular computer's user and folder layout.
</details>

### Question 32

A classmate's script reads the data like this:

```r
rm_canola <- read_csv("C:\\Users\\jordan\\Desktop\\stats stuff\\rm_canola_yields_1990_2025.csv")
```

It runs on their laptop and fails on everyone else's.

**(a)** In a comment: give two separate problems with this line (one about where the path points, one about the backslashes).

**(b)** Rewrite the line the way the module recommends, assuming the csv is in the project's `data/` folder.

**(c)** In a comment: the classmate asks "but with your short path, how does R know where to start looking?" Answer in one sentence.

**(d)** Your fixed line could itself fail with `Error: 'data/rm_canola_yields_1990_2025.csv' does not exist in current working directory (...)`. In a comment: what is the first thing to check?

<details><summary>Answer</summary>

- (a) The absolute path names one specific machine (user `jordan`, a Desktop folder), so it exists nowhere else; and backslashes mean something special inside an R string -- paths in R use forward slashes even on Windows.
- (b) `rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")`.
- (c) Opening the project folder in Positron (**File → Open Folder…**) makes that folder R's starting point, so relative paths begin from it.
- (d) That the folder opened in Positron is the project folder -- the one that contains `data/` with the csv inside. If some other folder is open, the relative path starts from the wrong place.
</details>

### Question 33

You organized the course as one big folder and opened `AREC_261/` in Positron:

```
AREC_261/
  module_2/
    data/
      rm_canola_yields_1990_2025.csv
    code/
  README.md
```

**(a)** Write the `read_csv()` line that works with this setup.

**(b)** In a comment: your line from (a) stops working after you re-organize and open `module_2/` itself in Positron. Why, and what does the path become?

**(c)** In a comment: starting from the opened `AREC_261/` folder, trace how R follows your path from (a) to the file, one folder at a time.

<details><summary>Answer</summary>

- (a) `rm_canola <- read_csv("module_2/data/rm_canola_yields_1990_2025.csv")`.
- (b) The starting point changed: relative paths begin at the folder you opened. With `module_2/` open, the path is `data/rm_canola_yields_1990_2025.csv`.
- (c) From `AREC_261`, R goes into `module_2`, then into `data`, and finds `rm_canola_yields_1990_2025.csv` -- each `/` steps into a subfolder.
</details>

### Question 34

**(a)** Run `glimpse()` on `rm_canola`. In a comment: which columns are text (`<chr>`) and which are numeric (`<dbl>`)?

**(b)** When `read_csv()` ran, it printed a note about the file. In a comment: what did that note report, and why does the module say to read it every time?

**(c)** `Year` came in as numeric (`<dbl>`). In a comment: which type of value is each of `2024` and `"2024"`? Why do we want a column we will compute with stored as numbers rather than as text?

<details><summary>Answer</summary>

- (a) `Crop` and `Unit` are `<chr>`; `Year`, `RM`, and `Yield` are `<dbl>`.
- (b) The row and column counts and its guess at each column's type (here `chr (2)` and `dbl (3)`). R guesses well but not perfectly, so the note is the first check that the file read the way you expect.
- (c) `2024` is numeric; `"2024"` is character (text). Numbers stored as text stop behaving like numbers -- you cannot do arithmetic with them, and sorting turns alphabetical -- so a column you will compute with should show `<dbl>`, not `<chr>`.
</details>

### Question 35

**(a)** Run `summary(rm_canola)`. From its output, report the minimum, median, and maximum of `Yield`, and the first and last `Year`.

**(b)** In a comment: work through the module's inspection questions for this file. Are the yields in a plausible range for canola in bu/ac? Are the years what the file name promises? Are there missing values?

**(c)** The minimum is below 2 bu/ac. In a comment: is a canola yield that low necessarily a data error? What kind of year could produce it?

<details><summary>Answer</summary>

- (a) Yield: min **1.9**, median **26.9**, max **61**. Years **1990** to **2025**.
- (b) Yes -- everything sits between about 2 and 61 bu/ac, sensible for canola; the years run 1990--2025 as advertised; and `summary()` prints an `NA's` count only when a column has missing values -- no such line appears, so there are none.
- (c) Not necessarily -- an RM average that low is what a severe drought or a widespread crop failure looks like. Extreme is not the same as wrong; it is a value to investigate, not delete.
</details>

### Question 36

**(a)** Compute the mean, median, standard deviation, and variance of `rm_canola$Yield`.

**(b)** Compute the 90th percentile of `Yield`.

**(c)** The mean sits a little above the median. In a comment: using Module 1's language about means and medians, what does that direction of gap suggest about the shape of canola yields?

<details><summary>Answer</summary>

```r
mean(rm_canola$Yield)
median(rm_canola$Yield)
sd(rm_canola$Yield)
var(rm_canola$Yield)
quantile(rm_canola$Yield, 0.9)
```

- (a) Mean **28.29**, median **26.9**, sd **10.06**, variance **101.21** (the sd squared, in squared units).
- (b) 90th percentile **42.5**.
- (c) A mean above the median suggests mild right skew: the very best RM-years pull the mean up more than the worst years pull it down.
</details>

### Question 37

Open `rm_canola_yields_1990_2025.csv` in Positron's data viewer (click the file in the Explorer pane).

**(a)** In a comment: according to the viewer's summary panel, does any column have missing values?

**(b)** In a comment: expand the `Yield` column's summary. Describe the shape of its histogram in one phrase, and say whether it matches your mean-versus-median reading from Question 36.

**(c)** A classmate says "the file is open in the viewer, so it's loaded into R." In a comment: correct them, and say what actually loads it.

<details><summary>Answer</summary>

- (a) No -- the summary panel shows 0% missing for every column.
- (b) A single hump around the mid-20s with a longer tail to the right -- consistent with the mean sitting a little above the median.
- (c) The viewer is Positron showing you the file; nothing exists in R until `read_csv()` runs and the result is assigned to an object.
</details>

### Question 38

**(a)** Compute the 25th and 75th percentiles of `Yield` in a single `quantile()` call.

**(b)** Compute the interquartile range twice: once with `IQR()`, once by subtracting your two percentiles.

**(c)** In a comment: say what the interquartile range from (b) means in words, for a reader who farms.

<details><summary>Answer</summary>

```r
quantile(rm_canola$Yield, c(0.25, 0.75))
IQR(rm_canola$Yield)
```

- (a) 25th **21**, 75th **35.4**.
- (b) Both give **14.4**.
- (c) The middle half of all RM-year canola yields falls between 21 and 35.4 bu/ac -- a typical RM in a typical year lands somewhere in that 14-bushel window.
</details>

### Question 39

A classmate's fresh R session runs `rm_yields <- read_csv("data/rm_yields_1990_2025.csv")` and gets:

```
Error in read_csv(...) : could not find function "read_csv"
```

**(a)** In a comment: what is the most likely cause, and what one line fixes it?

**(b)** In a comment: they object, "but I installed the tidyverse last week!" Explain the difference between installing and loading, and how often each is done.

**(c)** In a comment: state the quotes rule -- where do quotation marks go when installing, and when loading?

**(d)** The tidyverse is a collection of packages. In comments, name the four core ones from the module and the job each does.

<details><summary>Answer</summary>

- (a) The package is not loaded in this session; run `library(tidyverse)` first.
- (b) Installing downloads the package onto the computer -- once per computer. Loading switches it on -- once per script or session. Installation does not carry across sessions as loaded.
- (c) Quotes when you install (`install.packages("tidyverse")`), none when you load (`library(tidyverse)`).
- (d) `readr` reads data, `dplyr` manipulates it, `tidyr` reshapes it, `ggplot2` makes graphs.
</details>

### Question 40

You spot an obviously wrong yield value in the csv. A classmate suggests opening the file in Excel, fixing the cell, and saving. Separately, your download folder contains `RM Yields FINAL (2).csv`.

**(a)** In a comment: what is wrong with the classmate's suggestion, and where should the correction happen instead?

**(b)** In a comment: give two problems with the file name, and rename it following the module's two rules.

**(c)** In a comment: if a date belonged in the file name, how should it be written, and what is the advantage?

<details><summary>Answer</summary>

- (a) It overwrites the only copy of the raw data and leaves no record of what changed. Keep the file exactly as it arrived and make the correction in the R script, where it leaves a trail.
- (b) Spaces and punctuation, and a `FINAL (2)` tag that will only ever breed a `FINAL (3)`. Something like `rm_yields_long.csv`.
- (c) As YYYY-MM-DD, e.g. `2026-10-08_rm_yields.csv` -- alphabetical order is then also chronological order.
</details>

---

## Questions 41--50 · Manitoba Wheat Variety {.unnumbered}

*These questions use [`mb_wheat_reported_2020_2025.csv`](data/mb_wheat_reported_2020_2025.csv) -- spring wheat yields by variety and municipality, reported observations only.*

### Question 41

**(a)** In a comment at the top of your script, sketch your project's folder layout.

**(b)** Load the tidyverse and read the file into an object called `mb_wheat`, using a relative path.

**(c)** Report the number of rows, the number of columns, and the column names using functions.

**(d)** In a comment: what does one row of this dataset represent?

<details><summary>Answer</summary>

```r
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")
nrow(mb_wheat); ncol(mb_wheat); names(mb_wheat)
```

- (c) **2,398 rows**, **7 columns**: `Year`, `Municipality`, `Variety`, `Farms`, `Acres`, `Yield_bu_ac`, `Reported`.
- (d) One wheat variety grown in one municipality in one year -- its farm count, acres, and average yield.
</details>

### Question 42

A classmate's script reads the data like this:

```r
mb_wheat <- read_csv("/Users/sam/Downloads/mb wheat data.csv")
```

It works on their laptop and fails on everyone else's.

**(a)** In a comment: give two problems -- one about where the path points, one about the file's name and location.

**(b)** Rewrite the line the way the module recommends, assuming the file was saved as `mb_wheat_reported_2020_2025.csv` in the project's `data/` folder.

**(c)** In a comment: your line fails with `does not exist in current working directory`. What is the first thing to check?

<details><summary>Answer</summary>

- (a) The absolute path names one machine's user folder, so it exists nowhere else; and the file is loose in Downloads under a space-filled name instead of renamed and placed in the project's `data/` folder.
- (b) `mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")`.
- (c) That the folder opened in Positron is the project folder -- the one containing `data/`. Relative paths start from the opened folder.
</details>

### Question 43

Your course folder is organized as:

```
AREC_261/
  module_2/
    data/
      mb_wheat_reported_2020_2025.csv
    code/
  README.md
```

**(a)** Write the `read_csv()` line that works with `AREC_261/` open in Positron.

**(b)** In a comment: what does the path become if you open `module_2/` instead?

**(c)** In a comment: trace how R follows your path from (a) to the file, one folder at a time.

<details><summary>Answer</summary>

- (a) `mb_wheat <- read_csv("module_2/data/mb_wheat_reported_2020_2025.csv")`.
- (b) `data/mb_wheat_reported_2020_2025.csv` -- relative paths begin at the opened folder.
- (c) From `AREC_261`, R steps into `module_2`, then into `data`, and finds the csv; each `/` is one step down.
</details>

### Question 44

**(a)** Run `glimpse()` on `mb_wheat`. In a comment: which columns are text (`<chr>`), which numeric (`<dbl>`), and which logical (`<lgl>`)?

**(b)** In a comment: what values does the logical column hold in this file, and why is that no accident? (Look at the file's name.)

**(c)** In a comment: `Farms` holds whole numbers. Would `glimpse()` showing it as `<chr>` be a problem, and what would it tell you about the file?

<details><summary>Answer</summary>

- (a) `Municipality` and `Variety` are `<chr>`; `Year`, `Farms`, `Acres`, `Yield_bu_ac` are `<dbl>`; `Reported` is `<lgl>`.
- (b) All `TRUE` -- this is the reported-only file, with the suppressed (`FALSE`) rows removed.
- (c) Yes -- a numeric column arriving as text means some value in it could not be read as a number, and calculations like `mean()` would fail on it.
</details>

### Question 45

**(a)** Run `summary(mb_wheat)`. From its output, report the minimum, median, and maximum of `Yield_bu_ac`, and the first and last `Year`.

**(b)** In a comment: work through the module's inspection questions. Are the yields in a plausible range for spring wheat? Are there missing values?

**(c)** The minimum is **4.5 bu/ac**. In a comment: is a reported yield that low necessarily an error?

<details><summary>Answer</summary>

- (a) Yield: min **4.5**, median **62.2**, max **97.8**. Years **2020** to **2025**.
- (b) Plausible -- most values sit tens-of-bushels, sensible for Manitoba wheat; no `NA's` lines appear, so no missing values in this file.
- (c) Not necessarily -- a variety on a few drought- or flood-hit fields can genuinely average that low. Extreme is a value to investigate, not delete.
</details>

### Question 46

**(a)** Compute the mean, median, standard deviation, and variance of `mb_wheat$Yield_bu_ac`.

**(b)** In a comment: the mean sits a little *below* the median here. In Module 1's language, what does that direction of gap suggest, and what kind of observations would produce it?

**(c)** Compute the smallest and largest yields in one `range()` call.

<details><summary>Answer</summary>

```r
mean(mb_wheat$Yield_bu_ac)
median(mb_wheat$Yield_bu_ac)
sd(mb_wheat$Yield_bu_ac)
var(mb_wheat$Yield_bu_ac)
range(mb_wheat$Yield_bu_ac)
```

- (a) Mean **61.16**, median **62.2**, sd **12.38**, variance **153.34**.
- (b) Mild left skew: unusually *low* values pull the mean down more than the high values pull it up -- a tail of very poor variety-municipality observations mixed in with normal ones.
- (c) `[1]  4.5 97.8` -- `range()` returns both ends as one vector.
</details>

### Question 47

Open `mb_wheat_reported_2020_2025.csv` in Positron's data viewer.

**(a)** In a comment: according to the viewer's summary panel, does any column have missing values?

**(b)** In a comment: expand `Yield_bu_ac`. Describe the histogram's shape in one phrase, and say whether it matches your mean-versus-median reading from Question 46.

**(c)** A classmate says "it's open in the viewer, so it's loaded into R." In a comment: correct them.

<details><summary>Answer</summary>

- (a) No -- 0% missing in every column of this file.
- (b) A single hump around the low 60s with a longer tail stretching left -- consistent with the mean sitting below the median.
- (c) The viewer is Positron showing the file; nothing exists in R until `read_csv()` runs and the result is assigned to an object.
</details>

### Question 48

**(a)** Compute the 25th and 75th percentiles of `Yield_bu_ac` in one `quantile()` call.

**(b)** Compute the interquartile range twice: with `IQR()`, and by subtracting your percentiles.

**(c)** In a comment: say what the IQR means in words, for a reader who farms.

<details><summary>Answer</summary>

```r
quantile(mb_wheat$Yield_bu_ac, c(0.25, 0.75))
IQR(mb_wheat$Yield_bu_ac)
```

- (a) 25th **53.9**, 75th **69.6**.
- (b) Both give **15.7**.
- (c) The middle half of variety-municipality yields lands between about 54 and 70 bu/ac -- a typical observation falls in that 16-bushel window.
</details>

### Question 49

A classmate installed the tidyverse last week. Today they restart R, run `mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")` as their first line, and get `could not find function "read_csv"`.

**(a)** In a comment: what happened, and what one line fixes it?

**(b)** In a comment: how often is installing done, and how often is loading done?

**(c)** In a comment: state the quotes rule for the two commands.

<details><summary>Answer</summary>

- (a) Installing survives across sessions but loading does not -- each new session starts with no packages loaded. Run `library(tidyverse)` first.
- (b) Install once per computer; load once per script or session.
- (c) Quotes when you install (`install.packages("tidyverse")`), none when you load (`library(tidyverse)`).
</details>

### Question 50

You spot what looks like a data-entry slip in the csv, and your download folder contains `MB Wheat FINAL final(3).csv`.

**(a)** In a comment: a classmate suggests opening the csv in Excel, typing in a corrected yield, and saving. What is wrong with that, and where should any correction live?

**(b)** In a comment: give two problems with the file name and rename it following the module's rules.

**(c)** In a comment: if you wanted the download date in the name, how would you write it and why?

<details><summary>Answer</summary>

- (a) It overwrites the only copy of the raw data with an invented value and leaves no record. Raw data stays exactly as it arrived; corrections happen in the R script, where they are visible and reversible.
- (b) Spaces plus a `FINAL final(3)` tag that only breeds a `final(4)`. Something like `mb_wheat_yields_2020_2025.csv`.
- (c) As YYYY-MM-DD (e.g. `2026-10-08_mb_wheat.csv`) so alphabetical order is also chronological order.
</details>

## Questions 51--60 · Canada Field Crops {.unnumbered}

*These questions use [`statcan_spring_wheat_2015_2025.csv`](data/statcan_spring_wheat_2015_2025.csv) -- Statistics Canada spring wheat seeded acres and average yield by province.*

### Question 51

**(a)** In a comment at the top of your script, sketch your project's folder layout.

**(b)** Load the tidyverse and read the file into an object called `sw`, using a relative path.

**(c)** Report the number of rows, the number of columns, and the column names using functions.

**(d)** In a comment: what does one row of this dataset represent?

<details><summary>Answer</summary>

```r
library(tidyverse)
sw <- read_csv("data/statcan_spring_wheat_2015_2025.csv")
nrow(sw); ncol(sw); names(sw)
```

- (c) **94 rows**, **5 columns**: `Year`, `Province`, `Crop`, `Seeded_acres`, `Yield_bu_ac`.
- (d) One province in one year: its spring wheat seeded acres and average yield.
</details>

### Question 52

A classmate's script reads the data like this:

```r
sw <- read_csv("C:\\Users\\alex\\Desktop\\statcan spring wheat.csv")
```

**(a)** In a comment: give two problems with this line.

**(b)** Rewrite it the way the module recommends, with the file in the project's `data/` folder under its proper name.

**(c)** In a comment: the rewritten line still fails on your machine. What is the first thing to check?

<details><summary>Answer</summary>

- (a) The absolute path names one machine (user `alex`, a Desktop folder); and backslashes mean something special inside an R string -- write forward slashes even on Windows.
- (b) `sw <- read_csv("data/statcan_spring_wheat_2015_2025.csv")`.
- (c) That the folder opened in Positron is the project folder containing `data/` -- the relative path starts from whatever folder is open.
</details>

### Question 53

**(a)** With `AREC_261/` open in Positron and the csv at `AREC_261/module_2/data/`, write the working `read_csv()` line.

**(b)** In a comment: a classmate with the *same* folders gets `does not exist` from your exact line. What is different about their session?

**(c)** In a comment: trace how R follows your path from (a) to the file, one folder at a time.

<details><summary>Answer</summary>

- (a) `sw <- read_csv("module_2/data/statcan_spring_wheat_2015_2025.csv")`.
- (b) They opened a different folder (for example `module_2/` itself, or no folder at all) -- the relative path starts from the opened folder, so the same line works or fails depending on the starting point, not the code.
- (c) From `AREC_261`, R steps into `module_2`, then into `data`, and finds the csv; each `/` is one step down.
</details>

### Question 54

**(a)** Run `glimpse()` on `sw`. In a comment: which columns are text and which are numeric?

**(b)** In a comment: when `read_csv()` ran, what did its note report for this file?

**(c)** In a comment: Statistics Canada sometimes publishes a suppressed value as the letter `x`. If even one `Seeded_acres` value in the csv were an `x`, what type would the whole column arrive as, and what would break?

<details><summary>Answer</summary>

- (a) `Province` and `Crop` are `<chr>`; `Year`, `Seeded_acres`, and `Yield_bu_ac` are `<dbl>`.
- (b) 94 rows, 5 columns, and its type guesses: `chr (2)`, `dbl (3)`.
- (c) Text -- a column takes one type, and one value that cannot be read as a number turns the whole column `<chr>`. Every calculation on it (`mean()`, comparisons, sorting by size) would then fail, which is why the types line in `glimpse()` is worth a glance after every read.
</details>

### Question 55

**(a)** Run `summary(sw)`. From its output, report the minimum, median, and maximum of `Yield_bu_ac`, and the first and last `Year`.

**(b)** In a comment: run the module's inspection questions. Plausible ranges? Missing values? The variables you expected?

**(c)** Open the csv in Positron's data viewer and expand `Seeded_acres`. In a comment: describe its histogram's shape in one phrase.

<details><summary>Answer</summary>

- (a) Yield: min **33.7**, median **50.9**, max **73.9**. Years **2015** to **2025**.
- (b) Yields of 34--74 bu/ac are sensible for provincial spring wheat averages; no `NA's` lines appear, so nothing is missing; the five expected columns are present.
- (c) Almost everything piled near zero with a long tail stretching right -- a few prairie-sized values dwarf the rest.
</details>

### Question 56

**(a)** Compute the mean, median, and standard deviation of `sw$Yield_bu_ac`.

**(b)** In a comment: the mean and median here are nearly equal. What does that suggest about the shape of provincial wheat yields, in Module 1's language?

<details><summary>Answer</summary>

```r
mean(sw$Yield_bu_ac)
median(sw$Yield_bu_ac)
sd(sw$Yield_bu_ac)
```

- (a) Mean **51.32**, median **50.9**, sd **7.52**.
- (b) Roughly symmetric -- neither tail is pulling the mean away from the median.
</details>

### Question 57

**(a)** Compute the mean and the median of `sw$Seeded_acres`.

**(b)** The mean is about **2.06 million** acres; the median is about **115 thousand**. In a comment: that is a gap of nearly twenty to one -- what does it tell you, and what feature of Canadian wheat growing produces it?

**(c)** In a comment: a reporter wants one number for "the typical province's wheat acreage." Which of your two numbers do you give them?

<details><summary>Answer</summary>

```r
mean(sw$Seeded_acres)
median(sw$Seeded_acres)
```

- (a) Mean **2,063,932**; median **114,773**.
- (b) A mean far above the median flags a few enormous values pulling it up -- Saskatchewan and Alberta seed millions of acres while most provinces seed tens of thousands, so the distribution is heavily right-skewed.
- (c) The median -- the mean describes no actual province, sitting far above the typical one because of the prairie giants.
</details>

### Question 58

**(a)** Compute the 25th and 75th percentiles of `Yield_bu_ac` in one call, and the IQR twice (function, and subtraction).

**(b)** In a comment: say what the IQR means here in words.

<details><summary>Answer</summary>

```r
quantile(sw$Yield_bu_ac, c(0.25, 0.75))
IQR(sw$Yield_bu_ac)
```

- (a) 25th **47.3**, 75th **56.1**; IQR **8.8** both ways.
- (b) The middle half of province-year wheat yields falls in a nine-bushel window, from about 47 to 56 bu/ac.
</details>

### Question 59

A classmate on a brand-new laptop runs:

```r
install.packages(tidyverse)
```

and gets `Error: object 'tidyverse' not found`.

**(a)** In a comment: what rule from the module did they break, and what is the corrected line?

**(b)** In a comment: after installing correctly, what line still has to appear at the top of the script, and why?

**(c)** The tidyverse is a collection of packages. In comments, name the four core ones from the module and the job each does.

<details><summary>Answer</summary>

- (a) The quotes rule -- the package name is quoted when installing. Without quotes, R looks for an *object* called tidyverse and finds none. Corrected: `install.packages("tidyverse")`.
- (b) `library(tidyverse)` -- installing puts the package on the computer, but each script or session has to load it before its functions exist.
- (c) `readr` reads data, `dplyr` manipulates it, `tidyr` reshapes it, `ggplot2` makes graphs.
</details>

### Question 60

Your download folder contains `sk wheat data (new) FINAL.csv`.

**(a)** In a comment: list every violation of the module's file-naming rules in that name, and rename it properly.

**(b)** In a comment: why does the module say to save the file into `data/` untouched, even before you have looked at it?

**(c)** In a comment: if you wanted the download date in the name, how would you write it and why?

<details><summary>Answer</summary>

- (a) Spaces, parentheses, and a `FINAL` tag standing in for version control. Something like `statcan_spring_wheat_2015_2025.csv`.
- (b) `data/` holds raw files exactly as they arrived, so there is always an untouched original to go back to; every change happens in the script, where it leaves a record.
- (c) As YYYY-MM-DD (e.g. `2026-10-08_statcan_spring_wheat.csv`) so alphabetical order is also chronological order.
</details>

# Section 3 — Data Manipulation Functions

*Three blocks of ten, one per dataset. Each question uses two or three of the functions in sequence -- save each intermediate result to an object and work step by step (pipelines wait for Section 4).*

## Questions 61--70 · Saskatchewan RM Crop Yields {.unnumbered}

*These questions use the full eight-crop file: read [`rm_yields_1990_2025.csv`](data/rm_yields_1990_2025.csv) into an object called `rm_yields` at the start of this block.*

### Question 61

**(a)** Filter `rm_yields` to Canola in 2024, saving the result as `canola_2024`. How many rows are in the filtered data frame?

**(b)** From `canola_2024`, select only the `RM` and `Yield` columns, saving the result. Report its column names and its number of columns.

**(c)** In a comment: after (a) and (b), how many rows does `rm_yields` itself have, and why?

<details><summary>Answer</summary>

```r
canola_2024 <- filter(rm_yields, Crop == "Canola" & Year == 2024)
nrow(canola_2024)
canola_slim <- select(canola_2024, RM, Yield)
names(canola_slim)
ncol(canola_slim)
```

- (a) **293 rows**.
- (b) `RM`, `Yield`; **2 columns** (still 293 rows).
- (c) Still **71,104** -- `filter()` and `select()` return new data frames and leave the original untouched.
</details>

### Question 62

**(a)** Filter to rows where the crop is Oats **or** Barley, using `|`, saving the result. How many rows are in the filtered data frame?

**(b)** Write the same filter using `%in%` and confirm the row count matches.

**(c)** Sort your filtered data frame from highest yield to lowest. Which crop, RM, and year hold the top spot, and at what yield?

**(d)** In a comment: a classmate tried `filter(rm_yields, Crop == "Oats" & Crop == "Barley")` and got zero rows. Why?

<details><summary>Answer</summary>

```r
oats_barley <- filter(rm_yields, Crop == "Oats" | Crop == "Barley")
nrow(oats_barley)
oats_barley2 <- filter(rm_yields, Crop %in% c("Oats", "Barley"))
nrow(oats_barley2)
arrange(oats_barley, desc(Yield))
```

- (a, b) **19,709 rows** both ways.
- (c) A tie at **165.1 bu/ac**: Oats in RM **369** (2025) and Oats in RM **438** (2020) share the maximum, so either is a correct top row.
- (d) `&` requires both conditions true of the *same row*, and no single row's crop is both Oats and Barley. "Or" is the right connector.
</details>

### Question 63

**(a)** Filter to the Spring Wheat observations of at least 60 bu/ac (all years), using `>=` and saving the result. How many rows are in the filtered data frame?

**(b)** Filter again to keep only the 2023 rows of that result. How many rows are in this second data frame?

**(c)** From your 2023 data frame, select `RM` and `Yield` and sort best-first. Which RM tops the list, at what yield?

**(d)** In a comment: rewrite the condition "yield at least 60 and the crop is *not* Spring Wheat" using `!`.

<details><summary>Answer</summary>

```r
sw_60 <- filter(rm_yields, Crop == "Spring Wheat" & Yield >= 60)
nrow(sw_60)
sw_60_2023 <- filter(sw_60, Year == 2023)
nrow(sw_60_2023)
sw_best <- select(sw_60_2023, RM, Yield)
arrange(sw_best, desc(Yield))
```

- (a) **434 rows**. (b) **53 rows**.
- (c) RM **187**, at **74.7 bu/ac**.
- (d) `filter(rm_yields, Yield >= 60 & !Crop == "Spring Wheat")`.
</details>

### Question 64

**(a)** Make a version of the data with only `Year`, `RM`, `Crop`, and `Yield` -- once by *naming what you keep*, once by *dropping* what you don't want -- saving each, and confirm the two have the same column names. Then make a third version keeping the adjacent-column range `Year:Crop`, and report its column names.

**(b)** Rename `Yield` to `yield_per_acre` in the full data, saving the result, and show the new column names.

**(c)** Filter your renamed data frame to Peas in 2025. How many rows are in the filtered data frame?

<details><summary>Answer</summary>

```r
kept    <- select(rm_yields, Year, RM, Crop, Yield)
dropped <- select(rm_yields, -Unit)
names(kept)
names(dropped)
ranged  <- select(rm_yields, Year:Crop)
names(ranged)
renamed <- rename(rm_yields, yield_per_acre = Yield)
names(renamed)
peas_2025 <- filter(renamed, Crop == "Peas" & Year == 2025)
nrow(peas_2025)
```

- (a) Identical: `Year`, `RM`, `Crop`, `Yield` both ways; the range version keeps `Year`, `RM`, `Crop`.
- (b) `Year`, `RM`, `Crop`, `yield_per_acre`, `Unit`.
- (c) **285 rows**.
</details>

### Question 65

Lentil yields are in pounds per acre; one pound per acre is about **1.12 kg per hectare**.

**(a)** Filter to the lentil rows, saving as `lentils`. How many rows are in the filtered data frame?

**(b)** Use `mutate()` to add a `yield_kg_ha` column to `lentils`, saving the result back to `lentils`.

**(c)** Compute the mean of the new column, and the mean of the original `Yield` column, and check the ratio is the conversion factor.

<details><summary>Answer</summary>

```r
lentils <- filter(rm_yields, Crop == "Lentils")
nrow(lentils)
lentils <- mutate(lentils, yield_kg_ha = Yield * 1.12)
mean(lentils$yield_kg_ha)
mean(lentils$Yield)
```

- (a) **6,338 rows**.
- (c) Mean **1,353.4 kg/ha** vs **1,208.39 lb/ac** -- and 1353.4 / 1208.39 ≈ 1.12.
</details>

### Question 66

**(a)** Filter to Canola in 2024, then in one `mutate()` call add two columns: `yield_t_ha` (multiply by 0.0560) and `yield_kg_ha` (multiply by 56.0). Save each step.

**(b)** Compute the mean of each new column.

**(c)** Sort by `yield_t_ha`, best first. Which RM tops the table, and at what yield in t/ha?

**(d)** In a comment: without running anything, what is the mean of `yield_kg_ha` divided by the mean of `yield_t_ha`, and why?

<details><summary>Answer</summary>

```r
canola_2024 <- filter(rm_yields, Crop == "Canola" & Year == 2024)
canola_2024 <- mutate(canola_2024,
                      yield_t_ha  = Yield * 0.0560,
                      yield_kg_ha = Yield * 56.0)
mean(canola_2024$yield_t_ha)
mean(canola_2024$yield_kg_ha)
arrange(canola_2024, desc(yield_t_ha))
```

- (b) **1.76 t/ha** and **1,764.11 kg/ha**.
- (c) RM **493**, at **2.57 t/ha**.
- (d) 1,000 -- both columns are the same yields scaled by constants, and 56.0 / 0.0560 = 1000 (a tonne is 1,000 kg).
</details>

### Question 67

**(a)** Filter to Flax in 2025, saving as `flax_2025`. How many rows are in the filtered data frame?

**(b)** Sort it from highest yield to lowest, and report the RM and yield in the top row.

**(c)** Sort it from lowest to highest instead. What changes in the code?

**(d)** In a comment: after (b) and (c), is `flax_2025` itself now sorted? Why or why not?

<details><summary>Answer</summary>

```r
flax_2025 <- filter(rm_yields, Crop == "Flax" & Year == 2025)
nrow(flax_2025)
arrange(flax_2025, desc(Yield))
arrange(flax_2025, Yield)
```

- (a) **195 rows**.
- (b) RM **381** at **45.5 bu/ac**.
- (c) Drop the `desc()`.
- (d) No -- neither sorted result was assigned, so both were displayed and discarded; `flax_2025` is unchanged.
</details>

### Question 68

A script contains, in order:

```r
filter(rm_yields, Crop == "Peas")
peas <- filter(rm_yields, Crop == "Peas")
```

**(a)** In a comment: what does each line do, and what appears in the console after each?

**(b)** Report `nrow(peas)` and `nrow(rm_yields)`.

**(c)** In a comment: connect this to the same distinction you met with `x * 3` in the console back in Section 1.

<details><summary>Answer</summary>

- (a) Line 1 computes the pea rows and *prints* them -- nothing is saved. Line 2 computes the same thing and saves it as `peas`, printing nothing.
- (b) `nrow(peas)` is **9,270**; `nrow(rm_yields)` is still **71,104**.
- (c) It is displaying versus saving again: an expression's result appears in the console and vanishes unless assigned to an object.
</details>

### Question 69

**(a)** Filter to Canola in RM 232 (all years), saving the result. How many rows are in the filtered data frame?

**(b)** From that object, select only `Year` and `Yield`, saving again.

**(c)** Sort it so the best year is on top. Which year was it, and what was the yield?

<details><summary>Answer</summary>

```r
canola_232 <- filter(rm_yields, Crop == "Canola" & RM == 232)
nrow(canola_232)
canola_232 <- select(canola_232, Year, Yield)
arrange(canola_232, desc(Yield))
```

- (a) **31 rows** (some years are missing -- the crop was not grown or reported in that RM every year).
- (c) **2016**, at **43.9 bu/ac**.
</details>

### Question 70

**(a)** Using `head()`, display the first six rows of `rm_yields`.

**(b)** Sort the data by `Yield` in descending order into a new object, select its `Crop`, `Yield`, and `Unit` columns into another, and display the first six rows of that (leaving `rm_yields` unchanged throughout).

**(c)** In a comment: every row in (b) is the same crop. Which one, why those rows, and what would you check before comparing their values with the other crops?

**(d)** Now sort by `Crop` first and by descending `Yield` within each crop, into a new object, and display its first six rows. Which crop fills them, and why that crop?

<details><summary>Answer</summary>

```r
head(rm_yields)
sorted <- arrange(rm_yields, desc(Yield))
top_cols <- select(sorted, Crop, Yield, Unit)
head(top_cols)
by_crop <- arrange(rm_yields, Crop, desc(Yield))
head(by_crop)
```

- (b, c) All lentils -- their yields are recorded in pounds per acre, so they dominate the top of any sort on `Yield`. Before comparing across crops you would check the `Unit` column and convert to a common unit.
- (d) All Barley -- `Crop` sorts alphabetically first, so the display starts with Barley's own best yields (RM 380 in 2025 at 122 bu/ac on top).
</details>

---

## Questions 71--80 · Manitoba Wheat Variety {.unnumbered}

*These questions use the full [`mb_wheat_varieties.csv`](data/mb_wheat_varieties.csv) -- including the suppressed rows, where `Reported` is FALSE and the numeric columns are missing. Read it into an object called `mb` at the start of this block.*

### Question 71

**(a)** Filter `mb` to the reported rows (`Reported == TRUE`), saving as `mb_rep`. How many rows are in the filtered data frame?

**(b)** From `mb_rep`, select `Municipality`, `Variety`, and `Yield_bu_ac`, saving the result. Report its column names.

**(c)** In a comment: how many rows does `mb` itself still have, and what happened to the suppressed rows in `mb_rep`?

<details><summary>Answer</summary>

```r
mb_rep <- filter(mb, Reported == TRUE)
nrow(mb_rep)
mb_slim <- select(mb_rep, Municipality, Variety, Yield_bu_ac)
names(mb_slim)
```

- (a) **2,398 rows**.
- (b) `Municipality`, `Variety`, `Yield_bu_ac`.
- (c) `mb` still has **5,358** rows -- filtering returns a new data frame. The 2,960 suppressed rows are simply absent from `mb_rep`; they were not deleted from anything.
</details>

### Question 72

**(a)** From `mb_rep` (reported rows), filter to the varieties `AAC BRANDON (BW 932)` **or** `AAC STARBUCK <SECAN>` using `|`, saving the result. How many rows are in the filtered data frame?

**(b)** Write the same filter using `%in%` and confirm the count matches.

**(c)** In a comment: a classmate wrote the condition with `&` instead of `|` and got zero rows. Why?

<details><summary>Answer</summary>

```r
two_var <- filter(mb_rep, Variety == "AAC BRANDON (BW 932)" | Variety == "AAC STARBUCK <SECAN>")
nrow(two_var)
two_var2 <- filter(mb_rep, Variety %in% c("AAC BRANDON (BW 932)", "AAC STARBUCK <SECAN>"))
nrow(two_var2)
```

- (a, b) **893 rows** both ways.
- (c) `&` demands both conditions true of the same row, and no single row's variety is both names at once. "Or" is the right connector.
</details>

### Question 73

**(a)** Filter `mb_rep` to yields above 80 bu/ac, saving the result. How many rows are in the filtered data frame?

**(b)** Filter that result again to 2024 only. How many rows now?

**(c)** Sort your over-80 data frame from (a) best-first. Which municipality, variety, and year top the list, at what yield?

<details><summary>Answer</summary>

```r
high <- filter(mb_rep, Yield_bu_ac > 80)
nrow(high)
high_2024 <- filter(high, Year == 2024)
nrow(high_2024)
arrange(high, desc(Yield_bu_ac))
```

- (a) **97 rows**. (b) **16 rows**.
- (c) HILLSBURG-ROBLIN-SHELL RIVER, `BOLLES <SEED DEPOT>`, in **2022**, at **97.8 bu/ac**.
</details>

### Question 74

**(a)** From `mb_rep`, make a version with only `Year`, `Municipality`, `Variety`, and `Yield_bu_ac` -- once by naming what you keep, once by dropping what you don't (`Farms`, `Acres`, `Reported`) -- saving each, and confirm the two have the same column names. Then make a third version keeping the adjacent range `Year:Variety` and report its names.

**(b)** Rename `Yield_bu_ac` to `yield`, saving the result, and show the new column names.

<details><summary>Answer</summary>

```r
kept    <- select(mb_rep, Year, Municipality, Variety, Yield_bu_ac)
dropped <- select(mb_rep, -Farms, -Acres, -Reported)
names(kept)
names(dropped)
ranged  <- select(mb_rep, Year:Variety)
names(ranged)
renamed <- rename(mb_rep, yield = Yield_bu_ac)
names(renamed)
```

- (a) Identical four names both ways; the range version keeps `Year`, `Municipality`, `Variety`.
- (b) The same seven columns with `yield` in place of `Yield_bu_ac`.
</details>

### Question 75

**(a)** From `mb_rep`, use one `mutate()` to add `total_bu`, each observation's acres times its yield, saving the result.

**(b)** Compute the mean of `total_bu`.

**(c)** Sort by `total_bu`, largest first. Which municipality tops the table, and at roughly what total?

**(d)** In a comment: what does `total_bu` measure that `Yield_bu_ac` does not?

<details><summary>Answer</summary>

```r
mb_rep <- mutate(mb_rep, total_bu = Acres * Yield_bu_ac)
mean(mb_rep$total_bu)
arrange(mb_rep, desc(total_bu))
```

- (b) Mean **382,865 bu**.
- (c) PORTAGE LA PRAIRIE, at about **5.13 million bushels**.
- (d) Production -- how much grain the planting actually produced -- rather than how well each acre yielded. A modest yield on enormous acres can out-produce a great yield on few.
</details>

### Question 76

A bushel of wheat is 60 lb, so one bu/ac is about 0.0673 t/ha.

**(a)** Filter `mb_rep` to 2025, saving the result. How many rows are in the filtered data frame?

**(b)** Add a `yield_t_ha` column with `mutate()`, saving the result, and compute its mean.

**(c)** In a comment: the canola conversion factor in the chapters was 0.0560. Why is wheat's different?

<details><summary>Answer</summary>

```r
rep_2025 <- filter(mb_rep, Year == 2025)
nrow(rep_2025)
rep_2025 <- mutate(rep_2025, yield_t_ha = Yield_bu_ac * 0.0673)
mean(rep_2025$yield_t_ha)
```

- (a) **438 rows**.
- (b) Mean **4.58 t/ha**.
- (c) A bushel is a volume standardized by weight per crop -- wheat's bushel is 60 lb and canola's is 50 lb -- so the bu/ac-to-t/ha factor differs crop by crop.
</details>

### Question 77

**(a)** Sort `mb_rep` from highest yield to lowest, and report the municipality, variety, and year in the top row.

**(b)** Sort it from lowest to highest instead. What changes in the code?

**(c)** In a comment: after (a) and (b), is `mb_rep` itself now sorted? Why or why not?

<details><summary>Answer</summary>

```r
arrange(mb_rep, desc(Yield_bu_ac))
arrange(mb_rep, Yield_bu_ac)
```

- (a) HILLSBURG-ROBLIN-SHELL RIVER, `BOLLES <SEED DEPOT>`, **2022**, at **97.8 bu/ac**.
- (b) Drop the `desc()`.
- (c) No -- neither sorted result was assigned; both were displayed and discarded.
</details>

### Question 78

A script contains, in order:

```r
filter(mb, Reported == FALSE)
suppressed <- filter(mb, Reported == FALSE)
```

**(a)** In a comment: what does each line do, and what appears in the console after each?

**(b)** Report `nrow(suppressed)` and `nrow(mb)`.

**(c)** In a comment: the suppressed rows have no yields. What did MASC suppress them for, according to the dataset description?

<details><summary>Answer</summary>

- (a) Line 1 computes the suppressed rows and prints them, saving nothing. Line 2 computes the same thing and saves it as `suppressed`, printing nothing.
- (b) `nrow(suppressed)` is **2,960**; `nrow(mb)` is still **5,358**.
- (c) Too few farms grew that variety in that municipality -- the value is withheld to protect individual farms' privacy, not because it was never measured.
</details>

### Question 79

**(a)** From `mb_rep`, filter to the municipality ALEXANDER, saving the result. How many rows are in the filtered data frame?

**(b)** Select `Variety`, `Year`, and `Yield_bu_ac` from it, saving again.

**(c)** Sort so the best observation is on top. Which variety was it, at what yield?

<details><summary>Answer</summary>

```r
alexander <- filter(mb_rep, Municipality == "ALEXANDER")
nrow(alexander)
alexander <- select(alexander, Variety, Year, Yield_bu_ac)
arrange(alexander, desc(Yield_bu_ac))
```

- (a) **17 rows**.
- (c) `AAC ELIE(BW931)`, at **90.9 bu/ac**.
</details>

### Question 80

**(a)** Using `head()`, display the first six rows of `mb`.

**(b)** Sort `mb_rep` by `Acres` in descending order into a new object, select its `Municipality`, `Variety`, `Year`, and `Acres` columns into another, and display the first six rows.

**(c)** In a comment: what does the top row of (b) tell you, and why did we sort `mb_rep` rather than `mb`?

<details><summary>Answer</summary>

```r
head(mb)
by_acres <- arrange(mb_rep, desc(Acres))
big <- select(by_acres, Municipality, Variety, Year, Acres)
head(big)
```

- (b, c) The biggest single planting is `AAC BRANDON (BW 932)` in PORTAGE LA PRAIRIE, 2020, at **74,604 acres**. We sorted the reported rows because the suppressed rows have no acre values to sort on.
</details>

## Questions 81--90 · Canada Field Crops {.unnumbered}

*These questions use the full [`statcan_field_crops.csv`](data/statcan_field_crops.csv) -- thirteen crops by province, 2015--2025. Read it into an object called `crops` at the start of this block.*

### Question 81

**(a)** Filter `crops` to Saskatchewan, saving the result. How many rows are in the filtered data frame?

**(b)** From it, select `Year`, `Crop`, and `Yield_bu_ac`, saving the result, and report its column names.

**(c)** In a comment: how many rows does `crops` itself still have, and why?

<details><summary>Answer</summary>

```r
sask <- filter(crops, Province == "Saskatchewan")
nrow(sask)
sask_slim <- select(sask, Year, Crop, Yield_bu_ac)
names(sask_slim)
```

- (a) **137 rows**.
- (b) `Year`, `Crop`, `Yield_bu_ac`.
- (c) Still **1,067** -- the functions return new data frames and leave the original untouched.
</details>

### Question 82

**(a)** Filter to the three prairie provinces -- Alberta, Saskatchewan, Manitoba -- using `%in%`, saving the result. How many rows are in the filtered data frame?

**(b)** Write the same filter with two `|`s and confirm the count matches.

**(c)** In a comment: which version is easier to read, and when does the difference grow?

<details><summary>Answer</summary>

```r
prairie <- filter(crops, Province %in% c("Alberta", "Saskatchewan", "Manitoba"))
nrow(prairie)
prairie2 <- filter(crops, Province == "Alberta" | Province == "Saskatchewan" | Province == "Manitoba")
nrow(prairie2)
```

- (a, b) **410 rows** both ways.
- (c) `%in%` -- one condition instead of a chain, and the gap widens with every extra value in the set.
</details>

### Question 83

**(a)** Filter `crops` to Canola in 2025, saving the result. How many rows are in the filtered data frame?

**(b)** Sort it from highest yield to lowest. Which province tops the table, at what yield?

**(c)** In a comment: only seven provinces appear, not ten. What is the likely reason?

<details><summary>Answer</summary>

```r
canola_25 <- filter(crops, Crop == "Canola" & Year == 2025)
nrow(canola_25)
arrange(canola_25, desc(Yield_bu_ac))
```

- (a) **7 rows**.
- (b) Ontario, at **49.8 bu/ac**.
- (c) The other provinces grow too little canola for StatCan to publish an estimate -- absent rows usually mean not grown or not reported, not zero.
</details>

### Question 84

**(a)** Make a version of `crops` with only `Year`, `Province`, `Crop`, and `Yield_bu_ac` -- once by naming what you keep, once by dropping `Seeded_acres` -- saving each, and confirm the two match. Then keep the adjacent range `Year:Crop` and report its names.

**(b)** Rename `Seeded_acres` to `acres`, saving the result, and show the new column names.

<details><summary>Answer</summary>

```r
kept    <- select(crops, Year, Province, Crop, Yield_bu_ac)
dropped <- select(crops, -Seeded_acres)
names(kept)
names(dropped)
ranged  <- select(crops, Year:Crop)
names(ranged)
renamed <- rename(crops, acres = Seeded_acres)
names(renamed)
```

- (a) Identical four names both ways; the range keeps `Year`, `Province`, `Crop`.
- (b) `Year`, `Province`, `Crop`, `acres`, `Yield_bu_ac`.
</details>

### Question 85

**(a)** Filter `crops` to Spring wheat, saving the result. How many rows are in the filtered data frame?

**(b)** Add a `seeded_m_acres` column -- seeded acres in millions (divide by 1,000,000) -- with `mutate()`, saving the result.

**(c)** Compute the mean of the new column. In a comment: why might millions of acres be the better unit to report than raw acres?

<details><summary>Answer</summary>

```r
spring <- filter(crops, Crop == "Spring wheat")
nrow(spring)
spring <- mutate(spring, seeded_m_acres = Seeded_acres / 1000000)
mean(spring$seeded_m_acres)
```

- (a) **94 rows**.
- (c) Mean **2.06 million acres** -- a reader can hold "about two million acres" in their head; a nine-digit raw number they cannot.
</details>

### Question 86

**(a)** Filter to Saskatchewan spring wheat (all years), saving the result. How many rows are in the filtered data frame?

**(b)** Add a `total_bu` column, seeded acres times yield, with `mutate()`, saving the result.

**(c)** Sort by `total_bu`, largest first. Which year tops the table, and at roughly what production?

**(d)** In a comment: could a year top this table without having the best yield? How?

<details><summary>Answer</summary>

```r
sask_sw <- filter(crops, Province == "Saskatchewan" & Crop == "Spring wheat")
nrow(sask_sw)
sask_sw <- mutate(sask_sw, total_bu = Seeded_acres * Yield_bu_ac)
arrange(sask_sw, desc(total_bu))
```

- (a) **11 rows**.
- (c) **2025**, at about **490 million bushels**.
- (d) Yes -- production is acres times yield, so a big-acreage year with a middling yield can out-produce a small-acreage year with a great one.
</details>

### Question 87

**(a)** Sort `crops` by `Seeded_acres` in descending order into a new object, select `Province`, `Crop`, `Year`, and `Seeded_acres` into another, and display the first six rows.

**(b)** In a comment: which province-crop combination fills the top of the table, and at what peak acreage?

**(c)** In a comment: the crop list contains both `Wheat (all)` and `Spring wheat`. What does that mean for anyone tempted to add acreages across crops?

<details><summary>Answer</summary>

```r
by_acres <- arrange(crops, desc(Seeded_acres))
big <- select(by_acres, Province, Crop, Year, Seeded_acres)
head(big)
```

- (b) Saskatchewan's `Wheat (all)`, peaking at **14,178,100 acres** in 2024.
- (c) The categories overlap -- `Wheat (all)` already contains spring wheat, durum, and winter wheat -- so summing acreage across crop rows would double-count. Check what each category includes before adding.
</details>

### Question 88

A script contains, in order:

```r
filter(crops, Province == "Manitoba")
manitoba <- filter(crops, Province == "Manitoba")
```

**(a)** In a comment: what does each line do, and what appears in the console after each?

**(b)** Report `nrow(manitoba)` and `nrow(crops)`.

**(c)** In a comment: line 1's printed result has scrolled out of the console. Is it stored anywhere you can get it back from?

<details><summary>Answer</summary>

- (a) Line 1 computes Manitoba's rows and prints them, saving nothing; line 2 saves the same result as `manitoba` and prints nothing.
- (b) `nrow(manitoba)` is **137**; `nrow(crops)` is still **1,067**.
- (c) No -- a displayed result that was never assigned is gone once it scrolls past. Only line 2's `manitoba` can be brought back, which is the point of assigning.
</details>

### Question 89

**(a)** Filter `crops` to Flax, saving the result. How many rows are in the filtered data frame?

**(b)** Select `Province`, `Year`, and `Yield_bu_ac` from it, saving again, and sort best-first. Which province and year top the list, at what yield?

**(c)** In a comment: only 39 flax rows exist across eleven years. What does that tell you about where flax is grown?

<details><summary>Answer</summary>

```r
flax <- filter(crops, Crop == "Flax")
nrow(flax)
flax <- select(flax, Province, Year, Yield_bu_ac)
arrange(flax, desc(Yield_bu_ac))
```

- (a) **39 rows**.
- (b) Alberta in **2025**, at **40.8 bu/ac**.
- (c) Only three to six provinces (usually three) report it in any year -- flax is a prairie specialty crop, not grown countrywide.
</details>

### Question 90

**(a)** Using `head()`, display the first six rows of `crops`.

**(b)** Sort by `Yield_bu_ac` in descending order into a new object, select `Crop`, `Province`, `Year`, and `Yield_bu_ac` into another, and display the first six rows.

**(c)** In a comment: every top row is the same crop and province. Which, and why does a 180 bu/ac corn yield not mean corn "beats" wheat at 74?

<details><summary>Answer</summary>

```r
head(crops)
by_yield <- arrange(crops, desc(Yield_bu_ac))
top <- select(by_yield, Crop, Province, Year, Yield_bu_ac)
head(top)
```

- (b, c) All Ontario corn for grain, peaking at **180 bu/ac** in 2024. Corn is simply a higher-bulk crop -- and a corn bushel (56 lb) is not even the same quantity as a wheat bushel (60 lb), so cross-crop comparisons in bu/ac compare different things twice over.
</details>

# Section 4 — Pipelines and Grouped Summaries

*Three blocks of ten, one per dataset. Answer in the Section 4 block of your script. Write multi-step work as pipelines with `|>`, one step per line.*

## Questions 91--100 · Saskatchewan RM Crop Yields {.unnumbered}

*Still the full eight-crop file (`rm_yields`).*

### Question 91

**(a)** Write a pipeline, one step per line with a short comment on each step, that filters `rm_yields` to Durum in 2025, keeps only the `RM` and `Yield` columns, and sorts the result from highest yield to lowest.

**(b)** Run it. How many rows are in the resulting data frame, and which RM had the best durum yield in 2025, at what value?

**(c)** In a comment: read your pipeline aloud as a "take the data, then ..., then ..., then ..." sentence.

<details><summary>Answer</summary>

```r
rm_yields |>                              # take the data, THEN
  filter(Crop == "Durum" & Year == 2025) |>  # keep durum in 2025, THEN
  select(RM, Yield) |>                    # keep two columns, THEN
  arrange(desc(Yield))                    # sort best-first
```

- (b) **189 rows**; RM **369** at **105.6 bu/ac**.
- (c) Take `rm_yields`, then keep the 2025 durum rows, then keep the RM and Yield columns, then sort from highest yield to lowest.
</details>

### Question 92

**(a)** In one pipeline, compute the mean flax yield across 2020--2025 (`Year >= 2020`), calling the result column `mean_yield`.

**(b)** Add `n()` to the same `summarise()` to count how many observations that mean is based on.

**(c)** In a comment: why is reporting the count alongside the mean good practice?

<details><summary>Answer</summary>

```r
rm_yields |>
  filter(Crop == "Flax" & Year >= 2020) |>
  summarise(mean_yield = mean(Yield),
            n = n())
```

- (a, b) Mean **22.39 bu/ac** from **1,143** observations.
- (c) A mean based on a handful of rows deserves less trust than one based on a thousand; the count tells the reader which they are looking at.
</details>

### Question 93

**(a)** In one pipeline, filter to Canola and summarise three things: the mean yield, the standard deviation, and the number of observations.

**(b)** Report the three values.

**(c)** In a comment: your result is a data frame. How many rows does it have, and what would `mutate()` have produced instead of `summarise()` here?

<details><summary>Answer</summary>

```r
rm_yields |>
  filter(Crop == "Canola") |>
  summarise(mean_yield = mean(Yield),
            sd_yield = sd(Yield),
            n = n())
```

- (b) Mean **28.29**, sd **10.06**, n **10,039**.
- (c) One row. `mutate()` would have kept all 10,039 rows and repeated the overall mean on every one -- `summarise()` collapses, `mutate()` adds.
</details>

### Question 94

**(a)** In one pipeline, compute the mean yield and observation count for **each crop**, over the whole dataset.

**(b)** Which crop shows the largest mean, and roughly how far above the others is it?

**(c)** In a comment: why is that comparison meaningless as it stands, and which column proves it?

<details><summary>Answer</summary>

```r
rm_yields |>
  group_by(Crop) |>
  summarise(mean_yield = mean(Yield),
            n = n())
```

- (a) Eight rows, one per crop: Barley **51.8**, Canola **28.3**, Durum **33.6**, Flax **20.5**, Lentils **1,208.4**, Oats **64.7**, Peas **31.8**, Spring Wheat **35.3**.
- (b) Lentils, at nearly twenty times the next-highest crop.
- (c) Lentils are measured in pounds per acre and everything else in bushels -- the `Unit` column. The group means are each fine on their own terms; comparing them across the unit boundary is not.
</details>

### Question 95

**(a)** In one pipeline, compute the mean canola yield for **each year** from 2021 to 2025 (filter first, then group).

**(b)** Report the five means. Which year was worst, and which best?

**(c)** In a comment: 2021 was a severe drought year in Saskatchewan. Does the data agree?

<details><summary>Answer</summary>

```r
rm_yields |>
  filter(Crop == "Canola" & Year >= 2021) |>
  group_by(Year) |>
  summarise(mean_yield = mean(Yield))
```

- (a, b) 2021 **21.9**, 2022 **35.4**, 2023 **33.9**, 2024 **31.5**, 2025 **43.9**. Worst 2021, best 2025.
- (c) Yes -- the 2021 mean sits roughly a third below the surrounding years.
</details>

### Question 96

**(a)** In one pipeline, compute each RM's mean Spring Wheat yield across all years, and sort the result so the best wheat-growing RMs are on top.

**(b)** Report the top three RMs and their means.

**(c)** In a comment: your result has one row per RM. What did each row of the *input* represent, and what changed?

<details><summary>Answer</summary>

```r
rm_yields |>
  filter(Crop == "Spring Wheat") |>
  group_by(RM) |>
  summarise(mean_yield = mean(Yield)) |>
  arrange(desc(mean_yield))
```

- (b) RM **369** (47.6), RM **333** (47.2), RM **368** (47.2).
- (c) Each input row was one RM-year-crop observation; `group_by` + `summarise` collapsed the years, leaving one row per RM.
</details>

### Question 97

**(a)** In one pipeline: filter to Lentils, convert to kg/ha with `mutate()` (multiply by 1.12), then summarise the mean of the converted column.

**(b)** Report the mean in kg/ha.

**(c)** In a comment: in what order do your steps run, and why would putting the `summarise()` before the `mutate()` fail?

<details><summary>Answer</summary>

```r
rm_yields |>
  filter(Crop == "Lentils") |>
  mutate(yield_kg_ha = Yield * 1.12) |>
  summarise(mean_kg_ha = mean(yield_kg_ha))
```

- (b) **1,353.4 kg/ha**.
- (c) Top to bottom -- each step's output feeds the next. After `summarise()` the data has collapsed to one row and the `yield_kg_ha` column would not exist yet to summarise.
</details>

### Question 98

A classmate hands you this pipeline and asks what it does:

```r
rm_yields |>
  filter(Crop %in% c("Oats", "Barley") & Year >= 2015) |>
  group_by(Crop) |>
  summarise(mean_yield = mean(Yield),
            n = n())
```

**(a)** In a comment: explain the pipeline step by step, one sentence per step.

**(b)** Run it and report the output.

**(c)** In a comment: answer the classmate's real question in plain language -- what does the output say about oats and barley since 2015?

<details><summary>Answer</summary>

- (a) Take the data; keep only oat and barley rows from 2015 on; split those rows into the two crops; collapse each crop to its mean yield and its number of observations.
- (b) Barley **60.5 bu/ac** (n = 3,131); Oats **79.5 bu/ac** (n = 2,434).
- (c) Since 2015, oat yields have averaged about 19 bu/ac higher than barley yields across Saskatchewan RMs, based on several thousand observations of each.
</details>

### Question 99

This pipeline was meant to find low- and high-yielding canola observations -- below 15 **or** above 55 bu/ac -- but it returns zero rows:

```r
rm_yields |>
  filter(Crop == "Canola") |>
  filter(Yield < 15 & Yield > 55) |>
  arrange(desc(Yield))
```

**(a)** In a comment: apply the module's debugging habit -- which step do you run first, and what do you check at each step?

**(b)** In a comment: identify the broken step and explain *why* it returns nothing.

**(c)** Write the corrected pipeline.

**(d)** In a comment: your corrected pipeline works, but a classmate's copy of it still gives odd results even though their code now matches yours. What does the module say to try before hunting further?

<details><summary>Answer</summary>

- (a) Run the first step alone and check the result looks right (the canola rows), then add the second step and check again -- the first step where the data stops looking right is the problem.
- (b) The second filter: `&` demands both conditions true of the same row, and no yield is simultaneously below 15 and above 55. The question said "or".
- (c) Replace `&` with `|`: `filter(Yield < 15 | Yield > 55)`.
- (d) Restart R and run the script from the top in a fresh session -- leftover objects from the console, or stale versions of a data frame, disappear on a clean run.
</details>

### Question 100

**(a)** In one pipeline: for **2025 only**, compute each crop's mean yield and observation count, sorted from highest mean to lowest, and save the result as `crop_summary_2025`.

**(b)** Write the table to `output/crop_summary_2025.csv`.

**(c)** Report the two highest-yielding crops in the sorted table. In a comment: why does the crop in first place not mean what it appears to mean, and which crop is really the standout?

<details><summary>Answer</summary>

```r
crop_summary_2025 <- rm_yields |>
  filter(Year == 2025) |>
  group_by(Crop) |>
  summarise(mean_yield = mean(Yield),
            n = n()) |>
  arrange(desc(mean_yield))

write_csv(crop_summary_2025, "output/crop_summary_2025.csv")
```

- (a, c) Lentils top the table at **1,897** -- but that is pounds per acre, not bushels. Among the bushel crops, Oats lead at **97.5 bu/ac** (Barley next at 72.8). 2025 was a strong year across the board.
</details>

## Questions 101--110 · Manitoba Wheat Variety {.unnumbered}

*Full [`mb_wheat_varieties.csv`](data/mb_wheat_varieties.csv), read as `mb`, with the reported rows saved as `mb_rep <- filter(mb, Reported == TRUE)` at the start of this block. Build everything else as pipelines.*

### Question 101

**(a)** Write a pipeline, one step per line with a short comment on each, that filters `mb_rep` to 2025, keeps `Municipality`, `Variety`, and `Yield_bu_ac`, and sorts best-first.

**(b)** Run it. How many rows are in the resulting data frame, and which municipality and variety top it, at what yield?

**(c)** In a comment: read the pipeline aloud as a "take the data, then ..." sentence.

<details><summary>Answer</summary>

```r
mb_rep |>                                          # take the reported rows, THEN
  filter(Year == 2025) |>                          # keep 2025, THEN
  select(Municipality, Variety, Yield_bu_ac) |>    # keep three columns, THEN
  arrange(desc(Yield_bu_ac))                       # sort best-first
```

- (b) **438 rows**; LOUISE, growing SY MANNESS, at **96.3 bu/ac**.
- (c) Take the reported rows, then keep 2025, then keep three columns, then sort from highest yield to lowest.
</details>

### Question 102

**(a)** In one pipeline, compute the mean of `Yield_bu_ac` across all of `mb_rep`, calling it `mean_yield`, together with `n()`.

**(b)** Report both values.

**(c)** In a comment: why report the count alongside the mean?

<details><summary>Answer</summary>

```r
mb_rep |>
  summarise(mean_yield = mean(Yield_bu_ac),
            n = n())
```

- (b) Mean **61.16 bu/ac** from **2,398** observations.
- (c) A mean earns trust in proportion to what it is based on; the count tells the reader how much that is.
</details>

### Question 103

**(a)** In one pipeline, filter `mb_rep` to the variety `AAC STARBUCK <SECAN>` and summarise its mean yield, standard deviation, and observation count.

**(b)** Report the three values.

**(c)** In a comment: how many rows does your result have, and what would `mutate()` have produced in place of `summarise()`?

<details><summary>Answer</summary>

```r
mb_rep |>
  filter(Variety == "AAC STARBUCK <SECAN>") |>
  summarise(mean_yield = mean(Yield_bu_ac),
            sd_yield = sd(Yield_bu_ac),
            n = n())
```

- (b) Mean **62.7**, sd **11.28**, n **374**.
- (c) One row. `mutate()` would have kept all 374 rows with the summary repeated on each -- `summarise()` collapses, `mutate()` adds.
</details>

### Question 104

**(a)** In one pipeline, compute each variety's mean yield and observation count across `mb_rep`, sorted from highest mean to lowest.

**(b)** Report the top two varieties, their means, and their counts.

**(c)** In a comment: the second-place variety was observed exactly once. Which first-place claim would you trust more, and why?

<details><summary>Answer</summary>

```r
mb_rep |>
  group_by(Variety) |>
  summarise(mean_yield = mean(Yield_bu_ac),
            n = n()) |>
  arrange(desc(mean_yield))
```

- (b) AAC WESTKING, **79.2 bu/ac** from **17** observations; then CDC GO (BW781), **77.5** from **1**.
- (c) WESTKING's -- seventeen municipality-years is evidence; a single observation could be one lucky field in one good year.
</details>

### Question 105

**(a)** In one pipeline, compute the mean reported yield for each year, 2020 through 2025.

**(b)** Report the six means. Which year was worst, and which best?

**(c)** In a comment: the RM dataset showed Saskatchewan canola cratering in 2021. Does Manitoba wheat agree about that year?

<details><summary>Answer</summary>

```r
mb_rep |>
  group_by(Year) |>
  summarise(mean_yield = mean(Yield_bu_ac))
```

- (a, b) 2020 **62.4**, 2021 **49.6**, 2022 **59.8**, 2023 **61.4**, 2024 **64.7**, 2025 **68.0**. Worst 2021, best 2025.
- (c) Yes -- 2021 sits ten to fifteen bushels below every other year. The drought was prairie-wide, not a Saskatchewan quirk.
</details>

### Question 106

**(a)** In one pipeline, compute each municipality's mean reported yield, sorted best-first.

**(b)** Report the top three municipalities and their means.

**(c)** In a comment: what did each row of the input represent, and what does each row of the result represent?

<details><summary>Answer</summary>

```r
mb_rep |>
  group_by(Municipality) |>
  summarise(mean_yield = mean(Yield_bu_ac)) |>
  arrange(desc(mean_yield))
```

- (b) LOUISE (**72.9**), LAC DU BONNET (**71.5**), ALEXANDER (**71.4**).
- (c) Each input row was one variety in one municipality in one year; each result row is one municipality, its varieties and years collapsed into a single mean.
</details>

### Question 107

A bushel of wheat is 60 lb, so one bu/ac is about 0.0673 t/ha.

**(a)** In one pipeline: convert `mb_rep`'s yields to t/ha with `mutate()`, then compute the mean t/ha for each year.

**(b)** Report the 2021 and 2025 values.

**(c)** In a comment: why must the `mutate()` step come before the `summarise()` step?

<details><summary>Answer</summary>

```r
mb_rep |>
  mutate(yield_t_ha = Yield_bu_ac * 0.0673) |>
  group_by(Year) |>
  summarise(mean_t_ha = mean(yield_t_ha))
```

- (b) 2021 **3.34 t/ha**; 2025 **4.58 t/ha**.
- (c) Each step's output feeds the next -- after `summarise()` the data has collapsed to one row per year and the `yield_t_ha` column would not exist to average.
</details>

### Question 108

A classmate hands you this pipeline and asks what it does:

```r
mb_rep |>
  filter(Yield_bu_ac > 70) |>
  group_by(Variety) |>
  summarise(n = n()) |>
  arrange(desc(n))
```

**(a)** In a comment: explain it step by step, one sentence per step.

**(b)** Run it and report the top three rows.

**(c)** In a comment: answer the real question in plain language -- what does the output say?

<details><summary>Answer</summary>

- (a) Take the reported rows; keep only observations over 70 bu/ac; split them by variety; count each variety's rows; sort the counts largest-first.
- (b) `AAC STARBUCK <SECAN>` **102**, `AAC WHEATLAND <SECAN>` **82**, `AAC BRANDON (BW 932)` **73**.
- (c) When Manitoba wheat topped 70 bu/ac, it was most often Starbuck doing it -- these three varieties account for most of the high-yield observations.
</details>

### Question 109

This pipeline was meant to find AAC Brandon's reported observations, but it returns zero rows:

```r
mb_rep |>
  filter(Variety == "aac brandon (bw 932)") |>
  arrange(desc(Yield_bu_ac))
```

**(a)** In a comment: apply the debugging habit -- what do you run first, and what do you check?

**(b)** In a comment: identify the problem and say why R found nothing, with no error.

**(c)** Write the corrected pipeline.

<details><summary>Answer</summary>

- (a) Run the filter step alone and look at its result -- it returns zero rows, so the problem is there, not in the sort.
- (b) Text matching is exact, capitals included. The data stores the variety in capitals, so the lowercase string matches no row -- and a filter that matches nothing is not an error, just an empty result.
- (c) `mb_rep |> filter(Variety == "AAC BRANDON (BW 932)") |> arrange(desc(Yield_bu_ac))`.
</details>

### Question 110

**(a)** In one pipeline: for **2025 only**, compute each variety's mean reported yield and observation count, sorted best-first, and save the result as `variety_summary_2025`.

**(b)** Write the table to `output/variety_summary_2025.csv`.

**(c)** Report the top row. In a comment: the table also contains varieties observed only twice -- what should a reader keep in mind when comparing rows?

<details><summary>Answer</summary>

```r
variety_summary_2025 <- mb_rep |>
  filter(Year == 2025) |>
  group_by(Variety) |>
  summarise(mean_yield = mean(Yield_bu_ac),
            n = n()) |>
  arrange(desc(mean_yield))

write_csv(variety_summary_2025, "output/variety_summary_2025.csv")
```

- (c) AAC WESTKING, **79.2 bu/ac** from 17 observations. Rows with tiny counts (some varieties appear once or twice) carry far less evidence than their neighbours -- the `n` column is part of the answer, not decoration.
</details>

## Questions 111--120 · Canada Field Crops {.unnumbered}

*Full [`statcan_field_crops.csv`](data/statcan_field_crops.csv), read as `crops` at the start of this block. Build everything as pipelines.*

### Question 111

**(a)** Write a pipeline, one step per line with a short comment on each, that filters `crops` to Canola, keeps `Province`, `Year`, and `Yield_bu_ac`, and sorts best-first.

**(b)** Run it. How many rows, and which province-year tops the table, at what yield?

**(c)** In a comment: before declaring that province Canada's canola capital, what other column in the dataset would you look at, and why?

<details><summary>Answer</summary>

```r
crops |>                                    # take the data, THEN
  filter(Crop == "Canola") |>               # keep canola, THEN
  select(Province, Year, Yield_bu_ac) |>    # keep three columns, THEN
  arrange(desc(Yield_bu_ac))                # sort best-first
```

- (b) **75 rows**; New Brunswick in **2024**, at **54.2 bu/ac**. (One row, Prince Edward Island 2021, has a missing yield; `arrange()` puts missing values last, so the sort is unaffected.)
- (c) `Seeded_acres` -- a tiny acreage can post a great average that means little next to Saskatchewan's millions of acres. Best yield and biggest grower are different titles.
</details>

### Question 112

**(a)** In one pipeline, filter to Spring wheat and summarise the mean yield, the standard deviation, and the observation count.

**(b)** Report the three values.

**(c)** In a comment: say what the standard deviation means here, for a reader who farms.

<details><summary>Answer</summary>

```r
crops |>
  filter(Crop == "Spring wheat") |>
  summarise(mean_yield = mean(Yield_bu_ac),
            sd_yield = sd(Yield_bu_ac),
            n = n())
```

- (b) Mean **51.32 bu/ac**, sd **7.52**, from **94** province-years.
- (c) A typical province-year sits within about seven or eight bushels of the 51 bu/ac average -- the spread across provinces and years, not field-to-field variation.
</details>

### Question 113

**(a)** In one pipeline, filter to Saskatchewan spring wheat and summarise the mean yield and count.

**(b)** Report both values.

**(c)** In a comment: Saskatchewan's mean sits about five bushels below the all-province average of 51.3 bu/ac. How can the country's biggest wheat grower not have its best yields?

<details><summary>Answer</summary>

```r
crops |>
  filter(Province == "Saskatchewan" & Crop == "Spring wheat") |>
  summarise(mean_yield = mean(Yield_bu_ac),
            n = n())
```

- (b) Mean **46.6 bu/ac** from **11** years.
- (c) Acreage and yield measure different things -- Saskatchewan farms vast dryland acres, including land drier than anything a smaller, wetter province would put into wheat. Growing the most and yielding the most per acre are separate claims.
</details>

### Question 114

**(a)** In one pipeline, compute each province's mean spring wheat yield, with counts, sorted best-first.

**(b)** Report the top three provinces.

**(c)** In a comment: what does each row of the result represent, and how many years does each mean average over?

<details><summary>Answer</summary>

```r
crops |>
  filter(Crop == "Spring wheat") |>
  group_by(Province) |>
  summarise(mean_yield = mean(Yield_bu_ac),
            n = n()) |>
  arrange(desc(mean_yield))
```

- (b) Manitoba (**59.0**), Ontario (**56.6**), Alberta (**52.3**).
- (c) One row per province, its years collapsed into a single mean. Most provinces contribute all eleven years (n = 11), but not all -- Nova Scotia has spring wheat estimates for only six, which is what the `n` column is there to show.
</details>

### Question 115

**(a)** In one pipeline, compute the national mean spring wheat yield for each year, 2015 through 2025.

**(b)** Which year was worst, and which best, at what values?

**(c)** In a comment: both other datasets in this bank flagged the same worst year. What is the year, and what does agreement across three independent datasets tell you?

<details><summary>Answer</summary>

```r
crops |>
  filter(Crop == "Spring wheat") |>
  group_by(Year) |>
  summarise(mean_yield = mean(Yield_bu_ac))
```

- (b) Worst **2021** at **45.7**; best **2022** at **56.5**.
- (c) 2021 -- the prairie drought. When three datasets collected by different agencies point at the same year, the signal is in the world, not in any one file.
</details>

### Question 116

**(a)** In one pipeline, compute each province's mean spring wheat *seeded acres*, sorted largest-first.

**(b)** Report the top three.

**(c)** In a comment: on mean *yield*, Manitoba leads the provinces at 59.0 bu/ac and Saskatchewan trails the prairies. Compare that ordering with this table -- what changes, and what does that say about "best wheat province" as a phrase?

<details><summary>Answer</summary>

```r
crops |>
  filter(Crop == "Spring wheat") |>
  group_by(Province) |>
  summarise(mean_acres = mean(Seeded_acres)) |>
  arrange(desc(mean_acres))
```

- (b) Saskatchewan (**8.03M acres**), Alberta (**6.23M**), Manitoba (**2.99M**).
- (c) The order flips -- Manitoba leads on yield but is third on acres, Saskatchewan the reverse. "Best" depends entirely on whether you mean per-acre performance or total scale.
</details>

### Question 117

**(a)** In one pipeline: filter to Spring wheat, create `total_bu` (seeded acres times yield) with `mutate()`, then compute each province's mean `total_bu`, sorted largest-first.

**(b)** Report the top province and its rough value.

**(c)** In a comment: why must the `mutate()` come before the `group_by()`/`summarise()` pair?

<details><summary>Answer</summary>

```r
crops |>
  filter(Crop == "Spring wheat") |>
  mutate(total_bu = Seeded_acres * Yield_bu_ac) |>
  group_by(Province) |>
  summarise(mean_total_bu = mean(total_bu)) |>
  arrange(desc(mean_total_bu))
```

- (b) Saskatchewan, at about **377 million bushels** a year.
- (c) `summarise()` collapses the rows -- the per-row `total_bu` has to exist before the collapse, or there is nothing to average.
</details>

### Question 118

A classmate hands you this pipeline:

```r
crops |>
  filter(Province %in% c("Alberta", "Saskatchewan", "Manitoba") &
           Crop == "Spring wheat" & Year >= 2020) |>
  group_by(Province) |>
  summarise(mean_yield = mean(Yield_bu_ac),
            n = n())
```

**(a)** In a comment: explain it step by step.

**(b)** Run it and report the output.

**(c)** In a comment: answer in plain language -- what does it say about prairie wheat since 2020?

<details><summary>Answer</summary>

- (a) Take the data; keep prairie-province spring wheat rows from 2020 on; split them by province; report each province's mean yield and row count.
- (b) Alberta **51.2** (n = 6), Manitoba **60.9** (n = 6), Saskatchewan **47.3** (n = 6).
- (c) Since 2020, Manitoba's spring wheat has out-yielded Alberta's by about ten bushels an acre and Saskatchewan's by about thirteen -- each figure an average over six years.
</details>

### Question 119

This pipeline was meant to compute each province's average spring wheat production, but it stops with `object 'Seeded_acres' not found`:

```r
crops |>
  filter(Crop == "Spring wheat") |>
  group_by(Province) |>
  summarise(mean_yield = mean(Yield_bu_ac)) |>
  mutate(total_bu = Seeded_acres * Yield_bu_ac)
```

**(a)** In a comment: apply the debugging habit -- run the pipeline one step at a time. After which step does `Seeded_acres` stop existing, and how can you tell?

**(b)** In a comment: identify the bug and state the general rule it breaks.

**(c)** Write the corrected pipeline (its last step should report each province's mean `total_bu`).

<details><summary>Answer</summary>

- (a) After the `summarise()` step -- run the pipeline up to each `|>` and look at the columns. Once `summarise()` runs, the data has collapsed to `Province` and `mean_yield`; `Seeded_acres` and `Yield_bu_ac` are gone, so the `mutate()` has nothing to work with.
- (b) `mutate()` builds per-row columns, so it must come *before* the collapse -- the order of pipeline steps matters because each step works on the previous step's output.
- (c)

```r
crops |>
  filter(Crop == "Spring wheat") |>
  mutate(total_bu = Seeded_acres * Yield_bu_ac) |>
  group_by(Province) |>
  summarise(mean_total_bu = mean(total_bu))
```
</details>

### Question 120

**(a)** In one pipeline: for Spring wheat across all years, compute each province's mean yield and count, sorted best-first, and save the result as `wheat_by_province`.

**(b)** Write the table to `output/wheat_by_province.csv`.

**(c)** Report the top row. In a comment: a reader sees Manitoba first and concludes "Manitoba grows Canada's wheat." Saskatchewan seeds about 8 million acres of spring wheat a year to Manitoba's 3 million -- say precisely what is wrong with that sentence.

<details><summary>Answer</summary>

```r
wheat_by_province <- crops |>
  filter(Crop == "Spring wheat") |>
  group_by(Province) |>
  summarise(mean_yield = mean(Yield_bu_ac),
            n = n()) |>
  arrange(desc(mean_yield))

write_csv(wheat_by_province, "output/wheat_by_province.csv")
```

- (c) Manitoba, **59.0 bu/ac** (n = 11). Manitoba grows wheat *best per acre*; Saskatchewan *grows Canada's wheat*, seeding nearly three times Manitoba's acres. Yield tables and scale tables answer different questions, and this table is only the first kind.
</details>
