# Module 2 - Test bank {.unnumbered}

```{r}
#| include: false
#| eval: true
# console() renders a code listing plus the console session it produces.
source("../_console.R")
```

This bank holds **40 questions**: ten in each section.


- Section 1 covers R basics: vectors, objects, and scripts
- Section 2 covers reading and inspecting data
- Section 3 covers data manipulation functions
- Section 4 covers pipelines and grouped summaries

There are two datasets used in the test bank.  The datasets are:

- [**Saskatchewan RM Crop Yields**](data/rm_yields_1990_2025.csv) (1990--2025; [canola-only file](data/rm_canola_yields_1990_2025.csv) for Section 2). Average yield for eight crops in every Rural Municipality, in long form. Lentils are recorded in pounds per acre; every other crop is in bushels per acre -- the `Unit` column says which.
- [**Manitoba Wheat Variety**](data/mb_wheat_reported_2020_2025.csv) (2020--2025). Red spring wheat yield by variety and municipality, one row per municipality-variety-year.

Your module test will contain one question per section.  The test will only use one of the datasets, and the dataset will be provided in the test.

For each question, 20% of the mark is for presentation and the other 80% is for the correctness of your script.

You submit a single R script: a header block at the top, a clearly labelled section for each question, and your sentence answers written as comments.  The script should run from top to bottom in a project folder that has the csvs in `data/`.  Here is an example of what a test will look like and the expectation for what your script will look like:

::: {.callout-note collapse="true"}
## A sample test, with a full-marks answer script

A real test looks like this -- one question from each section:

**Question 1**

*Five canola fields yielded 52.3, 47.8, 55.1, 44.6, and 50.9 bu/ac, on 160, 320, 240, 130, and 200 acres.*

**(a)** Create a vector `yields` and a vector `acres` holding these values.

**(b)** Create a vector `yields_t_ha` converting the yields to tonnes per hectare (multiply by 0.0560) in one line.

**(c)** Compute the mean and standard deviation of `yields`.

**(d)** Compute total production in bushels, and the farm's average yield weighted by acres. In a comment: why does the weighted average differ from the plain mean?

**Question 2**

*The Saskatchewan canola file contains one RM-year average yield per row, measured in bu/ac.*

**(a)** Compute the mean, median, and standard deviation of `Yield`.

**(b)** Compute its 90th percentile.

**(c)** In a comment: the mean sits a little above the median -- what does that direction of gap suggest about the shape of canola yields?

**Question 3**

*The full RM file holds all eight crops; lentil yields are recorded in lb/ac.*

**(a)** Filter to the lentil rows, saving the result as `lentils`. How many rows are there?

**(b)** Add a `yield_kg_ha` column (multiply by 1.12), saving back to `lentils`.

**(c)** Compute the mean of the new column and of the original column, and check their ratio. In a comment: why is the ratio the conversion factor?

**Question 4**

*Using the full eight-crop RM file.*

**(a)** In one pipeline, compute each RM's mean Spring Wheat yield across all years, sorted best-first.

**(b)** Report the top three RMs in a comment.

**(c)** In a comment: what did each row of the input represent, and what does each row of the result represent?

And here is a script that would earn full marks, including all of the presentation component -- with the console session it produces shown beneath it:

```{r}
#| eval: true
#| echo: false
console('
# ---
# Title: Module 2 test
# Author: Jordan Field
# Date: 2026-10-15
# Description:
#   Answers to the four test questions, one block per question.
#   Sentence answers are written as comments below each result.
# ---

# Load packages
library(tidyverse)

# ============================================================
# Question 1
# ============================================================

# (a) Yields (bu/ac) and field sizes (acres)
yields <- c(52.3, 47.8, 55.1, 44.6, 50.9)
acres  <- c(160, 320, 240, 130, 200)

# (b) Convert every yield to tonnes per hectare in one step
yields_t_ha <- yields * 0.0560
yields_t_ha

# (c) Centre and spread of the yields
mean(yields)   # 50.14 bu/ac
sd(yields)     # 4.06 bu/ac

# (d) Total production, then the average weighted by acres
total_bu <- sum(yields * acres)
total_bu                 # 52,866 bu
total_bu / sum(acres)    # 50.35 bu/ac
# (d) The weighted average differs from the plain mean because the
# fields differ in size, so each yield should count by its acres.

# ============================================================
# Question 2
# ============================================================

# Read the canola file using a relative path from the project folder
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")

# (a) Centre and spread of the canola yields
mean(rm_canola$Yield)      # 28.29 bu/ac
median(rm_canola$Yield)    # 26.9 bu/ac
sd(rm_canola$Yield)        # 10.06 bu/ac

# (b) 90th percentile
quantile(rm_canola$Yield, 0.9)   # 42.5 bu/ac

# (c) The mean sits a little above the median, which suggests mild
# right skew: the best RM-years pull the mean up more than the
# worst years pull it down.

# ============================================================
# Question 3
# ============================================================

# Read the full eight-crop file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Keep only the lentil rows
lentils <- filter(rm_yields, Crop == "Lentils")
nrow(lentils)    # 6,338 rows

# (b) Add the converted column (1 lb/ac is about 1.12 kg/ha) and
# overwrite lentils with the version that has the extra column
lentils <- mutate(lentils, yield_kg_ha = Yield * 1.12)

# (c) Means of the new and original columns
mean(lentils$yield_kg_ha)   # 1,353.4 kg/ha
mean(lentils$Yield)         # 1,208.39 lb/ac
# (c) Their ratio is 1.12, the conversion factor -- multiplying every
# value by a constant multiplies the mean by the same constant.

# ============================================================
# Question 4
# ============================================================

# (a) Mean spring wheat yield by RM, best first
rm_yields |>                              # take the data, THEN
  filter(Crop == "Spring Wheat") |>       # keep spring wheat, THEN
  group_by(RM) |>                         # split the rows by RM, THEN
  summarise(mean_yield = mean(Yield)) |>  # one row per RM: its mean, THEN
  arrange(desc(mean_yield))               # best RMs on top

# (b) Top three: RM 369 (47.6), RM 333 (47.2), RM 368 (47.2) bu/ac.
# (c) Each input row was one RM-year observation of spring wheat; each
# result row is one RM, its years collapsed into a single mean.
')
```
:::

Procedure for test:

- At least an hour before the test, "Schedule an appointment" in the test for that day.  
- Bring your (charged) laptop to the test.
- Your test will be accessible at 11:30am.
- When you get to the room:

1. Close all applications other than your webbrowser (used to access the test) and VS Code.
2. Log into zoom and join the breakout room with your name on it.
3. In Zoom, share your screen (not an application window, but your whole screen) and record your session.  
4. Navigate to Canvas and access your test.
5. Download the dataset that you need
6. Complete the test in VS Code.
7. Upload the completed script to Canvas (while still logged into zoom).
8. Leave zoom, and upload the screenshot recording to Canvas.
9. Your done!!

---

# Section 1 — R Basics

### Question 1
Five canola fields yielded 52.3, 47.8, 55.1, 44.6, and 50.9 bu/ac, on 160, 320, 240, 130, and 200 acres.

**(a)** Create a vector `yields` and a vector `acres` holding these values.

**(b)** Using one line of code, create a vector `yields_t_ha` converting the yields to tonnes per hectare (multiply by 0.0560) in one line.

**(c)** Compute the mean and standard deviation of `yields`.

**(d)** Compute total production in bushels (each field's yield times its acres, summed), and the farm's average yield weighted by acres (total bushels divided by total acres).

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Create the yield and acreage vectors
yields <- c(52.3, 47.8, 55.1, 44.6, 50.9)
acres  <- c(160, 320, 240, 130, 200)

# (b) Convert every yield to tonnes per hectare in one line
yields_t_ha <- yields * 0.0560

# Print the converted yields
yields_t_ha

# (c) Mean and standard deviation of the yields
mean(yields)
sd(yields)

# (d) Total production in bushels
total_bu <- sum(yields * acres)
total_bu

# Average yield weighted by acres
total_bu / sum(acres)

## I find the mean to be 50.14 bu/ac and the sd 4.06.
## Total production is 52,866 bu and the acre-weighted average 50.35 bu/ac.
')
```

</details>

### Question 2
Six spring wheat fields yielded 51.2, 48.7, 53.4, 8.2, 50.9, and 49.5 bu/ac; the 8.2 field was hailed out in July.

**(a)** Save the yields and compute the mean and the median.

**(b)** In a comment: the two disagree by nearly seven bushels. Which one better describes a typical field on this farm, and why?

**(c)** Compute the mean of the five unhailed fields (a second vector is fine) and compare it with the median from part a. In a comment: what does the closeness of those two numbers tell you about the median?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Yields, including the hailed-out field
yields <- c(51.2, 48.7, 53.4, 8.2, 50.9, 49.5)
mean(yields)
median(yields)


## I find the mean to be 43.65 and the median 50.2.
## (b) The median describes a typical field better: five of the six fields
## sit near 50, and the single 8.2 drags the mean seven bushels below all
## of them while barely touching the median.

# (c) The five unhailed fields
unhailed <- c(51.2, 48.7, 53.4, 50.9, 49.5)
mean(unhailed)

## (c) The unhailed mean is 50.74, almost exactly the full-data median --
## the median was already ignoring the extreme value.
')
```

</details>

### Question 3
Five barley fields yielded 68.2, 71.5, 64.9, 74.1, and 66.3 bu/ac. Five lentil fields yielded 1450, 1720, 1280, 1610, and 1390 lb/ac.

**(a)** Compute the mean, the standard deviation, the variance, and the range (maximum minus minimum) of each crop's yields.

**(b)** The lentil standard deviation is far larger. In a comment: why can the two standard deviations not be compared directly?

**(c)** Compute the coefficient of variation (sd divided by mean) for each crop. Which crop's yields are more variable *relative to their own average*?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Yield vectors for each crop
barley  <- c(68.2, 71.5, 64.9, 74.1, 66.3)
lentils <- c(1450, 1720, 1280, 1610, 1390)

# Mean of each crop
mean(barley)
mean(lentils)

# Standard deviation of each crop
sd(barley)
sd(lentils)

# Variance of each crop
var(barley)
var(lentils)

# Range (maximum minus minimum) of each crop
max(barley) - min(barley)
max(lentils) - min(lentils)

## I find barley mean 69.0, sd 3.77, variance 14.25, range 9.2; lentils
## mean 1490, sd 175.36, variance 30,750, range 440.

## (b) The two standard deviations are in different units (bushels vs pounds
## per acre) and on very different scales, so the raw spreads are not
## comparable.

# (c) Coefficient of variation (sd divided by mean) for each crop
sd(barley) / mean(barley)
sd(lentils) / mean(lentils)

## (c) CV barley 0.055, lentils 0.118 -- lentil yields are more variable
## relative to their own average.
')
```

</details>

### Question 4
Here is a script exactly as saved:

```r
mean(yields)
yields <- c(31.7, 28.4, 33.9, 30.2)
```

**(a)** In a comment: what would happen if you ran this script top to bottom in a fresh session, and why?

**(b)** Rewrite the script in working order, with a header block and a comment on each line.  

**(c)** Compute the standard deviation and median of yields.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Run as saved, the script fails on line 1: mean(yields) is asked for
# before yields exists in the fresh session, so R stops with
# R reports that the object yields was not found.

# (b) The working order, with a header
# ---
# This script computes the mean of a yield vector.
# ---

# Create the yield vector first
yields <- c(31.7, 28.4, 33.9, 30.2)

# Then compute its mean
mean(yields)

## I find the mean to be 31.05.

# (c) Standard deviation and median of yields
sd(yields)
median(yields)

')
```

</details>

### Question 5
Suppose there are four fields:

  - Field 1 had canola, 160 acres, and was irrigated.
  - Field 2 had wheat, 320 acres, and was not irrigated.
  - Field 3 had peas, 240 acres, and was not irrigated.
  - Field 4 had barley, 130 acres, and was irrigated.

**(a)** Create three vectors of length four, one for the crop grown, one for the seeded acres, and one for whether the field was irrigated.

**(b)** Combine your three vectors into a data frame called `trial` and print it in the console.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Crop grown: character, so the values are quoted
crop      <- c("Canola", "Wheat", "Peas", "Barley")

# Seeded acres: numeric, no quotes
acres     <- c(160, 320, 240, 130)

# Irrigated or not: logical, no quotes
irrigated <- c(TRUE, FALSE, FALSE, TRUE)

# (b) Combine the three vectors into a data frame
trial <- data.frame(crop = crop,
                    acres = acres,
                    irrigated = irrigated)

# Print the data frame
trial
')
```

</details>

---

### Question 6
The same five canola fields were measured in two years. 2024: 38.2, 41.6, 35.9, 43.1, 39.7 bu/ac. 2025: 45.1, 44.8, 42.3, 49.6, 46.2 bu/ac.

**(a)** Save the two vectors and compute each field's change in yield in one line.

**(b)** Compute the mean change, and the mean percentage change (change over the 2024 yield, times 100).

**(c)** In a comment: every field improved, but which measure -- bushels gained or percent gained -- would you use to compare this farm against a lentil farm, and why?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) The two years, and each field change in one line
y2024 <- c(38.2, 41.6, 35.9, 43.1, 39.7)
y2025 <- c(45.1, 44.8, 42.3, 49.6, 46.2)
change <- y2025 - y2024
change

# (b) Mean change and mean percentage change
mean(change)
mean(change / y2024 * 100)

## I find a mean change of 5.9 bu/ac and a mean percentage change of 15.0%.
## (c) The percentage change: bushels gained are not comparable across crops
## measured on different scales (a lentil farm gains hundreds of lb/ac), but
## percent gained puts both farms on the same footing.
')
```

</details>

### Question 7
A farm's four barley fields: field IDs `"B1"` to `"B4"`, sizes 155, 240, 160, and 310 acres, yields 66.8, 71.2, 63.5, and 69.9 bu/ac.

**(a)** Build a data frame `barley_fields` holding the three variables.

**(b)** Compute each field's production (acres times yield) and save it as a vector.

**(c)** Compute total production, and the farm's average yield weighted by acres (total production over total acres).

**(d)** In a comment: compare the weighted average with `mean(barley_fields$yield)` and say why they differ.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) The barley fields as a data frame
barley_fields <- data.frame(
  field_id = c("B1", "B2", "B3", "B4"),
  acres    = c(155, 240, 160, 310),
  yield    = c(66.8, 71.2, 63.5, 69.9)
)

# (b) Production of each field
production <- barley_fields$acres * barley_fields$yield
production

# (c) Total production and the acre-weighted average yield
sum(production)
sum(production) / sum(barley_fields$acres)

# (d) The plain mean, for comparison
mean(barley_fields$yield)

## I find total production of 59,271 bu and a weighted average of 68.52.
## (d) The plain mean is 67.85: it counts each field once, while the
## weighted average lets the 310-acre field count for its share of the
## land, and that field yields above average, pulling the weighted figure up.
')
```

</details>

### Question 8
**(a)** Build a data frame called `durum_fields` with columns `field_id` (`"D1"` to `"D4"`), `region` (South, South, Central, North), and `yield` (28.3, 41.6, 35.2, 44.8).

**(b)** Report its number of rows, number of columns, and column names using functions.

**(c)** Using `$`, compute the mean yield and the range (maximum minus minimum).

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Build the data frame
durum_fields <- data.frame(
  field_id = c("D1", "D2", "D3", "D4"),
  region = c("South", "South", "Central", "North"),
  yield = c(28.3, 41.6, 35.2, 44.8)
)

# (b) Number of rows
nrow(durum_fields)

# Number of columns
ncol(durum_fields)

# Column names
names(durum_fields)

# (c) Mean yield, using $
mean(durum_fields$yield)

# Range (maximum minus minimum)
max(durum_fields$yield) - min(durum_fields$yield)

## (b) 4 rows, 3 columns; the columns are field_id, region, yield.
## (c) I find the mean to be 37.48 (the console shows 37.475) and the
## range 16.5.
')
```

</details>


### Question 9
Twelve canola fields yielded 24.1, 31.5, 28.8, 35.2, 22.7, 30.4, 27.9, 33.6, 25.5, 29.3, 32.8, and 26.4 bu/ac.

**(a)** Save the values as a vector and compute the mean.

**(b)** Compute the 10th, 50th, and 90th percentiles in a single `quantile()` call.

**(c)** In a comment: one of your three percentiles should equal the median. Which one, and does it?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Canola yields as a vector (bu/ac)
canola <- c(24.1, 31.5, 28.8, 35.2, 22.7, 30.4, 27.9, 33.6, 25.5, 29.3, 32.8, 26.4)

# Mean yield
mean(canola)

# (b) 10th, 50th, and 90th percentiles in one call
quantile(canola, c(0.1, 0.5, 0.9))

# (c) The median, to check against the 50th percentile
median(canola)

## I find the mean to be 29.02.
## (b) 10th 24.24, 50th 29.05, 90th 33.52.
## (c) The 50th percentile is the median by definition; median(canola)
## confirms 29.05.
')
```

</details>



### Question 10
Four oat fields yielded 84.3, 91.7, 77.2, and 88.5 bu/ac, and oats are selling at $4.15 per bushel.

**(a)** Save the yields as a vector and the price as a single object.

**(b)** Compute a revenue-per-acre vector (yield times price) in one line, using the price object rather than retyping the number.

**(c)** Compute the mean revenue per acre.

**(d)** In a comment: the price rises to $4.60. What value would you change, and which lines would you rerun to update the results? Why is that easier than typing the price into each calculation?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Oat yields (bu/ac) and the price ($/bu)
yields <- c(84.3, 91.7, 77.2, 88.5)
price  <- 4.15

# (b) Revenue per acre, using the price object
revenue <- yields * price

# Print the revenues
revenue

# (c) Mean revenue per acre
mean(revenue)

## (b) The revenues are 349.85, 380.56, 320.38, 367.28 $/ac, rounded to the
## cent (the console shows 349.845 and so on).
## (c) Mean revenue is $354.51 per acre.
## (d) Change price <- 4.60, then rerun the lines below it. The price lives
## in one object, so one edit updates everything -- the same reason a
## spreadsheet keeps a price in one cell.
')
```

</details>

# Section 2 — Reading and Inspecting Data

## Saskatchewan RM Canola {.unnumbered}

### Question 11
*This question uses: [rm_canola_yields_1990_2025.csv](data/rm_canola_yields_1990_2025.csv)*

**(a)** Run `summary(rm_canola)`. From its output, report the minimum, median, and maximum of `Yield`, and the first and last `Year`.

**(b)** In a comment: are most yields in a plausible range for canola, and do the extremes need investigation? Are the years what the file name promises? Are there missing values?

**(c)** The minimum is below 2 bu/ac. In a comment: is a canola yield that low necessarily a data error?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the canola file
library(tidyverse)
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")

# (a) Summary of every column
summary(rm_canola)

## (a) I find Yield has minimum 1.9, median 26.9, and maximum 61.0, and Year runs from 1990 to 2025.
## (b) Most yields are plausible for canola, but the values near 2 bu/ac
## deserve investigation. The years run 1990--2025 and there are no NAs.
## (c) Not necessarily. A severe drought or widespread crop failure could
## produce a very low RM average.
')
```

</details>


### Question 12
*This question uses: [rm_canola_yields_1990_2025.csv](data/rm_canola_yields_1990_2025.csv)*

**(a)** Compute the mean, median, standard deviation, and variance of `rm_canola$Yield`.

**(b)** Compute the 90th percentile of `Yield`.

**(c)** In a comment: what does that difference between the mean and median suggest about the shape of canola yields?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the canola file
library(tidyverse)
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")

# (a) Centre of the yields
mean(rm_canola$Yield)
median(rm_canola$Yield)

# Spread of the yields
sd(rm_canola$Yield)
var(rm_canola$Yield)

# (b) 90th percentile of the yields
quantile(rm_canola$Yield, probs = 0.9)

## (a) I find a mean of 28.29, a median of 26.9, a standard deviation of 10.06, and a variance of 101.21 -- the sd squared, in squared units.
## (b) The 90th percentile is 42.5 bu/ac.
## (c) A mean above the median suggests mild right skew: the very best RM-years pull the mean up more than the worst years pull it down.
')
```

</details>

### Question 13
*This question uses: [rm_canola_yields_1990_2025.csv](data/rm_canola_yields_1990_2025.csv)*

**(a)** Print the first six rows of `rm_canola` with `head()`.

**(b)** Report the number of rows in the data, and compute the first and last year in the data.

**(c)** There are 295 RMs in Saskatchewan.  Divide the row count by the number of years.  What does this tell you about the coverage of the dataset?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) First six rows
head(rm_canola)

# (b) Rows, and the span of years
nrow(rm_canola)
range(rm_canola$Year)

## (b) I find 10,039 rows spanning 1990 to 2025.

# (c) Rows per year
nrow(rm_canola) / 36

## (c) 10,039 / 36 is about 279 rows per year, against the 295 RMs --
## so not every RM reports canola every year; some RM-year combinations
## are absent from the file.
')
```

</details>

### Question 14
*This question uses: [rm_canola_yields_1990_2025.csv](data/rm_canola_yields_1990_2025.csv)*

**(a)** Using `$`, compute the mean, the standard deviation, and the 10th and 90th percentiles of `Yield` across the whole file.

**(b)** In a comment: an RM-year at the 10th percentile grew how much canola? Interpret the number for a grower.

**(c)** Compute the gap between the 90th and 10th percentiles. In a comment: what does this gap describe that the standard deviation also tries to describe?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Mean, sd, and the 10th and 90th percentiles
mean(rm_canola$Yield)
sd(rm_canola$Yield)
quantile(rm_canola$Yield, probs = c(0.10, 0.90))

## I find a mean of 28.29, an sd of 10.06, and percentiles of 16.1 and 42.5.

## (b) An RM-year at the 10th percentile grew 16.1 bu/ac -- nine in ten
## RM-years across the file were at or above that value.

# (c) The 90-10 gap
42.5 - 16.1

## (c) The gap is 26.4 bu/ac. Like the sd it describes how spread out
## yields are, but it reads directly as a bushel distance holding the
## middle 80% of RM-years, and it ignores the extremes entirely.
')
```

</details>


### Question 15
*This question uses: [rm_canola_yields_1990_2025.csv](data/rm_canola_yields_1990_2025.csv)*

**(a)** Compute the 25th, 50th, and 75th percentiles of `Yield` in one `quantile()` call.

**(b)** Compute the IQR twice: with `IQR()`, and as the difference between two of your percentiles. Confirm the two match.

**(c)** In a comment: which RM-years fall inside the span the IQR measures?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the canola file
library(tidyverse)
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")

# (a) The three quartiles in one call
quantile(rm_canola$Yield, probs = c(0.25, 0.50, 0.75))

# (b) The IQR, from the function
IQR(rm_canola$Yield)

# The IQR, as the 75th minus the 25th percentile
quantile(rm_canola$Yield, probs = 0.75) - quantile(rm_canola$Yield, probs = 0.25)

## I find quartiles of 21.0, 26.9, and 35.4 bu/ac, and an IQR of 14.4
## both ways.
## (c) The middle half of RM-years: everything above the bottom quarter
## and below the top quarter of yields.
')
```

</details>

## Manitoba Wheat Variety {.unnumbered}

### Question 16
*This question uses: [mb_wheat_reported_2020_2025.csv](data/mb_wheat_reported_2020_2025.csv)*

**(a)** Run `summary(mb_wheat)`. From its output, report the minimum, median, and maximum of `Yield_bu_ac`, and the first and last `Year`.

**(b)** In a comment: are most yields in a plausible range for spring wheat, and do the extremes need investigation? Are there missing values?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the wheat file
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) Summary of every column
summary(mb_wheat)

## (a) I find Yield_bu_ac has minimum 4.5, median 62.2, and maximum 97.8, and Year runs from 2020 to 2025.
## (b) Most yields are plausible for Manitoba spring wheat, but the lowest
## values deserve investigation. There are no missing values in this file.
')
```

</details>

### Question 17
*This question uses: [mb_wheat_reported_2020_2025.csv](data/mb_wheat_reported_2020_2025.csv)*

**(a)** Compute the mean, median, standard deviation, and variance of `mb_wheat$Yield_bu_ac`.

**(b)** In a comment: what does that difference between the mean and median suggest about the data?

**(c)** Compute the smallest and largest yields in the data.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the wheat file
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) Centre of the yields
mean(mb_wheat$Yield_bu_ac)
median(mb_wheat$Yield_bu_ac)

# Spread of the yields
sd(mb_wheat$Yield_bu_ac)
var(mb_wheat$Yield_bu_ac)

## (a) I find a mean of 61.16, a median of 62.2, a standard deviation of 12.38, and a variance of 153.34.

## (b) A mean below the median suggests mild left skew: unusually low values pull the mean down more than the high values pull it up -- a tail of very poor variety-municipality observations mixed in with the normal ones.

# (c) Smallest and largest yield
range(mb_wheat$Yield_bu_ac)

## (c) range() returns both ends as one vector: the smallest yield is 4.5 and the largest is 97.8 bu/ac.
')
```

</details>


### Question 18
*This question uses: [mb_wheat_reported_2020_2025.csv](data/mb_wheat_reported_2020_2025.csv)*

**(a)** Compute the 25th, 50th, and 75th percentiles of `Yield_bu_ac` in one `quantile()` call.

**(b)** Compute the IQR as the difference between two of your percentiles.

**(c)** Compare the IQR to the range. Which might give a better sense of the spread of the data?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the wheat file
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) The three quartiles in one call
quantile(mb_wheat$Yield_bu_ac, probs = c(0.25, 0.50, 0.75))

# (b) The IQR, as the 75th minus the 25th percentile
quantile(mb_wheat$Yield_bu_ac, probs = 0.75) - quantile(mb_wheat$Yield_bu_ac, probs = 0.25)

## I find quartiles of 53.9, 62.2, and 69.6 bu/ac, and an IQR of 15.7.

# (c) The range, for comparison
max(mb_wheat$Yield_bu_ac) - min(mb_wheat$Yield_bu_ac)

## (c) The range is 93.3 bu/ac against an IQR of 15.7. The IQR usually
## gives the better sense of spread: the range rests on the single best
## and single worst results, while the IQR is the width of the middle
## half of the data.
')
```

</details>

### Question 19
*This question uses: [mb_wheat_reported_2020_2025.csv](data/mb_wheat_reported_2020_2025.csv)*

**(a)** Compute the 5th and 95th percentiles of `Yield_bu_ac` in one call.

**(b)** In a comment: interpret the 95th percentile for a grower.

**(c)** Compute the gap between the two percentiles. In a comment: what fraction of all results does this span hold?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the wheat file
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) The 5th and 95th percentiles in one call
quantile(mb_wheat$Yield_bu_ac, probs = c(0.05, 0.95))

## (a) I find percentiles of 38.8 and 78.8 bu/ac.

## (b) A result at the 95th percentile yielded 78.8 bu/ac -- only one in
## twenty municipality-variety results beat that.

# (c) The gap between them
quantile(mb_wheat$Yield_bu_ac, probs = 0.95) - quantile(mb_wheat$Yield_bu_ac, probs = 0.05)

## (c) The gap is about 40 bu/ac, and it holds the middle 90% of results.
')
```

</details>

### Question 20
*This question uses: [mb_wheat_reported_2020_2025.csv](data/mb_wheat_reported_2020_2025.csv)*

**(a)** Compute the range of `Yield_bu_ac` (largest minus smallest) and the IQR.

**(b)** In a comment: the two numbers disagree by a lot. What does each one describe?

**(c)** In a comment: which of the two measures would change if we discovered that the highest value in the data was actually a data error?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the wheat file
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) The range
max(mb_wheat$Yield_bu_ac) - min(mb_wheat$Yield_bu_ac)

# The IQR
IQR(mb_wheat$Yield_bu_ac)

## I find a range of 93.3 bu/ac against an IQR of 15.7.
## (b) The range is the distance between the single best and single worst
## results; the IQR is the width of the middle half.
## (c) Only the range. The 97.8 is the maximum, so removing it moves the
## range directly, while the quartiles sit deep in the middle of 2,398
## results and barely notice.
')
```

</details>



# Section 3 — Data Manipulation Functions

## Saskatchewan RM Crop Yields {.unnumbered}

### Question 21
*This question uses: [rm_yields_1990_2025.csv](data/rm_yields_1990_2025.csv)*

**(a)** Filter to the Spring Wheat observations of at least 60 bu/ac (all years) and saving the result. How many rows are in the filtered data frame?

**(b)** Filter again to keep only the 2023 rows of that result. How many rows are in this second data frame?

**(c)** From your 2023 data frame, select `RM` and `Yield` and sort the data by yield (greatest to least). Which RM tops the list, at what yield?


<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Spring Wheat at 60 bu/ac or better, all years
sw_60 <- filter(rm_yields, Crop == "Spring Wheat" & Yield >= 60)
# Row count of the filtered data frame
nrow(sw_60)

## (a) I find 434 rows at 60 bu/ac or better.

# (b) Keep only the 2023 rows of that result
sw_60_2023 <- filter(sw_60, Year == 2023)
# Row count of the second data frame
nrow(sw_60_2023)

## (b) 53 of them are from 2023.

# (c) Just RM and Yield, sorted best first
sw_best <- select(sw_60_2023, RM, Yield)
arrange(sw_best, desc(Yield))

## (c) RM 187 tops the 2023 list, at 74.7 bu/ac.
')
```

</details>

### Question 22
*This question uses: [rm_yields_1990_2025.csv](data/rm_yields_1990_2025.csv)*

**(a)** Filter to rows where the crop is Oats **or** Barley, using `|`, saving the result. How many rows are in the filtered data frame?

**(b)** Write the same filter using `%in%` and confirm the row count matches.

**(c)** What would `filter(rm_yields, Crop == "Oats" & Crop == "Barley")` return? In a comment: explain why.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Oats or Barley, using the | connector
oats_barley <- filter(rm_yields, Crop == "Oats" | Crop == "Barley")
# Row count of the filtered data frame
nrow(oats_barley)

# (b) The same filter written with %in%
oats_barley2 <- filter(rm_yields, Crop %in% c("Oats", "Barley"))
# The count should match (a)
nrow(oats_barley2)

## I find 19,709 rows both ways.

## (c) It returns zero rows. The & version asks for rows whose crop is
## Oats and Barley at the same time, and no single row can satisfy both
## conditions. "Or" is the right connector, written |.
')
```

</details>

### Question 23
*This question uses: [rm_yields_1990_2025.csv](data/rm_yields_1990_2025.csv)*

**(a)** Filter to Canola in 2024, then in one `mutate()` call add two columns: `yield_t_ha` (multiply by 0.0560) and `yield_kg_ha` (multiply by 56.0). Save the result of each step.

**(b)** Compute the mean of each new column.

**(c)** In a comment: without running anything, what is the mean of `yield_kg_ha` divided by the mean of `yield_t_ha`, and why?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Canola in 2024
canola_2024 <- filter(rm_yields, Crop == "Canola" & Year == 2024)
# Add both converted columns in one mutate() call
canola_2024 <- mutate(canola_2024,
                      yield_t_ha  = Yield * 0.0560,  # tonnes per hectare
                      yield_kg_ha = Yield * 56.0)    # kilograms per hectare

# (b) Mean of each new column
mean(canola_2024$yield_t_ha)
mean(canola_2024$yield_kg_ha)

## (b) I find means of 1.76 t/ha and 1764.11 kg/ha.

## (c) The ratio of the means is 1000: both columns are the same yields scaled by constants, and 56.0 / 0.0560 = 1000 (a tonne is 1000 kg).
')
```

</details>

### Question 24
*This question uses: [rm_yields_1990_2025.csv](data/rm_yields_1990_2025.csv)*

Canola is selling at $14.20 per bushel.

**(a)** Filter to Canola in 2025 and, in one `mutate()`, add a `revenue_per_ac` column (yield times price). Save the result.

**(b)** Sort by revenue (greatest to least), and report the top RM and its revenue per acre.

**(c)** Compute the mean revenue per acre.


<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# Price object
price <- 14.20

# (a) 2025 canola with revenue per acre
canola_25 <- filter(rm_yields, Crop == "Canola" & Year == 2025)
canola_25 <- mutate(canola_25, revenue_per_ac = Yield * price)

# (b) Best revenue first
arrange(canola_25, desc(revenue_per_ac))

## (b) I find RM 287 on top at $866.20 per acre.

# (c) Mean revenue per acre
mean(canola_25$revenue_per_ac)

## (c) The mean is $623.94 per acre.
')
```

</details>

### Question 25
*This question uses: [rm_yields_1990_2025.csv](data/rm_yields_1990_2025.csv)*

**(a)** Filter `rm_yields` to Canola in 2024, saving the result as `canola_2024`. How many rows are in the filtered data frame?

**(b)** From `canola_2024`, select only the `RM` and `Yield` columns, saving the result. Report its column names and its number of columns.

**(c)** In a comment: after (a) and (b), how many rows does `rm_yields` itself have, and why?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Keep only the Canola rows from 2024
canola_2024 <- filter(rm_yields, Crop == "Canola" & Year == 2024)
# Row count of the filtered data frame
nrow(canola_2024)

# (b) Keep only the RM and Yield columns
canola_slim <- select(canola_2024, RM, Yield)
# Column names of the slimmed version
names(canola_slim)
# Number of columns
ncol(canola_slim)

# (c) The original data frame, after (a) and (b)
nrow(rm_yields)

## I find 293 rows of 2024 canola.
## The selected version has the columns RM and Yield, so 2 columns (and still 293 rows).
## rm_yields itself still has 71104 rows: filter() and select() return new data frames and leave the original untouched.
')
```

</details>


## Manitoba Wheat Variety {.unnumbered}

### Question 26
*This question uses: [mb_wheat_reported_2020_2025.csv](data/mb_wheat_reported_2020_2025.csv)*

**(a)** Filter to rows with at least 10 farms **and** a yield above 70 bu/ac, saving the result. How many rows are in it?

**(b)** Sort by yields (greatest to least) and report the top municipality, variety, and year.

**(c)** In a comment: compare this with filtering on yield alone without the minimum number of farms.  Comment on the trustworthiness of these results versus those in part b. 

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) At least 10 farms AND yield above 70
solid_high <- filter(mb_wheat, Farms >= 10 & Yield_bu_ac > 70)
nrow(solid_high)

# (b) Best first
arrange(solid_high, desc(Yield_bu_ac))

## I find 273 rows; LOUISE growing SY MANNESS in 2025 tops the list
## at 96.3 bu/ac.

# (c) The same filter on yield alone
high_only <- filter(mb_wheat, Yield_bu_ac > 70)
nrow(high_only)

## (c) Filtering on yield alone keeps 569 rows, and its top result is a
## 97.8 bu/ac average from just 5 farms. An average from a few farms is
## more sensitive to one unusual result, so the part b list is the more
## trustworthy ranking; the farm minimum gives every row a broader base.
')
```

</details>


### Question 27
*This question uses: [mb_wheat_reported_2020_2025.csv](data/mb_wheat_reported_2020_2025.csv)*

**(a)** Filter the data to the 2025 rows, saving the result. How many rows are in the filtered data frame?

**(b)** From the filtered data, select `Municipality`, `Variety`, and `Yield_bu_ac`, saving the result. Report its column names.

**(c)** In a comment: how many rows does the original data frame still have, and what happened to the other years' rows?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) Keep the 2025 rows only
wheat_2025 <- filter(mb_wheat, Year == 2025)
# Row count of the filtered data frame
nrow(wheat_2025)

# (b) Three columns from the 2025 rows
wheat_slim <- select(wheat_2025, Municipality, Variety, Yield_bu_ac)
# Its column names
names(wheat_slim)

# (c) The original data frame, after (a) and (b)
nrow(mb_wheat)

## I find 438 rows for 2025.
## The selected columns are Municipality, Variety, Yield_bu_ac.
## mb_wheat itself still has 2398 rows -- filtering returns a new data frame. The other years\' rows are simply absent from the filtered copy; they were not deleted from anything.
')
```

</details>

### Question 28
*This question uses: [mb_wheat_reported_2020_2025.csv](data/mb_wheat_reported_2020_2025.csv)*

**(a)** Filter the data to the varieties `AAC BRANDON (BW 932)` **or** `AAC STARBUCK <SECAN>` using `|`, saving the result. How many rows are in the filtered data frame?

**(b)** Write the same filter using `%in%` and confirm the count matches.

**(c)** If in part (a) you used `&` instead of `|`, what would the result be?  Explain.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) The two varieties, using the | connector
two_var <- filter(mb_wheat, Variety == "AAC BRANDON (BW 932)" | Variety == "AAC STARBUCK <SECAN>")
# Row count of the filtered data frame
nrow(two_var)

# (b) The same filter written with %in%
two_var2 <- filter(mb_wheat, Variety %in% c("AAC BRANDON (BW 932)", "AAC STARBUCK <SECAN>"))
# The count should match (a)
nrow(two_var2)

## I find 893 rows both ways.
## (c) It would return zero rows. The & version demands both conditions true of the same row, and no single row has a variety that is both names at once. "Or" is the right connector, written |.
')
```

</details>

### Question 29
*This question uses: [mb_wheat_reported_2020_2025.csv](data/mb_wheat_reported_2020_2025.csv)*

**(a)** Use one `mutate()` to add `acres_per_farm` (acres divided by farms), saving the result.

**(b)** Sort by `acres_per_farm`, largest first, and report the top municipality and variety.

**(c)** Compute the mean of `acres_per_farm`. In a comment, explain what this number means in the context of the data.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) Acres per farm
mb_wheat <- mutate(mb_wheat, acres_per_farm = Acres / Farms)

# (b) Largest first
arrange(mb_wheat, desc(acres_per_farm))

## (b) I find MOUNTAIN growing SY MANNESS on top, at 1,587 acres per farm.

# (c) The average across rows
mean(mb_wheat$acres_per_farm)

## (c) The mean is 370: in a typical municipality-variety-year result,
## each farm growing that variety seeded about 370 acres of it. It
## measures how concentrated a variety is in the hands of its growers --
## a different fact from how much land it covers in total.
')
```

</details>

### Question 30
*This question uses: [mb_wheat_reported_2020_2025.csv](data/mb_wheat_reported_2020_2025.csv)*

**(a)** Use one `mutate()` to add `total_bu`, each variety's total production in a municipality in bushels (acres times yield), saving the result.

**(b)** Sort by `total_bu`, largest first, and report the top municipality and variety.

**(c)** In a comment: why is the highest-yielding observation not the one with the greatest production?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) Total production: acres times the per-acre yield
mb_wheat <- mutate(mb_wheat, total_bu = Acres * Yield_bu_ac)

# (b) Largest first
arrange(mb_wheat, desc(total_bu))

## (b) I find PORTAGE LA PRAIRIE growing AAC BRANDON (BW 932) in 2020 on
## top, at about 5.1 million bushels.
## (c) Acreage. The top producer yielded a modest 68.8 bu/ac but on
## 74,604 acres, while the highest-yielding result (97.8 bu/ac) sat on
## under 2,000 acres. Total production rewards land as much as yield.
')
```

</details>

# Section 4 — Pipelines and Grouped Summaries

## Saskatchewan RM Crop Yields {.unnumbered}

### Question 31
*This question uses: [rm_yields_1990_2025.csv](data/rm_yields_1990_2025.csv)*

**(a)** In one pipeline: for **2025 only**, compute each crop's mean yield and observation count, sorted from highest mean to lowest, and save the result as `crop_summary_2025`.

**(b)** Write the table to `output/crop_summary_2025.csv`.

**(c)** Report the two highest-yielding crops in the sorted table. In a comment: check the Unit column -- why is the first-place crop not comparable to the rest, and which bushel crop is the real standout?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Each crop in 2025: mean yield and observation count, best first
crop_summary_2025 <- rm_yields |>
  filter(Year == 2025) |>                      # 2025 only
  group_by(Crop) |>                            # one group per crop
  summarise(mean_yield = mean(Yield),          # average yield
            n = n()) |>                        # observation count
  arrange(desc(mean_yield))                    # highest mean first

# Print the sorted table
crop_summary_2025

# (b) Write the table to a csv
write_csv(crop_summary_2025, "output/crop_summary_2025.csv")

## I find Lentils on top at 1897 and Oats second at 97.5.
## (c) The Unit column shows Lentils are recorded in pounds per acre, not
## bushels, so the first row is not comparable to the rest. Among the
## bushel crops, Oats are highest at 97.5 bu/ac.
')
```

</details>

### Question 32
*This question uses: [rm_yields_1990_2025.csv](data/rm_yields_1990_2025.csv)*

**(a)** In one pipeline: filter to Lentils from 2021 through 2025, convert from lb/acre to kg/ha with `mutate()` (multiply by 1.12), and compute the mean converted yield and `n()` for each year. Sort from highest mean to lowest and save the result as `lentil_summary`.

**(b)** Report the highest and lowest annual means and the counts behind them.

**(c)** In a comment: in what order do your steps run, and why would putting the `summarise()` before the `mutate()` fail?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Recent lentils, converted to kg/ha and summarised by year
lentil_summary <- rm_yields |>
  filter(Crop == "Lentils" & Year >= 2021) |>    # recent lentil rows
  mutate(yield_kg_ha = Yield * 1.12) |>          # convert lb/ac to kg/ha
  group_by(Year) |>                               # one group per year
  summarise(mean_kg_ha = mean(yield_kg_ha),       # annual average
            n = n()) |>                           # observations behind it
  arrange(desc(mean_kg_ha))                       # highest mean first

# Print the summary table
lentil_summary

## (b) The highest annual mean is 2025 at 2,124.4 kg/ha from 218
## observations. The lowest is 2021 at 1,057.5 kg/ha from 194.
## (c) The steps filter, convert, group, summarise, and sort in that order.
## Putting summarise() before mutate() would fail because yield_kg_ha would
## not exist when the summary tried to use it.
')
```

</details>

### Question 33
*This question uses: [rm_yields_1990_2025.csv](data/rm_yields_1990_2025.csv)*

Canola is selling at $14.20 per bushel.

**(a)** In one pipeline: filter to Canola observations in 2023-2025, add `revenue_per_ac` with `mutate()`, then compute the mean revenue for each year.

**(b)** Report the three yearly means.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# Price object
price <- 14.20

# (a) Mean canola revenue per acre by year, 2023-2025
rm_yields |>
  filter(Crop == "Canola" & Year >= 2023) |>      # recent canola rows
  mutate(revenue_per_ac = Yield * price) |>       # dollars per acre
  group_by(Year) |>                               # one group per year
  summarise(mean_revenue = mean(revenue_per_ac))  # average within each year

## (b) I find mean revenues of $481.11 in 2023, $447.33 in 2024, and
## $623.94 in 2025.
')
```

</details>

### Question 34
*This question uses: [rm_yields_1990_2025.csv](data/rm_yields_1990_2025.csv)*

**(a)** In one pipeline, compute the mean canola yield for **each year** from 2021 to 2025 (filter first, then group).

**(b)** 2021 was a severe drought year in Saskatchewan. Does the data agree?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Mean canola yield for each year from 2021 to 2025
rm_yields |>
  filter(Crop == "Canola" & Year >= 2021) |>   # canola, 2021 onward
  group_by(Year) |>                            # one group per year
  summarise(mean_yield = mean(Yield))          # average within each year

## (b) Yes, the data agrees: the 2021 mean sits roughly a third below the surrounding years.
')
```

</details>

### Question 35
*This question uses: [rm_yields_1990_2025.csv](data/rm_yields_1990_2025.csv)*

**(a)** In one pipeline: filter to Flax from 2020 through 2025, compute each RM's `mean_yield` and number of observations, retain RMs with at least four observations, and sort from highest mean to lowest. Save the result as `flax_summary`.

**(b)** Report the top RM, its mean, and its count.

**(c)** In a comment: why does the minimum-count rule improve the ranking, and why must `filter(n >= 4)` come after `summarise()`?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Recent flax means by RM, retaining RMs with at least four reports
flax_summary <- rm_yields |>
  filter(Crop == "Flax" & Year >= 2020) |>    # recent flax rows
  group_by(RM) |>                              # one group per RM
  summarise(mean_yield = mean(Yield),          # average within the RM
            n = n()) |>                        # observations behind it
  filter(n >= 4) |>                            # enough years to compare
  arrange(desc(mean_yield))                    # highest mean first

# Print the summary table
flax_summary

## (b) RM 336 is highest at 34.5 bu/ac from 4 observations.
## (c) The rule removes rankings based on only one or two years. The count
## n is created by summarise(), so it cannot be filtered before that step.
')
```

</details>

## Manitoba Wheat Variety {.unnumbered}

### Question 36
*This question uses: [mb_wheat_reported_2020_2025.csv](data/mb_wheat_reported_2020_2025.csv)*

**(a)** In one pipeline, compute each variety's total acres (`sum(Acres)`), sorted greatest to least.  

**(b)** In a comment: totalling acres makes sense, but what would totalling `Yield_bu_ac` produce, and why is that number meaningless?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) Total acres by variety, largest first
mb_wheat |>
  group_by(Variety) |>                    # one group per variety
  summarise(total_acres = sum(Acres)) |>  # acres are amounts, so they add
  arrange(desc(total_acres))              # biggest variety first


## (b) Totalling Yield_bu_ac would add per-acre rates across rows -- a
## number with no physical meaning, since a rate says how much each acre
## gave, not how much there was. Acres are an amount, so their total is a
## real quantity of land.
')
```

</details>

### Question 37
*This question uses: [mb_wheat_reported_2020_2025.csv](data/mb_wheat_reported_2020_2025.csv)*

**(a)** In one pipeline, compute each variety's mean yield and observation count, sorted by mean yield from greatest to least.

**(b)** Report the top two varieties, their means, and their counts.

**(c)** In a comment: Report the count of the top three varieties.  Do you find the average yield of each variety trustworthy?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) Each variety: mean yield and observation count, best first
mb_wheat |>
  group_by(Variety) |>                         # one group per variety
  summarise(mean_yield = mean(Yield_bu_ac),    # average yield
            n = n()) |>                        # observation count
  arrange(desc(mean_yield))                    # highest mean first

## (b) I find AAC WESTKING first at 79.2 bu/ac from 17 observations, then CDC GO (BW781) at 77.5 from a single observation.
## (c) The top three counts are 17 (AAC WESTKING), 1 (CDC GO), and
## 2 (AAC SPIKE). I would trust the AAC WESTKING mean: seventeen
## municipality-year results is a reasonable base. The other two rest on
## one or two results, so their averages could be one municipality having
## a good year.
')
```

</details>

### Question 38
*This question uses: [mb_wheat_reported_2020_2025.csv](data/mb_wheat_reported_2020_2025.csv)*

**(a)** In one pipeline: for **2025 only**, compute each variety's mean reported yield and observation count, sorted best-first, and save the result as `variety_summary_2025`.

**(b)** Write the table to `output/variety_summary_2025.csv`.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) Each variety in 2025: mean reported yield and count, best first
variety_summary_2025 <- mb_wheat |>
  filter(Year == 2025) |>                      # 2025 only
  group_by(Variety) |>                         # one group per variety
  summarise(mean_yield = mean(Yield_bu_ac),    # average yield
            n = n()) |>                        # observation count
  arrange(desc(mean_yield))                    # highest mean first

# Print the top of the sorted table
head(variety_summary_2025)

# (b) Write the table to a csv
write_csv(variety_summary_2025, "output/variety_summary_2025.csv")

')
```

</details>

### Question 39
*This question uses: [mb_wheat_reported_2020_2025.csv](data/mb_wheat_reported_2020_2025.csv)*

**(a)** In one pipeline, compute each municipality's mean yield and observation count, sorted from highest mean to lowest.

**(b)** Report the top municipality and how many observations back its mean.

**(c)** In a comment: a municipality mean pools every variety grown there. What two different things could make one municipality top this table?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) Mean yield and count by municipality, best first
mb_wheat |>
  group_by(Municipality) |>                 # one group per municipality
  summarise(mean_yield = mean(Yield_bu_ac), # average across its rows
            n = n()) |>                     # rows behind the average
  arrange(desc(mean_yield))                 # best municipality first

## (b) I find LOUISE on top at 72.86 bu/ac, backed by 26 observations.
## (c) Two different things could put a municipality here: genuinely better
## growing conditions, or a mix that happens to lean on high-yielding
## varieties. The mean pools varieties, so the ranking cannot tell the
## land from the lineup.
')
```

</details>

### Question 40
*This question uses: [mb_wheat_reported_2020_2025.csv](data/mb_wheat_reported_2020_2025.csv)*

**(a)** In one pipeline: filter the data to 2023, compute each variety's mean yield and count, sort best-first, and save the result.

**(b)** Report the top variety by mean and its count, and the largest count in the table and its variety.

**(c)** In a comment: are they the same variety? What do the two columns measure differently?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) 2023 variety means and counts, best first
variety_2023 <- mb_wheat |>
  filter(Year == 2023) |>                   # one year only
  group_by(Variety) |>                      # one group per variety
  summarise(mean_yield = mean(Yield_bu_ac), # average within the variety
            n = n()) |>                     # results behind it
  arrange(desc(mean_yield))                 # best mean first
variety_2023

## (b) I find AC BARRIE (BW 661) with the best 2023 mean, 67.3 bu/ac --
## from a single observation. The largest count is AAC BRANDON (BW 932),
## with 84.
## (c) They are not the same variety. The mean measures how well a variety
## did where it was grown; the count measures how widely it was grown. A
## chart-topping mean on n = 1 is one municipality having a good year.
')
```

</details>

