# Module 4 - Test bank {.unnumbered}

```{r}
#| include: false
#| eval: true
# console() renders a code listing plus the console session it produces.
source("../_console.R")
suppressMessages(library(tidyverse))
suppressMessages(library(janitor))
```

This bank holds **15 questions**: five in each section. Each section will grow to ten, so the question numbers leave room (Section 2 starts at Question 11, Section 3 at Question 21).

- Section 1 covers graphing principles
- Section 2 covers graphing in Excel
- Section 3 covers graphing in R

The bank uses the same datasets as the Module 1 bank:

- [**Saskatchewan RM Crop Yields**](data/rm_yields_1990plus.csv) (1990–2025). Average yield for eight crops in every Rural Municipality, one column per crop.
- [**Saskatchewan RM Crop Yields (Long)**](data/rm_yields_1990_2025.csv) (1990–2025). The same yields in long format: one row per RM-year-crop, with a `Unit` column. Yields are in bu/ac except lentils, which are in lb/ac.
- [**Manitoba Wheat Variety**](data/mb_wheat_reported_2020_2025.csv) (2020–2025). Red spring wheat by variety and municipality, with the number of farms, seeded acres and average yield.

Your module test will contain one question per section. The datasets you need will be provided in the test.

Answer Sections 1 and 2 in your workbook: one worksheet per question, labelled with the question number, with written answers typed into cells so they are visible when the workbook is opened. Answer Section 3 with an R script that has a header block, a clearly labelled section for each question, and your sentence answers written as comments, and it should run from top to bottom in a fresh session.

For each question, 20% of the mark is for presentation and the other 80% is for the correctness of your charts, workbook or script.

---

# Section 1 — Graphing principles

*These questions are about choosing and judging charts. Write your answers in worksheet cells. Where a question asks for a calculation, do it with a formula in a cell so the TA can see it.*

### Question 1

*In 2025 the provincial average spring wheat yield in Saskatchewan was 53.0 bu/ac and the average durum yield was 48.6 bu/ac.*

**(a)** Would you present these two numbers as a chart or as a sentence? Explain your choice.

**(b)** Describe a version of this comparison (more crops, more years, or more places) where a chart becomes the better choice, and name the chart type you would use.

<details><summary>Answer</summary>

**(a)** A sentence. Two values are easier to state than to plot: "spring wheat outyielded durum by about 4 bu/ac in 2025." A chart of two bars adds ink without adding understanding.

**(b)** Any answer that grows the comparison until a sentence can no longer hold it: the 2025 average yields of all eight crops (a bar chart), or wheat and durum yields every year from 1990 to 2025 (a line chart), or yields across the RMs (a map or a distribution).

</details>

### Question 2

*The table contains 2025 provincial average yields in bushels per acre. Copy it into a blank sheet.*

| Crop | Yield (bu/ac) |
|---|---:|
| Spring Wheat | 53.0 |
| Durum | 48.6 |
| Canola | 43.9 |
| Peas | 43.5 |
| Flax | 30.5 |

Suppose the yields are drawn as a bar chart whose value axis starts at 30 rather than at zero.

**(a)** Calculate the drawn length of the canola bar as a fraction of the spring wheat bar under each axis choice (starting at 30, and starting at zero).

**(b)** In a sentence: what would a reader who only glances at the truncated chart conclude about canola relative to spring wheat?

**(c)** State the rule for where the value axis of a bar chart should start, and why bar charts in particular need it.

<details><summary>Answer</summary>

**(a)** Starting at 30, the canola bar is (43.9 − 30) / (53.0 − 30) = **0.60** of the spring wheat bar. Starting at zero it is 43.9 / 53.0 = **0.83**.

**(b)** That canola yields are around 60% of spring wheat yields, when the true figure is 83% -- the truncated chart roughly doubles the apparent gap.

**(c)** The value axis of a bar chart starts at zero. A bar encodes its value as length, so cutting the base off every bar changes the lengths without changing the labels, and the visual comparison no longer matches the numbers.

</details>

### Question 3

*For each display below, name the chart type you would choose and justify it in one sentence.*

**(a)** How the provincial average canola yield changed from 1990 to 2025.

**(b)** The 2025 average yields of six crops.

**(c)** How Manitoba's 2025 spring wheat acres split across the four largest varieties and everything else.

**(d)** Whether RMs with high spring wheat yields in 2025 also had high barley yields.

<details><summary>Answer</summary>

**(a)** A line chart -- yield against a continuous time axis shows the trend and the year-to-year swings.

**(b)** A bar chart -- six categories compared by length, sorted so the ranking is visible.

**(c)** A pie chart (or a bar chart of shares) -- the five values are shares of one whole, and five slices is few enough to read.

**(d)** A scatter plot -- one point per RM, wheat yield on one axis and barley on the other, shows whether the two move together.

</details>

### Question 4

*A table holds the seeded acres of the three largest Manitoba wheat varieties in each of 2023, 2024 and 2025.*

**(a)** Which bar layout, stacked or grouped, shows how the total acreage of the three varieties changed from year to year?

**(b)** Which layout shows how each individual variety changed from year to year?

**(c)** What does a faceted chart (one small panel per variety) do that a single grouped chart does not?

<details><summary>Answer</summary>

**(a)** Stacked bars -- the height of each year's stack is the total, so the totals can be compared directly.

**(b)** Grouped bars -- every variety's bars sit on the axis baseline, so each variety can be tracked across years. In a stacked chart only the bottom segment has a common baseline.

**(c)** Faceting gives each variety its own panel with a shared scale, so overlapping or crowded series can be read separately while remaining comparable.

</details>

### Question 5

*The long yields file has a `Unit` column: every crop is recorded in bu/ac except lentils, which are in lb/ac (a 2025 provincial average of about 1,900).*

**(a)** Explain why a pie chart of the eight crops' 2025 average yields would be a poor chart, whatever the units.

**(b)** Name data in this module's files that a pie chart could reasonably show.

**(c)** A bar chart of the 2025 average yields puts lentils on the same axis as the bu/ac crops. Describe what the chart looks like and give one way to fix it.

<details><summary>Answer</summary>

**(a)** The slices of a pie must be shares of a whole. Eight average yields do not sum to anything meaningful -- there is no "total yield" the slices divide up -- so the chart implies a whole that does not exist.

**(b)** Shares of seeded acres: for example, each variety's share of Manitoba's 2025 spring wheat acres (grouped to a few slices), since acres do sum to a meaningful total.

**(c)** The lentils bar, at about 1,900, is roughly twenty times the next value, so the bu/ac crops compress into short, unreadable bars. Fixes: leave lentils off the chart (and say so), plot it in its own panel, or convert everything to a common unit.

</details>

---

# Section 2 — Graphing in Excel

*Build each chart on the worksheet for that question. Titles and axis labels are part of the marks in this section. The PivotTable steps are the ones from Module 1.*

### Question 11

*The file `mb_wheat_reported_2020_2025.csv` holds Manitoba red spring wheat records by variety, municipality and year: the number of farms, seeded acres and average yield.*

**(a)** Use a PivotTable to total the 2025 acres for each variety.

**(b)** Make a horizontal bar chart of the six most-planted varieties, largest at the top, with a title and a labelled acres axis.

**(c)** In a cell, explain why horizontal bars suit this chart better than vertical ones.

<details><summary>Answer</summary>

The six most-planted 2025 varieties: AAC Brandon (787,111 acres), AAC Starbuck (556,738), AAC Wheatland (400,416), SY Manness (259,696), AAC Hockley (250,423) and AAC Viewfield (140,772). Horizontal bars keep the long variety names readable without rotating them.

<!-- TODO: answer workbook -->

</details>

### Question 12

*The file `rm_yields_1990_2025.csv` holds the average yield of each crop in every RM and year, one row per RM-year-crop.*

**(a)** Use a PivotTable to calculate the provincial average canola yield in each year from 1990 to 2025.

**(b)** Make a line chart of the series with a title and labelled axes, and set the vertical axis to start at zero.

**(c)** Add a data label marking the year with the highest average yield.

<details><summary>Answer</summary>

A PivotTable with `Year` in Rows, `Yield` in Values (changed from Sum to Average), and `Crop` filtered to Canola. The series runs from about 21 bu/ac in 1990 to about 44 in 2025. The data label goes on the maximum point (select the point twice, then Add Data Label).

<!-- TODO: answer workbook -->

</details>

### Question 13

*The same file: one row per RM-year-crop.*

**(a)** Use a PivotTable to calculate the provincial average yields of spring wheat and canola in each year, with the two crops as columns.

**(b)** Plot both series on one line chart with a title, labelled axes and a legend.

**(c)** In a cell, explain what the legend is doing in this chart, and when a line chart can do without one.

<details><summary>Answer</summary>

The PivotTable has `Year` in Rows, `Crop` in Columns (filtered to the two crops) and average `Yield` in Values. The legend maps each colour to a crop; with only one series, or with the lines labelled directly at their right-hand ends, no legend is needed.

<!-- TODO: answer workbook -->

</details>

### Question 14

*The Manitoba wheat file again: variety, municipality, year, farms, acres, yield.*

**(a)** Insert a PivotChart showing the total acres of each variety in each year, 2020 to 2025, drawn as one line per variety.

**(b)** The chart has too many lines to read. Filter it to varieties with at least 300,000 total acres across all years.

**(c)** In a cell, explain what the filtered chart gains and what information it no longer shows.

<details><summary>Answer</summary>

The PivotChart has `Year` in the axis area, `Variety` in the legend area and `Acres` in Values. A value filter on the variety field (Value Filters > Greater Than, on the sum of acres) leaves eight varieties. The filtered chart makes the major varieties readable; it no longer shows the small varieties at all, or the total acreage across every variety.

<!-- TODO: answer workbook -->

</details>

### Question 15

*The Manitoba wheat file again.*

**(a)** Use a PivotTable to total the 2025 acres by variety.

**(b)** Beside the PivotTable, build a five-row block that keeps the four largest varieties and groups the rest into `Other` (copy the values out of the PivotTable rather than editing it). Make a pie chart of the block, with each slice labelled.

**(c)** In a cell, calculate the share of 2025 acres in the four largest varieties.

<details><summary>Answer</summary>

The four largest 2025 varieties (Brandon, Starbuck, Wheatland, Manness) total 2,003,961 acres of the province's 2,571,678: a **77.9%** share, so the `Other` slice holds the remaining 22.1%.

<!-- TODO: answer workbook -->

</details>

---

# Section 3 — Graphing in R

*Answer these in your R script. Load `tidyverse` and `janitor` once at the top, read each file with `read_csv()` and clean the column names with `clean_names()`. Written answers go in comments. The file paths below assume the data sits in your project's `data/` folder.*

### Question 21

*The file `mb_wheat_reported_2020_2025.csv` holds Manitoba red spring wheat records by variety, municipality and year: the number of farms, seeded acres and average yield.*

**(a)** Total the 2025 acres by variety and keep the varieties with more than 100,000 acres.

**(b)** Make a horizontal bar chart of those totals, ordered so the largest bar is at the top (this is `fct_reorder()`), with a title, a labelled acres axis and comma-formatted axis labels.

**(c)** In a comment: what does `fct_reorder()` change about the `variety` column, and what order would the bars take without it?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.0
# Read the Manitoba wheat file and clean the names
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv",
                     show_col_types = FALSE) |>
  clean_names()

# (a) Total 2025 acres by variety, keeping those over 100,000 acres
top_2025 <- mb_wheat |>
  filter(year == 2025) |>                          # 2025 only
  group_by(variety) |>                             # one group per variety
  summarise(acres = sum(acres, na.rm = TRUE)) |>   # total the acres
  filter(acres > 100000)                           # keep the large varieties

top_2025

# (b) Horizontal bar chart, largest at the top
top_2025 |>
  ggplot(aes(x = acres, y = fct_reorder(variety, acres))) +
  geom_col(fill = "DarkGreen") +
  scale_x_continuous(labels = scales::comma) +     # 400,000 not 4e+05
  labs(title = "Acres of Manitoba wheat varieties, 2025",
       x = "Acres", y = NULL) +
  theme_classic(base_size = 13)

# (c) fct_reorder() converts variety to a factor whose categories are
# ordered by acres, so the bars plot in that order. Without it the
# bars would plot alphabetically.
```

</details>

### Question 22

*The file `rm_yields_1990_2025.csv` holds the average yield of each crop in every RM and year, one row per RM-year-crop.*

**(a)** Calculate the provincial average canola yield in each year.

**(b)** Make a line chart of the series with a title, labelled axes and a y-axis that starts at zero (`limits = c(0, NA)`).

**(c)** In a comment: what does the `NA` in `limits = c(0, NA)` do?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.6
# Read the long yields file and clean the names
rm_yields <- read_csv("data/rm_yields_1990_2025.csv",
                      show_col_types = FALSE) |>
  clean_names()

# (a) Provincial average canola yield by year
canola_by_year <- rm_yields |>
  filter(crop == "Canola") |>                          # canola only
  group_by(year) |>                                    # one group per year
  summarise(mean_yield = mean(yield, na.rm = TRUE))    # average across RMs

canola_by_year

# (b) Line chart with a zero baseline
canola_by_year |>
  ggplot(aes(x = year, y = mean_yield)) +
  geom_line(colour = "DarkGreen", linewidth = 1) +
  scale_y_continuous(limits = c(0, NA)) +              # start the axis at zero
  labs(title = "Average canola yield in Saskatchewan RMs",
       x = "Year", y = "Yield (bu/ac)") +
  theme_classic(base_size = 13)

# (c) The NA leaves the upper limit for ggplot to choose from the data,
# so only the lower end of the axis is fixed.
```

</details>

### Question 23

*The same file: one row per RM-year-crop.*

**(a)** Filter to spring wheat and canola (`%in%`) and calculate each crop's provincial average yield in each year.

**(b)** Plot both series on one chart with one line per crop.

**(c)** In a comment: what does mapping `colour = crop` inside `aes()` do that setting a colour inside `geom_line()` does not?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.6
# Read the long yields file and clean the names
rm_yields <- read_csv("data/rm_yields_1990_2025.csv",
                      show_col_types = FALSE) |>
  clean_names()

# (a) Average yield by crop and year for the two crops
two_crops <- rm_yields |>
  filter(crop %in% c("Spring Wheat", "Canola")) |>   # keep either crop
  group_by(crop, year) |>                            # one group per crop-year
  summarise(mean_yield = mean(yield, na.rm = TRUE),
            .groups = "drop")

two_crops

# (b) One line per crop
two_crops |>
  ggplot(aes(x = year, y = mean_yield, colour = crop)) +   # one line per crop
  geom_line(linewidth = 1) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(title = "Average spring wheat and canola yields in Saskatchewan RMs",
       x = "Year", y = "Yield (bu/ac)", colour = NULL) +
  theme_classic(base_size = 13)

# (c) Mapping colour = crop inside aes() ties colour to the data: each
# value of crop gets its own colour and its own line, and a legend is
# added. A colour set inside geom_line() is a fixed appearance choice
# applied to everything, and draws one line.
```

</details>

### Question 24

*The Manitoba wheat file again: variety, municipality, year, farms, acres, yield.*

**(a)** Calculate each variety's total acres across all years and municipalities, and rename varieties with under 300,000 total acres to `"Other"` (`if_else()`).

**(b)** Total the acres by variety and year, and plot one line per variety.

**(c)** In a comment: why must the acres be re-totalled after the renaming before you plot?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.6
#| fig-height: 3.8
# Read the Manitoba wheat file and clean the names
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv",
                     show_col_types = FALSE) |>
  clean_names()

# (a) Attach each variety's all-years total, then rename the small ones
mb_grouped <- mb_wheat |>
  group_by(variety) |>                                     # one group per variety
  mutate(total_acres = sum(acres, na.rm = TRUE)) |>        # total on every row
  ungroup() |>
  mutate(variety = if_else(total_acres < 300000,           # under the threshold
                           "Other",                        # becomes Other
                           variety))                       # otherwise unchanged

# (b) Re-total by variety and year, then plot
wheat_by_year <- mb_grouped |>
  group_by(variety, year) |>
  summarise(acres = sum(acres, na.rm = TRUE), .groups = "drop")

wheat_by_year |>
  ggplot(aes(x = year, y = acres, colour = variety)) +
  geom_line(linewidth = 0.9) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(title = "Acres of Manitoba wheat varieties",
       x = "Year", y = "Total acres", colour = NULL) +
  theme_classic(base_size = 13)

# (c) After the renaming there are many rows labelled Other in each
# year (one per small variety and municipality). Without re-totalling,
# ggplot would draw a line through all of those points instead of one
# line for the combined Other acreage.
```

</details>

### Question 25

*The file `rm_yields_1990plus.csv` has one row per RM-year and one column per crop, with yields in bu/ac.*

**(a)** For 2025, make a scatter plot of spring wheat yield against barley yield across RMs, with a title and labelled axes.

**(b)** In a comment: describe the pattern in one sentence.

**(c)** Save the chart to `output/` as a PNG with `ggsave()`.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.0
# Read the wide yields file and clean the names
rm_wide <- read_csv("data/rm_yields_1990plus.csv",
                    show_col_types = FALSE) |>
  clean_names()

# (a) Scatter of the two 2025 yields, one point per RM
wheat_barley_scatter <- rm_wide |>
  filter(year == 2025) |>                       # 2025 only
  ggplot(aes(x = spring_wheat, y = barley)) +
  geom_point(colour = "DarkGreen", alpha = 0.6) +
  labs(title = "Spring wheat and barley yields in Saskatchewan RMs, 2025",
       x = "Spring wheat yield (bu/ac)", y = "Barley yield (bu/ac)") +
  theme_classic(base_size = 13)

wheat_barley_scatter

# (b) The points rise together: RMs with higher spring wheat yields
# generally also had higher barley yields in 2025.

# (c) Save the chart as a PNG
ggsave(
  "output/mod04_q25_scatter.png",
  plot = wheat_barley_scatter,
  width = 6.4,
  height = 4.0,
  dpi = 300
)
```

</details>
