# Module 2 — Test Bank {.unnumbered}

```{r}
#| include: false
#| eval: true
# console() renders a code listing plus the console session it produces.
source("../_console.R")
```

::: {.callout-note}
## How this test bank works

This bank holds **40 questions**, ten in each section. Section 1 gives you the data inside the question; Sections 2--4 each hold two blocks of five, one block per dataset (the datasets are described below). On the real test you receive **four questions -- one from each section** -- and you submit **a single, well-commented R script**: a header block at the top, a clearly labelled section for each question, and your sentence answers written as comments. The script should run from top to bottom in a project folder that has the csvs in `data/`. Each question is designed to take about **8--10 minutes**.

Expand each **Answer** block to check your work. Every value has been computed from the real data; if the source file is refreshed the exact numbers may shift slightly, but the method stays the same.
:::

**The data.** Two datasets, four files, all in your project's `data/` folder:

*Saskatchewan RM crop yields* -- average yields in every Saskatchewan Rural Municipality, one row per RM-year(-crop), columns `Year`, `RM`, `Crop`, `Yield`, `Unit`:

- [`rm_canola_yields_1990_2025.csv`](data/rm_canola_yields_1990_2025.csv) -- the **canola rows only**, for Section 2: one crop, one unit, so its summary statistics mean something.
- [`rm_yields_1990_2025.csv`](data/rm_yields_1990_2025.csv) -- all **eight crops**, for Sections 3--4. You worked with the wide version in Module 1; this is the long form, with `Crop` as a variable. **Lentils are recorded in pounds per acre; every other crop is in bushels per acre** -- the `Unit` column says which, and several questions turn on it.

*Manitoba wheat variety yields* -- MASC's variety-level spring wheat yields, one row per municipality-variety-year, 2020--2025, columns `Year`, `Municipality`, `Variety`, `Farms`, `Acres`, `Yield_bu_ac`, `Reported`. Where too few farms grew a variety, the row is **suppressed** for privacy: `Reported` is FALSE and the numeric columns are missing.

- [`mb_wheat_reported_2020_2025.csv`](data/mb_wheat_reported_2020_2025.csv) -- the **reported rows only**, for Section 2.
- [`mb_wheat_varieties.csv`](data/mb_wheat_varieties.csv) -- the **full file**, suppressed rows included, for Sections 3--4.

**Grading model.** **20%** for script presentation -- a header block, comments in the style of the module, clearly labelled question sections, and a script that runs from top to bottom in a fresh session. The remaining **80%** is split evenly across the four questions, and evenly across the parts within each question. Parts are marked on **follow-through**: if an early part is wrong but the later parts are correctly worked *from your own earlier answer*, the later parts still earn full marks. Where a part asks for a sentence, write it as a comment in your script.

> **Vocabulary note.** Everything here uses only Module 2 tools: `<-`, arithmetic on vectors, `c()`, `data.frame()`, `$`, `mean`, `median`, `sd`, `var`, `min`, `max`, `range`, `sum`, `quantile` (with `probs`), `IQR`, `nrow`, `ncol`, `names`, `head`, `print`, `glimpse`, `summary`, `library()`, `install.packages()`, `read_csv` with relative paths, `filter` (with `==`, `>`, `<`, `>=`, `&`, `|`, `!`, `%in%`), `select` (including `-` and `:`), `rename`, `mutate`, `arrange` and `desc`, the pipe `|>`, `summarise`, `n()`, `group_by`, and `write_csv`. Nothing beyond that is needed. The chapter's `summarise()` examples write `na.rm = TRUE` inside `mean()` and `sd()`; none of the data these questions filter to has missing yields, so answers with or without it get the same numbers and the same marks.

::: {.callout-note collapse="true"}
## A sample test, with a full-marks answer script

A real test looks like this -- one question from each section:

1. *Five canola fields yielded 52.3, 47.8, 55.1, 44.6, and 50.9 bu/ac, on 160, 320, 240, 130, and 200 acres. Create the two vectors; convert the yields to t/ha; compute the mean and sd of the yields; compute total production and the acreage-weighted average yield.*
2. *Compute the mean, median, and standard deviation of canola `Yield`, and its 90th percentile. In a comment: the mean sits a little above the median -- what does that direction of gap suggest about the shape of canola yields?*
3. *Filter to the lentil rows, saving as `lentils` (how many rows?); add a `yield_kg_ha` column (multiply by 1.12), saving back to `lentils`; compute the mean of the new column and of the original, and check their ratio.*
4. *In one pipeline, compute each RM's mean Spring Wheat yield across all years, sorted best-first. Report the top three RMs. In a comment: what did each row of the input represent, and what does each row of the result represent?*

And here is a script that would earn full marks, including all of the presentation component -- with the console session it produces shown beneath it:

```{r}
#| eval: true
#| echo: false
console('
# ---
# Title: Module 2 test
# Author: Jordan Field
# Date: 2026-10-15
# Description:
#   Answers to the four test questions, one section per block.
#   Sentence answers are written as comments below each result.
# ---

# Load packages
library(tidyverse)

# ============================================================
# Section 1: five canola fields
# ============================================================

# Yields (bu/ac) and field sizes (acres)
yields <- c(52.3, 47.8, 55.1, 44.6, 50.9)
acres  <- c(160, 320, 240, 130, 200)

# (a/b) Convert every yield to tonnes per hectare in one step
yields_t_ha <- yields * 0.0560
yields_t_ha

# (c) Centre and spread of the yields
mean(yields)   # 50.14 bu/ac
sd(yields)     # 4.06 bu/ac

# (d) Total production, then the average weighted by acres
total_bu <- sum(yields * acres)
total_bu                 # 52,866 bu
total_bu / sum(acres)    # 50.35 bu/ac
# The weighted average differs from the plain mean because the
# fields differ in size, so each yield should count by its acres.

# ============================================================
# Section 2: summary statistics for canola
# ============================================================

# Read the canola file using a relative path from the project folder
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")

# Centre and spread of the canola yields
mean(rm_canola$Yield)      # 28.29 bu/ac
median(rm_canola$Yield)    # 26.9 bu/ac
sd(rm_canola$Yield)        # 10.06 bu/ac

# 90th percentile
quantile(rm_canola$Yield, 0.9)   # 42.5 bu/ac

# The mean sits a little above the median, which suggests mild
# right skew: the best RM-years pull the mean up more than the
# worst years pull it down.

# ============================================================
# Section 3: converting the lentil yields
# ============================================================

# Read the full eight-crop file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# Keep only the lentil rows
lentils <- filter(rm_yields, Crop == "Lentils")
nrow(lentils)    # 6,338 rows

# Add the converted column (1 lb/ac is about 1.12 kg/ha) and
# overwrite lentils with the version that has the extra column
lentils <- mutate(lentils, yield_kg_ha = Yield * 1.12)

# Means of the new and original columns
mean(lentils$yield_kg_ha)   # 1,353.4 kg/ha
mean(lentils$Yield)         # 1,208.39 lb/ac
# Their ratio is 1.12, the conversion factor -- multiplying every
# value by a constant multiplies the mean by the same constant.

# ============================================================
# Section 4: best wheat-growing RMs
# ============================================================

rm_yields |>                              # take the data, THEN
  filter(Crop == "Spring Wheat") |>       # keep spring wheat, THEN
  group_by(RM) |>                         # split the rows by RM, THEN
  summarise(mean_yield = mean(Yield)) |>  # one row per RM: its mean, THEN
  arrange(desc(mean_yield))               # best RMs on top

# Top three: RM 369 (47.6), RM 333 (47.2), RM 368 (47.2) bu/ac.
# Each input row was one RM-year observation of spring wheat; each
# result row is one RM, its years collapsed into a single mean.
')
```
:::

---

# Section 1 — R Basics

*These questions are about the building blocks: objects, vectors, data frames, and functions. No data file is needed -- the numbers are in the question. Answer in the Section 1 block of your script, with sentence answers written as comments.*

### Question 1

Five canola fields yielded 52.3, 47.8, 55.1, 44.6, and 50.9 bu/ac, on 160, 320, 240, 130, and 200 acres.

**(a)** Create a vector `yields` and a vector `acres` holding these values.

**(b)** Create a vector `yields_t_ha` converting the yields to tonnes per hectare (multiply by 0.0560). in one line (R multiplies every element at once).

**(c)** Compute the mean and standard deviation of `yields`.

**(d)** Compute total production in bushels (each field's yield times its acres, summed), and the farm's average yield weighted by acres (total bushels divided by total acres).

<details><summary>Answer</summary>

```r
yields <- c(52.3, 47.8, 55.1, 44.6, 50.9)
acres  <- c(160, 320, 240, 130, 200)
yields_t_ha <- yields * 0.0560
mean(yields)
sd(yields)
total_bu <- sum(yields * acres)
total_bu / sum(acres)
```

- (b) 2.93, 2.68, 3.09, 2.50, 2.85 t/ha.
- (c) Mean **50.14**, sd **4.06**.
- (d) Total **52,866 bu**; weighted average **50.35 bu/ac**.
</details>

### Question 2

Five barley fields yielded 68.2, 71.5, 64.9, 74.1, and 66.3 bu/ac. Five lentil fields yielded 1450, 1720, 1280, 1610, and 1390 lb/ac.

**(a)** Compute the mean, the standard deviation, the variance, and the range (maximum minus minimum) of each crop's yields.

**(b)** The lentil standard deviation is far larger. In a comment: why can the two standard deviations not be compared directly?

**(c)** Compute the coefficient of variation (sd divided by mean) for each crop. Which crop's yields are more variable *relative to their own average*?

<details><summary>Answer</summary>

```r
barley  <- c(68.2, 71.5, 64.9, 74.1, 66.3)
lentils <- c(1450, 1720, 1280, 1610, 1390)
var(barley)
var(lentils)
max(barley) - min(barley)
max(lentils) - min(lentils)
sd(barley) / mean(barley)
sd(lentils) / mean(lentils)
```

- (a) Barley mean **69.0**, sd **3.77**, variance **14.25**, range **9.2**; lentils mean **1490**, sd **175.36**, variance **30,750**, range **440**.
- (b) They are in different units (bushels vs pounds per acre) and on very different scales, so the raw spreads are not comparable.
- (c) CV barley **0.055**, lentils **0.118** -- lentil yields are more variable relative to their own average.
</details>

### Question 3

A trial records, for each field: the crop grown, the seeded acres, and whether the field was irrigated.

**(a)** Create three vectors of length four holding made-up values for these three variables, choosing the appropriate type for each (character, numeric, logical).

**(b)** In a comment: which values needed quotation marks and which must not have them?

**(c)** Combine your three vectors into a data frame called `trial` and print it.

<details><summary>Answer</summary>

```r
crop      <- c("Canola", "Wheat", "Peas", "Canola")
acres     <- c(160, 320, 240, 130)
irrigated <- c(TRUE, FALSE, FALSE, TRUE)
trial <- data.frame(crop = crop,
                    acres = acres,
                    irrigated = irrigated)
trial
```

- (b) The character values (crop names) need quotes. Numbers must not have them (quoted numbers become text), and `TRUE`/`FALSE` must not have them -- quoted, they would be text rather than logical values.
</details>

---

### Question 4

Twelve canola fields yielded 24.1, 31.5, 28.8, 35.2, 22.7, 30.4, 27.9, 33.6, 25.5, 29.3, 32.8, and 26.4 bu/ac.

**(a)** Save the values as a vector and compute the mean.

**(b)** Compute the 10th, 50th, and 90th percentiles in a single `quantile()` call.

**(c)** In a comment: one of your three percentiles should equal the median. Which one, and does it?

<details><summary>Answer</summary>

```r
canola <- c(24.1, 31.5, 28.8, 35.2, 22.7, 30.4, 27.9, 33.6, 25.5, 29.3, 32.8, 26.4)
mean(canola)
quantile(canola, c(0.1, 0.5, 0.9))
```

- (a) Mean **29.02**.
- (b) 10th **24.24**, 50th **29.05**, 90th **33.52**.
- (c) The 50th percentile is the median by definition; `median(canola)` confirms **29.05**.
</details>

### Question 5

Consider these three lines:

```r
k <- 25
k == 30
mean(x = c(10, 20, 60))
```

**(a)** In a comment: say what each line does and what, if anything, it prints.

**(b)** In a comment: after the third line runs, does an object called `x` appear in the Variables pane? Why or why not?

**(c)** In a comment: which single symbol would you change to make line 2 *assign* 30 to `k`, and which does this book prefer for assignment?

<details><summary>Answer</summary>

- (a) Line 1 assigns 25 to `k` (prints nothing). Line 2 asks whether `k` equals 30 and prints `[1] FALSE`. Line 3 computes a mean and prints `[1] 30`.
- (b) No -- the `x =` inside the call names an argument for `mean()`; it does not create an object.
- (c) `==` would become `<-` (or `=`, which also assigns) -- and the book uses `<-` for assignment.
</details>

### Question 6

**(a)** `mean()` takes its data as an argument named `x`. Write two calls that compute the mean of 12, 18, and 30 -- one with the argument unnamed, one with it named -- and confirm they agree.

**(b)** In a comment: what do you type to open the documentation for `sd()`?

**(c)** A classmate writes `Mean(c(12, 18, 30))` and R replies `could not find function "Mean"`. In a comment: what went wrong?

**(d)** Compute the standard deviation of the same three values, with the argument named.

<details><summary>Answer</summary>

```r
mean(c(12, 18, 30))
mean(x = c(12, 18, 30))
sd(x = c(12, 18, 30))
```

- (a) Both print `[1] 20`.
- (b) `?sd`.
- (c) R is exact about names: `Mean` with a capital M is not the same function as `mean`.
- (d) `[1] 9.165151`.
</details>

### Question 7

Four oat fields yielded 84.3, 91.7, 77.2, and 88.5 bu/ac, and oats are selling at $4.15 per bushel.

**(a)** Save the yields as a vector and the price as a single object.

**(b)** Compute a revenue-per-acre vector (yield times price) in one line, using the price object rather than retyping the number.

**(c)** Compute the mean revenue per acre.

**(d)** In a comment: the price rises to $4.60. What two lines would you rerun to update every answer, and why is that easier than if you had typed the price into each formula?

<details><summary>Answer</summary>

```r
yields <- c(84.3, 91.7, 77.2, 88.5)
price  <- 4.15
revenue <- yields * price
revenue
mean(revenue)
```

- (b) 349.85, 380.56, 320.38, 367.28 $/ac, rounded to the cent (the console shows 349.845 and so on).
- (c) Mean **$354.51** per acre.
- (d) Change `price <- 4.60`, then rerun the lines below it. The price lives in one object, so one edit updates everything -- the same reason a spreadsheet keeps a price in one cell.
</details>

### Question 8

**(a)** Build a data frame called `durum_fields` with columns `field_id` (`"D1"` to `"D4"`), `region` (South, South, Central, North), and `yield` (28.3, 41.6, 35.2, 44.8).

**(b)** Report its number of rows, number of columns, and column names using functions.

**(c)** Using `$`, compute the mean yield and the range (maximum minus minimum).

<details><summary>Answer</summary>

```r
durum_fields <- data.frame(
  field_id = c("D1", "D2", "D3", "D4"),
  region = c("South", "South", "Central", "North"),
  yield = c(28.3, 41.6, 35.2, 44.8)
)
nrow(durum_fields); ncol(durum_fields); names(durum_fields)
mean(durum_fields$yield)
max(durum_fields$yield) - min(durum_fields$yield)
```

- (b) 4 rows, 3 columns; `field_id`, `region`, `yield`.
- (c) Mean **37.48**; range **16.5**.
</details>

### Question 9

A student runs these lines in the console:

```r
bins <- c(1450, 2200, 1875)
bins * 2
total <- sum(bins)
```

**(a)** In a comment: what exactly does the second line print, `[1]` included, and is that doubled vector saved anywhere?

**(b)** In a comment: what does the third line print, and what does the Variables pane show afterwards?

**(c)** The student now wants the doubled vector kept for later. Write the line that does it.

<details><summary>Answer</summary>

- (a) `[1] 2900 4400 3750` -- displayed only, saved nowhere.
- (b) Nothing prints (it is an assignment); the pane shows `bins` (the three values) and `total` = 5525.
- (c) `bins_doubled <- bins * 2`.
</details>

### Question 10

Here is a working script:

```r
v1 <- c(47.2, 49.8, 46.5, 50.1, 48.4)
v2 <- c(38.9, 55.6, 44.2, 58.3, 41.0)
mean(v1)
mean(v2)
```

**(a)** Rewrite it with a header and comments. The two vectors are canola yields from Farm A and Farm B.

**(b)** Add lines computing each farm's standard deviation and range.

**(c)** In a comment: the two farms have nearly the same mean. Using your numbers from (b), say in one sentence how they differ.

<details><summary>Answer</summary>

```r
# ---
# Title: Comparing two canola farms
# Author: Your Name
# Date: 2026-10-08
# Description:
#   Compares the centre and spread of canola yields
#   on Farm A and Farm B.
# ---

# Canola yields by field (bu/ac)
farm_a <- c(47.2, 49.8, 46.5, 50.1, 48.4)
farm_b <- c(38.9, 55.6, 44.2, 58.3, 41.0)

# Centre
mean(farm_a)
mean(farm_b)

# Spread
sd(farm_a)
sd(farm_b)
max(farm_a) - min(farm_a)
max(farm_b) - min(farm_b)
```

- (b) Farm A sd **1.57**, range **3.6**; Farm B sd **8.79**, range **19.4**.
- (c) The means are close (48.4 vs 47.6), but Farm B's yields swing far more from field to field -- its spread is five times Farm A's.
</details>

# Section 2 — Reading and Inspecting Data

*Three blocks of ten, one per dataset. Each block starts by naming its file; on a real test your Section 2 question comes from one of the three. Answer in the Section 2 block of your script.*

## Saskatchewan RM Canola {.unnumbered}

*These questions use the canola-only file, [`rm_canola_yields_1990_2025.csv`](data/rm_canola_yields_1990_2025.csv), in your project's `data/` folder.*

### Question 11

**(a)** In a comment at the top of your script, sketch your project's folder layout (the subfolders and where the csv sits).

**(b)** Load the tidyverse and read the canola file into an object called `rm_canola`, using a relative path.

**(c)** Report the number of rows, the number of columns, and the column names using functions.

**(d)** In a comment: why will your `read_csv()` line work on the grader's computer, when a full path starting `C:/Users/...` would not?

<details><summary>Answer</summary>

```r
library(tidyverse)
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")
nrow(rm_canola); ncol(rm_canola); names(rm_canola)
```

- (c) **10,039 rows**, **5 columns**: `Year`, `RM`, `Crop`, `Yield`, `Unit`.
- (d) The path is relative to the project folder, and the grader opens the same self-contained folder -- so `data/...` exists on their machine too. An absolute path names one particular computer's user and folder layout.
</details>

### Question 12

**(a)** Run `glimpse()` on `rm_canola`. In a comment: which columns are text (`<chr>`) and which are numeric (`<dbl>`)?

**(b)** When `read_csv()` ran, it printed a note about the file. In a comment: what did that note report, and why does the module say to read it every time?

**(c)** `Year` came in as numeric (`<dbl>`). In a comment: which type of value is each of `2024` and `"2024"`? Why do we want a column we will compute with stored as numbers rather than as text?

<details><summary>Answer</summary>

- (a) `Crop` and `Unit` are `<chr>`; `Year`, `RM`, and `Yield` are `<dbl>`.
- (b) The row and column counts and its guess at each column's type (here `chr (2)` and `dbl (3)`). R guesses well but not perfectly, so the note is the first check that the file read the way you expect.
- (c) `2024` is numeric; `"2024"` is character (text). Numbers stored as text stop behaving like numbers -- you cannot do arithmetic with them, and sorting turns alphabetical -- so a column you will compute with should show `<dbl>`, not `<chr>`.
</details>

### Question 13

**(a)** Run `summary(rm_canola)`. From its output, report the minimum, median, and maximum of `Yield`, and the first and last `Year`.

**(b)** In a comment: work through the module's inspection questions for this file. Are the yields in between roughly 10 and 60 bu/ac, a typical canola range? Are the years what the file name promises? Are there missing values?

**(c)** The minimum is below 2 bu/ac. In a comment: is a canola yield that low necessarily a data error? What kind of year could produce it?

<details><summary>Answer</summary>

- (a) Yield: min **1.9**, median **26.9**, max **61**. Years **1990** to **2025**.
- (b) Yes -- everything sits between about 2 and 61 bu/ac, sensible for canola; the years run 1990--2025 as advertised; and `summary()` prints an `NA's` count only when a column has missing values -- no such line appears, so there are none.
- (c) Not necessarily -- an RM average that low is what a severe drought or a widespread crop failure looks like. Extreme is not the same as wrong; it is a value to investigate, not delete.
</details>

### Question 14

**(a)** Compute the mean, median, standard deviation, and variance of `rm_canola$Yield`.

**(b)** Compute the 90th percentile of `Yield`.

**(c)** The mean sits a little above the median. In a comment: using Module 1's language about means and medians, what does that direction of gap suggest about the shape of canola yields?

<details><summary>Answer</summary>

```r
mean(rm_canola$Yield)
median(rm_canola$Yield)
sd(rm_canola$Yield)
var(rm_canola$Yield)
quantile(rm_canola$Yield, 0.9)
```

- (a) Mean **28.29**, median **26.9**, sd **10.06**, variance **101.21** (the sd squared, in squared units).
- (b) 90th percentile **42.5**.
- (c) A mean above the median suggests mild right skew: the very best RM-years pull the mean up more than the worst years pull it down.
</details>

### Question 15

A classmate's fresh R session runs `rm_yields <- read_csv("data/rm_yields_1990_2025.csv")` and gets:

```
Error in read_csv(...) : could not find function "read_csv"
```

**(a)** In a comment: what is the most likely cause, and what one line fixes it?

**(b)** In a comment: they object, "but I installed the tidyverse last week!" Explain the difference between installing and loading, and how often each is done.

**(c)** In a comment: state the quotes rule -- where do quotation marks go when installing, and when loading?

**(d)** The tidyverse is a collection of packages. In comments, name the four core ones from the module and the job each does.

<details><summary>Answer</summary>

- (a) The package is not loaded in this session; run `library(tidyverse)` first.
- (b) Installing downloads the package onto the computer -- once per computer. Loading switches it on -- once per script or session. Installation does not carry across sessions as loaded.
- (c) Quotes when you install (`install.packages("tidyverse")`), none when you load (`library(tidyverse)`).
- (d) `readr` reads data, `dplyr` manipulates it, `tidyr` reshapes it, `ggplot2` makes graphs.
</details>

## Manitoba Wheat Variety {.unnumbered}

*These questions use [`mb_wheat_reported_2020_2025.csv`](data/mb_wheat_reported_2020_2025.csv) -- spring wheat yields by variety and municipality, reported observations only.*

### Question 16

**(a)** In a comment at the top of your script, sketch your project's folder layout.

**(b)** Load the tidyverse and read the file into an object called `mb_wheat`, using a relative path.

**(c)** Report the number of rows, the number of columns, and the column names using functions.

**(d)** In a comment: what does one row of this dataset represent?

<details><summary>Answer</summary>

```r
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")
nrow(mb_wheat); ncol(mb_wheat); names(mb_wheat)
```

- (c) **2,398 rows**, **7 columns**: `Year`, `Municipality`, `Variety`, `Farms`, `Acres`, `Yield_bu_ac`, `Reported`.
- (d) One wheat variety grown in one municipality in one year -- its farm count, acres, and average yield.
</details>

### Question 17

A classmate's script reads the data like this:

```r
mb_wheat <- read_csv("/Users/sam/Downloads/mb wheat data.csv")
```

It works on their laptop and fails on everyone else's.

**(a)** In a comment: give two problems -- one about where the path points, one about the file's name and location.

**(b)** Rewrite the line the way the module recommends, assuming the file was saved as `mb_wheat_reported_2020_2025.csv` in the project's `data/` folder.

**(c)** In a comment: your line fails with `does not exist in current working directory`. What is the first thing to check?

<details><summary>Answer</summary>

- (a) The absolute path names one machine's user folder, so it exists nowhere else; and the file is loose in Downloads under a space-filled name instead of renamed and placed in the project's `data/` folder.
- (b) `mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")`.
- (c) That the folder opened in Positron is the project folder -- the one containing `data/`. Relative paths start from the opened folder.
</details>

### Question 18

**(a)** Run `glimpse()` on `mb_wheat`. In a comment: which columns are text (`<chr>`), which numeric (`<dbl>`), and which logical (`<lgl>`)?

**(b)** In a comment: what values does the logical column hold in this file, and how does the file's name explain this?

**(c)** In a comment: `Farms` holds whole numbers. Would `glimpse()` showing it as `<chr>` be a problem, and what would it tell you about the file?

<details><summary>Answer</summary>

- (a) `Municipality` and `Variety` are `<chr>`; `Year`, `Farms`, `Acres`, `Yield_bu_ac` are `<dbl>`; `Reported` is `<lgl>`.
- (b) All `TRUE` -- this is the reported-only file, with the suppressed (`FALSE`) rows removed.
- (c) Yes -- a numeric column arriving as text means some value in it could not be read as a number, and calculations like `mean()` would fail on it.
</details>

### Question 19

**(a)** Run `summary(mb_wheat)`. From its output, report the minimum, median, and maximum of `Yield_bu_ac`, and the first and last `Year`.

**(b)** In a comment: work through the module's inspection questions. Are the yields in mostly between 30 and 100 bu/ac, a typical spring wheat range? Are there missing values?

**(c)** The minimum is **4.5 bu/ac**. In a comment: is a reported yield that low necessarily an error?

<details><summary>Answer</summary>

- (a) Yield: min **4.5**, median **62.2**, max **97.8**. Years **2020** to **2025**.
- (b) Plausible -- most values sit tens-of-bushels, sensible for Manitoba wheat; no `NA's` lines appear, so no missing values in this file.
- (c) Not necessarily -- a variety on a few drought- or flood-hit fields can genuinely average that low. Extreme is a value to investigate, not delete.
</details>

### Question 20

**(a)** Compute the mean, median, standard deviation, and variance of `mb_wheat$Yield_bu_ac`.

**(b)** In a comment: the mean sits a little *below* the median here. In Module 1's language, what does that direction of gap suggest, and what kind of observations would produce it?

**(c)** Compute the smallest and largest yields in one `range()` call.

<details><summary>Answer</summary>

```r
mean(mb_wheat$Yield_bu_ac)
median(mb_wheat$Yield_bu_ac)
sd(mb_wheat$Yield_bu_ac)
var(mb_wheat$Yield_bu_ac)
range(mb_wheat$Yield_bu_ac)
```

- (a) Mean **61.16**, median **62.2**, sd **12.38**, variance **153.34**.
- (b) Mild left skew: unusually *low* values pull the mean down more than the high values pull it up -- a tail of very poor variety-municipality observations mixed in with normal ones.
- (c) `[1]  4.5 97.8` -- `range()` returns both ends as one vector.
</details>

# Section 3 — Data Manipulation Functions

*Three blocks of ten, one per dataset. Each question uses two or three of the functions in sequence -- save each intermediate result to an object and work step by step (pipelines wait for Section 4).*

## Saskatchewan RM Crop Yields {.unnumbered}

*These questions use the full eight-crop file: read [`rm_yields_1990_2025.csv`](data/rm_yields_1990_2025.csv) into an object called `rm_yields` at the start of this block.*

### Question 21

**(a)** Filter `rm_yields` to Canola in 2024, saving the result as `canola_2024`. How many rows are in the filtered data frame?

**(b)** From `canola_2024`, select only the `RM` and `Yield` columns, saving the result. Report its column names and its number of columns.

**(c)** In a comment: after (a) and (b), how many rows does `rm_yields` itself have, and why?

<details><summary>Answer</summary>

```r
canola_2024 <- filter(rm_yields, Crop == "Canola" & Year == 2024)
nrow(canola_2024)
canola_slim <- select(canola_2024, RM, Yield)
names(canola_slim)
ncol(canola_slim)
```

- (a) **293 rows**.
- (b) `RM`, `Yield`; **2 columns** (still 293 rows).
- (c) Still **71,104** -- `filter()` and `select()` return new data frames and leave the original untouched.
</details>

### Question 22

**(a)** Filter to rows where the crop is Oats **or** Barley, using `|`, saving the result. How many rows are in the filtered data frame?

**(b)** Write the same filter using `%in%` and confirm the row count matches.

**(c)** Sort your filtered data frame from highest yield to lowest. What crop, RM, and year appear in the top row, and at what yield?

**(d)** In a comment: a classmate tried `filter(rm_yields, Crop == "Oats" & Crop == "Barley")` and got zero rows. Why?

<details><summary>Answer</summary>

```r
oats_barley <- filter(rm_yields, Crop == "Oats" | Crop == "Barley")
nrow(oats_barley)
oats_barley2 <- filter(rm_yields, Crop %in% c("Oats", "Barley"))
nrow(oats_barley2)
arrange(oats_barley, desc(Yield))
```

- (a, b) **19,709 rows** both ways.
- (c) A tie at **165.1 bu/ac**: Oats in RM **369** (2025) and Oats in RM **438** (2020) share the maximum, so either is a correct top row.
- (d) `&` requires both conditions true of the *same row*, and no single row's crop is both Oats and Barley. "Or" is the right connector.
</details>

### Question 23

**(a)** Filter to the Spring Wheat observations of at least 60 bu/ac (all years), using `>=` and saving the result. How many rows are in the filtered data frame?

**(b)** Filter again to keep only the 2023 rows of that result. How many rows are in this second data frame?

**(c)** From your 2023 data frame, select `RM` and `Yield` and sort best-first. Which RM tops the list, at what yield?

**(d)** In a comment: rewrite the condition "yield at least 60 and the crop is *not* Spring Wheat" using `!`.

<details><summary>Answer</summary>

```r
sw_60 <- filter(rm_yields, Crop == "Spring Wheat" & Yield >= 60)
nrow(sw_60)
sw_60_2023 <- filter(sw_60, Year == 2023)
nrow(sw_60_2023)
sw_best <- select(sw_60_2023, RM, Yield)
arrange(sw_best, desc(Yield))
```

- (a) **434 rows**. (b) **53 rows**.
- (c) RM **187**, at **74.7 bu/ac**.
- (d) `filter(rm_yields, Yield >= 60 & !Crop == "Spring Wheat")`.
</details>

### Question 24

**(a)** Make a version of the data with only `Year`, `RM`, `Crop`, and `Yield` -- once by *naming what you keep*, once by *dropping* what you don't want -- saving each, and confirm the two have the same column names. Then make a third version keeping the adjacent-column range `Year:Crop`, and report its column names.

**(b)** Rename `Yield` to `yield_per_acre` in the full data, saving the result, and show the new column names.

**(c)** Filter your renamed data frame to Peas in 2025. How many rows are in the filtered data frame?

<details><summary>Answer</summary>

```r
kept    <- select(rm_yields, Year, RM, Crop, Yield)
dropped <- select(rm_yields, -Unit)
names(kept)
names(dropped)
ranged  <- select(rm_yields, Year:Crop)
names(ranged)
renamed <- rename(rm_yields, yield_per_acre = Yield)
names(renamed)
peas_2025 <- filter(renamed, Crop == "Peas" & Year == 2025)
nrow(peas_2025)
```

- (a) Identical: `Year`, `RM`, `Crop`, `Yield` both ways; the range version keeps `Year`, `RM`, `Crop`.
- (b) `Year`, `RM`, `Crop`, `yield_per_acre`, `Unit`.
- (c) **285 rows**.
</details>

### Question 25

**(a)** Filter to Canola in 2024, then in one `mutate()` call add two columns: `yield_t_ha` (multiply by 0.0560) and `yield_kg_ha` (multiply by 56.0). Save the result of each step.

**(b)** Compute the mean of each new column.

**(c)** Sort by `yield_t_ha`, best first. Which RM tops the table, and at what yield in t/ha?

**(d)** In a comment: without running anything, what is the mean of `yield_kg_ha` divided by the mean of `yield_t_ha`, and why?

<details><summary>Answer</summary>

```r
canola_2024 <- filter(rm_yields, Crop == "Canola" & Year == 2024)
canola_2024 <- mutate(canola_2024,
                      yield_t_ha  = Yield * 0.0560,
                      yield_kg_ha = Yield * 56.0)
mean(canola_2024$yield_t_ha)
mean(canola_2024$yield_kg_ha)
arrange(canola_2024, desc(yield_t_ha))
```

- (b) **1.76 t/ha** and **1,764.11 kg/ha**.
- (c) RM **493**, at **2.57 t/ha**.
- (d) 1,000 -- both columns are the same yields scaled by constants, and 56.0 / 0.0560 = 1000 (a tonne is 1,000 kg).
</details>

## Manitoba Wheat Variety {.unnumbered}

*These questions use the full [`mb_wheat_varieties.csv`](data/mb_wheat_varieties.csv) -- including the suppressed rows, where `Reported` is FALSE and the numeric columns are missing. Read it into an object called `mb` at the start of this block.*

### Question 26

**(a)** Filter `mb` to the reported rows (`Reported == TRUE`), saving as `mb_rep`. How many rows are in the filtered data frame?

**(b)** From `mb_rep`, select `Municipality`, `Variety`, and `Yield_bu_ac`, saving the result. Report its column names.

**(c)** In a comment: how many rows does `mb` itself still have, and what happened to the suppressed rows in `mb_rep`?

<details><summary>Answer</summary>

```r
mb_rep <- filter(mb, Reported == TRUE)
nrow(mb_rep)
mb_slim <- select(mb_rep, Municipality, Variety, Yield_bu_ac)
names(mb_slim)
```

- (a) **2,398 rows**.
- (b) `Municipality`, `Variety`, `Yield_bu_ac`.
- (c) `mb` still has **5,358** rows -- filtering returns a new data frame. The 2,960 suppressed rows are simply absent from `mb_rep`; they were not deleted from anything.
</details>

### Question 27

**(a)** From `mb_rep` (reported rows), filter to the varieties `AAC BRANDON (BW 932)` **or** `AAC STARBUCK <SECAN>` using `|`, saving the result. How many rows are in the filtered data frame?

**(b)** Write the same filter using `%in%` and confirm the count matches.

**(c)** In a comment: a classmate wrote the condition with `&` instead of `|` and got zero rows. Why?

<details><summary>Answer</summary>

```r
two_var <- filter(mb_rep, Variety == "AAC BRANDON (BW 932)" | Variety == "AAC STARBUCK <SECAN>")
nrow(two_var)
two_var2 <- filter(mb_rep, Variety %in% c("AAC BRANDON (BW 932)", "AAC STARBUCK <SECAN>"))
nrow(two_var2)
```

- (a, b) **893 rows** both ways.
- (c) `&` demands both conditions true of the same row, and no single row's variety is both names at once. "Or" is the right connector.
</details>

### Question 28

**(a)** Filter `mb_rep` to yields above 80 bu/ac, saving the result. How many rows are in the filtered data frame?

**(b)** Filter that result again to 2024 only. How many rows now?

**(c)** Sort your over-80 data frame from (a) best-first. Which municipality, variety, and year top the list, at what yield?

<details><summary>Answer</summary>

```r
high <- filter(mb_rep, Yield_bu_ac > 80)
nrow(high)
high_2024 <- filter(high, Year == 2024)
nrow(high_2024)
arrange(high, desc(Yield_bu_ac))
```

- (a) **97 rows**. (b) **16 rows**.
- (c) HILLSBURG-ROBLIN-SHELL RIVER, `BOLLES <SEED DEPOT>`, in **2022**, at **97.8 bu/ac**.
</details>

### Question 29

**(a)** From `mb_rep`, make a version with only `Year`, `Municipality`, `Variety`, and `Yield_bu_ac` -- once by naming what you keep, once by dropping what you don't (`Farms`, `Acres`, `Reported`) -- saving each, and confirm the two have the same column names. Then make a third version keeping the adjacent range `Year:Variety` and report its names.

**(b)** Rename `Yield_bu_ac` to `yield`, saving the result, and show the new column names.

<details><summary>Answer</summary>

```r
kept    <- select(mb_rep, Year, Municipality, Variety, Yield_bu_ac)
dropped <- select(mb_rep, -Farms, -Acres, -Reported)
names(kept)
names(dropped)
ranged  <- select(mb_rep, Year:Variety)
names(ranged)
renamed <- rename(mb_rep, yield = Yield_bu_ac)
names(renamed)
```

- (a) Identical four names both ways; the range version keeps `Year`, `Municipality`, `Variety`.
- (b) The same seven columns with `yield` in place of `Yield_bu_ac`.
</details>

### Question 30

**(a)** From `mb_rep`, use one `mutate()` to add `total_bu`, each observation's acres times its yield, saving the result.

**(b)** Compute the mean of `total_bu`.

**(c)** Sort by `total_bu`, largest first. Which municipality tops the table, and at roughly what total?

**(d)** In a comment: what does `total_bu` measure that `Yield_bu_ac` does not?

<details><summary>Answer</summary>

```r
mb_rep <- mutate(mb_rep, total_bu = Acres * Yield_bu_ac)
mean(mb_rep$total_bu)
arrange(mb_rep, desc(total_bu))
```

- (b) Mean **382,865 bu**.
- (c) PORTAGE LA PRAIRIE, at about **5.13 million bushels**.
- (d) Production -- how much grain the planting actually produced -- rather than how well each acre yielded. A modest yield on enormous acres can out-produce a great yield on few.
</details>

# Section 4 — Pipelines and Grouped Summaries

*Three blocks of ten, one per dataset. Answer in the Section 4 block of your script. Write multi-step work as pipelines with `|>`, one step per line.*

## Saskatchewan RM Crop Yields {.unnumbered}

*Still the full eight-crop file (`rm_yields`).*

### Question 31

**(a)** Write a pipeline, one step per line with a short comment on each step, that filters `rm_yields` to Durum in 2025, keeps only the `RM` and `Yield` columns, and sorts the result from highest yield to lowest.

**(b)** Run it. How many rows are in the resulting data frame, and which RM had the best durum yield in 2025, at what value?

**(c)** In a comment: read your pipeline aloud as a "take the data, then ..., then ..., then ..." sentence.

<details><summary>Answer</summary>

```r
rm_yields |>                              # take the data, THEN
  filter(Crop == "Durum" & Year == 2025) |>  # keep durum in 2025, THEN
  select(RM, Yield) |>                    # keep two columns, THEN
  arrange(desc(Yield))                    # sort best-first
```

- (b) **189 rows**; RM **369** at **105.6 bu/ac**.
- (c) Take `rm_yields`, then keep the 2025 durum rows, then keep the RM and Yield columns, then sort from highest yield to lowest.
</details>

### Question 32

**(a)** In one pipeline, compute the mean flax yield across 2020--2025 (`Year >= 2020`), calling the result column `mean_yield`.

**(b)** Add `n()` to the same `summarise()` to count how many observations that mean is based on.

**(c)** In a comment: why is reporting the count alongside the mean good practice?

<details><summary>Answer</summary>

```r
rm_yields |>
  filter(Crop == "Flax" & Year >= 2020) |>
  summarise(mean_yield = mean(Yield),
            n = n())
```

- (a, b) Mean **22.39 bu/ac** from **1,143** observations.
- (c) A mean based on a handful of rows deserves less trust than one based on a thousand; the count tells the reader which they are looking at.
</details>

### Question 33

**(a)** In one pipeline, compute the mean canola yield for **each year** from 2021 to 2025 (filter first, then group).

**(b)** Report the five means. Which year was worst, and which best?

**(c)** In a comment: 2021 was a severe drought year in Saskatchewan. Does the data agree?

<details><summary>Answer</summary>

```r
rm_yields |>
  filter(Crop == "Canola" & Year >= 2021) |>
  group_by(Year) |>
  summarise(mean_yield = mean(Yield))
```

- (a, b) 2021 **21.9**, 2022 **35.4**, 2023 **33.9**, 2024 **31.5**, 2025 **43.9**. Worst 2021, best 2025.
- (c) Yes -- the 2021 mean sits roughly a third below the surrounding years.
</details>

### Question 34

**(a)** In one pipeline: filter to Lentils, convert to kg/ha with `mutate()` (multiply by 1.12), then summarise the mean of the converted column.

**(b)** Report the mean in kg/ha.

**(c)** In a comment: in what order do your steps run, and why would putting the `summarise()` before the `mutate()` fail?

<details><summary>Answer</summary>

```r
rm_yields |>
  filter(Crop == "Lentils") |>
  mutate(yield_kg_ha = Yield * 1.12) |>
  summarise(mean_kg_ha = mean(yield_kg_ha))
```

- (b) **1,353.4 kg/ha**.
- (c) Top to bottom -- each step's output feeds the next. After `summarise()` the data has collapsed to one row and the `yield_kg_ha` column would not exist yet to summarise.
</details>

### Question 35

**(a)** In one pipeline: for **2025 only**, compute each crop's mean yield and observation count, sorted from highest mean to lowest, and save the result as `crop_summary_2025`.

**(b)** Write the table to `output/crop_summary_2025.csv`.

**(c)** Report the two highest-yielding crops in the sorted table. In a comment: check the Unit column: why is the first-place crop not comparable to the rest, and which bushel crop is the real standout?, and which crop is really the standout?

<details><summary>Answer</summary>

```r
crop_summary_2025 <- rm_yields |>
  filter(Year == 2025) |>
  group_by(Crop) |>
  summarise(mean_yield = mean(Yield),
            n = n()) |>
  arrange(desc(mean_yield))

write_csv(crop_summary_2025, "output/crop_summary_2025.csv")
```

- (a, c) Lentils top the table at **1,897** -- but that is pounds per acre, not bushels. Among the bushel crops, Oats lead at **97.5 bu/ac** (Barley next at 72.8). 2025 was a strong year across the board.
</details>

## Manitoba Wheat Variety {.unnumbered}

*Full [`mb_wheat_varieties.csv`](data/mb_wheat_varieties.csv), read as `mb`, with the reported rows saved as `mb_rep <- filter(mb, Reported == TRUE)` at the start of this block. Build everything else as pipelines.*

### Question 36

**(a)** Write a pipeline, one step per line with a short comment on each, that filters `mb_rep` to 2025, keeps `Municipality`, `Variety`, and `Yield_bu_ac`, and sorts best-first.

**(b)** Run it. How many rows are in the resulting data frame, and which municipality and variety top it, at what yield?

**(c)** In a comment: read the pipeline aloud as a "take the data, then ..." sentence.

<details><summary>Answer</summary>

```r
mb_rep |>                                          # take the reported rows, THEN
  filter(Year == 2025) |>                          # keep 2025, THEN
  select(Municipality, Variety, Yield_bu_ac) |>    # keep three columns, THEN
  arrange(desc(Yield_bu_ac))                       # sort best-first
```

- (b) **438 rows**; LOUISE, growing SY MANNESS, at **96.3 bu/ac**.
- (c) Take the reported rows, then keep 2025, then keep three columns, then sort from highest yield to lowest.
</details>

### Question 37

**(a)** In one pipeline, compute the mean of `Yield_bu_ac` across all of `mb_rep`, calling it `mean_yield`, together with `n()`.

**(b)** Report both values.

**(c)** In a comment: why report the count alongside the mean?

<details><summary>Answer</summary>

```r
mb_rep |>
  summarise(mean_yield = mean(Yield_bu_ac),
            n = n())
```

- (b) Mean **61.16 bu/ac** from **2,398** observations.
- (c) A mean earns trust in proportion to what it is based on; the count tells the reader how much that is.
</details>

### Question 38

**(a)** In one pipeline, compute each variety's mean yield and observation count across `mb_rep`, sorted from highest mean to lowest.

**(b)** Report the top two varieties, their means, and their counts.

**(c)** In a comment: the second-place variety was observed exactly once. Which of the two means would you trust more, and why?

<details><summary>Answer</summary>

```r
mb_rep |>
  group_by(Variety) |>
  summarise(mean_yield = mean(Yield_bu_ac),
            n = n()) |>
  arrange(desc(mean_yield))
```

- (b) AAC WESTKING, **79.2 bu/ac** from **17** observations; then CDC GO (BW781), **77.5** from **1**.
- (c) WESTKING's -- seventeen municipality-years is evidence; a single observation could be one lucky field in one good year.
</details>

### Question 39

A bushel of wheat is 60 lb, so one bu/ac is about 0.0673 t/ha.

**(a)** In one pipeline: convert `mb_rep`'s yields to t/ha with `mutate()`, then compute the mean t/ha for each year.

**(b)** Report the 2021 and 2025 values.

**(c)** In a comment: why must the `mutate()` step come before the `summarise()` step?

<details><summary>Answer</summary>

```r
mb_rep |>
  mutate(yield_t_ha = Yield_bu_ac * 0.0673) |>
  group_by(Year) |>
  summarise(mean_t_ha = mean(yield_t_ha))
```

- (b) 2021 **3.34 t/ha**; 2025 **4.58 t/ha**.
- (c) Each step's output feeds the next -- after `summarise()` the data has collapsed to one row per year and the `yield_t_ha` column would not exist to average.
</details>

### Question 40

**(a)** In one pipeline: for **2025 only**, compute each variety's mean reported yield and observation count, sorted best-first, and save the result as `variety_summary_2025`.

**(b)** Write the table to `output/variety_summary_2025.csv`.

**(c)** Report the top row. In a comment: the table also contains varieties observed only twice -- what should a reader keep in mind when comparing rows?

<details><summary>Answer</summary>

```r
variety_summary_2025 <- mb_rep |>
  filter(Year == 2025) |>
  group_by(Variety) |>
  summarise(mean_yield = mean(Yield_bu_ac),
            n = n()) |>
  arrange(desc(mean_yield))

write_csv(variety_summary_2025, "output/variety_summary_2025.csv")
```

- (c) AAC WESTKING, **79.2 bu/ac** from 17 observations. Rows with tiny counts (some varieties appear once or twice) carry far less evidence than their neighbours -- the `n` column is part of the answer, not decoration.
</details>
