# Module 2 — Test Bank {.unnumbered}

```{r}
#| include: false
#| eval: true
# console() renders a code listing plus the console session it produces.
source("../_console.R")
```

::: {.callout-note}
## How this test bank works

This bank holds **56 questions**, fourteen in each section. Section 1 gives you the data inside the question; Sections 2--4 each hold two blocks of seven, one block per dataset (the datasets are described below). On the real test you receive **four questions -- one from each section** -- and you submit **a single, well-commented R script**: a header block at the top, a clearly labelled section for each question, and your sentence answers written as comments. The script should run from top to bottom in a project folder that has the csvs in `data/`. Each question is designed to take about **8--10 minutes**.

Expand each **Answer** block to check your work. Every value has been computed from the real data; if the source file is refreshed the exact numbers may shift slightly, but the method stays the same.
:::

**The data.** Two datasets, four files, all in your project's `data/` folder:

*Saskatchewan RM crop yields* -- average yields in every Saskatchewan Rural Municipality, one row per RM-year(-crop), columns `Year`, `RM`, `Crop`, `Yield`, `Unit`:

- [`rm_canola_yields_1990_2025.csv`](data/rm_canola_yields_1990_2025.csv) -- the **canola rows only**, for Section 2: one crop, one unit, so its summary statistics mean something.
- [`rm_yields_1990_2025.csv`](data/rm_yields_1990_2025.csv) -- all **eight crops**, for Sections 3--4. You worked with the wide version in Module 1; this is the long form, with `Crop` as a variable. **Lentils are recorded in pounds per acre; every other crop is in bushels per acre** -- the `Unit` column says which, and several questions turn on it.

*Manitoba wheat variety yields* -- MASC's variety-level spring wheat yields, one row per municipality-variety-year, 2020--2025, columns `Year`, `Municipality`, `Variety`, `Farms`, `Acres`, `Yield_bu_ac`, `Reported`. Where too few farms grew a variety, the row is **suppressed** for privacy: `Reported` is FALSE and the numeric columns are missing.

- [`mb_wheat_reported_2020_2025.csv`](data/mb_wheat_reported_2020_2025.csv) -- the **reported rows only**, for Section 2.
- [`mb_wheat_varieties.csv`](data/mb_wheat_varieties.csv) -- the **full file**, suppressed rows included, for Sections 3--4.

**Grading model.** **20%** for script presentation -- a header block, comments in the style of the module, clearly labelled question sections, and a script that runs from top to bottom in a fresh session. The remaining **80%** is split evenly across the four questions, and evenly across the parts within each question. Parts are marked on **follow-through**: if an early part is wrong but the later parts are correctly worked *from your own earlier answer*, the later parts still earn full marks. Where a part asks for a sentence, write it as a comment in your script.

> **Vocabulary note.** Everything here uses only Module 2 tools: `<-`, arithmetic on vectors, `c()`, `data.frame()`, `$`, `mean`, `median`, `sd`, `var`, `min`, `max`, `range`, `sum`, `quantile` (with `probs`), `IQR`, `nrow`, `ncol`, `names`, `head`, `print`, `glimpse`, `summary`, `library()`, `install.packages()`, `read_csv` with relative paths, `filter` (with `==`, `>`, `<`, `>=`, `&`, `|`, `!`, `%in%`), those same comparisons applied to whole vectors (`sum()` of the TRUE/FALSE result counts matches), `select` (including `-` and `:`), `rename`, `mutate`, `arrange` and `desc`, the pipe `|>`, `summarise`, `n()`, `group_by`, and `write_csv`. Nothing beyond that is needed. The chapter's `summarise()` examples write `na.rm = TRUE` inside `mean()` and `sd()`; none of the data these questions filter to has missing yields, so answers with or without it get the same numbers and the same marks.

::: {.callout-note collapse="true"}
## A sample test, with a full-marks answer script

A real test looks like this -- one question from each section:

1. *Five canola fields yielded 52.3, 47.8, 55.1, 44.6, and 50.9 bu/ac, on 160, 320, 240, 130, and 200 acres. Create the two vectors; convert the yields to t/ha; compute the mean and sd of the yields; compute total production and the acreage-weighted average yield.*
2. *Compute the mean, median, and standard deviation of canola `Yield`, and its 90th percentile. In a comment: the mean sits a little above the median -- what does that direction of gap suggest about the shape of canola yields?*
3. *Filter to the lentil rows, saving as `lentils` (how many rows?); add a `yield_kg_ha` column (multiply by 1.12), saving back to `lentils`; compute the mean of the new column and of the original, and check their ratio.*
4. *In one pipeline, compute each RM's mean Spring Wheat yield across all years, sorted best-first. Report the top three RMs. In a comment: what did each row of the input represent, and what does each row of the result represent?*

And here is a script that would earn full marks, including all of the presentation component -- with the console session it produces shown beneath it:

```{r}
#| eval: true
#| echo: false
console('
# ---
# Title: Module 2 test
# Author: Jordan Field
# Date: 2026-10-15
# Description:
#   Answers to the four test questions, one section per block.
#   Sentence answers are written as comments below each result.
# ---

# Load packages
library(tidyverse)

# ============================================================
# Section 1: five canola fields
# ============================================================

# Yields (bu/ac) and field sizes (acres)
yields <- c(52.3, 47.8, 55.1, 44.6, 50.9)
acres  <- c(160, 320, 240, 130, 200)

# (a/b) Convert every yield to tonnes per hectare in one step
yields_t_ha <- yields * 0.0560
yields_t_ha

# (c) Centre and spread of the yields
mean(yields)   # 50.14 bu/ac
sd(yields)     # 4.06 bu/ac

# (d) Total production, then the average weighted by acres
total_bu <- sum(yields * acres)
total_bu                 # 52,866 bu
total_bu / sum(acres)    # 50.35 bu/ac
# The weighted average differs from the plain mean because the
# fields differ in size, so each yield should count by its acres.

# ============================================================
# Section 2: summary statistics for canola
# ============================================================

# Read the canola file using a relative path from the project folder
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")

# Centre and spread of the canola yields
mean(rm_canola$Yield)      # 28.29 bu/ac
median(rm_canola$Yield)    # 26.9 bu/ac
sd(rm_canola$Yield)        # 10.06 bu/ac

# 90th percentile
quantile(rm_canola$Yield, 0.9)   # 42.5 bu/ac

# The mean sits a little above the median, which suggests mild
# right skew: the best RM-years pull the mean up more than the
# worst years pull it down.

# ============================================================
# Section 3: converting the lentil yields
# ============================================================

# Read the full eight-crop file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# Keep only the lentil rows
lentils <- filter(rm_yields, Crop == "Lentils")
nrow(lentils)    # 6,338 rows

# Add the converted column (1 lb/ac is about 1.12 kg/ha) and
# overwrite lentils with the version that has the extra column
lentils <- mutate(lentils, yield_kg_ha = Yield * 1.12)

# Means of the new and original columns
mean(lentils$yield_kg_ha)   # 1,353.4 kg/ha
mean(lentils$Yield)         # 1,208.39 lb/ac
# Their ratio is 1.12, the conversion factor -- multiplying every
# value by a constant multiplies the mean by the same constant.

# ============================================================
# Section 4: best wheat-growing RMs
# ============================================================

rm_yields |>                              # take the data, THEN
  filter(Crop == "Spring Wheat") |>       # keep spring wheat, THEN
  group_by(RM) |>                         # split the rows by RM, THEN
  summarise(mean_yield = mean(Yield)) |>  # one row per RM: its mean, THEN
  arrange(desc(mean_yield))               # best RMs on top

# Top three: RM 369 (47.6), RM 333 (47.2), RM 368 (47.2) bu/ac.
# Each input row was one RM-year observation of spring wheat; each
# result row is one RM, its years collapsed into a single mean.
')
```
:::

---

# Section 1 — R Basics

### Question 1

Five canola fields yielded 52.3, 47.8, 55.1, 44.6, and 50.9 bu/ac, on 160, 320, 240, 130, and 200 acres.

**(a)** Create a vector `yields` and a vector `acres` holding these values.

**(b)** Create a vector `yields_t_ha` converting the yields to tonnes per hectare (multiply by 0.0560) in one line -- R multiplies every element at once.

**(c)** Compute the mean and standard deviation of `yields`.

**(d)** Compute total production in bushels (each field's yield times its acres, summed), and the farm's average yield weighted by acres (total bushels divided by total acres).

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Create the yield and acreage vectors
yields <- c(52.3, 47.8, 55.1, 44.6, 50.9)
acres  <- c(160, 320, 240, 130, 200)

# (b) Convert every yield to tonnes per hectare in one line
yields_t_ha <- yields * 0.0560

# Print the converted yields
yields_t_ha

# (c) Mean and standard deviation of the yields
mean(yields)
sd(yields)

# (d) Total production in bushels
total_bu <- sum(yields * acres)
total_bu

# (d) Average yield weighted by acres
total_bu / sum(acres)

## I find the mean to be 50.14 bu/ac and the sd 4.06.
## Total production is 52,866 bu and the acre-weighted average 50.35 bu/ac.
')
```

</details>

### Question 2

Six spring wheat fields yielded 51.2, 48.7, 53.4, 8.2, 50.9, and 49.5 bu/ac; the 8.2 field was hailed out in July.

**(a)** Save the yields and compute the mean and the median.

**(b)** In a comment: the two disagree by nearly seven bushels. Which one better describes a typical field on this farm, and why?

**(c)** Compute the mean of the five unhailed fields (a second vector is fine) and compare it with the median from part a. In a comment: what does the closeness of those two numbers tell you about the median?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Yields, including the hailed-out field
yields <- c(51.2, 48.7, 53.4, 8.2, 50.9, 49.5)
mean(yields)
median(yields)

# (c) The five unhailed fields
unhailed <- c(51.2, 48.7, 53.4, 50.9, 49.5)
mean(unhailed)

## I find the mean to be 43.65 and the median 50.2.
## (b) The median describes a typical field better: five of the six fields
## sit near 50, and the single 8.2 drags the mean seven bushels below all
## of them while barely touching the median.
## (c) The unhailed mean is 50.74, almost exactly the full-data median --
## the median was already ignoring the extreme value.
')
```

</details>

### Question 3

Five barley fields yielded 68.2, 71.5, 64.9, 74.1, and 66.3 bu/ac. Five lentil fields yielded 1450, 1720, 1280, 1610, and 1390 lb/ac.

**(a)** Compute the mean, the standard deviation, the variance, and the range (maximum minus minimum) of each crop's yields.

**(b)** The lentil standard deviation is far larger. In a comment: why can the two standard deviations not be compared directly?

**(c)** Compute the coefficient of variation (sd divided by mean) for each crop. Which crop's yields are more variable *relative to their own average*?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Yield vectors for each crop
barley  <- c(68.2, 71.5, 64.9, 74.1, 66.3)
lentils <- c(1450, 1720, 1280, 1610, 1390)

# (a) Mean of each crop
mean(barley)
mean(lentils)

# (a) Standard deviation of each crop
sd(barley)
sd(lentils)

# (a) Variance of each crop
var(barley)
var(lentils)

# (a) Range (maximum minus minimum) of each crop
max(barley) - min(barley)
max(lentils) - min(lentils)

# (c) Coefficient of variation (sd divided by mean) for each crop
sd(barley) / mean(barley)
sd(lentils) / mean(lentils)

## I find barley mean 69.0, sd 3.77, variance 14.25, range 9.2; lentils
## mean 1490, sd 175.36, variance 30,750, range 440.
## (b) The two standard deviations are in different units (bushels vs pounds
## per acre) and on very different scales, so the raw spreads are not
## comparable.
## (c) CV barley 0.055, lentils 0.118 -- lentil yields are more variable
## relative to their own average.
')
```

</details>

### Question 4

Here is a script exactly as saved:

```r
mean(yields)
yields <- c(31.7, 28.4, 33.9, 30.2)
```

**(a)** In a comment: what happens when this script runs top to bottom in a fresh session, and why?

**(b)** Rewrite the script in working order, with a header block and a comment on each line.

**(c)** In a comment: why does the order not matter when you type the same two lines interactively in a console session where `yields` already exists, and why is that a trap?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Run as saved, the script fails on line 1: mean(yields) is asked for
# before yields exists in the fresh session, so R stops with
# "object of type ... not found" -- object yields not found.

# (b) The working order, with a header
# ---
# Title: Canola yields
# Author: Your Name
# Description: Creates the yield vector and computes its mean.
# ---

# Create the yield vector first
yields <- c(31.7, 28.4, 33.9, 30.2)

# Then compute its mean
mean(yields)

## I find the mean to be 31.05.
## (c) Typed interactively into a session where yields already exists, the
## original order works because the object survives from earlier commands.
## That is the trap: the console remembers, the script does not, so a
## script can look tested while being unrunnable from the top.
')
```

</details>

### Question 5

A trial records, for each field: the crop grown, the seeded acres, and whether the field was irrigated.

**(a)** Create three vectors of length four holding made-up values for these three variables, choosing the appropriate type for each (character, numeric, logical).

**(b)** In a comment: which values needed quotation marks and which must not have them?

**(c)** Combine your three vectors into a data frame called `trial` and print it.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Crop grown: character, so the values are quoted
crop      <- c("Canola", "Wheat", "Peas", "Canola")

# (a) Seeded acres: numeric, no quotes
acres     <- c(160, 320, 240, 130)

# (a) Irrigated or not: logical, no quotes
irrigated <- c(TRUE, FALSE, FALSE, TRUE)

# (c) Combine the three vectors into a data frame
trial <- data.frame(crop = crop,
                    acres = acres,
                    irrigated = irrigated)

# Print the data frame
trial

## (b) The character values (crop names) need quotes. Numbers must not have
## them (quoted numbers become text), and TRUE/FALSE must not have them --
## quoted, they would be text rather than logical values.
')
```

</details>

---

### Question 6

The same five canola fields were measured in two years. 2024: 38.2, 41.6, 35.9, 43.1, 39.7 bu/ac. 2025: 45.1, 44.8, 42.3, 49.6, 46.2 bu/ac.

**(a)** Save the two vectors and compute each field's change in yield in one line.

**(b)** Compute the mean change, and the mean percentage change (change over the 2024 yield, times 100).

**(c)** In a comment: every field improved, but which measure -- bushels gained or percent gained -- would you use to compare this farm against a lentil farm, and why?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) The two years, and each field change in one line
y2024 <- c(38.2, 41.6, 35.9, 43.1, 39.7)
y2025 <- c(45.1, 44.8, 42.3, 49.6, 46.2)
change <- y2025 - y2024
change

# (b) Mean change and mean percentage change
mean(change)
mean(change / y2024 * 100)

## I find a mean change of 5.9 bu/ac and a mean percentage change of 15.0%.
## (c) The percentage change: bushels gained are not comparable across crops
## measured on different scales (a lentil farm gains hundreds of lb/ac), but
## percent gained puts both farms on the same footing.
')
```

</details>

### Question 7

A farm's four barley fields: field IDs `"B1"` to `"B4"`, sizes 155, 240, 160, and 310 acres, yields 66.8, 71.2, 63.5, and 69.9 bu/ac.

**(a)** Build a data frame `barley_fields` holding the three variables.

**(b)** Using `$`, compute each field's production (acres times yield) and save it as a vector.

**(c)** Compute total production, and the farm's average yield weighted by acres (total production over total acres).

**(d)** In a comment: compare the weighted average with `mean(barley_fields$yield)` and say why they differ.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) The barley fields as a data frame
barley_fields <- data.frame(
  field_id = c("B1", "B2", "B3", "B4"),
  acres    = c(155, 240, 160, 310),
  yield    = c(66.8, 71.2, 63.5, 69.9)
)

# (b) Production of each field
production <- barley_fields$acres * barley_fields$yield
production

# (c) Total production and the acre-weighted average yield
sum(production)
sum(production) / sum(barley_fields$acres)

# (d) The plain mean, for comparison
mean(barley_fields$yield)

## I find total production of 59,271 bu and a weighted average of 68.52.
## (d) The plain mean is 67.85: it counts each field once, while the
## weighted average lets the 310-acre field count for its share of the
## land, and that field yields above average, pulling the weighted figure up.
')
```

</details>

### Question 8

**(a)** Build a data frame called `durum_fields` with columns `field_id` (`"D1"` to `"D4"`), `region` (South, South, Central, North), and `yield` (28.3, 41.6, 35.2, 44.8).

**(b)** Report its number of rows, number of columns, and column names using functions.

**(c)** Using `$`, compute the mean yield and the range (maximum minus minimum).

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Build the data frame
durum_fields <- data.frame(
  field_id = c("D1", "D2", "D3", "D4"),
  region = c("South", "South", "Central", "North"),
  yield = c(28.3, 41.6, 35.2, 44.8)
)

# (b) Number of rows
nrow(durum_fields)

# (b) Number of columns
ncol(durum_fields)

# (b) Column names
names(durum_fields)

# (c) Mean yield, using $
mean(durum_fields$yield)

# (c) Range (maximum minus minimum)
max(durum_fields$yield) - min(durum_fields$yield)

## (b) 4 rows, 3 columns; the columns are field_id, region, yield.
## (c) I find the mean to be 37.48 (the console shows 37.475) and the
## range 16.5.
')
```

</details>

### Question 9

A student runs these lines in the console:

```r
bins <- c(1450, 2200, 1875)
bins * 2
total <- sum(bins)
```

**(a)** In a comment: what exactly does the second line print, `[1]` included, and is that doubled vector saved anywhere?

**(b)** In a comment: what does the third line print, and what does the Variables pane show afterwards?

**(c)** The student now wants the doubled vector kept for later. Write the line that does it.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Run the lines to see what each one prints

# Create the vector of bin weights
bins <- c(1450, 2200, 1875)

# Double it, without saving the result
bins * 2

# Sum, saved to total (an assignment prints nothing)
total <- sum(bins)

# (c) Keep the doubled vector for later
bins_doubled <- bins * 2

# Print it to confirm it was saved
bins_doubled

## (a) The second line prints [1] 2900 4400 3750 -- displayed only, saved
## nowhere.
## (b) The third line prints nothing (it is an assignment); afterwards the
## Variables pane shows bins (the three values) and total = 5525.
## (c) bins_doubled <- bins * 2 keeps the doubled vector.
')
```

</details>

### Question 10

Here is a working script:

```r
v1 <- c(47.2, 49.8, 46.5, 50.1, 48.4)
v2 <- c(38.9, 55.6, 44.2, 58.3, 41.0)
mean(v1)
mean(v2)
```

**(a)** Rewrite it with a header and comments. The two vectors are canola yields from Farm A and Farm B.

**(b)** Add lines computing each farm's standard deviation and range.

**(c)** In a comment: the two farms have nearly the same mean. Using your numbers from (b), say in one sentence how they differ.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# ---
# Title: Comparing two canola farms
# Author: Your Name
# Date: 2026-10-08
# Description:
#   Compares the centre and spread of canola yields
#   on Farm A and Farm B.
# ---

# (a) Canola yields by field (bu/ac)
farm_a <- c(47.2, 49.8, 46.5, 50.1, 48.4)
farm_b <- c(38.9, 55.6, 44.2, 58.3, 41.0)

# Centre: mean yield on each farm
mean(farm_a)
mean(farm_b)

# (b) Spread: standard deviation on each farm
sd(farm_a)
sd(farm_b)

# (b) Spread: range (maximum minus minimum) on each farm
max(farm_a) - min(farm_a)
max(farm_b) - min(farm_b)

## (b) Farm A sd 1.57, range 3.6; Farm B sd 8.79, range 19.4.
## (c) The means are close (48.4 vs 47.6), but Farm B yields swing far more
## from field to field -- the spread on Farm B is five times the spread on
## Farm A.
')
```

</details>

### Question 11

Twelve canola fields yielded 24.1, 31.5, 28.8, 35.2, 22.7, 30.4, 27.9, 33.6, 25.5, 29.3, 32.8, and 26.4 bu/ac.

**(a)** Save the values as a vector and compute the mean.

**(b)** Compute the 10th, 50th, and 90th percentiles in a single `quantile()` call.

**(c)** In a comment: one of your three percentiles should equal the median. Which one, and does it?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Canola yields as a vector (bu/ac)
canola <- c(24.1, 31.5, 28.8, 35.2, 22.7, 30.4, 27.9, 33.6, 25.5, 29.3, 32.8, 26.4)

# (a) Mean yield
mean(canola)

# (b) 10th, 50th, and 90th percentiles in one call
quantile(canola, c(0.1, 0.5, 0.9))

# (c) The median, to check against the 50th percentile
median(canola)

## I find the mean to be 29.02.
## (b) 10th 24.24, 50th 29.05, 90th 33.52.
## (c) The 50th percentile is the median by definition; median(canola)
## confirms 29.05.
')
```

</details>

### Question 12

Consider these three lines:

```r
k <- 25
k == 30
mean(x = c(10, 20, 60))
```

**(a)** In a comment: say what each line does and what, if anything, it prints.

**(b)** In a comment: after the third line runs, does an object called `x` appear in the Variables pane? Why or why not?

**(c)** In a comment: which single symbol would you change to make line 2 *assign* 30 to `k`, and which does this book prefer for assignment?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Run the three lines to see what each one prints

# Line 1: assign 25 to k
k <- 25

# Line 2: ask whether k equals 30
k == 30

# Line 3: compute a mean, with the argument named
mean(x = c(10, 20, 60))

## (a) Line 1 assigns 25 to k and prints nothing. Line 2 asks whether k
## equals 30 and prints [1] FALSE. Line 3 computes a mean and prints [1] 30.
## (b) No -- the x = inside the call names an argument for mean(); it does
## not create an object.
## (c) == would become <- (or =, which also assigns) -- and the book uses <-
## for assignment.
')
```

</details>

### Question 13

Six pea fields yielded 38.4, 41.2, 35.7, 44.9, 39.8, and 42.6 bu/ac. The buyer pays a premium on fields at or above 40.

**(a)** Save the yields as a vector.

**(b)** Write one comparison that returns TRUE or FALSE for every field at once (`>=` on the whole vector).

**(c)** Count the premium fields by calling `sum()` on your comparison.

**(d)** In a comment: why does `sum()` of a logical vector count the TRUEs?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Pea yields
yields <- c(38.4, 41.2, 35.7, 44.9, 39.8, 42.6)

# (b) One comparison for every field at once
yields >= 40

# (c) Count the premium fields
sum(yields >= 40)

## I find 3 premium fields.
## (d) A logical vector is stored as TRUE = 1 and FALSE = 0, so summing it
## adds a 1 for every field that met the condition -- the sum is the count.
')
```

</details>

### Question 14

Four oat fields yielded 84.3, 91.7, 77.2, and 88.5 bu/ac, and oats are selling at $4.15 per bushel.

**(a)** Save the yields as a vector and the price as a single object.

**(b)** Compute a revenue-per-acre vector (yield times price) in one line, using the price object rather than retyping the number.

**(c)** Compute the mean revenue per acre.

**(d)** In a comment: the price rises to $4.60. What two lines would you rerun to update every answer, and why is that easier than if you had typed the price into each formula?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Oat yields (bu/ac) and the price ($/bu)
yields <- c(84.3, 91.7, 77.2, 88.5)
price  <- 4.15

# (b) Revenue per acre, using the price object
revenue <- yields * price

# Print the revenues
revenue

# (c) Mean revenue per acre
mean(revenue)

## (b) The revenues are 349.85, 380.56, 320.38, 367.28 $/ac, rounded to the
## cent (the console shows 349.845 and so on).
## (c) Mean revenue is $354.51 per acre.
## (d) Change price <- 4.60, then rerun the lines below it. The price lives
## in one object, so one edit updates everything -- the same reason a
## spreadsheet keeps a price in one cell.
')
```

</details>

# Section 2 — Reading and Inspecting Data

## Saskatchewan RM Canola {.unnumbered}

### Question 15

**(a)** Run `summary(rm_canola)`. From its output, report the minimum, median, and maximum of `Yield`, and the first and last `Year`.

**(b)** In a comment: work through the module's inspection questions for this file. Are the yields roughly between 10 and 60 bu/ac, a typical canola range? Are the years what the file name promises? Are there missing values?

**(c)** The minimum is below 2 bu/ac. In a comment: is a canola yield that low necessarily a data error? What kind of year could produce it?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the canola file
library(tidyverse)
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")

# (a) Summary of every column
summary(rm_canola)

## (a) I find Yield has minimum 1.9, median 26.9, and maximum 61.0, and Year runs from 1990 to 2025.
## (b) The yields all sit between about 2 and 61 bu/ac, a sensible canola range; the years run 1990--2025 as the file name promises; and summary() prints an NA count under a column only when it has missing values -- no such line appears, so there are none.
## (c) Not necessarily -- an RM average that low is what a severe drought or a widespread crop failure looks like. Extreme is not the same as wrong; it is a value to investigate, not delete.
')
```

</details>

### Question 16

**(a)** In a comment at the top of your script, sketch your project's folder layout (the subfolders and where the csv sits).

**(b)** Load the tidyverse and read the canola file into an object called `rm_canola`, using a relative path.

**(c)** Report the number of rows, the number of columns, and the column names using functions.

**(d)** In a comment: why will your `read_csv()` line work on the grader's computer, when a full path starting `C:/Users/...` would not?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Sketch of the project folder layout:
# project/
#   analysis.R
#   data/
#     rm_canola_yields_1990_2025.csv

# (b) Load the tidyverse
library(tidyverse)

# (b) Read the canola file with a relative path
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")

# (c) Number of rows
nrow(rm_canola)

# (c) Number of columns
ncol(rm_canola)

# (c) Column names
names(rm_canola)

## (c) I find the file has 10039 rows and 5 columns: Year, RM, Crop, Yield, Unit.
## (d) The path is relative to the project folder, and the grader opens the same self-contained folder -- so "data/..." exists there too. A full path starting "C:/Users/..." names the user and folder layout of one particular computer, so it fails everywhere else.
')
```

</details>

### Question 17

**(a)** Compute the mean, median, standard deviation, and variance of `rm_canola$Yield`.

**(b)** Compute the 90th percentile of `Yield`.

**(c)** The mean sits a little above the median. In a comment: using Module 1's language about means and medians, what does that direction of gap suggest about the shape of canola yields?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the canola file
library(tidyverse)
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")

# (a) Centre of the yields
mean(rm_canola$Yield)
median(rm_canola$Yield)

# (a) Spread of the yields
sd(rm_canola$Yield)
var(rm_canola$Yield)

# (b) 90th percentile of the yields
quantile(rm_canola$Yield, probs = 0.9)

## (a) I find a mean of 28.29, a median of 26.9, a standard deviation of 10.06, and a variance of 101.21 -- the sd squared, in squared units.
## (b) The 90th percentile is 42.5 bu/ac.
## (c) A mean above the median suggests mild right skew: the very best RM-years pull the mean up more than the worst years pull it down.
')
```

</details>

### Question 18

**(a)** Print the first six rows of `rm_canola` with `head()`.

**(b)** In a comment: from those rows, say what one row of this dataset represents.

**(c)** Report the number of rows, and compute the first and last year in the file (`range()` on the Year column with `$`).

**(d)** Divide the row count by the number of years. In a comment: the result is close to, but not exactly, the number of Saskatchewan RMs (about 295) -- what does that tell you about the file?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) First six rows
head(rm_canola)

# (c) Rows, and the span of years
nrow(rm_canola)
range(rm_canola$Year)

# (d) Rows per year
nrow(rm_canola) / 36

## I find 10,039 rows spanning 1990 to 2025.
## (b) One row is one RM in one year: that RM average canola yield.
## (d) 10,039 / 36 is about 279 rows per year, against roughly 295 RMs --
## so not every RM reports canola every year; a handful of RM-years are
## missing from the file.
')
```

</details>

### Question 19

**(a)** Run `glimpse()` on `rm_canola`. In a comment: which columns are text (`<chr>`) and which are numeric (`<dbl>`)?

**(b)** When `read_csv()` ran, it printed a note about the file. In a comment: what did that note report, and why does the module say to read it every time?

**(c)** `Year` came in as numeric (`<dbl>`). In a comment: which type of value is each of `2024` and `"2024"`? Why do we want a column we will compute with stored as numbers rather than as text?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the canola file
library(tidyverse)
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")

# (a) Structure: one line per column, with its type
glimpse(rm_canola)

## (a) Crop and Unit are text (<chr>); Year, RM, and Yield are numeric (<dbl>).
## (b) The note reported the row and column counts and the type read_csv guessed for each column -- here chr (2) and dbl (3). R guesses well but not perfectly, so the note is the first check that the file read the way you expect.
## (c) 2024 is a numeric value; "2024" is a character (text) value. Numbers stored as text stop behaving like numbers -- you cannot do arithmetic with them, and sorting turns alphabetical -- so a column we will compute with should show <dbl>, not <chr>.
')
```

</details>

### Question 20

**(a)** Using `$`, compute the mean, the standard deviation, and the 10th and 90th percentiles of `Yield` across the whole file.

**(b)** In a comment: an RM-year at the 10th percentile grew how much canola? Interpret the number for a grower.

**(c)** Compute the gap between the 90th and 10th percentiles. In a comment: what does this gap describe that the standard deviation also tries to describe?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Mean, sd, and the 10th and 90th percentiles
mean(rm_canola$Yield)
sd(rm_canola$Yield)
quantile(rm_canola$Yield, probs = c(0.10, 0.90))

# (c) The 90-10 gap
42.5 - 16.1

## I find a mean of 28.29, an sd of 10.06, and percentiles of 16.1 and 42.5.
## (b) An RM-year at the 10th percentile grew 16.1 bu/ac -- nine in ten
## RM-years across the file did better than that.
## (c) The gap is 26.4 bu/ac. Like the sd it describes how spread out
## yields are, but it reads directly as a bushel distance holding the
## middle 80% of RM-years, and it ignores the extremes entirely.
')
```

</details>

### Question 21

In a fresh R session you run `rm_yields <- read_csv("data/rm_yields_1990_2025.csv")` and get:

```
Error in read_csv(...) : could not find function "read_csv"
```

**(a)** In a comment: what is the most likely cause, and what one line fixes it?

**(b)** In a comment: you installed the tidyverse only last week -- why does R still not find the function? Explain the difference between installing and loading, and how often each is done.

**(c)** In a comment: state the quotes rule -- where do quotation marks go when installing, and when loading?

**(d)** The tidyverse is a collection of packages. In comments, name the four core ones from the module and the job each does.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) The most likely cause: read_csv lives in a package that is not loaded
#     in this fresh session, so R cannot find the function. One line fixes it:
library(tidyverse)

# (a) With the package loaded, the same read line now works
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

## (a) The package that provides read_csv was not loaded in the session; running library(tidyverse) first fixes it.
## (b) Installing downloads the package onto the computer -- that happens once per computer. Loading switches it on -- that happens once per script or session. A package installed last week is still installed, but a fresh session starts with nothing loaded.
## (c) Quotes when you install, none when you load: install.packages("tidyverse") but library(tidyverse).
## (d) The four core packages: readr reads data, dplyr manipulates it, tidyr reshapes it, ggplot2 makes graphs.
')
```

</details>

## Manitoba Wheat Variety {.unnumbered}

### Question 22

**(a)** In a comment at the top of your script, sketch your project's folder layout.

**(b)** Load the tidyverse and read the file into an object called `mb_wheat`, using a relative path.

**(c)** Report the number of rows, the number of columns, and the column names using functions.

**(d)** In a comment: what does one row of this dataset represent?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Sketch of the project folder layout:
# project/
#   analysis.R
#   data/
#     mb_wheat_reported_2020_2025.csv

# (b) Load the tidyverse
library(tidyverse)

# (b) Read the wheat file with a relative path
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (c) Number of rows
nrow(mb_wheat)

# (c) Number of columns
ncol(mb_wheat)

# (c) Column names
names(mb_wheat)

## (c) I find the file has 2398 rows and 7 columns: Year, Municipality, Variety, Farms, Acres, Yield_bu_ac, Reported.
## (d) One row is one wheat variety grown in one municipality in one year -- its farm count, its acres, and its average yield.
')
```

</details>

### Question 23

**(a)** Run `glimpse()` on `mb_wheat`. In a comment: which columns are text (`<chr>`), which numeric (`<dbl>`), and which logical (`<lgl>`)?

**(b)** In a comment: what values does the logical column hold in this file, and how does the file's name explain this?

**(c)** In a comment: `Farms` holds whole numbers. Would `glimpse()` showing it as `<chr>` be a problem, and what would it tell you about the file?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the wheat file
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) Structure: one line per column, with its type
glimpse(mb_wheat)

# (b) The values the logical column holds
summary(mb_wheat$Reported)

## (a) Municipality and Variety are text (<chr>); Year, Farms, Acres, and Yield_bu_ac are numeric (<dbl>); Reported is logical (<lgl>).
## (b) Reported holds only TRUE -- all 2398 of them. This is the reported-only file, with the suppressed (FALSE) rows removed, which is what the file name promises.
## (c) Yes, it would be a problem -- a numeric column arriving as text means some value in it could not be read as a number, and calculations like mean() would fail on it.
')
```

</details>

### Question 24

**(a)** Run `summary(mb_wheat)`. From its output, report the minimum, median, and maximum of `Yield_bu_ac`, and the first and last `Year`.

**(b)** In a comment: work through the module's inspection questions. Are the yields mostly between 30 and 100 bu/ac, a typical spring wheat range? Are there missing values?

**(c)** The minimum is **4.5 bu/ac**. In a comment: is a reported yield that low necessarily an error?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the wheat file
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) Summary of every column
summary(mb_wheat)

## (a) I find Yield_bu_ac has minimum 4.5, median 62.2, and maximum 97.8, and Year runs from 2020 to 2025.
## (b) Plausible -- most yields sit in the tens of bushels, sensible for Manitoba spring wheat; and no NA count lines appear, so there are no missing values in this file.
## (c) Not necessarily -- a variety grown on a few drought- or flood-hit fields can genuinely average that low. Extreme is a value to investigate, not delete.
')
```

</details>

### Question 25

**(a)** Compute the mean, median, standard deviation, and variance of `mb_wheat$Yield_bu_ac`.

**(b)** In a comment: the mean sits a little *below* the median here. In Module 1's language, what does that direction of gap suggest, and what kind of observations would produce it?

**(c)** Compute the smallest and largest yields in one `range()` call.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the wheat file
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) Centre of the yields
mean(mb_wheat$Yield_bu_ac)
median(mb_wheat$Yield_bu_ac)

# (a) Spread of the yields
sd(mb_wheat$Yield_bu_ac)
var(mb_wheat$Yield_bu_ac)

# (c) Smallest and largest yield in one call
range(mb_wheat$Yield_bu_ac)

## (a) I find a mean of 61.16, a median of 62.2, a standard deviation of 12.38, and a variance of 153.34.
## (b) A mean below the median suggests mild left skew: unusually low values pull the mean down more than the high values pull it up -- a tail of very poor variety-municipality observations mixed in with the normal ones.
## (c) range() returns both ends as one vector: the smallest yield is 4.5 and the largest is 97.8 bu/ac.
')
```

</details>

### Question 26

Suppose your script reads the data like this:

```r
mb_wheat <- read_csv("/Users/sam/Downloads/mb wheat data.csv")
```

It works on your laptop and fails on everyone else's.

**(a)** In a comment: give two problems -- one about where the path points, one about the file's name and location.

**(b)** Rewrite the line the way the module recommends, assuming the file was saved as `mb_wheat_reported_2020_2025.csv` in the project's `data/` folder.

**(c)** In a comment: your line fails with `does not exist in current working directory`. What is the first thing to check?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# (b) The read line rewritten the way the module recommends: the file renamed,
#     moved into the data folder of the project, and read with a relative path
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

## (a) Two problems. First, the path is absolute: it names the Downloads folder of one user on one machine, so it exists nowhere else. Second, the file is loose in Downloads under a name with spaces in it, instead of renamed and placed in the data folder of the project.
## (c) Check that the folder opened in Positron is the project folder -- the one containing data/. Relative paths start from the opened folder.
')
```

</details>

### Question 27

**(a)** Using `$`, compute the mean and median of `Acres` across the file.

**(b)** In a comment: the mean is far above the median. In Module 1's language, what shape does that imply for the acreage distribution, and what kind of observations produce it?

**(c)** Compute the 90th percentile of `Acres`. In a comment: interpret it in one sentence.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Mean and median acres
mean(mb_wheat$Acres)
median(mb_wheat$Acres)

# (c) The 90th percentile of acres
quantile(mb_wheat$Acres, probs = 0.90)

## I find a mean of 6,118 acres against a median of 2,941.
## (b) A mean more than double the median is strong right skew: most
## variety-municipality rows cover a few thousand acres, while a small
## number of very large plantings pull the mean far above the middle row.
## (c) The 90th percentile is 15,065 acres -- one row in ten covers more
## than that, which is where the skew lives.
')
```

</details>

### Question 28

**(a)** Using `$`, compute the minimum, mean, and maximum of `Farms`.

**(b)** In a comment: what does one unit of this column mean -- what is being counted?

**(c)** In a comment: a variety-municipality average resting on the minimum number of farms is a different kind of number from one resting on fifty. Say why a careful reader would want the farm count beside a yield.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Smallest, average, and largest farm counts
min(mb_wheat$Farms)
mean(mb_wheat$Farms)
max(mb_wheat$Farms)

## I find farm counts running from 3 to 148, averaging 15.4.
## (b) One unit is one farm that grew that variety in that municipality
## that year -- the column counts the growers behind each row.
## (c) A yield resting on 3 farms can be moved a long way by one grower
## having a good or bad year; a yield resting on fifty is close to the
## municipal experience. The farm count tells the reader which kind of
## average they are looking at.
')
```

</details>

# Section 3 — Data Manipulation Functions

## Saskatchewan RM Crop Yields {.unnumbered}

### Question 29

**(a)** Filter to the Spring Wheat observations of at least 60 bu/ac (all years), using `>=` and saving the result. How many rows are in the filtered data frame?

**(b)** Filter again to keep only the 2023 rows of that result. How many rows are in this second data frame?

**(c)** From your 2023 data frame, select `RM` and `Yield` and sort best-first. Which RM tops the list, at what yield?

**(d)** In a comment: rewrite the condition "yield at least 60 and the crop is *not* Spring Wheat" using `!`.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Spring Wheat at 60 bu/ac or better, all years
sw_60 <- filter(rm_yields, Crop == "Spring Wheat" & Yield >= 60)
# Row count of the filtered data frame
nrow(sw_60)

# (b) Keep only the 2023 rows of that result
sw_60_2023 <- filter(sw_60, Year == 2023)
# Row count of the second data frame
nrow(sw_60_2023)

# (c) Just RM and Yield, sorted best first
sw_best <- select(sw_60_2023, RM, Yield)
arrange(sw_best, desc(Yield))

## I find 434 rows at 60 bu/ac or better, and 53 of them are from 2023.
## RM 187 tops the 2023 list, at 74.7 bu/ac.
## (d) filter(rm_yields, Yield >= 60 & !Crop == "Spring Wheat") -- the ! negates the crop condition.
')
```

</details>

### Question 30

**(a)** Filter to rows where the crop is Oats **or** Barley, using `|`, saving the result. How many rows are in the filtered data frame?

**(b)** Write the same filter using `%in%` and confirm the row count matches.

**(c)** Sort your filtered data frame from highest yield to lowest. What crop, RM, and year appear in the top row, and at what yield?

**(d)** What would `filter(rm_yields, Crop == "Oats" & Crop == "Barley")` return? In a comment: explain why.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Oats or Barley, using the | connector
oats_barley <- filter(rm_yields, Crop == "Oats" | Crop == "Barley")
# Row count of the filtered data frame
nrow(oats_barley)

# (b) The same filter written with %in%
oats_barley2 <- filter(rm_yields, Crop %in% c("Oats", "Barley"))
# The count should match (a)
nrow(oats_barley2)

# (c) Sort from highest yield to lowest
arrange(oats_barley, desc(Yield))

## I find 19709 rows both ways.
## The top of the sorted table is a tie at 165.1 bu/ac: Oats in RM 369 in 2025 and Oats in RM 438 in 2020 share the maximum, so either is a correct top row.
## (d) The & version asks for rows whose crop is Oats and Barley at the same time. No single row can satisfy both conditions, so zero rows come back. "Or" is the right connector, written |.
')
```

</details>

### Question 31

**(a)** Filter to Canola in 2024, then in one `mutate()` call add two columns: `yield_t_ha` (multiply by 0.0560) and `yield_kg_ha` (multiply by 56.0). Save the result of each step.

**(b)** Compute the mean of each new column.

**(c)** Sort by `yield_t_ha`, best first. Which RM tops the table, and at what yield in t/ha?

**(d)** In a comment: without running anything, what is the mean of `yield_kg_ha` divided by the mean of `yield_t_ha`, and why?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Canola in 2024
canola_2024 <- filter(rm_yields, Crop == "Canola" & Year == 2024)
# Add both converted columns in one mutate() call
canola_2024 <- mutate(canola_2024,
                      yield_t_ha  = Yield * 0.0560,  # tonnes per hectare
                      yield_kg_ha = Yield * 56.0)    # kilograms per hectare

# (b) Mean of each new column
mean(canola_2024$yield_t_ha)
mean(canola_2024$yield_kg_ha)

# (c) Sort by tonnes per hectare, best first
arrange(canola_2024, desc(yield_t_ha))

## I find means of 1.76 t/ha and 1764.11 kg/ha.
## RM 493 tops the table, at 2.57 t/ha.
## (d) The ratio of the means is 1000: both columns are the same yields scaled by constants, and 56.0 / 0.0560 = 1000 (a tonne is 1000 kg).
')
```

</details>

### Question 32

Canola is selling at $14.20 per bushel.

**(a)** Filter to Canola in 2025 and, in one `mutate()`, add a `revenue_per_ac` column (yield times price). Save the result.

**(b)** Sort by revenue, best first, and report the top RM and its revenue per acre.

**(c)** Compute the mean revenue per acre.

**(d)** In a comment: every RM's rank by revenue is identical to its rank by yield here. Why, and what would have to differ across RMs for the two rankings to disagree?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# Price object
price <- 14.20

# (a) 2025 canola with revenue per acre
canola_25 <- filter(rm_yields, Crop == "Canola" & Year == 2025)
canola_25 <- mutate(canola_25, revenue_per_ac = Yield * price)

# (b) Best revenue first
arrange(canola_25, desc(revenue_per_ac))

# (c) Mean revenue per acre
mean(canola_25$revenue_per_ac)

## I find RM 287 on top at $866.20 per acre, and a mean of $623.94.
## (d) Revenue is yield times the same price for every RM, so multiplying
## by a constant cannot reorder anyone. The rankings would only disagree if
## the price differed across RMs -- different grades, hauling deductions,
## or locally negotiated prices.
')
```

</details>

### Question 33

**(a)** Filter `rm_yields` to Canola in 2024, saving the result as `canola_2024`. How many rows are in the filtered data frame?

**(b)** From `canola_2024`, select only the `RM` and `Yield` columns, saving the result. Report its column names and its number of columns.

**(c)** In a comment: after (a) and (b), how many rows does `rm_yields` itself have, and why?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Keep only the Canola rows from 2024
canola_2024 <- filter(rm_yields, Crop == "Canola" & Year == 2024)
# Row count of the filtered data frame
nrow(canola_2024)

# (b) Keep only the RM and Yield columns
canola_slim <- select(canola_2024, RM, Yield)
# Column names of the slimmed version
names(canola_slim)
# Number of columns
ncol(canola_slim)

# (c) The original data frame, after (a) and (b)
nrow(rm_yields)

## I find 293 rows of 2024 canola.
## The selected version has the columns RM and Yield, so 2 columns (and still 293 rows).
## rm_yields itself still has 71104 rows: filter() and select() return new data frames and leave the original untouched.
')
```

</details>

### Question 34

**(a)** Make a version of the data with only `Year`, `RM`, `Crop`, and `Yield` -- once by *naming what you keep*, once by *dropping* what you don't want -- saving each, and confirm the two have the same column names. Then make a third version keeping the adjacent-column range `Year:Crop`, and report its column names.

**(b)** Rename `Yield` to `yield_per_acre` in the full data, saving the result, and show the new column names.

**(c)** Filter your renamed data frame to Peas in 2025. How many rows are in the filtered data frame?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Keep four columns by naming them
kept <- select(rm_yields, Year, RM, Crop, Yield)
# The same four columns by dropping Unit
dropped <- select(rm_yields, -Unit)
# Confirm the two have the same column names
names(kept)
names(dropped)
# The adjacent-column range Year:Crop
ranged <- select(rm_yields, Year:Crop)
# Its column names
names(ranged)

# (b) Rename Yield in the full data
renamed <- rename(rm_yields, yield_per_acre = Yield)
# The new column names
names(renamed)

# (c) Peas in 2025, from the renamed data frame
peas_2025 <- filter(renamed, Crop == "Peas" & Year == 2025)
# Row count of the filtered data frame
nrow(peas_2025)

## Both versions have the columns Year, RM, Crop, Yield; the range version keeps Year, RM, Crop.
## After the rename the columns are Year, RM, Crop, yield_per_acre, Unit.
## I find 285 rows of 2025 peas.
')
```

</details>

### Question 35

**(a)** Filter `rm_yields` to Canola from 2020 through 2025 in one `filter()` call (two conditions on `Year` joined with `&`), saving the result. How many rows are in it?

**(b)** Sort the result from highest yield to lowest. Which RM and year top the list, at what yield?

**(c)** What would the same filter written with `|` between the two year conditions return? In a comment: explain why.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Canola, 2020 through 2025
canola_recent <- filter(rm_yields, Crop == "Canola" & Year >= 2020 & Year <= 2025)
nrow(canola_recent)

# (b) Sorted best-first
arrange(canola_recent, desc(Yield))

## I find 1,743 rows; RM 287 in 2025 tops the list at 61.0 bu/ac.
## (c) With | between the year conditions every row passes -- every year is
## either at least 2020 or at most 2025 -- so the filter would keep all
## canola rows instead of the six recent years.
')
```

</details>

## Manitoba Wheat Variety {.unnumbered}

### Question 36

**(a)** From `mb_rep`, filter to rows with at least 10 farms **and** a yield above 70 bu/ac, saving the result. How many rows are in it?

**(b)** Sort best-first and report the top municipality, variety, and year.

**(c)** In a comment: compare this with filtering on yield alone -- what does the farm-count condition add to a claim that a variety "yields above 70 here"?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb <- read_csv("data/mb_wheat_varieties.csv")
mb_rep <- filter(mb, Reported == TRUE)

# (a) At least 10 farms AND yield above 70
solid_high <- filter(mb_rep, Farms >= 10 & Yield_bu_ac > 70)
nrow(solid_high)

# (b) Best first
arrange(solid_high, desc(Yield_bu_ac))

## I find 273 rows; LOUISE growing SY MANNESS in 2025 tops the list
## at 96.3 bu/ac.
## (c) Yield alone can be one grower and one lucky field. Requiring ten or
## more farms means every row here is a result many growers produced
## together, so "this variety yields above 70 here" rests on evidence
## rather than an anecdote.
')
```

</details>

### Question 37

**(a)** From `mb_rep`, use one `mutate()` to add `total_bu`, each observation's acres times its yield, saving the result.

**(b)** Compute the mean of `total_bu`.

**(c)** Sort by `total_bu`, largest first. Which municipality tops the table, and at roughly what total?

**(d)** In a comment: what does `total_bu` measure that `Yield_bu_ac` does not?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full Manitoba variety file, then keep the reported rows
mb <- read_csv("data/mb_wheat_varieties.csv")
mb_rep <- filter(mb, Reported == TRUE)

# (a) Total bushels: acres times yield, in one mutate() call
mb_rep <- mutate(mb_rep, total_bu = Acres * Yield_bu_ac)

# (b) Mean of the new column
mean(mb_rep$total_bu)

# (c) Sort by total production, largest first
arrange(mb_rep, desc(total_bu))

## I find a mean of about 382865 bu.
## PORTAGE LA PRAIRIE tops the table, at about 5.13 million bushels.
## (d) total_bu measures production -- how much grain the planting actually produced -- rather than how well each acre yielded. A modest yield on enormous acres can out-produce a great yield on few.
')
```

</details>

### Question 38

**(a)** From the full `mb` file, filter to the suppressed rows (`Reported == FALSE`), saving the result. How many rows are in it?

**(b)** Select `Year`, `Municipality`, and `Variety` from the result (the numeric columns are empty in these rows) and print its first six rows.

**(c)** In a comment: what do these rows represent, and why does keeping them in the file -- rather than deleting them -- tell an analyst something the reported rows alone cannot?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb <- read_csv("data/mb_wheat_varieties.csv")

# (a) The suppressed rows
suppressed <- filter(mb, Reported == FALSE)
nrow(suppressed)

# (b) Where and what they are
suppressed_ids <- select(suppressed, Year, Municipality, Variety)
head(suppressed_ids)

## I find 2,960 suppressed rows -- more than half the file.
## (c) Each one is a variety that was grown in that municipality that year
## by too few farms for the result to be published. Kept in the file, they
## show where wheat was grown thinly; the reported rows alone would make
## the thinly-grown varieties look like they were not grown at all.
')
```

</details>

### Question 39

**(a)** Filter `mb` to the reported rows (`Reported == TRUE`), saving as `mb_rep`. How many rows are in the filtered data frame?

**(b)** From `mb_rep`, select `Municipality`, `Variety`, and `Yield_bu_ac`, saving the result. Report its column names.

**(c)** In a comment: how many rows does `mb` itself still have, and what happened to the suppressed rows in `mb_rep`?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full Manitoba variety file, suppressed rows included
mb <- read_csv("data/mb_wheat_varieties.csv")

# (a) Keep the reported rows only
mb_rep <- filter(mb, Reported == TRUE)
# Row count of the filtered data frame
nrow(mb_rep)

# (b) Three columns from the reported rows
mb_slim <- select(mb_rep, Municipality, Variety, Yield_bu_ac)
# Its column names
names(mb_slim)

# (c) The full file, after (a) and (b)
nrow(mb)

## I find 2398 reported rows.
## The selected columns are Municipality, Variety, Yield_bu_ac.
## mb itself still has 5358 rows -- filtering returns a new data frame. The 2960 suppressed rows are simply absent from mb_rep; they were not deleted from anything.
')
```

</details>

### Question 40

**(a)** From `mb_rep` (reported rows), filter to the varieties `AAC BRANDON (BW 932)` **or** `AAC STARBUCK <SECAN>` using `|`, saving the result. How many rows are in the filtered data frame?

**(b)** Write the same filter using `%in%` and confirm the count matches.

**(c)** What would the same condition written with `&` instead of `|` return? In a comment: explain why.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full Manitoba variety file, then keep the reported rows
mb <- read_csv("data/mb_wheat_varieties.csv")
mb_rep <- filter(mb, Reported == TRUE)

# (a) The two varieties, using the | connector
two_var <- filter(mb_rep, Variety == "AAC BRANDON (BW 932)" | Variety == "AAC STARBUCK <SECAN>")
# Row count of the filtered data frame
nrow(two_var)

# (b) The same filter written with %in%
two_var2 <- filter(mb_rep, Variety %in% c("AAC BRANDON (BW 932)", "AAC STARBUCK <SECAN>"))
# The count should match (a)
nrow(two_var2)

## I find 893 rows both ways.
## (c) The & version demands both conditions true of the same row, and no single row has a variety that is both names at once. "Or" is the right connector, written |.
')
```

</details>

### Question 41

**(a)** From `mb_rep`, make a version with only `Year`, `Municipality`, `Variety`, and `Yield_bu_ac` -- once by naming what you keep, once by dropping what you don't (`Farms`, `Acres`, `Reported`) -- saving each, and confirm the two have the same column names. Then make a third version keeping the adjacent range `Year:Variety` and report its names.

**(b)** Rename `Yield_bu_ac` to `yield`, saving the result, and show the new column names.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full Manitoba variety file, then keep the reported rows
mb <- read_csv("data/mb_wheat_varieties.csv")
mb_rep <- filter(mb, Reported == TRUE)

# (a) Keep four columns by naming them
kept <- select(mb_rep, Year, Municipality, Variety, Yield_bu_ac)
# The same four columns by dropping the rest
dropped <- select(mb_rep, -Farms, -Acres, -Reported)
# Confirm the two have the same column names
names(kept)
names(dropped)
# The adjacent-column range Year:Variety
ranged <- select(mb_rep, Year:Variety)
# Its column names
names(ranged)

# (b) Rename the yield column
renamed <- rename(mb_rep, yield = Yield_bu_ac)
# The new column names
names(renamed)

## Both versions have the same four names, Year, Municipality, Variety, Yield_bu_ac; the range version keeps Year, Municipality, Variety.
## The renamed data frame has the same seven columns, with yield in place of Yield_bu_ac.
')
```

</details>

### Question 42

**(a)** From `mb_rep`, use one `mutate()` to add `acres_per_farm` (acres divided by farms), saving the result.

**(b)** Sort by `acres_per_farm`, largest first, and report the top municipality and variety.

**(c)** Compute the mean of `acres_per_farm`.

**(d)** In a comment: what does this column measure, and why is it a different fact about a variety than its total acres?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb <- read_csv("data/mb_wheat_varieties.csv")
mb_rep <- filter(mb, Reported == TRUE)

# (a) Acres per farm
mb_rep <- mutate(mb_rep, acres_per_farm = Acres / Farms)

# (b) Largest first
arrange(mb_rep, desc(acres_per_farm))

# (c) The average across rows
mean(mb_rep$acres_per_farm)

## I find MOUNTAIN growing SY MANNESS on top, at 1,587 acres per farm,
## against an average of 370.
## (d) It measures how concentrated a variety is in the hands of its
## growers -- few farms each seeding a lot, or many farms dabbling. Total
## acres says how much land; acres per farm says how it is held.
')
```

</details>

# Section 4 — Pipelines and Grouped Summaries

## Saskatchewan RM Crop Yields {.unnumbered}

### Question 43

**(a)** In one pipeline: for **2025 only**, compute each crop's mean yield and observation count, sorted from highest mean to lowest, and save the result as `crop_summary_2025`.

**(b)** Write the table to `output/crop_summary_2025.csv`.

**(c)** Report the two highest-yielding crops in the sorted table. In a comment: check the Unit column -- why is the first-place crop not comparable to the rest, and which bushel crop is the real standout?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Each crop in 2025: mean yield and observation count, best first
crop_summary_2025 <- rm_yields |>
  filter(Year == 2025) |>                      # 2025 only
  group_by(Crop) |>                            # one group per crop
  summarise(mean_yield = mean(Yield),          # average yield
            n = n()) |>                        # observation count
  arrange(desc(mean_yield))                    # highest mean first

# Print the sorted table
crop_summary_2025

# (b) Write the table to a csv
write_csv(crop_summary_2025, "output/crop_summary_2025.csv")

## I find Lentils on top at 1897 and Oats second at 97.5.
## (c) The Unit column shows Lentils are recorded in pounds per acre, not bushels, so the first place row is not comparable to the rest. Among the bushel crops, Oats are the real standout at 97.5 bu/ac, with Barley next at 72.8. 2025 was a strong year across the board.
')
```

</details>

### Question 44

**(a)** In one pipeline: filter to Lentils, convert to kg/ha with `mutate()` (multiply by 1.12), then summarise the mean of the converted column.

**(b)** Report the mean in kg/ha.

**(c)** In a comment: in what order do your steps run, and why would putting the `summarise()` before the `mutate()` fail?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Lentils, converted to kg/ha, then the mean of the converted column
lentil_summary <- rm_yields |>
  filter(Crop == "Lentils") |>                 # lentil rows only
  mutate(yield_kg_ha = Yield * 1.12) |>        # lb/ac times 1.12 gives kg/ha
  summarise(mean_kg_ha = mean(yield_kg_ha))    # average of the new column

# Print the summary table
lentil_summary

# The mean to more decimals than the tibble display shows
lentil_summary$mean_kg_ha

## (b) I find a mean of 1353.4 kg/ha.
## (c) The steps run top to bottom, each one feeding its output to the next. Putting summarise() before mutate() would fail because the data would already be collapsed to one row and the yield_kg_ha column would not exist yet.
')
```

</details>

### Question 45

Canola is selling at $14.20 per bushel.

**(a)** In one pipeline: filter to Canola 2023-2025, add `revenue_per_ac` with `mutate()`, then compute the mean revenue for each year.

**(b)** Report the three yearly means.

**(c)** In a comment: your pipeline has four steps. Say what each hands to the next, in order.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# Price object
price <- 14.20

# (a) Mean canola revenue per acre by year, 2023-2025
rm_yields |>
  filter(Crop == "Canola" & Year >= 2023) |>      # recent canola rows
  mutate(revenue_per_ac = Yield * price) |>       # dollars per acre
  group_by(Year) |>                               # one group per year
  summarise(mean_revenue = mean(revenue_per_ac))  # average within each year

## I find mean revenues of $481.11 in 2023, $447.33 in 2024, and $623.94
## in 2025.
## (c) filter hands the recent canola rows to mutate; mutate hands the same
## rows plus the revenue column to group_by; group_by hands the grouped
## rows to summarise, which collapses each year to one number.
')
```

</details>

### Question 46

**(a)** In one pipeline, compute the mean canola yield for **each year** from 2021 to 2025 (filter first, then group).

**(b)** Report the five means. Which year was worst, and which best?

**(c)** In a comment: 2021 was a severe drought year in Saskatchewan. Does the data agree?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Mean canola yield for each year from 2021 to 2025
rm_yields |>
  filter(Crop == "Canola" & Year >= 2021) |>   # canola, 2021 onward
  group_by(Year) |>                            # one group per year
  summarise(mean_yield = mean(Yield))          # average within each year

## I find mean canola yields of 21.9 in 2021, 35.4 in 2022, 33.9 in 2023, 31.5 in 2024, and 43.9 in 2025. The worst year was 2021 and the best was 2025.
## (c) Yes, the data agrees: the 2021 mean sits roughly a third below the surrounding years.
')
```

</details>

### Question 47

**(a)** In one pipeline, compute the mean flax yield across 2020--2025 (`Year >= 2020`), calling the result column `mean_yield`.

**(b)** Add `n()` to the same `summarise()` to count how many observations that mean is based on.

**(c)** In a comment: why is reporting the count alongside the mean good practice?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a, b) Mean flax yield across 2020 to 2025, with the observation count
flax_summary <- rm_yields |>
  filter(Crop == "Flax" & Year >= 2020) |>     # flax from 2020 onward
  summarise(mean_yield = mean(Yield),          # average yield
            n = n())                           # rows behind the mean

# Print the summary table
flax_summary

# The mean to more decimals than the tibble display shows
flax_summary$mean_yield

## I find a mean of 22.39 bu/ac from 1143 observations.
## (c) A mean based on a handful of rows deserves less trust than one based on a thousand; the count tells the reader which they are looking at.
')
```

</details>

### Question 48

**(a)** Write a pipeline, one step per line with a short comment on each step, that filters `rm_yields` to Durum in 2025, keeps only the `RM` and `Yield` columns, and sorts the result from highest yield to lowest.

**(b)** Run it. How many rows are in the resulting data frame, and which RM had the best durum yield in 2025, at what value?

**(c)** In a comment: read your pipeline aloud as a "take the data, then ..., then ..., then ..." sentence.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Durum in 2025, two columns, best first
durum_2025 <- rm_yields |>                     # take the data, THEN
  filter(Crop == "Durum" & Year == 2025) |>    # keep durum in 2025, THEN
  select(RM, Yield) |>                         # keep two columns, THEN
  arrange(desc(Yield))                         # sort best first

# (b) How many rows the result has
nrow(durum_2025)

# The top of the sorted table
head(durum_2025)

# The top yields to one decimal, since the tibble display rounds
head(durum_2025$Yield)

## I find 189 rows; RM 369 had the best durum yield in 2025, at 105.6 bu/ac.
## (c) Take rm_yields, then keep the 2025 durum rows, then keep the RM and Yield columns, then sort from highest yield to lowest.
')
```

</details>

### Question 49

**(a)** In one pipeline: filter to Spring Wheat in 2025, compute the mean, the standard deviation, and `n()` in one `summarise()`.

**(b)** Report the three values.

**(c)** In a comment: what does the standard deviation describe here -- variation across what units, over what period?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) 2025 spring wheat: mean, sd, and count in one summarise
rm_yields |>
  filter(Crop == "Spring Wheat" & Year == 2025) |>   # 2025 spring wheat rows
  summarise(mean_yield = mean(Yield),                # average across RMs
            sd_yield = sd(Yield),                    # spread across RMs
            n = n())                                 # observations behind them

## I find a mean of 52.96 and an sd of 14.67 from 283 RMs.
## (c) The sd describes how much RM average yields varied from RM to RM
## across Saskatchewan in that one year -- geography, not time.
')
```

</details>

## Manitoba Wheat Variety {.unnumbered}

### Question 50

**(a)** In one pipeline, compute each variety's total acres from `mb_rep` (`sum(Acres)`), sorted largest first.

**(b)** Report the top variety and its total.

**(c)** In a comment: totalling acres makes sense, but what would totalling `Yield_bu_ac` produce, and why is that number meaningless?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb <- read_csv("data/mb_wheat_varieties.csv")
mb_rep <- filter(mb, Reported == TRUE)

# (a) Total acres by variety, largest first
mb_rep |>
  group_by(Variety) |>                    # one group per variety
  summarise(total_acres = sum(Acres)) |>  # acres are amounts, so they add
  arrange(desc(total_acres))              # biggest variety first

## I find AAC BRANDON (BW 932) on top at about 6.74 million acres, with
## AAC STARBUCK second at about 2.50 million.
## (c) Totalling Yield_bu_ac would add per-acre rates across rows -- a
## number with no physical meaning, since a rate says how much each acre
## gave, not how much there was. Acres are an amount, so their total is a
## real quantity of land.
')
```

</details>

### Question 51

**(a)** In one pipeline, compute each variety's mean yield and observation count across `mb_rep`, sorted from highest mean to lowest.

**(b)** Report the top two varieties, their means, and their counts.

**(c)** In a comment: the second-place variety was observed exactly once. Which of the two means would you trust more, and why?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full variety file
mb <- read_csv("data/mb_wheat_varieties.csv")

# Keep only the reported rows
mb_rep <- filter(mb, Reported == TRUE)

# (a) Each variety: mean yield and observation count, best first
mb_rep |>
  group_by(Variety) |>                         # one group per variety
  summarise(mean_yield = mean(Yield_bu_ac),    # average yield
            n = n()) |>                        # observation count
  arrange(desc(mean_yield))                    # highest mean first

## (b) I find AAC WESTKING first at 79.2 bu/ac from 17 observations, then CDC GO (BW781) at 77.5 from a single observation.
## (c) I would trust the AAC WESTKING mean more: seventeen municipality-years is evidence, while a single observation could be one lucky field in one good year.
')
```

</details>

### Question 52

**(a)** Write a pipeline, one step per line with a short comment on each, that filters `mb_rep` to 2025, keeps `Municipality`, `Variety`, and `Yield_bu_ac`, and sorts best-first.

**(b)** Run it. How many rows are in the resulting data frame, and which municipality and variety top it, at what yield?

**(c)** In a comment: read the pipeline aloud as a "take the data, then ..." sentence.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full variety file
mb <- read_csv("data/mb_wheat_varieties.csv")

# Keep only the reported rows
mb_rep <- filter(mb, Reported == TRUE)

# (a) Reported 2025 rows, three columns, best first
wheat_2025 <- mb_rep |>                              # take the reported rows, THEN
  filter(Year == 2025) |>                            # keep 2025, THEN
  select(Municipality, Variety, Yield_bu_ac) |>      # keep three columns, THEN
  arrange(desc(Yield_bu_ac))                         # sort best first

# (b) How many rows the result has
nrow(wheat_2025)

# The top of the sorted table
head(wheat_2025)

## I find 438 rows; LOUISE, growing SY MANNESS, tops the table at 96.3 bu/ac.
## (c) Take the reported rows, then keep 2025, then keep three columns, then sort from highest yield to lowest.
')
```

</details>

### Question 53

**(a)** In one pipeline: for **2025 only**, compute each variety's mean reported yield and observation count, sorted best-first, and save the result as `variety_summary_2025`.

**(b)** Write the table to `output/variety_summary_2025.csv`.

**(c)** Report the top row. In a comment: the table also contains varieties observed only twice -- what should a reader keep in mind when comparing rows?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full variety file
mb <- read_csv("data/mb_wheat_varieties.csv")

# Keep only the reported rows
mb_rep <- filter(mb, Reported == TRUE)

# (a) Each variety in 2025: mean reported yield and count, best first
variety_summary_2025 <- mb_rep |>
  filter(Year == 2025) |>                      # 2025 only
  group_by(Variety) |>                         # one group per variety
  summarise(mean_yield = mean(Yield_bu_ac),    # average yield
            n = n()) |>                        # observation count
  arrange(desc(mean_yield))                    # highest mean first

# Print the top of the sorted table
head(variety_summary_2025)

# (b) Write the table to a csv
write_csv(variety_summary_2025, "output/variety_summary_2025.csv")

## (c) I find AAC WESTKING on top at 79.2 bu/ac from 17 observations. Rows with tiny counts carry far less evidence than their neighbours; some varieties appear only once or twice, so the n column is part of the answer, not decoration.
')
```

</details>

### Question 54

**(a)** In one pipeline, compute each municipality's mean yield and observation count from `mb_rep`, sorted from highest mean to lowest.

**(b)** Report the top municipality and how many observations back its mean.

**(c)** In a comment: a municipality mean pools every variety grown there. What two different things could make one municipality top this table?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb <- read_csv("data/mb_wheat_varieties.csv")
mb_rep <- filter(mb, Reported == TRUE)

# (a) Mean yield and count by municipality, best first
mb_rep |>
  group_by(Municipality) |>                 # one group per municipality
  summarise(mean_yield = mean(Yield_bu_ac), # average across its rows
            n = n()) |>                     # rows behind the average
  arrange(desc(mean_yield))                 # best municipality first

## I find LOUISE on top at 72.86 bu/ac, backed by 26 observations.
## (c) Two different things could put a municipality here: genuinely better
## growing conditions, or a mix that happens to lean on high-yielding
## varieties. The mean pools varieties, so the ranking cannot tell the
## land from the lineup.
')
```

</details>

### Question 55

**(a)** In one pipeline: filter `mb_rep` to 2023, compute each variety's mean yield and count, sort best-first, and save the result.

**(b)** Report the top variety by mean and its count, and the largest count in the table and its variety.

**(c)** In a comment: are they the same variety? What do the two columns measure differently?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb <- read_csv("data/mb_wheat_varieties.csv")
mb_rep <- filter(mb, Reported == TRUE)

# (a) 2023 variety means and counts, best first
variety_2023 <- mb_rep |>
  filter(Year == 2023) |>                   # one year only
  group_by(Variety) |>                      # one group per variety
  summarise(mean_yield = mean(Yield_bu_ac), # average within the variety
            n = n()) |>                     # results behind it
  arrange(desc(mean_yield))                 # best mean first
variety_2023

## I find AC BARRIE (BW 661) with the best 2023 mean, 67.3 bu/ac -- from a
## single observation. The largest count is AAC BRANDON (BW 932), with 84.
## (c) They are not the same variety. The mean measures how well a variety
## did where it was grown; the count measures how widely it was grown. A
## chart-topping mean on n = 1 is one municipality having a good year.
')
```

</details>

### Question 56

A bushel of wheat is 60 lb, so one bu/ac is about 0.0673 t/ha.

**(a)** In one pipeline: convert yields to t/ha with `mutate()`, compute each municipality's mean t/ha and count, keep municipalities and sort best-first, and save the result as `muni_summary`.

**(b)** Write the table to `output/muni_summary.csv`.

**(c)** Report the top municipality. In a comment: its mean rests on how many observations, and how does that shape your confidence in the ranking?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the data
library(tidyverse)
mb <- read_csv("data/mb_wheat_varieties.csv")
mb_rep <- filter(mb, Reported == TRUE)

# (a) Municipality means in t/ha with counts, best first
muni_summary <- mb_rep |>
  mutate(yield_t_ha = Yield_bu_ac * 0.0673) |>   # convert the rate
  group_by(Municipality) |>                      # one group per municipality
  summarise(mean_t_ha = mean(yield_t_ha),        # average within it
            n = n()) |>                          # rows behind the average
  arrange(desc(mean_t_ha))                       # best first
muni_summary

# (b) Write the table to the output folder
write_csv(muni_summary, "output/muni_summary.csv")

## I find LOUISE on top at 4.90 t/ha, resting on 26 observations.
## (c) Twenty-six results is a reasonable base -- enough that one variety
## or one odd field is not driving the ranking, though thinner than the
## biggest municipalities; the n column is what lets a reader make that
## judgment for every row.
')
```

</details>

