# Module 2 — Test Bank {.unnumbered}

```{r}
#| include: false
#| eval: true
# console() renders a code listing plus the console session it produces.
source("../_console.R")
```

::: {.callout-note}
## How this test bank works

This bank holds **40 questions**, ten in each section. Section 1 gives you the data inside the question; Sections 2--4 each hold two blocks of five, one block per dataset (the datasets are described below). On the real test you receive **four questions -- one from each section** -- and you submit **a single, well-commented R script**: a header block at the top, a clearly labelled section for each question, and your sentence answers written as comments. The script should run from top to bottom in a project folder that has the csvs in `data/`. Each question is designed to take about **8--10 minutes**.

Expand each **Answer** block to check your work. Every value has been computed from the real data; if the source file is refreshed the exact numbers may shift slightly, but the method stays the same.
:::

**The data.** Two datasets, four files, all in your project's `data/` folder:

*Saskatchewan RM crop yields* -- average yields in every Saskatchewan Rural Municipality, one row per RM-year(-crop), columns `Year`, `RM`, `Crop`, `Yield`, `Unit`:

- [`rm_canola_yields_1990_2025.csv`](data/rm_canola_yields_1990_2025.csv) -- the **canola rows only**, for Section 2: one crop, one unit, so its summary statistics mean something.
- [`rm_yields_1990_2025.csv`](data/rm_yields_1990_2025.csv) -- all **eight crops**, for Sections 3--4. You worked with the wide version in Module 1; this is the long form, with `Crop` as a variable. **Lentils are recorded in pounds per acre; every other crop is in bushels per acre** -- the `Unit` column says which, and several questions turn on it.

*Manitoba wheat variety yields* -- MASC's variety-level spring wheat yields, one row per municipality-variety-year, 2020--2025, columns `Year`, `Municipality`, `Variety`, `Farms`, `Acres`, `Yield_bu_ac`, `Reported`. Where too few farms grew a variety, the row is **suppressed** for privacy: `Reported` is FALSE and the numeric columns are missing.

- [`mb_wheat_reported_2020_2025.csv`](data/mb_wheat_reported_2020_2025.csv) -- the **reported rows only**, for Section 2.
- [`mb_wheat_varieties.csv`](data/mb_wheat_varieties.csv) -- the **full file**, suppressed rows included, for Sections 3--4.

**Grading model.** **20%** for script presentation -- a header block, comments in the style of the module, clearly labelled question sections, and a script that runs from top to bottom in a fresh session. The remaining **80%** is split evenly across the four questions, and evenly across the parts within each question. Parts are marked on **follow-through**: if an early part is wrong but the later parts are correctly worked *from your own earlier answer*, the later parts still earn full marks. Where a part asks for a sentence, write it as a comment in your script.

> **Vocabulary note.** Everything here uses only Module 2 tools: `<-`, arithmetic on vectors, `c()`, `data.frame()`, `$`, `mean`, `median`, `sd`, `var`, `min`, `max`, `range`, `sum`, `quantile` (with `probs`), `IQR`, `nrow`, `ncol`, `names`, `head`, `print`, `glimpse`, `summary`, `library()`, `install.packages()`, `read_csv` with relative paths, `filter` (with `==`, `>`, `<`, `>=`, `&`, `|`, `!`, `%in%`), `select` (including `-` and `:`), `rename`, `mutate`, `arrange` and `desc`, the pipe `|>`, `summarise`, `n()`, `group_by`, and `write_csv`. Nothing beyond that is needed. The chapter's `summarise()` examples write `na.rm = TRUE` inside `mean()` and `sd()`; none of the data these questions filter to has missing yields, so answers with or without it get the same numbers and the same marks.

::: {.callout-note collapse="true"}
## A sample test, with a full-marks answer script

A real test looks like this -- one question from each section:

1. *Five canola fields yielded 52.3, 47.8, 55.1, 44.6, and 50.9 bu/ac, on 160, 320, 240, 130, and 200 acres. Create the two vectors; convert the yields to t/ha; compute the mean and sd of the yields; compute total production and the acreage-weighted average yield.*
2. *Compute the mean, median, and standard deviation of canola `Yield`, and its 90th percentile. In a comment: the mean sits a little above the median -- what does that direction of gap suggest about the shape of canola yields?*
3. *Filter to the lentil rows, saving as `lentils` (how many rows?); add a `yield_kg_ha` column (multiply by 1.12), saving back to `lentils`; compute the mean of the new column and of the original, and check their ratio.*
4. *In one pipeline, compute each RM's mean Spring Wheat yield across all years, sorted best-first. Report the top three RMs. In a comment: what did each row of the input represent, and what does each row of the result represent?*

And here is a script that would earn full marks, including all of the presentation component -- with the console session it produces shown beneath it:

```{r}
#| eval: true
#| echo: false
console('
# ---
# Title: Module 2 test
# Author: Jordan Field
# Date: 2026-10-15
# Description:
#   Answers to the four test questions, one section per block.
#   Sentence answers are written as comments below each result.
# ---

# Load packages
library(tidyverse)

# ============================================================
# Section 1: five canola fields
# ============================================================

# Yields (bu/ac) and field sizes (acres)
yields <- c(52.3, 47.8, 55.1, 44.6, 50.9)
acres  <- c(160, 320, 240, 130, 200)

# (a/b) Convert every yield to tonnes per hectare in one step
yields_t_ha <- yields * 0.0560
yields_t_ha

# (c) Centre and spread of the yields
mean(yields)   # 50.14 bu/ac
sd(yields)     # 4.06 bu/ac

# (d) Total production, then the average weighted by acres
total_bu <- sum(yields * acres)
total_bu                 # 52,866 bu
total_bu / sum(acres)    # 50.35 bu/ac
# The weighted average differs from the plain mean because the
# fields differ in size, so each yield should count by its acres.

# ============================================================
# Section 2: summary statistics for canola
# ============================================================

# Read the canola file using a relative path from the project folder
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")

# Centre and spread of the canola yields
mean(rm_canola$Yield)      # 28.29 bu/ac
median(rm_canola$Yield)    # 26.9 bu/ac
sd(rm_canola$Yield)        # 10.06 bu/ac

# 90th percentile
quantile(rm_canola$Yield, 0.9)   # 42.5 bu/ac

# The mean sits a little above the median, which suggests mild
# right skew: the best RM-years pull the mean up more than the
# worst years pull it down.

# ============================================================
# Section 3: converting the lentil yields
# ============================================================

# Read the full eight-crop file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# Keep only the lentil rows
lentils <- filter(rm_yields, Crop == "Lentils")
nrow(lentils)    # 6,338 rows

# Add the converted column (1 lb/ac is about 1.12 kg/ha) and
# overwrite lentils with the version that has the extra column
lentils <- mutate(lentils, yield_kg_ha = Yield * 1.12)

# Means of the new and original columns
mean(lentils$yield_kg_ha)   # 1,353.4 kg/ha
mean(lentils$Yield)         # 1,208.39 lb/ac
# Their ratio is 1.12, the conversion factor -- multiplying every
# value by a constant multiplies the mean by the same constant.

# ============================================================
# Section 4: best wheat-growing RMs
# ============================================================

rm_yields |>                              # take the data, THEN
  filter(Crop == "Spring Wheat") |>       # keep spring wheat, THEN
  group_by(RM) |>                         # split the rows by RM, THEN
  summarise(mean_yield = mean(Yield)) |>  # one row per RM: its mean, THEN
  arrange(desc(mean_yield))               # best RMs on top

# Top three: RM 369 (47.6), RM 333 (47.2), RM 368 (47.2) bu/ac.
# Each input row was one RM-year observation of spring wheat; each
# result row is one RM, its years collapsed into a single mean.
')
```
:::

---

# Section 1 — R Basics

*These questions are about the building blocks: objects, vectors, data frames, and functions. No data file is needed -- the numbers are in the question. Answer in the Section 1 block of your script, with sentence answers written as comments.*

### Question 1

Five canola fields yielded 52.3, 47.8, 55.1, 44.6, and 50.9 bu/ac, on 160, 320, 240, 130, and 200 acres.

**(a)** Create a vector `yields` and a vector `acres` holding these values.

**(b)** Create a vector `yields_t_ha` converting the yields to tonnes per hectare (multiply by 0.0560) in one line -- R multiplies every element at once.

**(c)** Compute the mean and standard deviation of `yields`.

**(d)** Compute total production in bushels (each field's yield times its acres, summed), and the farm's average yield weighted by acres (total bushels divided by total acres).

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Create the yield and acreage vectors
yields <- c(52.3, 47.8, 55.1, 44.6, 50.9)
acres  <- c(160, 320, 240, 130, 200)

# (b) Convert every yield to tonnes per hectare in one line
yields_t_ha <- yields * 0.0560

# Print the converted yields
yields_t_ha

# (c) Mean and standard deviation of the yields
mean(yields)
sd(yields)

# (d) Total production in bushels
total_bu <- sum(yields * acres)
total_bu

# (d) Average yield weighted by acres
total_bu / sum(acres)

## I find the mean to be 50.14 bu/ac and the sd 4.06.
## Total production is 52,866 bu and the acre-weighted average 50.35 bu/ac.
')
```

</details>

### Question 2

Five barley fields yielded 68.2, 71.5, 64.9, 74.1, and 66.3 bu/ac. Five lentil fields yielded 1450, 1720, 1280, 1610, and 1390 lb/ac.

**(a)** Compute the mean, the standard deviation, the variance, and the range (maximum minus minimum) of each crop's yields.

**(b)** The lentil standard deviation is far larger. In a comment: why can the two standard deviations not be compared directly?

**(c)** Compute the coefficient of variation (sd divided by mean) for each crop. Which crop's yields are more variable *relative to their own average*?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Yield vectors for each crop
barley  <- c(68.2, 71.5, 64.9, 74.1, 66.3)
lentils <- c(1450, 1720, 1280, 1610, 1390)

# (a) Mean of each crop
mean(barley)
mean(lentils)

# (a) Standard deviation of each crop
sd(barley)
sd(lentils)

# (a) Variance of each crop
var(barley)
var(lentils)

# (a) Range (maximum minus minimum) of each crop
max(barley) - min(barley)
max(lentils) - min(lentils)

# (c) Coefficient of variation (sd divided by mean) for each crop
sd(barley) / mean(barley)
sd(lentils) / mean(lentils)

## I find barley mean 69.0, sd 3.77, variance 14.25, range 9.2; lentils
## mean 1490, sd 175.36, variance 30,750, range 440.
## (b) The two standard deviations are in different units (bushels vs pounds
## per acre) and on very different scales, so the raw spreads are not
## comparable.
## (c) CV barley 0.055, lentils 0.118 -- lentil yields are more variable
## relative to their own average.
')
```

</details>

### Question 3

A trial records, for each field: the crop grown, the seeded acres, and whether the field was irrigated.

**(a)** Create three vectors of length four holding made-up values for these three variables, choosing the appropriate type for each (character, numeric, logical).

**(b)** In a comment: which values needed quotation marks and which must not have them?

**(c)** Combine your three vectors into a data frame called `trial` and print it.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Crop grown: character, so the values are quoted
crop      <- c("Canola", "Wheat", "Peas", "Canola")

# (a) Seeded acres: numeric, no quotes
acres     <- c(160, 320, 240, 130)

# (a) Irrigated or not: logical, no quotes
irrigated <- c(TRUE, FALSE, FALSE, TRUE)

# (c) Combine the three vectors into a data frame
trial <- data.frame(crop = crop,
                    acres = acres,
                    irrigated = irrigated)

# Print the data frame
trial

## (b) The character values (crop names) need quotes. Numbers must not have
## them (quoted numbers become text), and TRUE/FALSE must not have them --
## quoted, they would be text rather than logical values.
')
```

</details>

---

### Question 4

Twelve canola fields yielded 24.1, 31.5, 28.8, 35.2, 22.7, 30.4, 27.9, 33.6, 25.5, 29.3, 32.8, and 26.4 bu/ac.

**(a)** Save the values as a vector and compute the mean.

**(b)** Compute the 10th, 50th, and 90th percentiles in a single `quantile()` call.

**(c)** In a comment: one of your three percentiles should equal the median. Which one, and does it?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Canola yields as a vector (bu/ac)
canola <- c(24.1, 31.5, 28.8, 35.2, 22.7, 30.4, 27.9, 33.6, 25.5, 29.3, 32.8, 26.4)

# (a) Mean yield
mean(canola)

# (b) 10th, 50th, and 90th percentiles in one call
quantile(canola, c(0.1, 0.5, 0.9))

# (c) The median, to check against the 50th percentile
median(canola)

## I find the mean to be 29.02.
## (b) 10th 24.24, 50th 29.05, 90th 33.52.
## (c) The 50th percentile is the median by definition; median(canola)
## confirms 29.05.
')
```

</details>

### Question 5

Consider these three lines:

```r
k <- 25
k == 30
mean(x = c(10, 20, 60))
```

**(a)** In a comment: say what each line does and what, if anything, it prints.

**(b)** In a comment: after the third line runs, does an object called `x` appear in the Variables pane? Why or why not?

**(c)** In a comment: which single symbol would you change to make line 2 *assign* 30 to `k`, and which does this book prefer for assignment?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Run the three lines to see what each one prints

# Line 1: assign 25 to k
k <- 25

# Line 2: ask whether k equals 30
k == 30

# Line 3: compute a mean, with the argument named
mean(x = c(10, 20, 60))

## (a) Line 1 assigns 25 to k and prints nothing. Line 2 asks whether k
## equals 30 and prints [1] FALSE. Line 3 computes a mean and prints [1] 30.
## (b) No -- the x = inside the call names an argument for mean(); it does
## not create an object.
## (c) == would become <- (or =, which also assigns) -- and the book uses <-
## for assignment.
')
```

</details>

### Question 6

**(a)** `mean()` takes its data as an argument named `x`. Write two calls that compute the mean of 12, 18, and 30 -- one with the argument unnamed, one with it named -- and confirm they agree.

**(b)** In a comment: what do you type to open the documentation for `sd()`?

**(c)** A classmate writes `Mean(c(12, 18, 30))` and R replies `could not find function "Mean"`. In a comment: what went wrong?

**(d)** Compute the standard deviation of the same three values, with the argument named.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Mean with the argument unnamed
mean(c(12, 18, 30))

# (a) Mean with the argument named
mean(x = c(12, 18, 30))

# (c) What the classmate typed: Mean with a capital M
Mean(c(12, 18, 30))

# (d) Standard deviation with the argument named
sd(x = c(12, 18, 30))

## (a) Both calls print [1] 20, so they agree.
## (b) ?sd opens the documentation.
## (c) R is exact about names: Mean with a capital M is not the same function
## as mean.
## (d) The standard deviation is 9.165151.
')
```

</details>

### Question 7

Four oat fields yielded 84.3, 91.7, 77.2, and 88.5 bu/ac, and oats are selling at $4.15 per bushel.

**(a)** Save the yields as a vector and the price as a single object.

**(b)** Compute a revenue-per-acre vector (yield times price) in one line, using the price object rather than retyping the number.

**(c)** Compute the mean revenue per acre.

**(d)** In a comment: the price rises to $4.60. What two lines would you rerun to update every answer, and why is that easier than if you had typed the price into each formula?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Oat yields (bu/ac) and the price ($/bu)
yields <- c(84.3, 91.7, 77.2, 88.5)
price  <- 4.15

# (b) Revenue per acre, using the price object
revenue <- yields * price

# Print the revenues
revenue

# (c) Mean revenue per acre
mean(revenue)

## (b) The revenues are 349.85, 380.56, 320.38, 367.28 $/ac, rounded to the
## cent (the console shows 349.845 and so on).
## (c) Mean revenue is $354.51 per acre.
## (d) Change price <- 4.60, then rerun the lines below it. The price lives
## in one object, so one edit updates everything -- the same reason a
## spreadsheet keeps a price in one cell.
')
```

</details>

### Question 8

**(a)** Build a data frame called `durum_fields` with columns `field_id` (`"D1"` to `"D4"`), `region` (South, South, Central, North), and `yield` (28.3, 41.6, 35.2, 44.8).

**(b)** Report its number of rows, number of columns, and column names using functions.

**(c)** Using `$`, compute the mean yield and the range (maximum minus minimum).

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Build the data frame
durum_fields <- data.frame(
  field_id = c("D1", "D2", "D3", "D4"),
  region = c("South", "South", "Central", "North"),
  yield = c(28.3, 41.6, 35.2, 44.8)
)

# (b) Number of rows
nrow(durum_fields)

# (b) Number of columns
ncol(durum_fields)

# (b) Column names
names(durum_fields)

# (c) Mean yield, using $
mean(durum_fields$yield)

# (c) Range (maximum minus minimum)
max(durum_fields$yield) - min(durum_fields$yield)

## (b) 4 rows, 3 columns; the columns are field_id, region, yield.
## (c) I find the mean to be 37.48 (the console shows 37.475) and the
## range 16.5.
')
```

</details>

### Question 9

A student runs these lines in the console:

```r
bins <- c(1450, 2200, 1875)
bins * 2
total <- sum(bins)
```

**(a)** In a comment: what exactly does the second line print, `[1]` included, and is that doubled vector saved anywhere?

**(b)** In a comment: what does the third line print, and what does the Variables pane show afterwards?

**(c)** The student now wants the doubled vector kept for later. Write the line that does it.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Run the lines to see what each one prints

# Create the vector of bin weights
bins <- c(1450, 2200, 1875)

# Double it, without saving the result
bins * 2

# Sum, saved to total (an assignment prints nothing)
total <- sum(bins)

# (c) Keep the doubled vector for later
bins_doubled <- bins * 2

# Print it to confirm it was saved
bins_doubled

## (a) The second line prints [1] 2900 4400 3750 -- displayed only, saved
## nowhere.
## (b) The third line prints nothing (it is an assignment); afterwards the
## Variables pane shows bins (the three values) and total = 5525.
## (c) bins_doubled <- bins * 2 keeps the doubled vector.
')
```

</details>

### Question 10

Here is a working script:

```r
v1 <- c(47.2, 49.8, 46.5, 50.1, 48.4)
v2 <- c(38.9, 55.6, 44.2, 58.3, 41.0)
mean(v1)
mean(v2)
```

**(a)** Rewrite it with a header and comments. The two vectors are canola yields from Farm A and Farm B.

**(b)** Add lines computing each farm's standard deviation and range.

**(c)** In a comment: the two farms have nearly the same mean. Using your numbers from (b), say in one sentence how they differ.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# ---
# Title: Comparing two canola farms
# Author: Your Name
# Date: 2026-10-08
# Description:
#   Compares the centre and spread of canola yields
#   on Farm A and Farm B.
# ---

# (a) Canola yields by field (bu/ac)
farm_a <- c(47.2, 49.8, 46.5, 50.1, 48.4)
farm_b <- c(38.9, 55.6, 44.2, 58.3, 41.0)

# Centre: mean yield on each farm
mean(farm_a)
mean(farm_b)

# (b) Spread: standard deviation on each farm
sd(farm_a)
sd(farm_b)

# (b) Spread: range (maximum minus minimum) on each farm
max(farm_a) - min(farm_a)
max(farm_b) - min(farm_b)

## (b) Farm A sd 1.57, range 3.6; Farm B sd 8.79, range 19.4.
## (c) The means are close (48.4 vs 47.6), but Farm B yields swing far more
## from field to field -- the spread on Farm B is five times the spread on
## Farm A.
')
```

</details>

# Section 2 — Reading and Inspecting Data

*Three blocks of ten, one per dataset. Each block starts by naming its file; on a real test your Section 2 question comes from one of the three. Answer in the Section 2 block of your script.*

## Saskatchewan RM Canola {.unnumbered}

*These questions use the canola-only file, [`rm_canola_yields_1990_2025.csv`](data/rm_canola_yields_1990_2025.csv), in your project's `data/` folder.*

### Question 11

**(a)** In a comment at the top of your script, sketch your project's folder layout (the subfolders and where the csv sits).

**(b)** Load the tidyverse and read the canola file into an object called `rm_canola`, using a relative path.

**(c)** Report the number of rows, the number of columns, and the column names using functions.

**(d)** In a comment: why will your `read_csv()` line work on the grader's computer, when a full path starting `C:/Users/...` would not?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Sketch of the project folder layout:
# project/
#   analysis.R
#   data/
#     rm_canola_yields_1990_2025.csv

# (b) Load the tidyverse
library(tidyverse)

# (b) Read the canola file with a relative path
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")

# (c) Number of rows
nrow(rm_canola)

# (c) Number of columns
ncol(rm_canola)

# (c) Column names
names(rm_canola)

## (c) I find the file has 10039 rows and 5 columns: Year, RM, Crop, Yield, Unit.
## (d) The path is relative to the project folder, and the grader opens the same self-contained folder -- so "data/..." exists there too. A full path starting "C:/Users/..." names the user and folder layout of one particular computer, so it fails everywhere else.
')
```

</details>

### Question 12

**(a)** Run `glimpse()` on `rm_canola`. In a comment: which columns are text (`<chr>`) and which are numeric (`<dbl>`)?

**(b)** When `read_csv()` ran, it printed a note about the file. In a comment: what did that note report, and why does the module say to read it every time?

**(c)** `Year` came in as numeric (`<dbl>`). In a comment: which type of value is each of `2024` and `"2024"`? Why do we want a column we will compute with stored as numbers rather than as text?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the canola file
library(tidyverse)
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")

# (a) Structure: one line per column, with its type
glimpse(rm_canola)

## (a) Crop and Unit are text (<chr>); Year, RM, and Yield are numeric (<dbl>).
## (b) The note reported the row and column counts and the type read_csv guessed for each column -- here chr (2) and dbl (3). R guesses well but not perfectly, so the note is the first check that the file read the way you expect.
## (c) 2024 is a numeric value; "2024" is a character (text) value. Numbers stored as text stop behaving like numbers -- you cannot do arithmetic with them, and sorting turns alphabetical -- so a column we will compute with should show <dbl>, not <chr>.
')
```

</details>

### Question 13

**(a)** Run `summary(rm_canola)`. From its output, report the minimum, median, and maximum of `Yield`, and the first and last `Year`.

**(b)** In a comment: work through the module's inspection questions for this file. Are the yields in between roughly 10 and 60 bu/ac, a typical canola range? Are the years what the file name promises? Are there missing values?

**(c)** The minimum is below 2 bu/ac. In a comment: is a canola yield that low necessarily a data error? What kind of year could produce it?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the canola file
library(tidyverse)
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")

# (a) Summary of every column
summary(rm_canola)

## (a) I find Yield has minimum 1.9, median 26.9, and maximum 61.0, and Year runs from 1990 to 2025.
## (b) The yields all sit between about 2 and 61 bu/ac, a sensible canola range; the years run 1990--2025 as the file name promises; and summary() prints an NA count under a column only when it has missing values -- no such line appears, so there are none.
## (c) Not necessarily -- an RM average that low is what a severe drought or a widespread crop failure looks like. Extreme is not the same as wrong; it is a value to investigate, not delete.
')
```

</details>

### Question 14

**(a)** Compute the mean, median, standard deviation, and variance of `rm_canola$Yield`.

**(b)** Compute the 90th percentile of `Yield`.

**(c)** The mean sits a little above the median. In a comment: using Module 1's language about means and medians, what does that direction of gap suggest about the shape of canola yields?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the canola file
library(tidyverse)
rm_canola <- read_csv("data/rm_canola_yields_1990_2025.csv")

# (a) Centre of the yields
mean(rm_canola$Yield)
median(rm_canola$Yield)

# (a) Spread of the yields
sd(rm_canola$Yield)
var(rm_canola$Yield)

# (b) 90th percentile of the yields
quantile(rm_canola$Yield, probs = 0.9)

## (a) I find a mean of 28.29, a median of 26.9, a standard deviation of 10.06, and a variance of 101.21 -- the sd squared, in squared units.
## (b) The 90th percentile is 42.5 bu/ac.
## (c) A mean above the median suggests mild right skew: the very best RM-years pull the mean up more than the worst years pull it down.
')
```

</details>

### Question 15

A classmate's fresh R session runs `rm_yields <- read_csv("data/rm_yields_1990_2025.csv")` and gets:

```
Error in read_csv(...) : could not find function "read_csv"
```

**(a)** In a comment: what is the most likely cause, and what one line fixes it?

**(b)** In a comment: they object, "but I installed the tidyverse last week!" Explain the difference between installing and loading, and how often each is done.

**(c)** In a comment: state the quotes rule -- where do quotation marks go when installing, and when loading?

**(d)** The tidyverse is a collection of packages. In comments, name the four core ones from the module and the job each does.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) The most likely cause: read_csv lives in a package that is not loaded
#     in this fresh session, so R cannot find the function. One line fixes it:
library(tidyverse)

# (a) With the package loaded, the same read line now works
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

## (a) The package that provides read_csv was not loaded in the session; running library(tidyverse) first fixes it.
## (b) Installing downloads the package onto the computer -- that happens once per computer. Loading switches it on -- that happens once per script or session. A package installed last week is still installed, but a fresh session starts with nothing loaded.
## (c) Quotes when you install, none when you load: install.packages("tidyverse") but library(tidyverse).
## (d) The four core packages: readr reads data, dplyr manipulates it, tidyr reshapes it, ggplot2 makes graphs.
')
```

</details>

## Manitoba Wheat Variety {.unnumbered}

*These questions use [`mb_wheat_reported_2020_2025.csv`](data/mb_wheat_reported_2020_2025.csv) -- spring wheat yields by variety and municipality, reported observations only.*

### Question 16

**(a)** In a comment at the top of your script, sketch your project's folder layout.

**(b)** Load the tidyverse and read the file into an object called `mb_wheat`, using a relative path.

**(c)** Report the number of rows, the number of columns, and the column names using functions.

**(d)** In a comment: what does one row of this dataset represent?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# (a) Sketch of the project folder layout:
# project/
#   analysis.R
#   data/
#     mb_wheat_reported_2020_2025.csv

# (b) Load the tidyverse
library(tidyverse)

# (b) Read the wheat file with a relative path
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (c) Number of rows
nrow(mb_wheat)

# (c) Number of columns
ncol(mb_wheat)

# (c) Column names
names(mb_wheat)

## (c) I find the file has 2398 rows and 7 columns: Year, Municipality, Variety, Farms, Acres, Yield_bu_ac, Reported.
## (d) One row is one wheat variety grown in one municipality in one year -- its farm count, its acres, and its average yield.
')
```

</details>

### Question 17

A classmate's script reads the data like this:

```r
mb_wheat <- read_csv("/Users/sam/Downloads/mb wheat data.csv")
```

It works on their laptop and fails on everyone else's.

**(a)** In a comment: give two problems -- one about where the path points, one about the file's name and location.

**(b)** Rewrite the line the way the module recommends, assuming the file was saved as `mb_wheat_reported_2020_2025.csv` in the project's `data/` folder.

**(c)** In a comment: your line fails with `does not exist in current working directory`. What is the first thing to check?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# (b) The read line rewritten the way the module recommends: the file renamed,
#     moved into the data folder of the project, and read with a relative path
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

## (a) Two problems. First, the path is absolute: it names the Downloads folder of one user on one machine, so it exists nowhere else. Second, the file is loose in Downloads under a name with spaces in it, instead of renamed and placed in the data folder of the project.
## (c) Check that the folder opened in Positron is the project folder -- the one containing data/. Relative paths start from the opened folder.
')
```

</details>

### Question 18

**(a)** Run `glimpse()` on `mb_wheat`. In a comment: which columns are text (`<chr>`), which numeric (`<dbl>`), and which logical (`<lgl>`)?

**(b)** In a comment: what values does the logical column hold in this file, and how does the file's name explain this?

**(c)** In a comment: `Farms` holds whole numbers. Would `glimpse()` showing it as `<chr>` be a problem, and what would it tell you about the file?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the wheat file
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) Structure: one line per column, with its type
glimpse(mb_wheat)

# (b) The values the logical column holds
summary(mb_wheat$Reported)

## (a) Municipality and Variety are text (<chr>); Year, Farms, Acres, and Yield_bu_ac are numeric (<dbl>); Reported is logical (<lgl>).
## (b) Reported holds only TRUE -- all 2398 of them. This is the reported-only file, with the suppressed (FALSE) rows removed, which is what the file name promises.
## (c) Yes, it would be a problem -- a numeric column arriving as text means some value in it could not be read as a number, and calculations like mean() would fail on it.
')
```

</details>

### Question 19

**(a)** Run `summary(mb_wheat)`. From its output, report the minimum, median, and maximum of `Yield_bu_ac`, and the first and last `Year`.

**(b)** In a comment: work through the module's inspection questions. Are the yields in mostly between 30 and 100 bu/ac, a typical spring wheat range? Are there missing values?

**(c)** The minimum is **4.5 bu/ac**. In a comment: is a reported yield that low necessarily an error?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the wheat file
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) Summary of every column
summary(mb_wheat)

## (a) I find Yield_bu_ac has minimum 4.5, median 62.2, and maximum 97.8, and Year runs from 2020 to 2025.
## (b) Plausible -- most yields sit in the tens of bushels, sensible for Manitoba spring wheat; and no NA count lines appear, so there are no missing values in this file.
## (c) Not necessarily -- a variety grown on a few drought- or flood-hit fields can genuinely average that low. Extreme is a value to investigate, not delete.
')
```

</details>

### Question 20

**(a)** Compute the mean, median, standard deviation, and variance of `mb_wheat$Yield_bu_ac`.

**(b)** In a comment: the mean sits a little *below* the median here. In Module 1's language, what does that direction of gap suggest, and what kind of observations would produce it?

**(c)** Compute the smallest and largest yields in one `range()` call.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the wheat file
library(tidyverse)
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv")

# (a) Centre of the yields
mean(mb_wheat$Yield_bu_ac)
median(mb_wheat$Yield_bu_ac)

# (a) Spread of the yields
sd(mb_wheat$Yield_bu_ac)
var(mb_wheat$Yield_bu_ac)

# (c) Smallest and largest yield in one call
range(mb_wheat$Yield_bu_ac)

## (a) I find a mean of 61.16, a median of 62.2, a standard deviation of 12.38, and a variance of 153.34.
## (b) A mean below the median suggests mild left skew: unusually low values pull the mean down more than the high values pull it up -- a tail of very poor variety-municipality observations mixed in with the normal ones.
## (c) range() returns both ends as one vector: the smallest yield is 4.5 and the largest is 97.8 bu/ac.
')
```

</details>

# Section 3 — Data Manipulation Functions

*Three blocks of ten, one per dataset. Each question uses two or three of the functions in sequence -- save each intermediate result to an object and work step by step (pipelines wait for Section 4).*

## Saskatchewan RM Crop Yields {.unnumbered}

*These questions use the full eight-crop file: read [`rm_yields_1990_2025.csv`](data/rm_yields_1990_2025.csv) into an object called `rm_yields` at the start of this block.*

### Question 21

**(a)** Filter `rm_yields` to Canola in 2024, saving the result as `canola_2024`. How many rows are in the filtered data frame?

**(b)** From `canola_2024`, select only the `RM` and `Yield` columns, saving the result. Report its column names and its number of columns.

**(c)** In a comment: after (a) and (b), how many rows does `rm_yields` itself have, and why?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Keep only the Canola rows from 2024
canola_2024 <- filter(rm_yields, Crop == "Canola" & Year == 2024)
# Row count of the filtered data frame
nrow(canola_2024)

# (b) Keep only the RM and Yield columns
canola_slim <- select(canola_2024, RM, Yield)
# Column names of the slimmed version
names(canola_slim)
# Number of columns
ncol(canola_slim)

# (c) The original data frame, after (a) and (b)
nrow(rm_yields)

## I find 293 rows of 2024 canola.
## The selected version has the columns RM and Yield, so 2 columns (and still 293 rows).
## rm_yields itself still has 71104 rows: filter() and select() return new data frames and leave the original untouched.
')
```

</details>

### Question 22

**(a)** Filter to rows where the crop is Oats **or** Barley, using `|`, saving the result. How many rows are in the filtered data frame?

**(b)** Write the same filter using `%in%` and confirm the row count matches.

**(c)** Sort your filtered data frame from highest yield to lowest. What crop, RM, and year appear in the top row, and at what yield?

**(d)** In a comment: a classmate tried `filter(rm_yields, Crop == "Oats" & Crop == "Barley")` and got zero rows. Why?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Oats or Barley, using the | connector
oats_barley <- filter(rm_yields, Crop == "Oats" | Crop == "Barley")
# Row count of the filtered data frame
nrow(oats_barley)

# (b) The same filter written with %in%
oats_barley2 <- filter(rm_yields, Crop %in% c("Oats", "Barley"))
# The count should match (a)
nrow(oats_barley2)

# (c) Sort from highest yield to lowest
arrange(oats_barley, desc(Yield))

## I find 19709 rows both ways.
## The top of the sorted table is a tie at 165.1 bu/ac: Oats in RM 369 in 2025 and Oats in RM 438 in 2020 share the maximum, so either is a correct top row.
## (d) The & version asks for rows whose crop is Oats and Barley at the same time. No single row can satisfy both conditions, so zero rows come back. "Or" is the right connector, written |.
')
```

</details>

### Question 23

**(a)** Filter to the Spring Wheat observations of at least 60 bu/ac (all years), using `>=` and saving the result. How many rows are in the filtered data frame?

**(b)** Filter again to keep only the 2023 rows of that result. How many rows are in this second data frame?

**(c)** From your 2023 data frame, select `RM` and `Yield` and sort best-first. Which RM tops the list, at what yield?

**(d)** In a comment: rewrite the condition "yield at least 60 and the crop is *not* Spring Wheat" using `!`.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Spring Wheat at 60 bu/ac or better, all years
sw_60 <- filter(rm_yields, Crop == "Spring Wheat" & Yield >= 60)
# Row count of the filtered data frame
nrow(sw_60)

# (b) Keep only the 2023 rows of that result
sw_60_2023 <- filter(sw_60, Year == 2023)
# Row count of the second data frame
nrow(sw_60_2023)

# (c) Just RM and Yield, sorted best first
sw_best <- select(sw_60_2023, RM, Yield)
arrange(sw_best, desc(Yield))

## I find 434 rows at 60 bu/ac or better, and 53 of them are from 2023.
## RM 187 tops the 2023 list, at 74.7 bu/ac.
## (d) filter(rm_yields, Yield >= 60 & !Crop == "Spring Wheat") -- the ! negates the crop condition.
')
```

</details>

### Question 24

**(a)** Make a version of the data with only `Year`, `RM`, `Crop`, and `Yield` -- once by *naming what you keep*, once by *dropping* what you don't want -- saving each, and confirm the two have the same column names. Then make a third version keeping the adjacent-column range `Year:Crop`, and report its column names.

**(b)** Rename `Yield` to `yield_per_acre` in the full data, saving the result, and show the new column names.

**(c)** Filter your renamed data frame to Peas in 2025. How many rows are in the filtered data frame?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Keep four columns by naming them
kept <- select(rm_yields, Year, RM, Crop, Yield)
# The same four columns by dropping Unit
dropped <- select(rm_yields, -Unit)
# Confirm the two have the same column names
names(kept)
names(dropped)
# The adjacent-column range Year:Crop
ranged <- select(rm_yields, Year:Crop)
# Its column names
names(ranged)

# (b) Rename Yield in the full data
renamed <- rename(rm_yields, yield_per_acre = Yield)
# The new column names
names(renamed)

# (c) Peas in 2025, from the renamed data frame
peas_2025 <- filter(renamed, Crop == "Peas" & Year == 2025)
# Row count of the filtered data frame
nrow(peas_2025)

## Both versions have the columns Year, RM, Crop, Yield; the range version keeps Year, RM, Crop.
## After the rename the columns are Year, RM, Crop, yield_per_acre, Unit.
## I find 285 rows of 2025 peas.
')
```

</details>

### Question 25

**(a)** Filter to Canola in 2024, then in one `mutate()` call add two columns: `yield_t_ha` (multiply by 0.0560) and `yield_kg_ha` (multiply by 56.0). Save the result of each step.

**(b)** Compute the mean of each new column.

**(c)** Sort by `yield_t_ha`, best first. Which RM tops the table, and at what yield in t/ha?

**(d)** In a comment: without running anything, what is the mean of `yield_kg_ha` divided by the mean of `yield_t_ha`, and why?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Canola in 2024
canola_2024 <- filter(rm_yields, Crop == "Canola" & Year == 2024)
# Add both converted columns in one mutate() call
canola_2024 <- mutate(canola_2024,
                      yield_t_ha  = Yield * 0.0560,  # tonnes per hectare
                      yield_kg_ha = Yield * 56.0)    # kilograms per hectare

# (b) Mean of each new column
mean(canola_2024$yield_t_ha)
mean(canola_2024$yield_kg_ha)

# (c) Sort by tonnes per hectare, best first
arrange(canola_2024, desc(yield_t_ha))

## I find means of 1.76 t/ha and 1764.11 kg/ha.
## RM 493 tops the table, at 2.57 t/ha.
## (d) The ratio of the means is 1000: both columns are the same yields scaled by constants, and 56.0 / 0.0560 = 1000 (a tonne is 1000 kg).
')
```

</details>

## Manitoba Wheat Variety {.unnumbered}

*These questions use the full [`mb_wheat_varieties.csv`](data/mb_wheat_varieties.csv) -- including the suppressed rows, where `Reported` is FALSE and the numeric columns are missing. Read it into an object called `mb` at the start of this block.*

### Question 26

**(a)** Filter `mb` to the reported rows (`Reported == TRUE`), saving as `mb_rep`. How many rows are in the filtered data frame?

**(b)** From `mb_rep`, select `Municipality`, `Variety`, and `Yield_bu_ac`, saving the result. Report its column names.

**(c)** In a comment: how many rows does `mb` itself still have, and what happened to the suppressed rows in `mb_rep`?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full Manitoba variety file, suppressed rows included
mb <- read_csv("data/mb_wheat_varieties.csv")

# (a) Keep the reported rows only
mb_rep <- filter(mb, Reported == TRUE)
# Row count of the filtered data frame
nrow(mb_rep)

# (b) Three columns from the reported rows
mb_slim <- select(mb_rep, Municipality, Variety, Yield_bu_ac)
# Its column names
names(mb_slim)

# (c) The full file, after (a) and (b)
nrow(mb)

## I find 2398 reported rows.
## The selected columns are Municipality, Variety, Yield_bu_ac.
## mb itself still has 5358 rows -- filtering returns a new data frame. The 2960 suppressed rows are simply absent from mb_rep; they were not deleted from anything.
')
```

</details>

### Question 27

**(a)** From `mb_rep` (reported rows), filter to the varieties `AAC BRANDON (BW 932)` **or** `AAC STARBUCK <SECAN>` using `|`, saving the result. How many rows are in the filtered data frame?

**(b)** Write the same filter using `%in%` and confirm the count matches.

**(c)** In a comment: a classmate wrote the condition with `&` instead of `|` and got zero rows. Why?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full Manitoba variety file, then keep the reported rows
mb <- read_csv("data/mb_wheat_varieties.csv")
mb_rep <- filter(mb, Reported == TRUE)

# (a) The two varieties, using the | connector
two_var <- filter(mb_rep, Variety == "AAC BRANDON (BW 932)" | Variety == "AAC STARBUCK <SECAN>")
# Row count of the filtered data frame
nrow(two_var)

# (b) The same filter written with %in%
two_var2 <- filter(mb_rep, Variety %in% c("AAC BRANDON (BW 932)", "AAC STARBUCK <SECAN>"))
# The count should match (a)
nrow(two_var2)

## I find 893 rows both ways.
## (c) The & version demands both conditions true of the same row, and no single row has a variety that is both names at once. "Or" is the right connector, written |.
')
```

</details>

### Question 28

**(a)** Filter `mb_rep` to yields above 80 bu/ac, saving the result. How many rows are in the filtered data frame?

**(b)** Filter that result again to 2024 only. How many rows now?

**(c)** Sort your over-80 data frame from (a) best-first. Which municipality, variety, and year top the list, at what yield?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full Manitoba variety file, then keep the reported rows
mb <- read_csv("data/mb_wheat_varieties.csv")
mb_rep <- filter(mb, Reported == TRUE)

# (a) Reported yields above 80 bu/ac
high <- filter(mb_rep, Yield_bu_ac > 80)
# Row count of the filtered data frame
nrow(high)

# (b) The 2024 rows of that result
high_2024 <- filter(high, Year == 2024)
# Row count now
nrow(high_2024)

# (c) Sort the over-80 rows best first
arrange(high, desc(Yield_bu_ac))

## I find 97 rows above 80 bu/ac, 16 of them from 2024.
## HILLSBURG-ROBLIN-SHELL RIVER tops the list, with BOLLES <SEED DEPOT> in 2022, at 97.8 bu/ac.
')
```

</details>

### Question 29

**(a)** From `mb_rep`, make a version with only `Year`, `Municipality`, `Variety`, and `Yield_bu_ac` -- once by naming what you keep, once by dropping what you don't (`Farms`, `Acres`, `Reported`) -- saving each, and confirm the two have the same column names. Then make a third version keeping the adjacent range `Year:Variety` and report its names.

**(b)** Rename `Yield_bu_ac` to `yield`, saving the result, and show the new column names.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full Manitoba variety file, then keep the reported rows
mb <- read_csv("data/mb_wheat_varieties.csv")
mb_rep <- filter(mb, Reported == TRUE)

# (a) Keep four columns by naming them
kept <- select(mb_rep, Year, Municipality, Variety, Yield_bu_ac)
# The same four columns by dropping the rest
dropped <- select(mb_rep, -Farms, -Acres, -Reported)
# Confirm the two have the same column names
names(kept)
names(dropped)
# The adjacent-column range Year:Variety
ranged <- select(mb_rep, Year:Variety)
# Its column names
names(ranged)

# (b) Rename the yield column
renamed <- rename(mb_rep, yield = Yield_bu_ac)
# The new column names
names(renamed)

## Both versions have the same four names, Year, Municipality, Variety, Yield_bu_ac; the range version keeps Year, Municipality, Variety.
## The renamed data frame has the same seven columns, with yield in place of Yield_bu_ac.
')
```

</details>

### Question 30

**(a)** From `mb_rep`, use one `mutate()` to add `total_bu`, each observation's acres times its yield, saving the result.

**(b)** Compute the mean of `total_bu`.

**(c)** Sort by `total_bu`, largest first. Which municipality tops the table, and at roughly what total?

**(d)** In a comment: what does `total_bu` measure that `Yield_bu_ac` does not?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full Manitoba variety file, then keep the reported rows
mb <- read_csv("data/mb_wheat_varieties.csv")
mb_rep <- filter(mb, Reported == TRUE)

# (a) Total bushels: acres times yield, in one mutate() call
mb_rep <- mutate(mb_rep, total_bu = Acres * Yield_bu_ac)

# (b) Mean of the new column
mean(mb_rep$total_bu)

# (c) Sort by total production, largest first
arrange(mb_rep, desc(total_bu))

## I find a mean of about 382865 bu.
## PORTAGE LA PRAIRIE tops the table, at about 5.13 million bushels.
## (d) total_bu measures production -- how much grain the planting actually produced -- rather than how well each acre yielded. A modest yield on enormous acres can out-produce a great yield on few.
')
```

</details>

# Section 4 — Pipelines and Grouped Summaries

*Three blocks of ten, one per dataset. Answer in the Section 4 block of your script. Write multi-step work as pipelines with `|>`, one step per line.*

## Saskatchewan RM Crop Yields {.unnumbered}

*Still the full eight-crop file (`rm_yields`).*

### Question 31

**(a)** Write a pipeline, one step per line with a short comment on each step, that filters `rm_yields` to Durum in 2025, keeps only the `RM` and `Yield` columns, and sorts the result from highest yield to lowest.

**(b)** Run it. How many rows are in the resulting data frame, and which RM had the best durum yield in 2025, at what value?

**(c)** In a comment: read your pipeline aloud as a "take the data, then ..., then ..., then ..." sentence.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Durum in 2025, two columns, best first
durum_2025 <- rm_yields |>                     # take the data, THEN
  filter(Crop == "Durum" & Year == 2025) |>    # keep durum in 2025, THEN
  select(RM, Yield) |>                         # keep two columns, THEN
  arrange(desc(Yield))                         # sort best first

# (b) How many rows the result has
nrow(durum_2025)

# The top of the sorted table
head(durum_2025)

# The top yields to one decimal, since the tibble display rounds
head(durum_2025$Yield)

## I find 189 rows; RM 369 had the best durum yield in 2025, at 105.6 bu/ac.
## (c) Take rm_yields, then keep the 2025 durum rows, then keep the RM and Yield columns, then sort from highest yield to lowest.
')
```

</details>

### Question 32

**(a)** In one pipeline, compute the mean flax yield across 2020--2025 (`Year >= 2020`), calling the result column `mean_yield`.

**(b)** Add `n()` to the same `summarise()` to count how many observations that mean is based on.

**(c)** In a comment: why is reporting the count alongside the mean good practice?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a, b) Mean flax yield across 2020 to 2025, with the observation count
flax_summary <- rm_yields |>
  filter(Crop == "Flax" & Year >= 2020) |>     # flax from 2020 onward
  summarise(mean_yield = mean(Yield),          # average yield
            n = n())                           # rows behind the mean

# Print the summary table
flax_summary

# The mean to more decimals than the tibble display shows
flax_summary$mean_yield

## I find a mean of 22.39 bu/ac from 1143 observations.
## (c) A mean based on a handful of rows deserves less trust than one based on a thousand; the count tells the reader which they are looking at.
')
```

</details>

### Question 33

**(a)** In one pipeline, compute the mean canola yield for **each year** from 2021 to 2025 (filter first, then group).

**(b)** Report the five means. Which year was worst, and which best?

**(c)** In a comment: 2021 was a severe drought year in Saskatchewan. Does the data agree?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Mean canola yield for each year from 2021 to 2025
rm_yields |>
  filter(Crop == "Canola" & Year >= 2021) |>   # canola, 2021 onward
  group_by(Year) |>                            # one group per year
  summarise(mean_yield = mean(Yield))          # average within each year

## I find mean canola yields of 21.9 in 2021, 35.4 in 2022, 33.9 in 2023, 31.5 in 2024, and 43.9 in 2025. The worst year was 2021 and the best was 2025.
## (c) Yes, the data agrees: the 2021 mean sits roughly a third below the surrounding years.
')
```

</details>

### Question 34

**(a)** In one pipeline: filter to Lentils, convert to kg/ha with `mutate()` (multiply by 1.12), then summarise the mean of the converted column.

**(b)** Report the mean in kg/ha.

**(c)** In a comment: in what order do your steps run, and why would putting the `summarise()` before the `mutate()` fail?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Lentils, converted to kg/ha, then the mean of the converted column
lentil_summary <- rm_yields |>
  filter(Crop == "Lentils") |>                 # lentil rows only
  mutate(yield_kg_ha = Yield * 1.12) |>        # lb/ac times 1.12 gives kg/ha
  summarise(mean_kg_ha = mean(yield_kg_ha))    # average of the new column

# Print the summary table
lentil_summary

# The mean to more decimals than the tibble display shows
lentil_summary$mean_kg_ha

## (b) I find a mean of 1353.4 kg/ha.
## (c) The steps run top to bottom, each one feeding its output to the next. Putting summarise() before mutate() would fail because the data would already be collapsed to one row and the yield_kg_ha column would not exist yet.
')
```

</details>

### Question 35

**(a)** In one pipeline: for **2025 only**, compute each crop's mean yield and observation count, sorted from highest mean to lowest, and save the result as `crop_summary_2025`.

**(b)** Write the table to `output/crop_summary_2025.csv`.

**(c)** Report the two highest-yielding crops in the sorted table. In a comment: check the Unit column: why is the first-place crop not comparable to the rest, and which bushel crop is the real standout?, and which crop is really the standout?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the long yield file
rm_yields <- read_csv("data/rm_yields_1990_2025.csv")

# (a) Each crop in 2025: mean yield and observation count, best first
crop_summary_2025 <- rm_yields |>
  filter(Year == 2025) |>                      # 2025 only
  group_by(Crop) |>                            # one group per crop
  summarise(mean_yield = mean(Yield),          # average yield
            n = n()) |>                        # observation count
  arrange(desc(mean_yield))                    # highest mean first

# Print the sorted table
crop_summary_2025

# (b) Write the table to a csv
write_csv(crop_summary_2025, "output/crop_summary_2025.csv")

## I find Lentils on top at 1897 and Oats second at 97.5.
## (c) The Unit column shows Lentils are recorded in pounds per acre, not bushels, so the first place row is not comparable to the rest. Among the bushel crops, Oats are the real standout at 97.5 bu/ac, with Barley next at 72.8. 2025 was a strong year across the board.
')
```

</details>

## Manitoba Wheat Variety {.unnumbered}

*Full [`mb_wheat_varieties.csv`](data/mb_wheat_varieties.csv), read as `mb`, with the reported rows saved as `mb_rep <- filter(mb, Reported == TRUE)` at the start of this block. Build everything else as pipelines.*

### Question 36

**(a)** Write a pipeline, one step per line with a short comment on each, that filters `mb_rep` to 2025, keeps `Municipality`, `Variety`, and `Yield_bu_ac`, and sorts best-first.

**(b)** Run it. How many rows are in the resulting data frame, and which municipality and variety top it, at what yield?

**(c)** In a comment: read the pipeline aloud as a "take the data, then ..." sentence.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full variety file
mb <- read_csv("data/mb_wheat_varieties.csv")

# Keep only the reported rows
mb_rep <- filter(mb, Reported == TRUE)

# (a) Reported 2025 rows, three columns, best first
wheat_2025 <- mb_rep |>                              # take the reported rows, THEN
  filter(Year == 2025) |>                            # keep 2025, THEN
  select(Municipality, Variety, Yield_bu_ac) |>      # keep three columns, THEN
  arrange(desc(Yield_bu_ac))                         # sort best first

# (b) How many rows the result has
nrow(wheat_2025)

# The top of the sorted table
head(wheat_2025)

## I find 438 rows; LOUISE, growing SY MANNESS, tops the table at 96.3 bu/ac.
## (c) Take the reported rows, then keep 2025, then keep three columns, then sort from highest yield to lowest.
')
```

</details>

### Question 37

**(a)** In one pipeline, compute the mean of `Yield_bu_ac` across all of `mb_rep`, calling it `mean_yield`, together with `n()`.

**(b)** Report both values.

**(c)** In a comment: why report the count alongside the mean?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full variety file
mb <- read_csv("data/mb_wheat_varieties.csv")

# Keep only the reported rows
mb_rep <- filter(mb, Reported == TRUE)

# (a) Overall mean yield and count across the reported rows
overall_summary <- mb_rep |>
  summarise(mean_yield = mean(Yield_bu_ac),    # average yield
            n = n())                           # rows behind the mean

# Print the summary table
overall_summary

# The mean to more decimals than the tibble display shows
overall_summary$mean_yield

## (b) I find a mean of 61.16 bu/ac from 2398 observations.
## (c) A mean earns trust in proportion to what it is based on; the count tells the reader how much that is.
')
```

</details>

### Question 38

**(a)** In one pipeline, compute each variety's mean yield and observation count across `mb_rep`, sorted from highest mean to lowest.

**(b)** Report the top two varieties, their means, and their counts.

**(c)** In a comment: the second-place variety was observed exactly once. Which of the two means would you trust more, and why?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full variety file
mb <- read_csv("data/mb_wheat_varieties.csv")

# Keep only the reported rows
mb_rep <- filter(mb, Reported == TRUE)

# (a) Each variety: mean yield and observation count, best first
mb_rep |>
  group_by(Variety) |>                         # one group per variety
  summarise(mean_yield = mean(Yield_bu_ac),    # average yield
            n = n()) |>                        # observation count
  arrange(desc(mean_yield))                    # highest mean first

## (b) I find AAC WESTKING first at 79.2 bu/ac from 17 observations, then CDC GO (BW781) at 77.5 from a single observation.
## (c) I would trust the AAC WESTKING mean more: seventeen municipality-years is evidence, while a single observation could be one lucky field in one good year.
')
```

</details>

### Question 39

A bushel of wheat is 60 lb, so one bu/ac is about 0.0673 t/ha.

**(a)** In one pipeline: convert `mb_rep`'s yields to t/ha with `mutate()`, then compute the mean t/ha for each year.

**(b)** Report the 2021 and 2025 values.

**(c)** In a comment: why must the `mutate()` step come before the `summarise()` step?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full variety file
mb <- read_csv("data/mb_wheat_varieties.csv")

# Keep only the reported rows
mb_rep <- filter(mb, Reported == TRUE)

# (a) Convert to t/ha, then the mean for each year
mb_rep |>
  mutate(yield_t_ha = Yield_bu_ac * 0.0673) |> # bu/ac times 0.0673 gives t/ha
  group_by(Year) |>                            # one group per year
  summarise(mean_t_ha = mean(yield_t_ha))      # average within each year

## (b) I find a mean of 3.34 t/ha in 2021 and 4.58 t/ha in 2025.
## (c) Each step feeds its output to the next; after summarise() the data has collapsed to one row per year and the yield_t_ha column would not exist to average.
')
```

</details>

### Question 40

**(a)** In one pipeline: for **2025 only**, compute each variety's mean reported yield and observation count, sorted best-first, and save the result as `variety_summary_2025`.

**(b)** Write the table to `output/variety_summary_2025.csv`.

**(c)** Report the top row. In a comment: the table also contains varieties observed only twice -- what should a reader keep in mind when comparing rows?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
library(tidyverse)

# Read the full variety file
mb <- read_csv("data/mb_wheat_varieties.csv")

# Keep only the reported rows
mb_rep <- filter(mb, Reported == TRUE)

# (a) Each variety in 2025: mean reported yield and count, best first
variety_summary_2025 <- mb_rep |>
  filter(Year == 2025) |>                      # 2025 only
  group_by(Variety) |>                         # one group per variety
  summarise(mean_yield = mean(Yield_bu_ac),    # average yield
            n = n()) |>                        # observation count
  arrange(desc(mean_yield))                    # highest mean first

# Print the top of the sorted table
head(variety_summary_2025)

# (b) Write the table to a csv
write_csv(variety_summary_2025, "output/variety_summary_2025.csv")

## (c) I find AAC WESTKING on top at 79.2 bu/ac from 17 observations. Rows with tiny counts carry far less evidence than their neighbours; some varieties appear only once or twice, so the n column is part of the answer, not decoration.
')
```

</details>
