# Module 4 - Test bank {.unnumbered}

```{r}
#| include: false
#| eval: true
# console() renders a code listing plus the console session it produces.
source("../_console.R")
suppressMessages(library(tidyverse))
suppressMessages(library(janitor))
```

This bank holds **45 questions**: fifteen in each section.

- Section 1 (Questions 1–15) covers graphing principles
- Section 2 (Questions 16–30) covers graphing in Excel
- Section 3 (Questions 31–45) covers graphing in R

The bank uses the same datasets as the Module 1 bank -- two data sources, with the Saskatchewan yields in two shapes:

- [**Saskatchewan RM Crop Yields**](data/rm_yields_1990plus.csv) (1990–2025). Average yield for eight crops in every Rural Municipality, one column per crop.
- [**Saskatchewan RM Crop Yields (Long)**](data/rm_yields_1990_2025.csv) (1990–2025). The same yields in long format: one row per RM-year-crop, with a `Unit` column. Yields are in bu/ac except lentils, which are in lb/ac.
- [**Manitoba Wheat Variety**](data/mb_wheat_reported_2020_2025.csv) (2020–2025). Red spring wheat by variety and municipality, with the number of farms, seeded acres and average yield.

Your module test will contain one question per section. The datasets you need will be provided in the test. Each question is written to be answerable in fifteen minutes or less.

Answer Sections 1 and 2 in your workbook: one worksheet per question, labelled with the question number, with written answers typed into cells so they are visible when the workbook is opened. Answer Section 3 with an R script that has a header block, a clearly labelled section for each question, and your sentence answers written as comments, and it should run from top to bottom in a fresh session.

For each question, 20% of the mark is for presentation and the other 80% is for the correctness of your charts, workbook or script.

---

# Section 1 — Graphing principles

*These questions are about choosing and judging charts. Write your answers in worksheet cells. Where a question asks for a calculation, do it with a formula in a cell so the TA can see it.*

### Question 1

*In 2025 the provincial average spring wheat yield in Saskatchewan was 53.0 bu/ac and the average durum yield was 48.6 bu/ac.*

**(a)** Would you present these two numbers as a chart or as a sentence? Explain your choice.

**(b)** Write the sentence you would put in a report, quantifying the difference.

**(c)** Describe a version of this comparison (more crops, more years, or more places) where a chart becomes the better choice, and name the chart type you would use.

<details><summary>Answer</summary>

**(a)** A sentence. Two values are easier to state than to plot; a chart of two bars adds ink without adding understanding.

**(b)** Any sentence that carries the numbers: "Spring wheat averaged 53.0 bu/ac in Saskatchewan in 2025, about 4 bu/ac (9%) above durum."

**(c)** Any answer that grows the comparison until a sentence can no longer hold it: the 2025 average yields of all eight crops (a bar chart), or wheat and durum yields every year from 1990 to 2025 (a line chart), or yields across the RMs (a map or a distribution).

</details>

### Question 2

*The table contains 2025 provincial average yields in bushels per acre. Copy it into a blank sheet.*

| Crop | Yield (bu/ac) |
|---|---:|
| Spring Wheat | 53.0 |
| Durum | 48.6 |
| Canola | 43.9 |
| Peas | 43.5 |
| Flax | 30.5 |

Suppose the yields are drawn as a bar chart whose value axis starts at 30 rather than at zero.

**(a)** Calculate the drawn length of the canola bar as a fraction of the spring wheat bar under each axis choice (starting at 30, and starting at zero).

**(b)** In a sentence: what would a reader who only glances at the truncated chart conclude about canola relative to spring wheat?

**(c)** State the rule for where the value axis of a bar chart should start, and why bar charts in particular need it.

<details><summary>Answer</summary>

**(a)** Starting at 30, the canola bar is (43.9 − 30) / (53.0 − 30) = **0.60** of the spring wheat bar. Starting at zero it is 43.9 / 53.0 = **0.83**.

**(b)** That canola yields are around 60% of spring wheat yields, when the true figure is 83% -- the truncated chart roughly doubles the apparent gap.

**(c)** The value axis of a bar chart starts at zero. A bar encodes its value as length, so cutting the base off every bar changes the lengths without changing the labels, and the visual comparison no longer matches the numbers.

</details>

### Question 3

*For each display below, name the chart type you would choose and justify it in one sentence.*

**(a)** How the provincial average canola yield changed from 1990 to 2025.

**(b)** The 2025 average yields of six crops.

**(c)** How Manitoba's 2025 spring wheat acres split across the four largest varieties and everything else.

**(d)** Whether RMs with high spring wheat yields in 2025 also had high barley yields.

<details><summary>Answer</summary>

**(a)** A line chart -- yield against a continuous time axis shows the trend and the year-to-year swings.

**(b)** A bar chart -- six categories compared by length, sorted so the ranking is visible.

**(c)** A pie chart (or a bar chart of shares) -- the five values are shares of one whole, and five slices is few enough to read.

**(d)** A scatter plot -- one point per RM, wheat yield on one axis and barley on the other, shows whether the two move together.

</details>

### Question 4

*A table holds the seeded acres of the three largest Manitoba wheat varieties in each of 2023, 2024 and 2025.*

**(a)** Which bar layout, stacked or grouped, shows how the total acreage of the three varieties changed from year to year?

**(b)** Which layout shows how each individual variety changed from year to year?

**(c)** What does a faceted chart (one small panel per variety) do that a single grouped chart does not?

<details><summary>Answer</summary>

**(a)** Stacked bars -- the height of each year's stack is the total, so the totals can be compared directly.

**(b)** Grouped bars -- every variety's bars sit on the axis baseline, so each variety can be tracked across years. In a stacked chart only the bottom segment has a common baseline.

**(c)** Faceting gives each variety its own panel with a shared scale, so overlapping or crowded series can be read separately while remaining comparable.

</details>

### Question 5

*The long yields file has a `Unit` column: every crop is recorded in bu/ac except lentils, which are in lb/ac (a 2025 provincial average of about 1,900).*

**(a)** Explain why a pie chart of the eight crops' 2025 average yields would be a poor chart, whatever the units.

**(b)** Name data in this module's files that a pie chart could reasonably show.

**(c)** A bar chart of the 2025 average yields puts lentils on the same axis as the bu/ac crops. Describe what the chart looks like and give one way to fix it.

<details><summary>Answer</summary>

**(a)** The slices of a pie must be shares of a whole. Eight average yields do not sum to anything meaningful -- there is no "total yield" the slices divide up -- so the chart implies a whole that does not exist.

**(b)** Shares of seeded acres: for example, each variety's share of Manitoba's 2025 spring wheat acres (grouped to a few slices), since acres do sum to a meaningful total.

**(c)** The lentils bar, at about 1,900, is roughly twenty times the next value, so the bu/ac crops compress into short, unreadable bars. Fixes: leave lentils off the chart (and say so), plot it in its own panel, or convert everything to a common unit.

</details>

### Question 6

*A line chart shows the provincial average durum yield in Saskatchewan for every year from 1990 to 2025. Its title is "Chart 1", its vertical axis is labelled "Value" and its horizontal axis is labelled "Year".*

**(a)** Write a descriptive title for the chart.

**(b)** Write a vertical axis label, including the units.

**(c)** In a sentence: what does a good title tell the reader that the axis labels cannot?

<details><summary>Answer</summary>

**(a)** Something that names the what, where and when: "Average durum yield in Saskatchewan RMs, 1990–2025."

**(b)** "Yield (bu/ac)".

**(c)** The axis labels name the variables; the title states what the chart is about as a whole -- the crop, the place, the period, and (at its best) the point the chart is making.

</details>

### Question 7

*A line chart of six Manitoba wheat varieties' acres over 2020–2025 has: horizontal and vertical gridlines, patterned fills behind the plot, a heavy border around the plot area, bold tick marks on both axes, and value labels on every point of every line.*

**(a)** List the elements you would remove.

**(b)** State the principle that guides what stays and what goes.

**(c)** Name one element you would keep, and why.

<details><summary>Answer</summary>

**(a)** The patterned fills, the heavy border, the vertical gridlines, the bold tick marks, and the value labels on every point. Light horizontal gridlines can stay or go.

**(b)** Every element should help the reader see the data; anything that decorates without informing is clutter and comes off.

**(c)** Reasonable choices: light horizontal gridlines (they help read values off six lines) or the axis labels. The justification matters more than the choice.

</details>

### Question 8

*A chart shows six Manitoba wheat varieties as six coloured lines over 2020–2025, with a legend box on the right mapping each colour to a variety name.*

**(a)** What is the advantage of labelling each line directly at its right-hand end instead of using the legend?

**(b)** When is a legend still the better choice for a line chart?

**(c)** The six lines end at these 2025 values: Brandon 787,111 acres, Starbuck 556,738, Wheatland 400,416, Manness 259,696, Hockley 250,423, Viewfield 140,772. Which direct labels risk colliding, and what would you do about it?

<details><summary>Answer</summary>

**(a)** The reader no longer has to bounce between the plot and the legend, matching colours from memory -- the name sits where the eye already is when it finishes tracing the line.

**(b)** When direct labels will not fit: many lines ending close together or crossing at the right edge, or series that end at different times, so labels would overlap.

**(c)** Manness and Hockley end about 9,000 acres apart -- on an axis running to 800,000 their labels sit almost on top of each other. Nudge one label up and one down (keeping each next to its line), stack the two names, or shorten them.

</details>

### Question 9

*A line chart shows the provincial average canola yield in each year from 1990 to 2025.*

**(a)** A horizontal reference line is added at the average yield across all 36 years. What question can the reader now answer at a glance that they could not before?

**(b)** Suggest one other reference line that could be useful on a yield chart, and say who would use it.

<details><summary>Answer</summary>

**(a)** Which years were above or below normal, and by how much -- each year's distance from the long-run average is now directly visible.

**(b)** Examples: a crop insurance coverage level (a producer sees which years would have triggered a claim), a break-even yield at current prices, or a rolling five-year average showing whether recent years beat the recent past.

</details>

### Question 10

*The average reported Manitoba wheat yield by year: 62.4 bu/ac in 2020, 49.6 in 2021, 59.8 in 2022, 61.4 in 2023, 64.7 in 2024 and 68.0 in 2025. The 2021 drop reflects the drought that year.*

**(a)** You are annotating a line chart of this series. Write the annotation you would place on it, and say where it goes.

**(b)** Why is annotating only the 2021 movement better than annotating every year's movement?

<details><summary>Answer</summary>

**(a)** A short note at the 2021 trough, along the lines of "2021 drought" -- the unusual movement gets the explanation, placed next to the point it explains.

**(b)** An annotation directs attention; six annotations compete for it. The other years' movements are ordinary variation the line already shows, so labelling them adds reading work without adding understanding.

</details>

### Question 11

*You want to plot the provincial average canola yield (bu/ac, roughly 16–44) and lentil yield (lb/ac, roughly 700–1,900) over 1990–2025 on one chart.*

**(a)** What happens if both series go on a single value axis?

**(b)** A second vertical axis on the right side fixes that. What is the danger of a dual-axis chart?

**(c)** Suggest one alternative to the dual axis.

<details><summary>Answer</summary>

**(a)** The lentil series, roughly forty times larger, sets the scale; the canola series compresses into a nearly flat line along the bottom.

**(b)** The right-hand scale is a free choice, and different choices make the two series appear to track each other, diverge, or cross -- the visual story depends on an arbitrary decision rather than the data.

**(c)** Two panels stacked with a shared year axis, or index both series to a common starting year (1990 = 100) so one axis serves both.

</details>

### Question 12

*A scatter plot shows one point per Manitoba wheat variety for 2025: total acres on the x-axis and average yield on the y-axis. The file also records the number of farms growing each variety.*

**(a)** How can the number of farms be added to the chart without adding an axis?

**(b)** What becomes harder for the reader when that information is added?

**(c)** When is this kind of chart worth it anyway?

<details><summary>Answer</summary>

**(a)** Map farms to the size of each point (a bubble chart). Colour is the other common channel, better kept for categories.

**(b)** Judging the sizes: readers compare areas poorly, so the farms values can only be read approximately, and large bubbles can overlap and hide points.

**(c)** When the extra variable is context rather than the main comparison -- the reader needs "this variety is grown by many farms, that one by few" rather than exact counts.

</details>

### Question 13

*The table holds the seeded acres of two Manitoba wheat varieties.*

| Year | AAC Brandon | AAC Starbuck |
|---:|---:|---:|
| 2020 | 1,644,428 | 4,929 |
| 2021 | 1,257,062 | 286,941 |
| 2022 | 1,123,634 | 505,262 |
| 2023 | 1,024,022 | 599,098 |
| 2024 | 904,855 | 550,970 |
| 2025 | 787,111 | 556,738 |

**(a)** In one sentence, state the story a chart of this table should tell.

**(b)** Name the chart type that tells it, and one design choice that would make the story easier to see.

<details><summary>Answer</summary>

**(a)** Something like: "Brandon's acreage has fallen by more than half since 2020 while Starbuck grew from almost nothing to become its main rival."

**(b)** A line chart with one line per variety. Helpful choices: direct labels at the line ends, a zero baseline, or highlighting the two lines in distinct colours. Two grouped bars per year is defensible but harder to read as a trend.

</details>

### Question 14

*You have the 2025 provincial average yields of the eight crops. One reader needs the exact values to put in a report; another wants to see at a glance which crops yield most.*

**(a)** Which presentation serves each reader, and why?

**(b)** How can one display serve both readers at once?

<details><summary>Answer</summary>

**(a)** The report writer is served by a table -- exact values, easy to copy. The comparing reader is served by a sorted bar chart -- ranking and relative size are visible without reading numbers.

**(b)** A sorted bar chart with the value written as a data label at the end of each bar carries the exact numbers and the visual comparison in one display.

</details>

### Question 15

*A line chart shows nine Manitoba wheat variety series over 2020–2025, one colour per variety, and you want the reader to focus on AAC Brandon.*

**(a)** Describe a colour scheme that puts the attention on Brandon.

**(b)** Why do nine distinct bright colours fail here?

<details><summary>Answer</summary>

**(a)** Draw Brandon in one strong colour and every other variety in the same muted grey. The context lines stay readable but recede; the highlighted line carries the story.

**(b)** Nine saturated colours compete equally for attention, so nothing stands out; several will also be hard to tell apart (and harder still for colour-blind readers), turning reading the chart into a legend-matching exercise.

</details>

---

# Section 2 — Graphing in Excel

*Build each chart on the worksheet for that question. Titles and axis labels are part of the marks in this section. The PivotTable steps are the ones from Module 1.*

### Question 16

*The file `mb_wheat_reported_2020_2025.csv` holds Manitoba red spring wheat records by variety, municipality and year: the number of farms, seeded acres and average yield.*

**(a)** Use a PivotTable to total the 2025 acres for each variety.

**(b)** Make a horizontal bar chart of the six most-planted varieties, largest at the top, with a title and a labelled acres axis.

**(c)** In a cell, explain why horizontal bars suit this chart better than vertical ones.

<details><summary>Answer</summary>

The six most-planted 2025 varieties: AAC Brandon (787,111 acres), AAC Starbuck (556,738), AAC Wheatland (400,416), SY Manness (259,696), AAC Hockley (250,423) and AAC Viewfield (140,772). Horizontal bars keep the long variety names readable without rotating them.

<!-- TODO: answer workbook -->

</details>

### Question 17

*The file `rm_yields_1990_2025.csv` holds the average yield of each crop in every RM and year, one row per RM-year-crop.*

**(a)** Use a PivotTable to calculate the provincial average canola yield in each year from 1990 to 2025.

**(b)** Make a line chart of the series with a title and labelled axes, and set the vertical axis to start at zero.

**(c)** Add a data label marking the year with the highest average yield.

<details><summary>Answer</summary>

A PivotTable with `Year` in Rows, `Yield` in Values (changed from Sum to Average), and `Crop` filtered to Canola. The series runs from about 21 bu/ac in 1990 to about 44 in 2025. The data label goes on the maximum point (select the point twice, then Add Data Label).

<!-- TODO: answer workbook -->

</details>

### Question 18

*The same file: one row per RM-year-crop.*

**(a)** Use a PivotTable to calculate the provincial average yields of spring wheat and canola in each year, with the two crops as columns.

**(b)** Plot both series on one line chart with a title, labelled axes and a legend.

**(c)** In a cell, explain what the legend is doing in this chart, and when a line chart can do without one.

<details><summary>Answer</summary>

The PivotTable has `Year` in Rows, `Crop` in Columns (filtered to the two crops) and average `Yield` in Values. The legend maps each colour to a crop; with only one series, or with the lines labelled directly at their right-hand ends, no legend is needed.

<!-- TODO: answer workbook -->

</details>

### Question 19

*The Manitoba wheat file again: variety, municipality, year, farms, acres, yield.*

**(a)** Insert a PivotChart showing the total acres of each variety in each year, 2020 to 2025, drawn as one line per variety.

**(b)** The chart has too many lines to read. Filter it to the major varieties -- I kept the eight whose acres summed over 2020–2025 exceed 300,000 (ticking just those in the PivotChart's field list is fine).

**(c)** In a cell, explain what the filtered chart gains and what information it no longer shows.

<details><summary>Answer</summary>

The PivotChart has `Year` in the axis area, `Variety` in the legend area and `Acres` in Values. Eight varieties have more than 300,000 acres summed over the six years; keep them by unticking the others in the field list (a value filter on the variety field does the same thing faster). The filtered chart makes the major varieties readable; it no longer shows the small varieties at all, or the total acreage across every variety.

<!-- TODO: answer workbook -->

</details>

### Question 20

*The Manitoba wheat file again.*

**(a)** Use a PivotTable to total the 2025 acres by variety.

**(b)** Beside the PivotTable, build a five-row block that keeps the four largest varieties and groups the rest into `Other` (copy the values out of the PivotTable rather than editing it). Make a pie chart of the block, with each slice labelled.

**(c)** In a cell, calculate the share of 2025 acres in the four largest varieties.

<details><summary>Answer</summary>

The four largest 2025 varieties (Brandon, Starbuck, Wheatland, Manness) total 2,003,961 acres of the province's 2,571,678: a **77.9%** share, so the `Other` slice holds the remaining 22.1%.

<!-- TODO: answer workbook -->

</details>

### Question 21

*The file `rm_yields_1990_2025.csv`: one row per RM-year-crop, with lentils in lb/ac and everything else in bu/ac.*

**(a)** Use a PivotTable to calculate the 2025 provincial average yield of each crop, leaving lentils out (its units don't belong on this axis).

**(b)** Make a bar chart of the averages, change it to horizontal bars, and sort so the highest-yielding crop is at the top (watch which sort order puts the largest bar on top).

**(c)** Add a title and a labelled axis.

<details><summary>Answer</summary>

`Crop` in Rows (filtered to exclude Lentils), average `Yield` in Values, `Year` filtered to 2025. The 2025 averages: Oats 97.5, Barley 72.8, Spring Wheat 53.0, Durum 48.6, Canola 43.9, Peas 43.5, Flax 30.5. Horizontal bars put the first row at the bottom, so the table is sorted ascending to get the largest bar on top.

<!-- TODO: answer workbook -->

</details>

### Question 22

*The same file. In 2025 there is one canola yield for each of about 290 RMs.*

**(a)** Get the 2025 canola yields into a column (a filter on the long table, or a PivotTable with RM in Rows whose values you copy out as plain values -- Excel's statistic charts will not draw directly from a PivotTable).

**(b)** Make a histogram of the yields.

**(c)** In a cell, describe the shape of the distribution in one sentence: where the yields centre and how spread out they are.

<details><summary>Answer</summary>

An Insert > Histogram chart on the yield column. The distribution centres around 44 bu/ac; most RMs fall roughly between 30 and 55, with a tail of lower-yielding RMs. Any sentence naming the centre and the spread earns the marks.

<!-- TODO: answer workbook -->

</details>

### Question 23

*The Manitoba wheat file.*

**(a)** Use a PivotTable to total the 2025 acres of the four largest varieties, and make a bar chart of them.

**(b)** Add data labels showing the acres at the end of each bar, formatted with a thousands separator.

**(c)** In a cell, explain what the data labels let you remove from the chart.

<details><summary>Answer</summary>

Data labels are added through Add Chart Element > Data Labels and formatted (Number, with a 1000 separator) in the Format Data Labels pane. With exact values printed on the bars, the gridlines and even the value axis can go -- the labels carry the numbers, the bars carry the comparison.

<!-- TODO: answer workbook -->

</details>

### Question 24

*The file `rm_yields_1990_2025.csv` again. The `Unit` column shows lentils in lb/ac.*

**(a)** Insert a PivotChart of the provincial average yield of every crop by year, one line per crop.

**(b)** One line makes the rest unreadable. Identify it, explain why (look at the `Unit` column), filter it out, and retitle the chart to say what it now shows.

<details><summary>Answer</summary>

The lentils line sits in the thousands (lb/ac) and flattens the bu/ac crops against the axis -- a units problem, not a yields problem. After removing lentils, any descriptive title for the remaining bu/ac crops earns the marks.

<!-- TODO: answer workbook -->

</details>

### Question 25

*The Manitoba wheat file.*

**(a)** Use a PivotTable to total the acres of AAC Brandon and AAC Starbuck in each year, with the two varieties as columns.

**(b)** Plot both series on one line chart with a title, labelled axes and a legend (or direct labels).

**(c)** In a cell, state in one sentence the story the chart tells.

<details><summary>Answer</summary>

Brandon falls from 1,644,428 acres in 2020 to 787,111 in 2025 while Starbuck rises from 4,929 to 556,738 -- the story is an old leading variety being displaced by a new one.

<!-- TODO: answer workbook -->

</details>

### Question 26

*The file `rm_yields_1990plus.csv`: one row per RM-year, one column per crop.*

**(a)** Filter the table to RM 1.

**(b)** Make a line chart of RM 1's canola yield from 1990 to 2025, with a title and labelled axes.

**(c)** In a cell: this chart shows one RM; the provincial average line in the textbook is much smoother. Why?

<details><summary>Answer</summary>

Select the `Year` and `Canola` columns with the filter applied and insert a line chart. A single RM keeps all its local weather luck, so the line swings hard year to year; averaging across ~300 RMs cancels much of that noise, which is why the provincial series is smoother.

<!-- TODO: answer workbook -->

</details>

### Question 27

*The Manitoba wheat file.*

**(a)** Use a PivotTable to total the acres of AAC Brandon and AAC Starbuck in each year, and make a clustered (grouped) column chart from it.

**(b)** Change the chart type to stacked columns.

**(c)** In a cell: which question does each version answer better -- "how did the combined acreage of the two varieties change?" and "how did each variety change?"

<details><summary>Answer</summary>

Stacked columns answer the combined-acreage question: the stack height is the total. Grouped columns answer the per-variety question: every bar starts at zero, so each variety can be tracked across years.

<!-- TODO: answer workbook -->

</details>

### Question 28

*The file `rm_yields_1990plus.csv`. RM 1 grew durum in some years but not others, so its `Durum` column has blank cells.*

**(a)** Filter to RM 1 and make a line chart of the durum yield over 1990–2025.

**(b)** Use **Select Data > Hidden and Empty Cells** to control how the blanks are drawn, and choose the setting you would publish.

**(c)** In a cell: why would filling the blank years with zero be worse than leaving gaps?

<details><summary>Answer</summary>

Showing empty cells as **Gaps** is the honest setting: the line breaks where there is no data. A zero draws the line down to the floor and claims a yield of 0 bu/ac in years when durum simply wasn't grown or reported -- a missing value is not a zero.

<!-- TODO: answer workbook -->

</details>

### Question 29

*The Manitoba wheat file.*

**(a)** Make a bar chart of the 2025 acres of the six most-planted varieties (a PivotTable to total the acres, then sort).

**(b)** Format it so AAC Starbuck stands out: all bars in one muted colour, Starbuck in a strong one (clicking a bar twice selects just that bar). Narrow the gap between the bars.

**(c)** In a cell: when is highlighting one bar the right choice, and when is it misleading?

<details><summary>Answer</summary>

Single-bar fill and gap width both live in the Format Data Series pane. Highlighting is right when the chart's story is about that category ("here is where Starbuck sits among the leaders"); it misleads when the highlight implies importance the data doesn't support, or hides that another bar is larger.

<!-- TODO: answer workbook -->

</details>

### Question 30

*The Manitoba wheat file.*

**(a)** Make a line chart of the total reported acres by year, 2020 to 2025 (a PivotTable with `Year` in Rows and `Acres` in Values).

**(b)** Strip the chart for a report: remove the gridlines and the chart border, and increase the font size of the title and axes until they are readable at a glance.

**(c)** In a cell: which default Excel chart elements did you remove, and what rule decided it?

<details><summary>Answer</summary>

Gridlines and border go (plus the legend if Excel added one -- a single series needs none); larger fonts stay because the defaults are too small to read. The rule: keep what helps the reader see the data, remove what only decorates.

<!-- TODO: answer workbook -->

</details>

---

# Section 3 — Graphing in R

*Answer these in your R script. Load `tidyverse` and `janitor` once at the top, read each file with `read_csv()` and clean the column names with `clean_names()`. Written answers go in comments. The file paths below assume the data sits in your project's `data/` folder.*

### Question 31

*The file `mb_wheat_reported_2020_2025.csv` holds Manitoba red spring wheat records by variety, municipality and year: the number of farms, seeded acres and average yield.*

**(a)** Total the 2025 acres by variety and keep the varieties with more than 100,000 acres.

**(b)** Make a horizontal bar chart of those totals, ordered so the largest bar is at the top (this is `fct_reorder()`), with a title, a labelled acres axis and comma-formatted axis labels.

**(c)** In a comment: what does `fct_reorder()` change about the `variety` column, and what order would the bars take without it?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.0
# Read the Manitoba wheat file and clean the names
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv",
                     show_col_types = FALSE) |>
  clean_names()

# (a) Total 2025 acres by variety, keeping those over 100,000 acres
top_2025 <- mb_wheat |>
  filter(year == 2025) |>                          # 2025 only
  group_by(variety) |>                             # one group per variety
  summarise(acres = sum(acres, na.rm = TRUE)) |>   # total the acres
  filter(acres > 100000)                           # keep the large varieties

top_2025

# (b) Horizontal bar chart, largest at the top
top_2025 |>
  ggplot(aes(x = acres, y = fct_reorder(variety, acres))) +
  geom_col(fill = "DarkGreen") +
  scale_x_continuous(labels = scales::comma) +     # 400,000 not 4e+05
  labs(title = "Acres of Manitoba wheat varieties, 2025",
       x = "Acres", y = NULL) +
  theme_classic(base_size = 13)

# (c) fct_reorder() converts variety to a factor whose categories are
# ordered by acres, so the bars plot in that order. Without it the
# bars would plot alphabetically.
```

</details>

### Question 32

*The file `rm_yields_1990_2025.csv` holds the average yield of each crop in every RM and year, one row per RM-year-crop.*

**(a)** Calculate the provincial average canola yield in each year.

**(b)** Make a line chart of the series with a title, labelled axes and a y-axis that starts at zero (`limits = c(0, NA)`).

**(c)** In a comment: what does the `NA` in `limits = c(0, NA)` do?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.6
# Read the long yields file and clean the names
rm_yields <- read_csv("data/rm_yields_1990_2025.csv",
                      show_col_types = FALSE) |>
  clean_names()

# (a) Provincial average canola yield by year
canola_by_year <- rm_yields |>
  filter(crop == "Canola") |>                          # canola only
  group_by(year) |>                                    # one group per year
  summarise(mean_yield = mean(yield, na.rm = TRUE))    # average across RMs

canola_by_year

# (b) Line chart with a zero baseline
canola_by_year |>
  ggplot(aes(x = year, y = mean_yield)) +
  geom_line(colour = "DarkGreen", linewidth = 1) +
  scale_y_continuous(limits = c(0, NA)) +              # start the axis at zero
  labs(title = "Average canola yield in Saskatchewan RMs",
       x = "Year", y = "Yield (bu/ac)") +
  theme_classic(base_size = 13)

# (c) The NA leaves the upper limit for ggplot to choose from the data,
# so only the lower end of the axis is fixed.
```

</details>

### Question 33

*The same file: one row per RM-year-crop.*

**(a)** Filter to spring wheat and canola (`%in%`) and calculate each crop's provincial average yield in each year.

**(b)** Plot both series on one chart with one line per crop.

**(c)** In a comment: what does mapping `colour = crop` inside `aes()` do that setting a colour inside `geom_line()` does not?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.6
# Read the long yields file and clean the names
rm_yields <- read_csv("data/rm_yields_1990_2025.csv",
                      show_col_types = FALSE) |>
  clean_names()

# (a) Average yield by crop and year for the two crops
two_crops <- rm_yields |>
  filter(crop %in% c("Spring Wheat", "Canola")) |>   # keep either crop
  group_by(crop, year) |>                            # one group per crop-year
  summarise(mean_yield = mean(yield, na.rm = TRUE),
            .groups = "drop")

two_crops

# (b) One line per crop
two_crops |>
  ggplot(aes(x = year, y = mean_yield, colour = crop)) +   # one line per crop
  geom_line(linewidth = 1) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(title = "Average spring wheat and canola yields in Saskatchewan RMs",
       x = "Year", y = "Yield (bu/ac)", colour = NULL) +
  theme_classic(base_size = 13)

# (c) Mapping colour = crop inside aes() ties colour to the data: each
# value of crop gets its own colour and its own line, and a legend is
# added. A colour set inside geom_line() is a fixed appearance choice
# applied to everything, and draws one line.
```

</details>

### Question 34

*The Manitoba wheat file again: variety, municipality, year, farms, acres, yield.*

**(a)** Calculate each variety's total acres across all years and municipalities, and rename varieties with under 300,000 total acres to `"Other"` (`if_else()`).

**(b)** Total the acres by variety and year, and plot one line per variety.

**(c)** In a comment: why must the acres be re-totalled after the renaming before you plot?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.6
#| fig-height: 3.8
# Read the Manitoba wheat file and clean the names
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv",
                     show_col_types = FALSE) |>
  clean_names()

# (a) Attach each variety's all-years total, then rename the small ones
mb_grouped <- mb_wheat |>
  group_by(variety) |>                                     # one group per variety
  mutate(total_acres = sum(acres, na.rm = TRUE)) |>        # total on every row
  ungroup() |>
  mutate(variety = if_else(total_acres < 300000,           # under the threshold
                           "Other",                        # becomes Other
                           variety))                       # otherwise unchanged

# (b) Re-total by variety and year, then plot
wheat_by_year <- mb_grouped |>
  group_by(variety, year) |>
  summarise(acres = sum(acres, na.rm = TRUE), .groups = "drop")

wheat_by_year |>
  ggplot(aes(x = year, y = acres, colour = variety)) +
  geom_line(linewidth = 0.9) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(title = "Acres of Manitoba wheat varieties",
       x = "Year", y = "Total acres", colour = NULL) +
  theme_classic(base_size = 13)

# (c) After the renaming there are many rows labelled Other in each
# year (one per small variety and municipality). Without re-totalling,
# ggplot would draw a line through all of those points instead of one
# line for the combined Other acreage.
```

</details>

### Question 35

*The file `rm_yields_1990plus.csv` has one row per RM-year and one column per crop, with yields in bu/ac.*

**(a)** For 2025, make a scatter plot of spring wheat yield against barley yield across RMs, with a title and labelled axes.

**(b)** In a comment: describe the pattern in one sentence.

**(c)** Save the chart to `output/` as a PNG with `ggsave()`.

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.0
# Read the wide yields file and clean the names
rm_wide <- read_csv("data/rm_yields_1990plus.csv",
                    show_col_types = FALSE) |>
  clean_names()

# (a) Scatter of the two 2025 yields, one point per RM
wheat_barley_scatter <- rm_wide |>
  filter(year == 2025) |>                       # 2025 only
  ggplot(aes(x = spring_wheat, y = barley)) +
  geom_point(colour = "DarkGreen", alpha = 0.6) +
  labs(title = "Spring wheat and barley yields in Saskatchewan RMs, 2025",
       x = "Spring wheat yield (bu/ac)", y = "Barley yield (bu/ac)") +
  theme_classic(base_size = 13)

wheat_barley_scatter

# (b) The points rise together: RMs with higher spring wheat yields
# generally also had higher barley yields in 2025.

# (c) Save the chart as a PNG
ggsave(
  "output/mod04_q35_scatter.png",
  plot = wheat_barley_scatter,
  width = 6.4,
  height = 4.0,
  dpi = 300
)
```

</details>

### Question 36

*The Manitoba wheat file. Riding Mountain West (recorded as `RIDING MOUNTAIN WEST`) reported eleven varieties in 2025.*

**(a)** Filter to Riding Mountain West in 2025 (two conditions in one `filter()`) and total the acres by variety.

**(b)** Make a horizontal bar chart of the varieties ordered by acres, with a title naming the municipality.

**(c)** In a comment: your filter used two conditions -- what does the comma between them mean?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.0
# Read the Manitoba wheat file and clean the names
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv",
                     show_col_types = FALSE) |>
  clean_names()

# (a) One municipality, one year
rmw_2025 <- mb_wheat |>
  filter(municipality == "RIDING MOUNTAIN WEST",
         year == 2025) |>                         # both conditions must hold
  group_by(variety) |>
  summarise(acres = sum(acres, na.rm = TRUE))

rmw_2025

# (b) Ordered horizontal bars
rmw_2025 |>
  ggplot(aes(x = acres, y = fct_reorder(variety, acres))) +
  geom_col(fill = "DarkGreen") +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Wheat variety acres in Riding Mountain West, 2025",
       x = "Acres", y = NULL) +
  theme_classic(base_size = 13)

# (c) The comma is an AND: a row is kept only when it matches the
# municipality and the year.
```

</details>

### Question 37

*The file `rm_yields_1990_2025.csv`. Lentils are recorded in lb/ac; the other crops in bu/ac.*

**(a)** Calculate the 2025 provincial average yield of each crop except lentils (filter it out, and say why in a comment).

**(b)** Make a bar chart of the averages, ordered from highest to lowest.

**(c)** In a comment: what would the chart look like if lentils stayed in?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.8
# Read the long yields file and clean the names
rm_yields <- read_csv("data/rm_yields_1990_2025.csv",
                      show_col_types = FALSE) |>
  clean_names()

# (a) 2025 averages, without lentils -- its lb/ac values would sit on
# a bu/ac axis as if they were comparable
avg_2025 <- rm_yields |>
  filter(year == 2025,
         crop != "Lentils") |>
  group_by(crop) |>
  summarise(mean_yield = mean(yield, na.rm = TRUE))

avg_2025

# (b) Ordered bars
avg_2025 |>
  ggplot(aes(x = mean_yield, y = fct_reorder(crop, mean_yield))) +
  geom_col(fill = "DarkGreen") +
  labs(title = "Average 2025 crop yields in Saskatchewan RMs",
       x = "Yield (bu/ac)", y = NULL) +
  theme_classic(base_size = 13)

# (c) The lentils bar, near 1,900, would stretch the axis twenty-fold
# and squash the bu/ac crops into unreadable slivers -- a units
# problem drawn as if it were a yields story.
```

</details>

### Question 38

*The Manitoba wheat file.*

**(a)** Total AAC Brandon's acres in each year (the variety is recorded as `AAC BRANDON (BW 932)`).

**(b)** Make a line chart of the series with a title, labelled axes, comma-formatted axis labels and a y-axis starting at zero.

**(c)** In a comment: with the zero baseline, roughly what fraction of its 2020 acreage does Brandon keep by 2025? Would a truncated axis make the decline look smaller or larger?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.6
# Read the Manitoba wheat file and clean the names
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv",
                     show_col_types = FALSE) |>
  clean_names()

# (a) Brandon's total acres by year
brandon <- mb_wheat |>
  filter(variety == "AAC BRANDON (BW 932)") |>
  group_by(year) |>
  summarise(acres = sum(acres, na.rm = TRUE))

brandon

# (b) Line chart with a zero baseline
brandon |>
  ggplot(aes(x = year, y = acres)) +
  geom_line(colour = "DarkGreen", linewidth = 1) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(title = "Acres of AAC Brandon wheat in Manitoba",
       x = "Year", y = "Acres") +
  theme_classic(base_size = 13)

# (c) About half: 787,111 of 1,644,428 acres, or 48%. A truncated
# axis would make the decline look larger -- the line would fall
# almost the full height of the chart.
```

</details>

### Question 39

*The Manitoba wheat file.*

**(a)** Total the 2025 acres by variety, keeping the four largest varieties and grouping the rest into `"Other"` (`if_else()` on the totals).

**(b)** Make a pie chart of the five values (`geom_col()` plus `coord_polar()`), with a title.

**(c)** In a comment: what makes acres suitable for a pie chart when yields are not?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.4
#| fig-height: 3.4
# Read the Manitoba wheat file and clean the names
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv",
                     show_col_types = FALSE) |>
  clean_names()

# (a) 2025 totals, then keep the four largest varieties and group the
# rest as Other; the threshold sits between the 4th- and 5th-largest
pie_data <- mb_wheat |>
  filter(year == 2025) |>
  group_by(variety) |>
  summarise(acres = sum(acres, na.rm = TRUE)) |>
  mutate(variety = if_else(acres > 255000, variety, "Other")) |>
  group_by(variety) |>
  summarise(acres = sum(acres))

pie_data

# (b) Pie chart
pie_data |>
  ggplot(aes(x = "", y = acres, fill = variety)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  labs(title = "2025 Manitoba wheat acres by variety",
       x = NULL, y = NULL, fill = NULL) +
  theme_void(base_size = 13)

# (c) The acres sum to the province's total seeded acres, so each
# slice is a genuine share of a whole. Yields do not sum to anything
# meaningful, so slices of yield have no interpretation.
```

</details>

### Question 40

*The file `rm_yields_1990_2025.csv`.*

**(a)** Make a bar chart of the 2025 provincial average yield of each crop except lentils, ordered from highest to lowest, and assign the chart to an object.

**(b)** Save it twice with `ggsave()`: once as a PNG and once as a PDF, both 6.4 inches wide and 4 inches tall.

**(c)** In a comment: when is the PDF the better file to hand over?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.0
# Read the long yields file and clean the names
rm_yields <- read_csv("data/rm_yields_1990_2025.csv",
                      show_col_types = FALSE) |>
  clean_names()

# (a) The chart as an object
yield_bar <- rm_yields |>
  filter(year == 2025,
         crop != "Lentils") |>
  group_by(crop) |>
  summarise(mean_yield = mean(yield, na.rm = TRUE)) |>
  ggplot(aes(x = mean_yield, y = fct_reorder(crop, mean_yield))) +
  geom_col(fill = "DarkGreen") +
  labs(title = "Average 2025 crop yields in Saskatchewan RMs",
       x = "Yield (bu/ac)", y = NULL) +
  theme_classic(base_size = 13)

yield_bar

# (b) Save as PNG and as PDF
ggsave("output/mod04_q40_yields.png", plot = yield_bar,
       width = 6.4, height = 4.0, dpi = 300)
ggsave("output/mod04_q40_yields.pdf", plot = yield_bar,
       width = 6.4, height = 4.0)

# (c) The PDF keeps lines and text as vectors, so it stays sharp at
# any size -- the right choice for print and publication. The PNG is
# for slides and web pages.
```

</details>

### Question 41

*The Manitoba wheat file. The code below draws a chart of the total reported acres by year, but it is unfinished.*

```r
mb_wheat |>
  group_by(year) |>
  summarise(acres = sum(acres, na.rm = TRUE)) |>
  ggplot(aes(x = year, y = acres)) +
  geom_line()
```

**(a)** Finish the chart: a title, labelled axes, comma-formatted axis labels, a y-axis starting at zero, and a cleaner theme with a readable font size.

**(b)** In a comment: which of your additions changes what the reader concludes from the chart (rather than just how it looks)?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.6
# Read the Manitoba wheat file and clean the names
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv",
                     show_col_types = FALSE) |>
  clean_names()

# (a) The finished chart
mb_wheat |>
  group_by(year) |>
  summarise(acres = sum(acres, na.rm = TRUE)) |>
  ggplot(aes(x = year, y = acres)) +
  geom_line(colour = "DarkGreen", linewidth = 1) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(title = "Total reported wheat acres in Manitoba",
       x = "Year", y = "Acres") +
  theme_classic(base_size = 13)

# (b) The zero baseline is the one that changes the conclusion: with
# ggplot's default limits the small year-to-year movements fill the
# whole panel and look dramatic; from zero they are visibly modest.
# The title, labels, commas and theme change readability, not meaning.
```

</details>

### Question 42

*The Manitoba wheat file: in 2025 it holds 23 varieties.*

**(a)** For 2025, calculate each variety's total acres and average yield, and make a scatter plot with acres on the x-axis (comma labels) and yield on the y-axis.

**(b)** In a comment: is the most-planted variety also the highest-yielding in this data?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.0
# Read the Manitoba wheat file and clean the names
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv",
                     show_col_types = FALSE) |>
  clean_names()

# (a) One point per variety: total acres and average reported yield
varieties_2025 <- mb_wheat |>
  filter(year == 2025) |>
  group_by(variety) |>
  summarise(acres = sum(acres, na.rm = TRUE),
            yield = mean(yield_bu_ac, na.rm = TRUE))

varieties_2025 |>
  ggplot(aes(x = acres, y = yield)) +
  geom_point(colour = "DarkGreen", size = 2) +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Manitoba wheat varieties, 2025: acres and yield",
       x = "Total acres", y = "Average yield (bu/ac)") +
  theme_classic(base_size = 13)

# (b) No. AAC Brandon has by far the most acres (about 787,000) with
# an average yield near 66 bu/ac, while smaller varieties such as AAC
# Viewfield average over 73. The most-planted variety is not the
# top-yielding one in this data.
```

</details>

### Question 43

*The file `rm_yields_1990plus.csv`: one row per RM-year, one column per crop. To draw one line per crop in ggplot, the crops need to be in one column.*

**(a)** For canola and barley, calculate the provincial average yield in each year (this is easiest after `pivot_longer()` on the two crop columns).

**(b)** Plot the two series with one line per crop.

**(c)** In a comment: why did the wide table need reshaping before ggplot could colour by crop?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.6
# Read the wide yields file and clean the names
rm_wide <- read_csv("data/rm_yields_1990plus.csv",
                    show_col_types = FALSE) |>
  clean_names()

# (a) Two crop columns into one, then average by crop and year
two_long <- rm_wide |>
  pivot_longer(cols = c(canola, barley),      # the two crops
               names_to = "crop",
               values_to = "yield") |>
  group_by(crop, year) |>
  summarise(mean_yield = mean(yield, na.rm = TRUE),
            .groups = "drop")

two_long

# (b) One line per crop
two_long |>
  ggplot(aes(x = year, y = mean_yield, colour = crop)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(title = "Average canola and barley yields in Saskatchewan RMs",
       x = "Year", y = "Yield (bu/ac)", colour = NULL) +
  theme_classic(base_size = 13)

# (c) colour = crop needs a crop column to map from. In the wide
# table the crop lives in the column names, not in a column, so
# there is nothing for aes() to point at until the table is long.
```

</details>

### Question 44

*The Manitoba wheat file.*

**(a)** Make a vertical bar chart of the 2025 acres of the varieties with more than 100,000 acres -- the six largest (variety on the x-axis).

**(b)** Make the same chart with horizontal bars (swap the axes in `aes()`), ordered by acres.

**(c)** In a comment: which version is easier to read here, and what feature of this data decides it?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.0
# Read the Manitoba wheat file and clean the names
mb_wheat <- read_csv("data/mb_wheat_reported_2020_2025.csv",
                     show_col_types = FALSE) |>
  clean_names()

# The six largest 2025 varieties
top_six <- mb_wheat |>
  filter(year == 2025) |>
  group_by(variety) |>
  summarise(acres = sum(acres, na.rm = TRUE)) |>
  filter(acres > 100000)        # keeps the six largest

# (a) Vertical bars: the variety names collide on the x-axis
top_six |>
  ggplot(aes(x = variety, y = acres)) +
  geom_col(fill = "DarkGreen") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Acres of Manitoba wheat varieties, 2025",
       x = NULL, y = "Acres") +
  theme_classic(base_size = 13)

# (b) Horizontal bars, ordered
top_six |>
  ggplot(aes(x = acres, y = fct_reorder(variety, acres))) +
  geom_col(fill = "DarkGreen") +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Acres of Manitoba wheat varieties, 2025",
       x = "Acres", y = NULL) +
  theme_classic(base_size = 13)

# (c) Horizontal. The variety names are long, so on the x-axis they
# overlap or need rotating; on the y-axis they read naturally.
```

</details>

### Question 45

*The file `rm_yields_1990_2025.csv`.*

**(a)** Calculate the provincial average canola yield by year and plot the full 1990–2025 series as a line chart with a zero baseline.

**(b)** Make a second chart of the same series restricted to 2015–2025 (`filter(year >= 2015)`).

**(c)** In a comment: what does the short window make more visible, and what does it hide?

<details><summary>Answer</summary>

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.6
# Read the long yields file and clean the names
rm_yields <- read_csv("data/rm_yields_1990_2025.csv",
                      show_col_types = FALSE) |>
  clean_names()

# (a) The full series
canola_by_year <- rm_yields |>
  filter(crop == "Canola") |>
  group_by(year) |>
  summarise(mean_yield = mean(yield, na.rm = TRUE))

canola_by_year |>
  ggplot(aes(x = year, y = mean_yield)) +
  geom_line(colour = "DarkGreen", linewidth = 1) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(title = "Average canola yield in Saskatchewan RMs, 1990-2025",
       x = "Year", y = "Yield (bu/ac)") +
  theme_classic(base_size = 13)

# (b) The last decade only
canola_by_year |>
  filter(year >= 2015) |>
  ggplot(aes(x = year, y = mean_yield)) +
  geom_line(colour = "DarkGreen", linewidth = 1) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(title = "Average canola yield in Saskatchewan RMs, 2015-2025",
       x = "Year", y = "Yield (bu/ac)") +
  theme_classic(base_size = 13)

# (c) The short window makes recent movements easier to see -- the
# 2021 drought trough (about 22 bu/ac) and the recovery to about 44
# by 2025. It hides the long-run context: over 36 years yields have
# roughly doubled, and swings like 2021 have happened before.
```

</details>
