# Merging Data {#sec-merging-data}

```{r}
#| include: false
#| eval: true
source("_console.R")
```

Data for one question often arrives in several tables. The Saskatchewan yield data from @sec-direct-download identifies rural municipalities and years. The weather data identifies stations and years. A third table says which weather station is closest to each RM.

A **merge**, also called a **join**, matches rows that refer to the same thing and puts their columns together. In this chapter we will merge the three tables so that every RM-year has its crop yields and May to August precipitation in one row.

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Identify the key that connects two tables.
2. Distinguish one-to-one, many-to-one and many-to-many relationships.
3. Merge tables in Excel with `XLOOKUP`, including a lookup that uses two columns.
4. Merge tables in R with `left_join()` and related functions.
5. Check for unmatched rows and unexpected changes in the row count.
:::

## The Three Tables {#sec-three-tables}

The files are:

- [`rm_yields_2015_2024.csv`](practice/data/rm_yields_2015_2024.csv), with one row per RM per year. It contains 2,950 rows for ten years and six crops. Yields are measured in bushels per acre.
- [`station_precip.csv`](practice/data/station_precip.csv), with one row per weather station per year. It contains May to August precipitation for sixteen stations and a count of days with no reading.
- [`rm_lookup.csv`](practice/data/rm_lookup.csv), with one row per RM. It contains the RM name, census division, 2021 population and the closest of the sixteen weather stations.

The yields table and rain table have no column in common. The lookup table connects them. We first attach a weather station to each yield row, then attach the rain recorded at that station in that year.

The yields are the Government of Saskatchewan data described in @sec-direct-download, restricted to 2015–2024. The other two files were assembled for this course from public sources. `station_precip.csv` sums Environment and Climate Change Canada daily station observations from May 1 through August 31; `days_missing` records dates with no observation, and the total sums the days that were available. `rm_lookup.csv` combines RM names and 2021 population from Statistics Canada with the nearest station among the sixteen selected for the course. `station_km` records the resulting distance. These construction choices are part of the data's provenance, not facts supplied by either original file.

### Keys {.unnumbered}

A **key** is a column, or a set of columns, used to identify and match rows. The key follows from what one row represents:

- A row in the yields table is one RM in one year, so its key is `RM` and `Year` together.
- A row in the lookup table is one RM, so its key is `RM_Number`.
- A row in the rain table is one station in one year, so its key is `Station` and `Year` together.

The first merge uses RM number. The second uses station and year. Station alone is not enough because each station appears once in every year.

### One-to-one, many-to-one and many-to-many {.unnumbered}

The number of rows sharing a key determines what a join returns. @fig-join-types uses a table of student grades to show the three cases.

![Three kinds of merge. Each line connects a grades row to the row it matches.](images/join_types.svg){#fig-join-types}

- **One-to-one (1:1).** Each key appears once in both tables. Grades and attendance both have one row per student-class, so the merged table has the same number of rows as either input.
- **Many-to-one (n:1).** The key repeats in one table and is unique in the other. Several grade rows can share one professor. In our data, each RM has ten yield rows but only one lookup row. A many-to-one join should have the same number of rows as the table on the many side.
- **Many-to-many (n:n).** The key repeats in both tables. Each row on one side matches every row with the same key on the other. If a class has two TAs, each grade row appears twice. If we join yields to rain by station but leave out year, each yield row matches ten years of rain.

Before joining, decide which relationship you expect. Then compare the row counts before and after.

## Merging in Excel {#sec-merge-excel}

In Excel, `XLOOKUP` can bring a column from one table into another (@sec-lookup). The **Merge** sheet in the workbook below contains thirty yield rows and the two lookup columns beside them. The `FORMULATEXT` column shows the precipitation formula, and the checks below the data count `#N/A` results.

The first lookup matches the RM number and returns its station. The second lookup needs station and year, so it joins the two values on each side with `& "|" &` to make a composite key. The shaded column deliberately leaves out year. It returns the first rain value for that station in the rain sheet—2015 here—for every year, without producing an error. These formulas require Microsoft 365's `XLOOKUP`, which University of Saskatchewan students can use through the university.

::: {.workbook-callout}
<!-- Public view-only link expires 2026-11-20 under the University of Saskatchewan 90-day policy. -->

::: {.excel-embed}
<iframe width="402" height="700" frameborder="0" scrolling="no" src="https://usaskca1-my.sharepoint.com/personal/pjs998_usask_ca/_layouts/15/Doc.aspx?sourcedoc=%7BE4286D57-D90E-4E60-9E9A-C493B5099E8E%7D&action=embedview&wdAllowInteractivity=False&wdDownloadButton=True&wdInConfigurator=True&ActiveCell=%27Merge%27!A1"></iframe>
:::

[Download this workbook](textbook_examples/mod03_merge_excel.xlsx) · [Open online](https://usaskca1-my.sharepoint.com/:x:/g/personal/pjs998_usask_ca/IQBXbSjkDtlgTp6axJO1CZ6OAa9jBZn7UhVyZRjoYXeXy3Q?e=AcDC7T) (File → Save a Copy to view formulas and edit)
:::

An unmatched `XLOOKUP` returns `#N/A`, which is easy to count. An incomplete key can be harder to find because the formula returns a value. Before a lookup, confirm that the key on the lookup side is unique: `XLOOKUP` silently returns the first duplicate. Check a few rows against the source tables, especially where the year or another part of the key changes. Excel lookups ignore case but not stray spaces, and the number `1` does not match the text `"1"`; if an entire column returns `#N/A`, check the key types and use `TRIM()` to remove extra spaces.

For a merge that must be repeated, Power Query records the source, key columns and join type. Import each table with **Data > Get Data > From Text/CSV > Transform Data**. In the editor choose **Home > Merge Queries**, select the key columns in the same order on both sides using Ctrl-click (Command-click on a Mac), and keep **Left Outer** as the join kind. The dialog reports how many rows match. After clicking **OK**, expand the new Table column with the double-arrow icon in its header. Power Query also offers Inner, Full Outer and Left Anti joins, corresponding to the R joins below.

## Merging in R {#sec-merge-r}

Read the three files and inspect their columns:

```{r}
#| eval: true
#| echo: false
console('
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

# Read the three tables and clean their column names
yields <- read_csv(
  "practice/data/rm_yields_2015_2024.csv",
  show_col_types = FALSE
) |> clean_names()
rms <- read_csv(
  "practice/data/rm_lookup.csv",
  show_col_types = FALSE
) |> clean_names()
rain <- read_csv(
  "practice/data/station_precip.csv",
  show_col_types = FALSE
) |> clean_names()

glimpse(yields)
glimpse(rms)
glimpse(rain)
')
```

`left_join()` takes two data frames and a `by` argument that identifies the key. If a key has different names in the two tables, write `join_by(left_name == right_name)`; a column with the same name on both sides is written once. The older `by = c("left_name" = "right_name")` form still appears in other people's code.

### Four joins {.unnumbered}

The join functions differ in which rows they keep:

- **`left_join(a, b)`** keeps every row of `a`. New columns contain `NA` where `b` has no match.
- **`inner_join(a, b)`** keeps only rows that matched in both tables.
- **`full_join(a, b)`** keeps every row from both tables, with `NA` where either side has no match.
- **`anti_join(a, b)`** returns the rows of `a` that have no match in `b`. It is useful for finding failed matches before a join.

We will use `left_join()` because the yield rows are the observations we want to keep.

Before joining, check that the columns expected to identify one row are actually unique:

```{r}
#| eval: true
#| echo: false
console('
rms |>
  count(rm_number) |>
  filter(n > 1)

rain |>
  count(station, year) |>
  filter(n > 1)
')
```

Both checks return empty tibbles. A key is only a key if each combination occurs no more than once on the side where uniqueness is expected.

### First join: yields and the RM lookup {.unnumbered}

The RM column is called `rm` in the yields table and `rm_number` in the lookup. Check for unmatched RMs, perform the join and compare the row counts:

```{r}
#| eval: true
#| echo: false
console('
# List RM numbers that have no lookup row
yields |>
  anti_join(rms, by = join_by(rm == rm_number)) |>
  distinct(rm)

# Add the RM information and nearest station
with_station <- yields |>
  left_join(rms,
            by = join_by(rm == rm_number),
            relationship = "many-to-one")

nrow(yields)
nrow(with_station)
')
```

The `anti_join()` returns an empty tibble, so every RM has a match. `yields` has 2,950 rows and so does `with_station`, which is what a many-to-one join should give. The `relationship = "many-to-one"` argument records what we expect and makes `left_join()` stop if the lookup contains more than one row for an RM.

In R, text keys are sensitive to both case and spaces. For example, `"canola"`, `"Canola"` and `"Canola "` are three different keys. These differences have to be reconciled before the join (@sec-categories).

### A failed match on purpose {.unnumbered}

The complete rain table matches every station-year. To see how the joins differ when data are missing, temporarily remove 2021:

```{r}
#| eval: true
#| echo: false
console('
rain_short <- rain |>
  filter(year != 2021)

# Which yield rows no longer have rain?
with_station |>
  anti_join(
    rain_short,
    by = join_by(weather_station == station, year)
  ) |>
  count(weather_station)

# A left join keeps all 2,950 yield rows,
# including 295 rows with missing rain.
with_station |>
  left_join(
    rain_short,
    by = join_by(weather_station == station, year)
  ) |>
  filter(is.na(precip_may_aug_mm)) |>
  nrow()

# An inner join drops those 295 rows.
with_station |>
  inner_join(
    rain_short,
    by = join_by(weather_station == station, year)
  ) |>
  nrow()

# A full join also keeps a right-side row
# that has no yield-side match.
rain_extra <- rain_short |>
  add_row(
    station = "Course example",
    year = 2025,
    precip_may_aug_mm = 200,
    days_missing = 0
  )

with_station |>
  full_join(
    rain_extra,
    by = join_by(weather_station == station, year)
  ) |>
  filter(weather_station == "Course example") |>
  select(year, weather_station, rm, canola)
')
```

`anti_join()` identifies the failed keys, `left_join()` makes the missing matches visible as `NA`, and `inner_join()` silently removes them. The final `full_join()` also retains the deliberately added rain row that has no yield-side match, filling its yield columns with `NA`.

### Second join: add the rain {.unnumbered}

The second key has two parts. The station column is called `weather_station` on the left and `station` on the right; `year` has the same name in both tables.

```{r}
#| eval: true
#| echo: false
console('
# Find station-years with no precipitation row
with_station |>
  anti_join(rain,
            by = join_by(weather_station == station, year)) |>
  distinct(weather_station, year)

# Add precipitation for the matching station and year
merged <- with_station |>
  left_join(rain,
            by = join_by(weather_station == station, year),
            relationship = "many-to-one")

nrow(merged)
')
```

Again there are no unmatched keys and the row count remains 2,950.

### The row-count trap {.unnumbered}

Now leave `year` out of the key:

```{r}
#| eval: true
#| echo: false
console('
# This key is incomplete: every row matches ten years of rain
with_station |>
  left_join(rain,
            by = join_by(weather_station == station)) |>
  nrow()

# State the expected relationship to turn
# the warning into an error.
try(
  with_station |>
    left_join(
      rain,
      by = join_by(weather_station == station),
      relationship = "many-to-one"
    )
)
')
```

Each yield row matches all ten years for its station, so 2,950 rows become 29,500. By default, `dplyr` warns about the unexpected many-to-many relationship. When `relationship = "many-to-one"` states what was intended, the same mistake becomes an error and the analysis stops. Excel's station-only `XLOOKUP` makes the mistake differently: it returns only the first matching year and gives no warning.

## What the Merge Gives You {#sec-merge-result}

The merged table has one row per RM-year, with yield and precipitation beside one another:

```{r}
#| eval: true
#| echo: false
console('
# Display the columns used in the analysis
merged |>
  select(year, rm, rm_name, canola,
         weather_station, precip_may_aug_mm)

# Calculate annual mean yield and precipitation
merged |>
  filter(!is.na(canola)) |>
  summarise(
    canola_bu_ac = mean(canola),
    rain_mm = mean(precip_may_aug_mm),
    max_days_missing = max(days_missing),
    .by = year
  )

# Count yield values excluded from those means
merged |>
  summarise(missing_canola = sum(is.na(canola)))

# Describe association, not causation
merged |>
  summarise(
    yield_rain_correlation = cor(
      canola,
      precip_may_aug_mm,
      use = "complete.obs"
    )
  )
')
```

Across RM-years, wetter observations tend to have higher canola yield, but the relationship is loose: the correlation is about 0.41. The annual table also resists a simple rainfall story. Both 2021 and 2024 have low average yields despite ordinary or above-average rainfall. These data cannot say why. Testing whether heat contributed would require temperature data; explaining the remaining differences would require other variables as well.

The annual means exclude 46 RM-years with no published canola yield. They also average rainfall over RM rows, so a station serving more RMs receives more weight than a station serving fewer RMs. The merge has further limits that should be recorded in the project's README. The median RM is 62 km from its assigned station, one quarter are more than 85 km away, and the maximum is 161 km, so station rainfall is a coarse proxy for rainfall across an RM. Some station-years also have many missing daily observations—up to 25 days in this file—and their seasonal totals sum only the available days, which biases those totals downward.
