# Working with Spatial Data in R {#sec-spatial-r}

```{r}
#| include: false
#| eval: true
source("_console.R")
# Chunks in this chapter run with practice/ as the working directory, so the
# student-facing paths (data/..., output/...) resolve at render time.
knitr::opts_knit$set(root.dir = "practice")
suppressMessages({library(tidyverse); library(sf); library(terra)})
```

@sec-spatial-data treated spatial data as something to look at. This chapter makes it something to work with. The **sf** package handles vector data and the **terra** package handles rasters, and both give you objects that the Module 2 verbs mostly work on unchanged.

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Read vector data with `st_read()` and interpret the header it prints.
2. Turn a table of coordinates into spatial points.
3. Check and transform a coordinate reference system, and measure area in a projected one.
4. Join ordinary data to boundaries, and find which polygon contains a point.
5. Read a raster with terra, and crop and mask it with a polygon.
6. Write spatial results to a file.
:::

The two working files are the RM boundaries from @sec-spatial-data and a slice of the 2023 Annual Crop Inventory covering two RMs, [`aci_rm_slice_2023.tif`](practice/data/aci_rm_slice_2023.tif).

## Reading vector data

`st_read()` reads most vector formats -- GeoJSON, GeoPackage, and shapefiles, which arrive as a folder of files that travel together.

```{r}
#| eval: true
#| echo: false
console('
# Load the package for vector spatial data
library(sf)

# Read the RM boundaries
rms <- st_read("data/sask_rms.geojson")
')
```

The banner is worth reading before the data: 297 features, geometry type `MULTIPOLYGON`, a bounding box, and the CRS, `WGS 84`. Those are the questions @sec-spatial-data taught, answered before you look at a single row.

```{r}
#| eval: true
#| echo: false
console('
# The first rows: a data frame with a geometry column
head(rms, 3)
')
```

Everything from Module 2 applies. `filter()` keeps rows, `select()` keeps columns, and the geometry column follows along.

## Points from coordinates

A table with longitude and latitude columns becomes spatial with `st_as_sf()`. The `coords` argument names the two columns, longitude first, and `crs = 4326` states what the numbers mean.

```{r}
#| eval: true
#| echo: false
console('
# Four cities as a plain table
cities <- tibble(
  city = c("Saskatoon", "Regina", "Swift Current", "Yorkton"),
  lon  = c(-106.67, -104.61, -107.79, -102.46),
  lat  = c(52.13, 50.45, 50.28, 51.21)
)

# Turn the coordinate columns into point geometry
cities_sf <- st_as_sf(cities, coords = c("lon", "lat"), crs = 4326)

# Print the spatial version
cities_sf
')
```

Both objects are now layers that can share a map:

```{r}
#| label: fig-points-on-rms
#| eval: true
#| echo: true
#| message: false
#| fig-width: 5.5
#| fig-height: 5.8
#| fig-cap: "Two vector layers on one map: the RM polygons and four cities as points."
#| fig-alt: "A map of the southern Saskatchewan RM grid with four labelled points."
# The polygons, then the points on top
ggplot() +
  geom_sf(data = rms, fill = "grey97", colour = "grey60", linewidth = 0.1) +
  geom_sf(data = cities_sf, colour = "#4A7C59", size = 2) +
  theme_void()
```

## Transforming coordinate systems

`st_crs()` reports what a layer's coordinates mean, and `st_transform()` converts them. Transforming the RMs to UTM Zone 13 turns degrees into metres:

```{r}
#| eval: true
#| echo: false
console('
# The CRS the file arrived in
st_crs(rms)$Name

# Convert from degrees to UTM Zone 13 metres
rms_utm <- st_transform(rms, 32613)

# The bounding box before and after
st_bbox(rms)
st_bbox(rms_utm)
')
```

Same province, different numbers: the first box is in degrees, the second in metres. Measurement is the reason to care. With the coordinates in metres, `st_area()` gives a real area -- here, the RM of Corman Park in square kilometres:

```{r}
#| eval: true
#| echo: false
console('
# Area of RM 344, Corman Park, in square kilometres
rms_utm |>
  filter(RMNO == "344") |>
  st_area() |>
  as.numeric() / 1e6      # drop the m^2 label, then scale
')
```

## Attaching a table to boundaries

The RM boundaries carry no yields, and the RM yield table carries no shapes. The RM number links them. Read the yield file and keep the 2023 canola column:

```{r}
#| eval: true
#| echo: false
console('
# Read the wide RM yield file
rm_yields <- read_csv("data/rm_yields_1990plus.csv", show_col_types = FALSE)

# One row per RM: 2023 canola
canola_2023 <- rm_yields |>
  filter(Year == 2023) |>
  select(RM, Canola)

# Join the yields onto the boundaries
rm_canola <- left_join(rms, canola_2023, by = c("RMNO" = "RM"))
')
```

The join refuses to run: `RMNO` came out of the boundary file as text and `RM` is a number. The fix is to make the types agree before joining, and then to check what the join produced:

```{r}
#| eval: true
#| echo: false
console('
# Make the key numeric, then join
rm_canola <- rms |>
  mutate(RMNO = as.numeric(RMNO)) |>
  left_join(canola_2023, by = c("RMNO" = "RM"))

# Same 297 rows as the boundaries
nrow(rm_canola)

# RMs with no reported canola in 2023
sum(is.na(rm_canola$Canola))
')
```

A left join keeps every boundary, so the row count matching `rms` is the first check, and the count of missing yields is the second -- those RMs reported no canola that year, and @sec-maps has to decide how to colour them. Mapping this object is where that chapter starts.

## Which polygon holds a point

A **spatial join** matches rows by location instead of by key. `st_join()` attaches to each point the attributes of the polygon it falls in:

```{r}
#| eval: true
#| echo: false
console('
# Which RM is each city in?
st_join(cities_sf, rms) |>
  select(city, RMNO, RMNM)
')
```

Counting many points per polygon is the same join followed by `count()`. Module 8 uses exactly that to summarize thousands of machinery points inside a field boundary (@sec-combining-spatial).

## Reading raster data

The **terra** package does for rasters what sf does for vectors. `rast()` reads the file, and printing the object answers the @sec-spatial-data questions -- how many cells, how big each one is, how much ground they cover, and in what CRS:

```{r}
#| eval: true
#| echo: false
console('
# Load the package for raster data
library(terra)

# Read the crop inventory slice
aci <- rast("data/aci_rm_slice_2023.tif")

# Size, resolution, extent and CRS in one printout
aci
')
```

The slice covers the RMs of Wood Creek and McCraney at 30 metres, and its CRS is already UTM Zone 13. @fig-aci-slice puts the boundaries over the cells.

```{r}
#| label: fig-aci-slice
#| eval: true
#| echo: true
#| message: false
#| fig-width: 7.0
#| fig-height: 4.0
#| fig-cap: "The 2023 Annual Crop Inventory over the RMs of Wood Creek (281, right) and McCraney (282, left), with the RM boundaries as a vector layer on top."
#| fig-alt: "A wide map of coloured crop cells with two black polygon outlines over them."
# Shrink for display: each 8-by-8 block of cells becomes one cell
aci_small <- aggregate(aci, fact = 8, fun = "modal")

# The class table pairs each value with a crop name and a colour
classes <- read_csv("data/aci_crop_classes.csv", show_col_types = FALSE)

# Raster to a data frame of cell centres and values
aci_df <- as.data.frame(aci_small, xy = TRUE) |>
  rename(value = 3) |>                     # the third column holds the class
  left_join(classes, by = "value")         # attach names and colours

# The named palette for the classes present
pal <- aci_df |>
  distinct(crop, colour) |>
  deframe()

# The two RM boundaries, in the raster's CRS
pair <- rms_utm |>
  filter(RMNO %in% c("281", "282"))

ggplot() +
  geom_raster(data = aci_df, aes(x, y, fill = crop)) +
  geom_sf(data = pair, fill = NA, colour = "black", linewidth = 0.5) +
  scale_fill_manual(values = pal) +
  theme_void() +
  # A compact legend leaves more room for the map
  theme(legend.key.size = unit(4, "mm"),
        legend.text = element_text(size = 8)) +
  labs(fill = NULL)
```

The overlay works because both layers are in the same CRS. Had the boundaries stayed in degrees, the outlines would land nowhere near the cells -- the rule from @sec-spatial-data, in raster form.

## Cutting a raster with a polygon

`crop()` trims a raster to a polygon's bounding box and `mask()` blanks the cells outside the polygon itself. Together they cut one RM out of the slice, and `freq()` counts what is in it:

```{r}
#| eval: true
#| echo: false
console('
# Wood Creek, in the raster CRS
wood_creek <- rms_utm |>
  filter(RMNO == "281")

# Trim to the bounding box, then blank the cells outside the RM
aci_wc <- crop(aci, vect(wood_creek)) |>
  mask(vect(wood_creek))

# Count the cells in each class and turn counts into shares
freq(aci_wc) |>
  left_join(classes, by = "value") |>       # attach the crop names
  mutate(share = count / sum(count)) |>     # share of the RM
  select(crop, count, share) |>
  arrange(desc(share)) |>
  head(6)
')
```

This is the aggregation direction from @sec-spatial-data in action: a million 30-metre cells collapse to a handful of numbers about one RM. Each 30-metre cell is about 0.22 acres, so multiplying a count by 0.22 gives the class's area in acres.

## Writing spatial data

`st_write()` saves a vector layer and `writeRaster()` saves a raster. For vector output use GeoPackage rather than shapefile -- a shapefile truncates column names to ten characters and splits into several files.

```{r}
#| eval: true
#| echo: false
console('
# Save the joined boundaries-plus-yields layer
st_write(rm_canola, "output/rm_canola_2023.gpkg", delete_dsn = TRUE, quiet = TRUE)

# Save the masked raster
writeRaster(aci_wc, "output/aci_wood_creek_2023.tif", overwrite = TRUE)
')
```

The saved GeoPackage is what @sec-maps reads back to make its first map.
