# Merging Data {#sec-merging-data}

```{r}
#| include: false
#| eval: true
source("_console.R")
```

The data you need to answer one question often arrive in several tables. For example, suppose we wanted to understand the relationship between RM yields and precipitation. The yield data would come in one table, and the precipitation data would likely come in another. To answer our question, we would need to **merge** or **join** these tables.

Throughout this section, we will work with three datasets:

- [`rm_yields_2015_2024.csv`](practice/data/rm_yields_2015_2024.csv), with one row per RM per year. It contains 2,950 rows for ten years and six crops. Yields are measured in bushels per acre.
- [`station_precip.csv`](practice/data/station_precip.csv), with one row per weather station per year. It contains May to August precipitation for sixteen stations and a count of days with no reading.
- [`rm_lookup.csv`](practice/data/rm_lookup.csv), with one row per RM. It contains the RM name, census division, 2021 population and the closest of the sixteen weather stations.

We will do two merges:

1. Merge the RM yields with the RM lookup. This will give us a dataset containing the closest weather station for each yield observation.
2. Merge the dataset from step 1 with the weather station data. This will give us the growing-season precipitation for each observation.

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Identify the key that connects two tables.
2. Distinguish one-to-one, many-to-one and many-to-many relationships.
3. Merge tables in Excel with Power Query, including a merge that uses two key columns.
4. Merge tables in R with `left_join()` and related functions.
5. Check for unmatched rows and unexpected changes in the row count.
:::


### Keys {.unnumbered}

A key is a column, or set of columns, used to identify and match rows across datasets. When we merge, we say that we merge on the key. The key must be present in both datasets.


In the three tables above, the keys for the two merges are:

1. RM yields to RM lookup: RM.
2. The resulting dataset to weather-station data: weather station and year.

### One-to-one, many-to-one and many-to-many {.unnumbered}

There are three types of merges:


![Three kinds of merge. Each line connects a grades row to the row it matches.](images/join_types.svg){#fig-join-types}

- **One-to-one (1:1).** Each key value appears once in both tables. Grades and attendance both have one row per student-class, so the merged table has the same number of rows as either input.
- **Many-to-one (n:1).** Values in the key column may repeat in one dataset but are unique in the other. In this example, students take many classes, so each student appears multiple times in the table on the left. Each class has a single professor, however, so each class appears only once in the table on the right.
- **Many-to-many (n:n).** Values in the key column repeat in both datasets. In the figure, students take multiple classes (the table on the left), and each class can have multiple teaching assistants (the table on the right).

Before joining, decide which relationship you expect. Then compare the row counts before and after.

Both of our merges will be many-to-one:

1. The RM yields dataset contains many observations for the same RM, but the RM lookup contains only one row for each RM.
2. The merged dataset from step 1 contains many observations with the same closest weather station and year, but the weather-station dataset contains only one observation for each station and year.

## Merging in Excel {#sec-merge-excel}

In Chapter 2, we used `XLOOKUP` to bring values from one table into another (@sec-lookup). That approach is useful for a quick lookup. Here, we will use Power Query because it records the merge as a repeatable series of steps and can match on several columns directly.

### First merge: yields and the RM lookup {.unnumbered}

Begin by importing `rm_yields_2015_2024.csv` and `rm_lookup.csv` with **Data > Get Data > From Text/CSV > Transform Data**. In the Power Query editor:

1. Select the RM yields query.
2. Choose **Home > Merge Queries**.
3. Select the RM lookup as the second table.
4. Select `rm` in the yields table and `rm_number` in the lookup table.
5. Keep **Left Outer** as the join kind. This keeps every yield observation and adds information wherever the lookup contains a matching RM.
6. Check the number of matching rows reported in the dialog, then choose **OK**.
7. Expand the new table column using the double-arrow icon in its header. Select the RM name, census division, population and closest weather station.

This is a many-to-one merge: the yields table contains many rows for each RM, while the lookup contains one row for each RM. The number of yield rows should not change.

### Second merge: matching on station and year {.unnumbered}

Next, import `station_precip.csv`. Each yield observation must match precipitation from both the correct weather station and the correct year. Unlike the helper-column method used with `XLOOKUP` in Chapter 2, Power Query can select both keys directly:

1. Choose **Home > Merge Queries** from the merged yields query.
2. Select the closest-weather-station column, then hold Ctrl (Command on a Mac) and select `year`.
3. In the precipitation table, select `station` first and `year` second. The numbered labels beside the column headings confirm the key order.
4. Keep **Left Outer** as the join kind and check the number of matches reported in the dialog.
5. Choose **OK**, then expand the new table column and select the precipitation variables you need.

The column order matters: the first selected column in one table is matched to the first selected column in the other, and the second is matched to the second. Selecting station and year prevents Power Query from returning the first year it finds for a station.

### Checking the merge {.unnumbered}

After each merge, compare the number of rows with the original yields table. A many-to-one left merge should retain the same number of rows. Filter the expanded columns for `null` to find observations that did not match, and check several rows against the source tables.

Power Query also offers **Inner**, **Full Outer** and **Left Anti** joins. These correspond to the R joins introduced below. A Left Anti join is particularly useful for listing rows that have no match before completing the merge.

## Merging in R {#sec-merge-r}

In tidyverse we merge data with join functions.  There are four different types of joins depending on which data we want to keep in the merged file.  Suppose we have two datasets `y` and `x`:


- **`left_join(x, y, by)`** keeps every row of `x` whether it has a match in `y` or not.  But does not keep rows in `y` that don't have a match. 
- **`inner_join(x, y, by)`** keeps only rows that matched in both tables.
- **`full_join(x, y, by)`** keeps every row from both tables, regardless of whether there is a match.
- **`anti_join(x, y, by)`** returns the rows of `x` that have no match in `y`. It is useful for finding failed matches before a join.

The `by` argument specifies the keys we merge on.  It can be a single column name or a vector of names.


Read the three files and inspect their columns:

```{r}
#| eval: true
#| echo: false
console('

## Load pacakges
library(tidyverse)

# Read the three tables
yields <- read_csv(
  "practice/data/rm_yields_2015_2024.csv",
  show_col_types = FALSE
)

rms <- read_csv(
  "practice/data/rm_lookup.csv",
  show_col_types = FALSE
)

rain <- read_csv(
  "practice/data/station_precip.csv",
  show_col_types = FALSE
)

')
```

We can now use `left_join()` takes two data frames and a `by` argument that identifies the key. If a key has different names in the two tables, write `join_by(left_name == right_name)`; a column with the same name on both sides is written once. The older `by = c("left_name" = "right_name")` form still appears in other people's code.



Before joining, check that the columns expected to identify one row are actually unique:

```{r}
#| eval: true
#| echo: false
console('
rms |>
  count(rm_number) |>
  filter(n > 1)

rain |>
  count(station, year) |>
  filter(n > 1)
')
```

Both checks return empty tibbles. A key is only a key if each combination occurs no more than once on the side where uniqueness is expected.

### First join: yields and the RM lookup {.unnumbered}

The RM column is called `rm` in the yields table and `rm_number` in the lookup. Check for unmatched RMs, perform the join and compare the row counts:

```{r}
#| eval: true
#| echo: false
console('
# List RM numbers that have no lookup row
yields |>
  anti_join(rms, by = join_by(rm == rm_number)) |>
  distinct(rm)

# Add the RM information and nearest station
with_station <- yields |>
  left_join(rms,
            by = join_by(rm == rm_number),
            relationship = "many-to-one")

nrow(yields)
nrow(with_station)
')
```

The `anti_join()` returns an empty tibble, so every RM has a match. `yields` has 2,950 rows and so does `with_station`, which is what a many-to-one join should give. The `relationship = "many-to-one"` argument records what we expect and makes `left_join()` stop if the lookup contains more than one row for an RM.

In R, text keys are sensitive to both case and spaces. For example, `"canola"`, `"Canola"` and `"Canola "` are three different keys. These differences have to be reconciled before the join (@sec-categories).

### A failed match on purpose {.unnumbered}

The complete rain table matches every station-year. To see how the joins differ when data are missing, temporarily remove 2021:

```{r}
#| eval: true
#| echo: false
console('
rain_short <- rain |>
  filter(year != 2021)

# Which yield rows no longer have rain?
with_station |>
  anti_join(
    rain_short,
    by = join_by(weather_station == station, year)
  ) |>
  count(weather_station)

# A left join keeps all 2,950 yield rows,
# including 295 rows with missing rain.
with_station |>
  left_join(
    rain_short,
    by = join_by(weather_station == station, year)
  ) |>
  filter(is.na(precip_may_aug_mm)) |>
  nrow()

# An inner join drops those 295 rows.
with_station |>
  inner_join(
    rain_short,
    by = join_by(weather_station == station, year)
  ) |>
  nrow()

# A full join also keeps a right-side row
# that has no yield-side match.
rain_extra <- rain_short |>
  add_row(
    station = "Course example",
    year = 2025,
    precip_may_aug_mm = 200,
    days_missing = 0
  )

with_station |>
  full_join(
    rain_extra,
    by = join_by(weather_station == station, year)
  ) |>
  filter(weather_station == "Course example") |>
  select(year, weather_station, rm, canola)
')
```

`anti_join()` identifies the failed keys, `left_join()` makes the missing matches visible as `NA`, and `inner_join()` silently removes them. The final `full_join()` also retains the deliberately added rain row that has no yield-side match, filling its yield columns with `NA`.

### Second join: add the rain {.unnumbered}

The second key has two parts. The station column is called `weather_station` on the left and `station` on the right; `year` has the same name in both tables.

```{r}
#| eval: true
#| echo: false
console('
# Find station-years with no precipitation row
with_station |>
  anti_join(rain,
            by = join_by(weather_station == station, year)) |>
  distinct(weather_station, year)

# Add precipitation for the matching station and year
merged <- with_station |>
  left_join(rain,
            by = join_by(weather_station == station, year),
            relationship = "many-to-one")

nrow(merged)
')
```

Again there are no unmatched keys and the row count remains 2,950.

### The row-count trap {.unnumbered}

Now leave `year` out of the key:

```{r}
#| eval: true
#| echo: false
console('
# This key is incomplete: every row matches ten years of rain
with_station |>
  left_join(rain,
            by = join_by(weather_station == station)) |>
  nrow()

# State the expected relationship to turn
# the warning into an error.
try(
  with_station |>
    left_join(
      rain,
      by = join_by(weather_station == station),
      relationship = "many-to-one"
    )
)
')
```

Each yield row matches all ten years for its station, so 2,950 rows become 29,500. By default, `dplyr` warns about the unexpected many-to-many relationship. When `relationship = "many-to-one"` states what was intended, the same mistake becomes an error and the analysis stops. Excel's station-only `XLOOKUP` makes the mistake differently: it returns only the first matching year and gives no warning.

## What the Merge Gives You {#sec-merge-result}

The merged table has one row per RM-year, with yield and precipitation beside one another:

```{r}
#| eval: true
#| echo: false
console('
# Display the columns used in the analysis
merged |>
  select(year, rm, rm_name, canola,
         weather_station, precip_may_aug_mm)

# Calculate annual mean yield and precipitation
merged |>
  filter(!is.na(canola)) |>
  summarise(
    canola_bu_ac = mean(canola),
    rain_mm = mean(precip_may_aug_mm),
    max_days_missing = max(days_missing),
    .by = year
  )

# Count yield values excluded from those means
merged |>
  summarise(missing_canola = sum(is.na(canola)))

# Describe association, not causation
merged |>
  summarise(
    yield_rain_correlation = cor(
      canola,
      precip_may_aug_mm,
      use = "complete.obs"
    )
  )
')
```

Across RM-years, wetter observations tend to have higher canola yield, but the relationship is loose: the correlation is about 0.41. The annual table also resists a simple rainfall story. Both 2021 and 2024 have low average yields despite ordinary or above-average rainfall. These data cannot say why. Testing whether heat contributed would require temperature data; explaining the remaining differences would require other variables as well.

The annual means exclude 46 RM-years with no published canola yield. They also average rainfall over RM rows, so a station serving more RMs receives more weight than a station serving fewer RMs. The merge has further limits that should be recorded in the project's README. The median RM is 62 km from its assigned station, one quarter are more than 85 km away, and the maximum is 161 km, so station rainfall is a coarse proxy for rainfall across an RM. Some station-years also have many missing daily observations—up to 25 days in this file—and their seasonal totals sum only the available days, which biases those totals downward.
