# The `dplyr` Verbs {#sec-dplyr-verbs}

```{r}
#| include: false
#| eval: true
source("_console.R")
suppressMessages({library(dplyr); library(readr); library(tidyr)})
yields <- readr::read_csv("practice/data/field_yields.csv", show_col_types = FALSE)
```

With a dataset in R and inspected, we can start doing things to it. Most data analysis
involves repeating a surprisingly small number of operations: keep certain rows, keep
certain columns, create or change a column, sort the rows, calculate summary statistics,
and do those calculations separately for different groups.

The tidyverse package `dplyr` gives us a function for each of these jobs. They are called
**verbs** because each describes an action performed on a data frame. None of the
operations are new -- each is the R form of something from the Excel modules:

| Question                                | `dplyr` verb  | Excel counterpart        |
|-----------------------------------------|---------------|--------------------------|
| Which rows do I want?                   | `filter()`    | filtering                |
| Which columns do I want?                | `select()`    | keeping certain columns  |
| What new column do I want?              | `mutate()`    | a formula in a new column|
| In what order do I want the rows?       | `arrange()`   | sorting                  |
| What summary do I want?                 | `summarise()` | summary formulas         |
| That summary separately for groups?     | `group_by()`  | a PivotTable             |

Once you understand these operations, most everyday data manipulation in R is a matter of
combining them.

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Use `filter()`, `select()`, `mutate()`, and `arrange()` to manipulate a data frame.
2. Chain several operations together using the pipe, `|>`.
3. Use `summarise()` to calculate statistics from a data frame.
4. Use `group_by()` with `summarise()` to calculate statistics separately for groups.
5. Read a longer pipeline from top to bottom and explain what each step does.
6. Save a result to a CSV file.
:::

## The Data {.unnumbered}

The examples use a new dataset, [`field_yields.csv`](practice/data/field_yields.csv) -- 60
canola fields with columns `field_id`, `region`, `variety`, `acres`, `yield_bu_acre`,
`year`, `seeding_date`, and `units`. Download it into your project's `data/` folder and
start your script with:

```r
library(tidyverse)

yields <- read_csv("data/field_yields.csv")
```

Before the verbs, one principle to understand. Suppose we run:

```{r}
#| eval: true
#| echo: false
console('
filter(yields, yield_bu_acre > 50)
')
```

R displays the fields with yields above 50 bu/ac, but it does not remove the other fields
from `yields` -- check `nrow(yields)` and all 60 rows are still there. `filter()` takes
`yields` as input and returns a *new* data frame as output; if we only display that output,
R does not save it. To keep the result, assign it to an object:

```r
high_yield <- filter(yields, yield_bu_acre > 50)
```

Now there are two objects: `yields`, the original 60 fields, and `high_yield`, the filtered
subset. This is a useful feature -- you can experiment with transformations without changing
the original data. Throughout the chapter, an example without `<-` is just displaying a
result; add a name in front when you want to keep it.

## The Four Row-and-Column Verbs {#sec-verbs-basic}

Two of the four verbs mainly affect the rows: `filter()` chooses rows, and `arrange()`
changes their order. Two mainly affect the columns: `select()` chooses columns, and
`mutate()` creates or changes them. Each is a function whose first argument is the data
frame, followed by what to do with it.

### `filter()`: Keep Certain Rows {.unnumbered}

`filter()` keeps the rows that satisfy a condition:

```r
filter(yields, yield_bu_acre > 50)
```

Read this as: from `yields`, keep the rows where `yield_bu_acre` is greater than 50. Text
variables work too:

```r
filter(yields, region == "South")
```

Note the **two** equals signs: `==` asks whether two things are equal, so
`region == "South"` asks "is `region` equal to `"South"`?". A single `=` has a different job
in R -- naming arguments, as in `mean(x, na.rm = TRUE)` -- so writing `region = "South"`
inside `filter()` will not do what you want. Also note the quotation marks: `"South"` with
quotes is text; `South` without quotes would be read as the name of an object, which
probably doesn't exist and would cause an error.

You can combine conditions with `&` (and -- both must be true), `|` (or -- at least one must
be true), and `!` (not):

```r
filter(yields, yield_bu_acre > 50 & region == "South")
filter(yields, region == "South" | region == "Central")
filter(yields, !is.na(yield_bu_acre))  # is.na() asks if a value is missing; ! keeps the non-missing
```

When you want to match several possible values of the same variable, `%in%` is easier to
read:

```r
filter(yields, region %in% c("South", "Central"))
```

You do not need to memorize every comparison operator. The idea is that `filter()` chooses
observations by asking a TRUE/FALSE question about each row.

### `select()`: Keep Certain Columns {.unnumbered}

`select()` chooses variables rather than observations:

```{r}
#| eval: true
#| echo: false
console('
select(yields, field_id, region, yield_bu_acre)
')
```

The result has the same rows but only those three columns -- especially useful with datasets
containing dozens or hundreds of variables. You can also drop a column by putting `-` in
front of its name, or keep a range of adjacent columns:

```r
select(yields, -units)                   # everything except units
select(yields, field_id:yield_bu_acre)   # the columns from field_id through yield_bu_acre
```

A relative of `select()` is `rename()`, which changes a column's name without dropping anything: `rename(yields, yield = yield_bu_acre)`.

Keep the pair straight: `filter()` works on rows, `select()` on columns.

### `mutate()`: Create or Change Columns {.unnumbered}

`mutate()` creates a new variable or changes an existing one. Suppose we want yield in
tonnes per hectare rather than bushels per acre:

```{r}
#| eval: true
#| echo: false
console('
mutate(yields, yield_t_ha = yield_bu_acre * 0.0560)
')
```

The result contains all the original columns plus a new `yield_t_ha`. Read the expression
as: create `yield_t_ha` and set it equal to `yield_bu_acre` × 0.0560. This is the R version
of adding a formula column in Excel.

You can create several variables in one call, and later expressions can use variables
created earlier in the same `mutate()`:

```r
mutate(yields,
       yield_t_ha = yield_bu_acre * 0.0560,
       hectares = acres * 0.4047,
       total_tonnes = yield_t_ha * hectares)
```

### `arrange()`: Sort Rows {.unnumbered}

`arrange()` changes the order of the rows. `arrange(yields, yield_bu_acre)` sorts from
lowest yield to highest; wrap the column in `desc()` to sort the other way:

```{r}
#| eval: true
#| echo: false
console('
arrange(yields, desc(yield_bu_acre))
')
```

You can sort by more than one variable -- `arrange(yields, region, desc(yield_bu_acre))`
sorts by region and then, within each region, from highest yield to lowest. Like the other
verbs, `arrange()` does not permanently rearrange `yields` unless you save the result.

## The Pipe `|>` {#sec-pipe}

Real questions usually take more than one verb. Suppose we want to know: among fields in the
South region, what are the field IDs and yields, ordered from highest to lowest? We know
every operation we need -- filter to the South, select the two columns, arrange by
descending yield. Without a pipe, each step's output becomes the next step's first argument,
and the calls nest:

```r
arrange(select(filter(yields, region == "South"), field_id, yield_bu_acre),
        desc(yield_bu_acre))
```

That works, but you have to start reading in the middle and work outward. The **pipe**,
`|>`, lets us write the analysis in the order we think about it. The pipe takes whatever is
on its *left* and feeds it as the first argument to the function on its *right*, so
`yields |> filter(region == "South")` is exactly `filter(yields, region == "South")`. Read
`|>` as the word "then":

```{r}
#| eval: true
#| echo: false
console('
yields |>
  filter(region == "South") |>
  select(field_id, yield_bu_acre) |>
  arrange(desc(yield_bu_acre))
')
```

Take `yields`, *then* keep the South fields, *then* keep two columns, *then* sort. Once we
use the pipe, we no longer repeat the data-frame name in every function, and a multi-step
analysis reads top to bottom. From here on the examples use the pipe. (Older code uses `%>%`
for the same idea; `|>` is the modern built-in version.)

To keep a result, assign the whole pipeline to an object:

```r
south_yields <- yields |>
  filter(region == "South") |>
  select(field_id, yield_bu_acre) |>
  arrange(desc(yield_bu_acre))
```

### Debugging a pipeline {.unnumbered}

Pipelines also give you a way to debug. When a pipeline produces a result that looks wrong,
do not stare at the whole thing and guess. Run it one step at a time: run the first step
alone, check the result, add the second step, check again, and keep going. The first point
where the data stops looking the way you expect is where the problem is. This works because
every verb does one thing -- and the same habit is how you check a pipeline an AI assistant
wrote for you.

## `summarise()` and `group_by()` {#sec-summarise-group}

The four verbs so far rearrange or add information while keeping observations as rows.
Sometimes we want to *collapse* the observations into summary statistics. That is the job of
`summarise()`.

### `summarise()`: Many Rows Become One {.unnumbered}

```{r}
#| eval: true
#| echo: false
console('
yields |>
  summarise(
    mean_yield = mean(yield_bu_acre, na.rm = TRUE),
    sd_yield = sd(yield_bu_acre, na.rm = TRUE),
    n_fields = n()
  )
')
```

The names on the left (`mean_yield`, `sd_yield`, `n_fields`) become column names in the
result; the expressions on the right calculate the values. `n()` counts the rows.

The contrast with `mutate()` matters. `mutate()` keeps the observations and adds
information; `summarise()` collapses the observations into summaries. So
`yields |> mutate(mean_yield = mean(yield_bu_acre, na.rm = TRUE))` still has 60 rows, with
the overall mean repeated on every row, while the `summarise()` version returns one row.

### `group_by()`: Calculate Separately for Groups {.unnumbered}

The overall mean may not be what we want. Suppose we want the mean yield for each region.
`group_by()` doesn't change the data -- it tells subsequent operations to apply *per group*:

```{r}
#| eval: true
#| echo: false
console('
yields |>
  group_by(region) |>
  summarise(
    mean_yield = mean(yield_bu_acre, na.rm = TRUE),
    sd_yield = sd(yield_bu_acre, na.rm = TRUE),
    n_fields = n()
  )
')
```

Take `yields`, *then* divide the observations into regions, *then* calculate the statistics
separately for each region. The result has one row per region. This is the R equivalent of a
PivotTable.

You can group by more than one variable:

```r
yields |>
  group_by(region, variety) |>
  summarise(mean_yield = mean(yield_bu_acre, na.rm = TRUE),
            n_fields = n(),
            .groups = "drop")
```

Now one row is a region-variety combination. Grouping is state that can affect later
operations: the `.groups = "drop"` argument tells `summarise()` to return an ordinary,
ungrouped data frame. You do not need it on every simple grouped summary, but it is useful
when you plan to keep manipulating the result. (There is also an `ungroup()` function, which
appears later in this chapter in a situation where it genuinely matters.)

The central pattern -- `data |> group_by(group_variable) |> summarise(...)` -- appears
throughout the rest of the book.

## Unit Conversion: A Worked Example {#sec-unit-conversion}

Canadian ag data often comes in mixed units. Canola yields are traditionally reported in bushels per acre (bu/ac) in Canada, but metric tonnes per hectare (t/ha) in global datasets. The conversion for canola is approximately:

$$
\text{yield (t/ha)} = \text{yield (bu/ac)} \times 0.0560
$$

This factor comes from two facts: a bushel of canola weighs about 22.68 kg (canola's standard 50-lb bushel), and an acre is about 0.4047 ha. So one bu/ac is $22.68 \div 0.4047 \div 1000 \approx 0.0560$ t/ha. (The exact factor depends on the crop's bushel weight -- a wheat or corn bushel is 60 or 56 lb and gives a different number -- so in a real analysis you would look up the right bushel weight for your crop and cite your source.)

If every observation is in bu/ac, the transformation is exactly what `mutate()` is for:

```r
yields_converted <- yields |>
  mutate(yield_t_ha = yield_bu_acre * 0.0560)

yields_converted |>
  select(field_id, yield_bu_acre, yield_t_ha)
```

The original `yield_bu_acre` stays in the data alongside the new `yield_t_ha`, which makes
the transformation easy to check.

### An extension: different rules for different rows {.unnumbered}

Sometimes a transformation depends on a condition. Suppose some rows are in bu/ac and some
already in t/ha, with a `units` column telling you which. `case_when` handles it -- read
this one to understand it rather than to memorize it:

```r
yields_converted <- yields |>
  mutate(
    yield_t_ha = case_when(
      units == "bu/ac" ~ yield_bu_acre * 0.0560,   # convert the bu/ac rows
      units == "t/ha"  ~ yield_bu_acre,             # already metric: leave as-is
      TRUE             ~ NA_real_                    # anything else becomes missing
    )
  )
```

Each line reads "if this condition is true, use this value". The first matching condition
wins, and the `TRUE ~ ...` line is the catch-all: anything that matched neither condition
becomes a missing numeric value (`NA_real_` is R's way of writing one). `case_when` is the R
analogue of nested `IF`s in Excel, and it comes back when we clean data in @sec-cleaning.
The point for now is that `mutate()` can create variables using rules, not just arithmetic.

## Writing Output {#sec-writing-output}

Results so far have existed only inside R. To save one as a file that can be opened in Excel
or sent to someone else, pass the object and a filename to `write_csv`. With the project
layout from @sec-project-folder, results go in `output/`:

```r
region_summary <- yields |>
  group_by(region) |>
  summarise(mean_yield = mean(yield_bu_acre, na.rm = TRUE),
            n_fields = n(),
            .groups = "drop")

write_csv(region_summary, "output/region_summary.csv")
```

`read_csv` brings data into R; `write_csv` sends a data frame out to a file. Because the
output is created by the script, you can always recreate it by running the script again.

## Combining It All {#sec-dplyr-combining}

We now have enough tools to answer a realistic question: for each region and variety in
2025, what was the mean yield in t/ha, and how many fields were observed? Show the
highest-yielding variety first within each region.

```{r}
#| eval: true
#| echo: false
console('
region_variety_summary <- yields |>
  filter(year == 2025) |>
  mutate(yield_t_ha = yield_bu_acre * 0.0560) |>
  group_by(region, variety) |>
  summarise(mean_yield_t_ha = mean(yield_t_ha, na.rm = TRUE),
            n_fields = n(),
            .groups = "drop") |>
  arrange(region, desc(mean_yield_t_ha))
region_variety_summary
')
```

Do not read it as one giant command; read it one verb at a time:

1. **Filter** to 2025.
2. **Compute** metric yield.
3. **Group** the observations into region-variety combinations.
4. **Summarise** each group down to one row: its mean yield and its number of fields.
5. **Sort** by region and, within each region, from highest mean yield to lowest.

Each line does one simple thing, and that is the reason for pipes: a good pipeline makes the
logic of an analysis visible. Saving the table is one more line:
`write_csv(region_variety_summary, "output/region_variety_summary.csv")`.

### A slightly harder example {.unnumbered}

Which fields yielded above the average for their own region? This needs one new idea:
`mutate()` can operate within groups.

```{r}
#| eval: true
#| echo: false
console('
above_region_average <- yields |>
  group_by(region) |>
  mutate(region_mean = mean(yield_bu_acre, na.rm = TRUE)) |>
  ungroup() |>
  filter(yield_bu_acre > region_mean) |>
  select(field_id, region, yield_bu_acre, region_mean) |>
  arrange(region, desc(yield_bu_acre))
above_region_average
')
```

The important step is `group_by(region)` followed by `mutate()`. Because the data is grouped
by region, `mean()` is calculated separately within each region -- and unlike `summarise()`,
`mutate()` keeps every field and adds its region's mean as a new column on each one. This is
also where `ungroup()` earns its place: we are finished calculating within regions, so we
remove the grouping and the later steps treat the result as an ordinary data frame.

You do not need to produce this second pipeline from memory. You should be able to read it
from top to bottom and explain what each step is doing. As analyses become more complicated
-- and especially once AI helps generate the code -- that is the skill that matters: looking
at a sequence of operations and deciding whether it actually answers the question you
intended to ask.

::: {.callout-tip collapse="true"}
## Other resources — dplyr

- **R for Data Science (2nd ed.)** -- [Chapter 3, "Data transformation"](https://r4ds.hadley.nz/data-transform.html): the verbs, the pipe, and grouped summaries.
- **dplyr documentation** -- the [dplyr reference site](https://dplyr.tidyverse.org/), including the one-page [cheatsheet](https://rstudio.github.io/cheatsheets/html/data-transformation.html).
- **Video** -- Equitable Equations, [*The five main dplyr verbs*](https://www.youtube.com/watch?v=b2ODCVAsXlw) -- filter, select, mutate, arrange, and summarise in one sitting.
:::
