# Graphing in R {#sec-graphing-r}

```{r}
#| include: false
#| eval: true
source("_console.R")
```

The Excel chapter built a bar chart, scatter plot and histogram from actual Saskatchewan variety-yield records. This chapter reproduces those examples with `ggplot2`, then adds value labels, a reference line, direct variety labels and two displays that reveal more than the Excel summaries.

The data come from [`sask_variety_yields.csv`](practice/data/sask_variety_yields.csv), which contains Saskatchewan Crop Insurance Corporation variety yields by risk zone from 2021 to 2025 [@scic_smp]. Suppressed observations have missing acres and yield.

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
suppressPackageStartupMessages(library(tidyverse))

# Read the actual Saskatchewan variety-yield file
variety_yields <- read_csv(
  "practice/data/sask_variety_yields.csv",
  show_col_types = FALSE
)

# Keep canola records, then keep reported values for calculations
canola <- variety_yields |>
  filter(Crop == "Canola/Rapeseed")

reported_canola <- canola |>
  filter(
    !is.na(Acres),
    !is.na(Yield),
    Acres > 0
  )

# Calculate an acre-weighted provincial result for each variety and year
provincial_variety <- reported_canola |>
  group_by(Variety, Year) |>
  summarise(
    acres = sum(Acres),
    weighted_yield = sum(Acres * Yield) / sum(Acres),
    zones = n(),
    .groups = "drop"
  )

# Select the ten varieties with the most reported acres in 2025
leaders_2025 <- provincial_variety |>
  filter(Year == 2025) |>
  slice_max(acres, n = 10, with_ties = FALSE) |>
  arrange(weighted_yield)

# Inspect the chart-driving table
leaders_2025 |>
  select(Variety, acres, weighted_yield, zones)
')
```

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Build a plot with `ggplot()`, `aes()` and a geom.
2. Reproduce the Excel bar chart, scatter plot and histogram.
3. Change a plot's title, axis labels, limits and tick marks.
4. Calculate and graph an acre-weighted yield.
5. Distinguish mapping an aesthetic to a variable from setting it to a constant.
6. Add value labels, a reference line and direct series labels.
7. Compare risk-zone distributions and save a finished plot.
:::

## Calculate the values before drawing {#sec-r-weighted-summary}

The source contains one row per risk-zone, crop, variety and year combination. It does not contain one provincial yield per variety.

The `summarise()` call first multiplies each reported risk-zone yield by its acres. It then divides the total by the reported acres:

$$
\text{weighted yield} = \frac{\sum(\text{Acres} \times \text{Yield})}{\sum \text{Acres}}.
$$

This is the same calculation performed by `SUMIFS` in the Excel workbook. A simple `mean(Yield)` would give every risk zone equal influence regardless of its acreage.

The code excludes rows with missing acres or yield. It does not turn suppressed observations into zeroes.

## Build the bar chart {#sec-r-bar-example}

The Excel chart used a ten-row summary. `leaders_2025` is the equivalent R data frame.

```{r}
#| label: fig-r-variety-bars-basic
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.0
#| fig-height: 4.0
#| fig-cap: "Acre-weighted 2025 yield for the ten canola varieties with the most reported acres."
#| fig-alt: "Ten horizontal green bars range from about 44 to 49 bushels per acre. L350PC has the longest bar and L233P the shortest."
leaders_2025 |>
  ggplot(
    aes(
      x = weighted_yield,
      y = fct_reorder(Variety, weighted_yield)
    )
  ) +
  geom_col(fill = "#3F6F52")
```

`ggplot()` names the data and maps columns to horizontal and vertical position. `geom_col()` draws bars whose lengths use the already-calculated `weighted_yield` values. `geom_bar()` would count rows instead, which answers a different question.

The selected varieties range from 44.3 to 49.0 bu/ac. L340PC has by far the most reported acres, but it does not have the highest weighted yield in this selected group. The chart therefore compares yield; the selection rule is acreage.

## Change the title and axes {#sec-r-title-axes}

`labs()` supplies the title, subtitle, caption and axis text. `scale_x_continuous()` controls the displayed range and tick marks. These are separate from the calculation, so changing the scale does not change the weighted yields.

```{r}
#| label: fig-r-variety-bars-polished
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.3
#| fig-cap: "A finished horizontal bar chart with a zero baseline and direct value labels."
#| fig-alt: "A labelled horizontal bar chart shows weighted yields from 44.3 to 49.0 bushels per acre for ten canola varieties."
# Store the finished plot so it can be saved later
bar_plot <- leaders_2025 |>
  ggplot(
    aes(
      x = weighted_yield,
      y = fct_reorder(Variety, weighted_yield)
    )
  ) +
  # Set one colour for the whole series
  geom_col(
    width = 0.7,
    fill = "#3F6F52"
  ) +
  # Map formatted yield values to text
  geom_text(
    aes(label = sprintf("%.1f", weighted_yield)),
    hjust = -0.25,
    size = 3.3
  ) +
  # Match the Excel baseline and tick spacing
  scale_x_continuous(
    limits = c(0, 55),
    breaks = seq(0, 50, by = 10),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(
    title = "2025 yield of leading canola varieties",
    subtitle = "Ten varieties with the largest reported acreage",
    x = "Acre-weighted yield (bu/ac)",
    y = NULL,
    caption = "Source: SCIC Sask Management Plus variety-yield data"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )

# Print the stored plot
bar_plot
```

`limits = c(0, 55)` gives the bars a zero baseline. `breaks = seq(0, 50, by = 10)` places the labelled ticks. The labels remain inside the 55 bu/ac limit because the largest value is about 49.

The `fill` argument is outside `aes()` because every bar receives the same colour. Moving `fill = Variety` inside `aes()` would map colour to a column and create ten colours plus a redundant legend.

## Compare 2024 and 2025 {#sec-r-scatter-example}

The scatter plot needs one 2024 value and one 2025 value on each row. `pivot_wider()` places the two years in separate columns, and `drop_na()` keeps only varieties reported in both years.

```{r}
#| eval: true
#| echo: false
console('
# Put the two annual yields on the same row
year_pair <- provincial_variety |>
  filter(Year %in% c(2024, 2025)) |>
  select(Variety, Year, acres, weighted_yield) |>
  pivot_wider(
    names_from = Year,
    values_from = c(acres, weighted_yield),
    names_sep = "_"
  ) |>
  drop_na(weighted_yield_2024, weighted_yield_2025)

# Select the most widely grown 2025 variety for a direct label
largest_variety <- year_pair |>
  slice_max(acres_2025, n = 1, with_ties = FALSE)

tibble(
  varieties_plotted = nrow(year_pair),
  correlation = cor(
    year_pair$weighted_yield_2024,
    year_pair$weighted_yield_2025
  ),
  labelled_variety = largest_variety$Variety
)
')
```

```{r}
#| label: fig-r-variety-scatter
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.0
#| fig-height: 4.0
#| fig-cap: "Acre-weighted canola variety yields in 2024 and 2025."
#| fig-alt: "A scatter plot of 72 varieties has a weak upward pattern. A dashed equality line is shown and L340PC is labelled directly."
ggplot(
  year_pair,
  aes(
    x = weighted_yield_2024,
    y = weighted_yield_2025
  )
) +
  # The equality line marks no change between years
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    colour = "grey55"
  ) +
  geom_point(
    colour = "#2878A5",
    alpha = 0.7,
    size = 2
  ) +
  # Label only the variety with the most 2025 acres
  geom_label(
    data = largest_variety,
    aes(label = Variety),
    nudge_x = -3,
    nudge_y = 2,
    size = 3,
    label.size = 0.15,
    fill = "white"
  ) +
  scale_x_continuous(
    limits = c(0, 55),
    breaks = seq(0, 50, by = 10)
  ) +
  scale_y_continuous(
    limits = c(20, 55),
    breaks = seq(20, 50, by = 10)
  ) +
  labs(
    title = "Canola variety yield: 2024 versus 2025",
    x = "2024 acre-weighted yield (bu/ac)",
    y = "2025 acre-weighted yield (bu/ac)"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title.position = "plot")
```

The plot contains 72 varieties and the correlation is about 0.29. A point above the dashed equality line has a higher weighted yield in 2025 than in 2024. The line is a comparison reference, not a fitted model.

`geom_label()` directly identifies L340PC without creating a legend or labelling all 72 points. Direct labels are most useful when the text singles out a small number of observations for a stated reason.

## Show the 2025 yield distribution {#sec-r-histogram-example}

The histogram uses the 674 reported 2025 risk-zone-by-variety yields, not the 97 provincial variety summaries. That distinction determines what each bar counts.

```{r}
#| label: fig-r-canola-histogram
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.0
#| fig-height: 3.7
#| fig-cap: "Distribution of 674 reported 2025 canola yields using 5 bu/ac bins."
#| fig-alt: "A histogram peaks between 45 and 50 bushels per acre, with 222 reported risk-zone-by-variety rows in that interval."
canola_2025 <- reported_canola |>
  filter(Year == 2025)

ggplot(canola_2025, aes(x = Yield)) +
  # Match the Excel width and boundary
  geom_histogram(
    binwidth = 5,
    boundary = 15,
    closed = "left",
    fill = "#D8A940",
    colour = "white"
  ) +
  scale_x_continuous(
    limits = c(15, 70),
    breaks = seq(15, 70, by = 5),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(
    title = "Reported 2025 canola yields",
    subtitle = "Five-bushel-per-acre intervals",
    x = "Risk-zone variety yield (bu/ac)",
    y = "Reported rows"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor = element_blank()
  )
```

The eleven bin counts add to 674. The tallest interval is 45 to 50 bu/ac, with 222 rows. Changing `binwidth` or `boundary` can change the apparent shape, so those settings are part of the analysis rather than decoration.

## Show the observations behind five varieties {#sec-r-variety-distributions}

A provincial weighted yield hides the range of reported risk-zone yields. The next plot shows one point per reporting risk zone for the five varieties with the most 2025 acres.

```{r}
#| label: fig-r-variety-box-jitter
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.2
#| fig-height: 3.8
#| fig-cap: "Risk-zone yield distributions for the five most widely grown 2025 canola varieties."
#| fig-alt: "Five horizontal box plots and jittered points show that reported risk-zone yields vary considerably within each canola variety."
leading_five <- leaders_2025 |>
  slice_max(acres, n = 5, with_ties = FALSE) |>
  pull(Variety)

zone_yields <- canola_2025 |>
  filter(Variety %in% leading_five)

ggplot(
  zone_yields,
  aes(
    x = Yield,
    y = fct_reorder(Variety, Yield, median)
  )
) +
  geom_boxplot(
    width = 0.55,
    fill = "#E4EEE7",
    colour = "#2F5940",
    outlier.shape = NA
  ) +
  geom_jitter(
    width = 0,
    height = 0.12,
    colour = "#2878A5",
    alpha = 0.55,
    size = 1.6
  ) +
  labs(
    title = "Risk-zone yields behind the provincial summaries",
    x = "Reported 2025 yield (bu/ac)",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title.position = "plot",
    panel.grid.major.y = element_blank()
  )
```

Each point receives equal visual weight in this distribution plot, while the provincial summaries weight the points by acres. The box plot answers a question about the spread across reporting zones; the bar chart answers a question about acre-weighted provincial yield.

## Directly label several trend lines {#sec-r-direct-line-labels}

L340PC, L345PC and L233P have reported provincial results in all five years. A line chart is appropriate because year supplies an ordered sequence.

```{r}
#| label: fig-r-direct-line-labels
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.3
#| fig-height: 3.8
#| fig-cap: "Five-year acre-weighted yield trends with variety names printed at the line ends."
#| fig-alt: "Three coloured lines show annual weighted yield from 2021 to 2025. L340PC, L345PC and L233P are labelled directly at 2025."
trend_varieties <- provincial_variety |>
  filter(Variety %in% c("L340PC", "L345PC", "L233P"))

trend_ends <- trend_varieties |>
  filter(Year == 2025) |>
  mutate(
    label_y = case_when(
      Variety == "L345PC" ~ weighted_yield + 0.5,
      Variety == "L233P" ~ weighted_yield - 0.5,
      TRUE ~ weighted_yield
    )
  )

ggplot(
  trend_varieties,
  aes(
    x = Year,
    y = weighted_yield,
    colour = Variety
  )
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  geom_text(
    data = trend_ends,
    aes(y = label_y, label = Variety),
    hjust = -0.12,
    size = 3.2,
    show.legend = FALSE
  ) +
  scale_colour_manual(
    values = c(
      "L340PC" = "#3F6F52",
      "L345PC" = "#2878A5",
      "L233P" = "#A85B3C"
    )
  ) +
  scale_x_continuous(
    breaks = 2021:2025,
    expand = expansion(mult = c(0.03, 0.20))
  ) +
  labs(
    title = "Acre-weighted canola yield over time",
    x = NULL,
    y = "Acre-weighted yield (bu/ac)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title.position = "plot",
    legend.position = "none",
    panel.grid.minor = element_blank()
  )
```

Printing the names beside the final points lets the reader identify the lines without repeatedly looking between the plot and a legend. Direct labels work best with a small number of clearly separated series.

## Save the finished chart {#sec-saving-plots}

Assigning a plot to a name makes the object passed to `ggsave()` explicit. Width and height are in inches unless another unit is supplied.

```{r}
#| eval: true
#| echo: false
console('
# Create the output folder if needed
dir.create("output", showWarnings = FALSE)

# Save the finished bar chart
ggsave(
  "output/canola_variety_bar.png",
  plot = bar_plot,
  width = 6.4,
  height = 4.3,
  dpi = 300
)

# Confirm that the file was written
file.info(
  "output/canola_variety_bar.png"
)["size"]
')
```

PNG works for slides and web pages. PDF preserves vector lines and text for publication. Stating the plot object, dimensions and resolution in code makes the export reproducible.

::: {.callout-tip collapse="true"}
## Other resources -- ggplot2

- [ggplot2 documentation](https://ggplot2.tidyverse.org/)
- [Data Visualization chapter in R for Data Science](https://r4ds.hadley.nz/data-visualize)
:::
