# Cleaning Data {#sec-cleaning}

```{r}
#| include: false
#| eval: true
source("_console.R")
```

A file that opens without an error may still contain duplicate rows, inconsistent names, mixed units and missing values recorded as ordinary text. Data cleaning is a large part of any data analysis project -- for many projects it takes the majority of the time. Unfortuantely, it is also the dullest part of the project.  Hopeflly, this chapter helps us be more efficient and transparent in our data cleaning.  

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Explain what tidy data means and compare long and wide forms of the same data.
2. Reshape data in Excel with Power Query and in R with `pivot_longer()` and `pivot_wider()`.
3. Conduct an initial data-quality check using summaries, visualizations and relationships among variables.
4. Clean column names with `rename()` and `janitor::clean_names()`.
5. Find exact and conflicting duplicates.
6. Find impossible values, inconsistent categories and mixed units.
7. Recognize common codes for missing values and decide when observations can be dropped.
8. Identify dates that were imported as text.
:::

## Data shape {#sec-tidy-data}

The group of R pacakges we are using for data cleaning is called the tidyverse, becasue they are centered on the idea of tidy data.  The concept of tidy data is also relevant for Excel and any other data analysis project.  A dataset is considered tidy when:

1. Every variable has its own column.
2. Every observation has its own row.
3. Every value has its own cell.

This begs a kind of philosophical question about what a vairable is.  Consider the RM yield data set that we have worked with.  The raw data comes in the form below, where an observation refers to a RM and year, and the yields of each crop are a column.  

INSERT DATA

We could alternatively make this a wider dataset where every row is an RM and every column is the yield of a crop in a year.

INSERT DATA

Or we could make this a longer dataset where every observation is a RM-year-crop combination 

INSERT DATA

None of these are definitively right or wrong.  However, data analysts generally prefer the final option which is the longer version of the data.  In Excel this allows us to create PivotTables that ... . In R it allows us to more easily filter by crop when doing analysis.  For example,...  

When you open a table, write down what one row represents and what each column represents. This usually makes the shape problem visible.

::: {.callout-tip collapse="true"}
## Other resources -- tidy data

- **R for Data Science (2nd ed.)** -- [Chapter 5, "Data tidying"](https://r4ds.hadley.nz/data-tidy) develops the tidy-data idea and shows how to reshape datasets between long and wide forms.
- **Video** -- Riffomonas Project, [*Reshaping data to be long or wide with pivot_longer and pivot_wider*](https://www.youtube.com/watch?v=ZVddiwrbrr4).
:::

### Data shape in Excel {#sec-unpivot}

The RM yields file (@sec-direct-download) is wide, with one column for each crop. Power Query calls the wide-to-long operation **Unpivot**. Import the CSV with **Data > Get Data > From Text/CSV**, choose **Transform Data**, select `Year` and `RM`, and then choose **Transform > Unpivot Columns > Unpivot Other Columns**. Rename the two new columns `crop` and `yield_bu_ac` before loading the result.

Power Query returns 15,478 rows because **Unpivot Other Columns** drops source cells containing `null`. The R example below keeps those cells as `NA` and therefore returns 17,700 rows; add `values_drop_na = TRUE` to `pivot_longer()` when you want the two results to agree. The reverse operation is **Pivot Column** in Power Query or a PivotTable on a worksheet.

INSERT SCREENSHOTS HERE

### Data shape in R {#sec-reshaping}

`pivot_longer()` does the same job as Unpivot. It takes several columns, puts their names in one column and puts their values in another. Here is the RM yields table:

```{r}
#| eval: true
#| echo: false
console('
# Load packages
library(tidyverse)

# Read the wide table
yields_wide <- read_csv(
  "practice/data/rm_yields_2015_2024.csv",
  show_col_types = FALSE
)

# Print the wide data
yields_wide

# Put the crop names and yields into columns
yields_long <- yields_wide |>
  pivot_longer(
    cols = -c(Year, RM),
    names_to = "crop",
    values_to = "yield_bu_ac"
  )

# Print the long data
yields_long

nrow(yields_long)
')
```

The column selection `-c(Year, RM)` means every column except `Year` and `RM`. Those two columns continue to identify the observation, while the six crop columns become values in the new `crop` column. The result has one row per RM-year-crop and six times as many rows as the wide table.

`pivot_wider()` goes in the other direction. `names_from` supplies the new column names and `values_from` supplies the values placed under them:

```{r}
#| eval: true
#| echo: false
console('
# Return the crop column to six separate crop columns
yields_long |>
  pivot_wider(
    names_from = crop,
    values_from = yield_bu_ac
  )
')
```

Again, these is no one correct shape for the data.  Typically, analysis is easier on long data.  However, wider data is often better for presentation.

## Data cleaning {#sec-data-cleaning}

### Checking data quality {#sec-data-quality .unnumbered}

Before analysing a new dataset, check what arrived. A file can open without an error and still contain the wrong units, missing records, duplicate observations or values that do not mean what their column names suggest. A short inspection at the beginning can save a great deal of confusion later.

These checks can find problems, but they cannot prove that every value is correct. A yield of 55 bu/ac may be perfectly plausible and still belong to the wrong field. We need to understand the source as well as inspect the values.

#### Start with what you expect {.unnumbered}

Use the source documentation, data dictionary or README to answer:

- What does one row represent?
- What population and time period does the file cover?
- What units are used?
- Which columns should identify a unique observation?
- How are missing values and categories coded?
- Has the file already been filtered, cleaned or aggregated?

Without this information, it can be difficult to decide whether an unusual value is a problem.

#### Take a first look {.unnumbered}

Before calculating anything, check the basic structure:

- Count the rows and columns.
- Inspect the column names and data types.
- Look at several rows from the beginning and end of the file.
- Check whether numbers or dates were imported as text.
- Count repeated values of any column, or combination of columns, that should identify one row.

An unexpected row count or repeated identifier may mean that records are missing, duplicated or stored at a different level than you expected.

#### Check each variable {.unnumbered}

For a **numeric variable**:

- Count the observed and missing values.
- Examine the minimum, quartiles, median, mean and maximum.
- Plot a histogram or box plot.
- Look for impossible values, extreme observations, several clusters and suspicious concentrations at zero.

The useful limits depend on the subject. A negative crop yield is impossible, while a yield of 100 bu/ac may be unusual but real. An extreme value should be investigated, not automatically deleted.

For a **categorical variable**:

- Count the distinct values and their frequencies.
- Look for inconsistent capitalization, spelling and extra spaces, such as `Canola`, `canola` and `Canola `.
- Check whether blank strings or codes such as `-99`, `N/A` and `unknown` represent missing values.
- Investigate rare categories rather than assuming that they are mistakes.

For a **date variable**:

- Check the earliest and latest dates.
- Look for dates outside the period covered by the data.
- Check for gaps, repeated dates and dates stored as text.

#### Check variables together {.unnumbered}

Some problems only appear when two or more variables are compared:

- Seeding should occur before harvest.
- Percentages should normally fall between 0 and 100.
- Reported production should be consistent with acres and yield.
- Category and location codes should agree with their documentation.

A scatter plot can reveal an odd relationship between two numeric variables. A line plot can expose gaps or abrupt changes over time. Summaries grouped by year, region or source can show that missing values or unusual measurements are concentrated in one part of the dataset.

#### Investigate before changing {.unnumbered}

When something looks wrong:

1. Return to the source record or documentation if possible.
2. Flag the observation rather than immediately deleting it.
3. Keep the raw file unchanged.
4. Record any correction, conversion or exclusion in a reproducible cleaning step.
5. If the correct treatment remains uncertain, report that uncertainty and compare the results with and without the questionable observations.

### Data cleaning in Excel {#sec-cleaning-excel}

Excel offers three approaches. Formulas can clean a file once while leaving the raw columns visible. Power Query can reshape the table and records each cleaning step. Find & Replace changes cells directly and leaves little evidence of what was changed.

#### Cleaning with formulas {.unnumbered}

The workbook below keeps the raw delivery data in columns A through F and places cleaning formulas beside it. The formulas standardize crop names, rescale moisture values entered as decimals, count repeated ticket numbers and flag impossible weights. `FORMULATEXT` cells show the formulas, and summary checks appear below the table.

::: {.workbook-callout}
<!-- Public view-only link expires 2026-11-20 under the University of Saskatchewan 90-day policy. -->

::: {.excel-embed}
<iframe width="402" height="700" frameborder="0" scrolling="no" src="https://usaskca1-my.sharepoint.com/personal/pjs998_usask_ca/_layouts/15/Doc.aspx?sourcedoc=%7B6F136BC8-96DB-499F-AE4B-A04CFDF86DED%7D&action=embedview&wdAllowInteractivity=False&wdDownloadButton=True&wdInConfigurator=True&ActiveCell=%27Cleaning%27!A1"></iframe>
:::

[Download this workbook](textbook_examples/mod03_cleaning_excel.xlsx) · [Open online](https://usaskca1-my.sharepoint.com/:x:/g/personal/pjs998_usask_ca/IQDIaxNv25afSa5LoEz9-G3tAXoBxYJcoWjP4fMOie4JVrQ?e=PsWLYM) (File → Save a Copy to view formulas and edit)
:::

The filter dropdown is useful when checking categories because it lists each distinct value. `MIN`, `MAX`, `COUNTBLANK` and a count of repeated identifiers cover many numerical checks. Excel's summary functions silently ignore text in a numeric range, so add a separate text check such as `=SUMPRODUCT(--ISTEXT(D5:D23))`; a weight written as `38.2 t` otherwise disappears from the summary without an error. Keep the raw columns beside the cleaned ones so that a changed value can be traced back.

#### Cleaning with Power Query {.unnumbered}

Power Query has recorded steps for the same work:

- **Remove Top Rows** and **Use First Row as Headers** deal with files that begin with a title or notes.
- The type icon in a column header sets a column to Text, Whole Number, Decimal or Date. A value that cannot be converted becomes an error that can be inspected.
- **Transform > Format > Trim** removes extra spaces, and **Capitalize Each Word** standardizes capitalization.
- **Replace Values** handles synonyms and missing-value codes. Power Query stores a missing value as `null`.
- **Remove Duplicates** drops exact duplicates. To find a repeated identifier with different values, use **Group By** to count rows for each identifier and filter to counts above one.
- **Split Column > By Delimiter** separates values that were stored together.
- **Fill > Down** copies a category into blank cells below it when a spreadsheet records the label only once for a group of rows.

For dates, set `delivery_date` to Date and inspect conversion errors. If the whole source uses another convention, choose **Change Type > Using Locale** and state that convention. Opening the CSV directly lets Excel guess each cell: an ambiguous value such as `03/11/2025` can become the wrong date without an error. Power Query Editor requires a current desktop version of Excel 365 and is not available in Excel for the web.

Each operation appears in the **Applied Steps** list. Clicking a step shows the table at that point. When a new version of the source file arrives, **Refresh** runs the steps again.

### Data cleaning in R {#sec-cleaning-r}

The R examples use [`grain_deliveries_messy.csv`](practice/data/grain_deliveries_messy.csv), a synthetic file constructed for this chapter to contain the problems discussed below. Read it without letting `read_csv()` trim spaces so that we can see the spaces in some crop names:

```{r}
#| eval: true
#| echo: false
console('
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

# Read the file and identify common missing-value codes
deliveries <- read_csv(
  "practice/data/grain_deliveries_messy.csv",
  na = c("", "NA", "N/A", "-", "."),
  trim_ws = FALSE
)

glimpse(deliveries)
')
```

The output already shows two problems. `weight_tonnes` was read as text because one value contains a unit, and `delivery_date` is text because the file mixes date formats. `read_csv()` normally removes leading and trailing spaces; `trim_ws = FALSE` is used here only to keep that problem visible. `str_squish()` remains useful for doubled internal spaces and for text imported through other routes.

#### Awkward Column Names {#sec-clean-names}

Column names such as `Yield (bu/ac)`, `REGION_NAME` and `Field ID Number` are cumbersome in code. When columns name have spaces, you need to use back ticks when referring to them (e.g. `data$`Yield (bu/ac)`` [CORRECT BY BACK TICKS TO BE VISIBLE]).  It is almost always best to use only lower cases, as R (and other software programs) are case-sensitive. 

This book uses **snake case**: lower-case words separated by underscores, such as `yield_bu_ac`, `region_name` and `field_id`.   Some people prefer using dots intead of underscores (e.g. `yield.bu.ac`). Personally, I find that underscores look a bit nicer.

Use `rename()` when only a few names need changing. The new name goes on the left and the old name on the right:

```{r}
#| eval: true
#| echo: false
console('
# Make a small table with awkward names
awkward_names <- tibble(
  `Weight (t)` = c(42.1, 37.8),
  CROP_NAME = c("Canola", "Wheat")
)

awkward_names |>
  rename(
    weight_tonnes = `Weight (t)`,
    crop = CROP_NAME
  )
')
```

Another option is to use the `clean_names()` function from the `janitor` package, which converts every column name in a data frame to snake case. 

```r
# Install janitor package (if not installed already)
install.packages("janitor")

```

```{r}
#| eval: true
#| echo: false
console('
library(janitor)

# Clean every column name in one call
awkward_names |>
  clean_names() |>
  names()

deliveries <- deliveries |> clean_names()
')
```






#### Text in a Numeric Column {#sec-text-numbers}

The first `glimpse()` reported `weight_tonnes` as `<chr>`, or character text. One row contains `"38.2 t"`, so `read_csv()` could not treat the whole column as numeric. Find the values containing something other than a number, decimal point or minus sign:

```{r}
#| eval: true
#| echo: false
console('
# Find weights containing a unit or other text
deliveries |>
  filter(str_detect(weight_tonnes, "[^0-9.-]")) |>
  select(ticket_id, weight_tonnes)
')
```

The file was constructed with every weight in tonnes, so the suffix can be removed. In real work, that fact must come from a data dictionary or the person who made the file. `parse_number()` keeps the numeric part:

```{r}
#| eval: true
#| echo: false
console('
# Convert the weight column after checking its units
deliveries <- deliveries |>
  mutate(weight_tonnes = parse_number(weight_tonnes))

glimpse(deliveries)
')
```

If the suffix indicated a different unit, removing it would give the wrong answer. Convert the value to the common unit before discarding the label.

#### Duplicates {#sec-duplicates}

An **exact duplicate** repeats the entire row. `duplicated()` finds the later copy, while `distinct()` keeps one copy of every unique row:

```{r}
#| eval: true
#| echo: false
console('
# Show exact duplicates
deliveries |>
  get_dupes()

nrow(deliveries)

# Keep one copy of each unique row
deliveries <- deliveries |> distinct()
nrow(deliveries)
')
```

A **conflicting duplicate** repeats an identifier but not the whole row. Ticket `T0030` appears with two different weights, so `distinct()` keeps both. Count the identifier to find it:

```{r}
#| eval: true
#| echo: false
console('
# Find and inspect repeated ticket numbers
deliveries |>
  get_dupes(ticket_id)
')
```

The data alone does not say which row is correct. Check the source record. If that is impossible, report how the conflict was handled.

#### Impossible Values {#sec-impossible-values}

Every numeric variable has a possible range. A delivery weight cannot be negative and a moisture percentage cannot be 250. `summary()` is a quick way to inspect the minimum and maximum:

```{r}
#| eval: true
#| echo: false
console('
# Inspect the ranges of the numeric measurements
deliveries |>
  select(weight_tonnes, moisture_pct) |>
  summary()

# Display values outside their possible ranges
deliveries |>
  filter(weight_tonnes <= 0 | moisture_pct > 100) |>
  select(ticket_id, weight_tonnes, moisture_pct)
')
```

A negative weight may be a sign error or a reversed transaction. A moisture value of 250 may have lost its decimal point. Those are possible explanations, not facts. Correct a value only when the source supports the correction.

Here the synthetic file provides no defensible replacement values. Record that decision by treating the three impossible measurements as missing rather than silently carrying them into later summaries:

```{r}
#| eval: true
#| echo: false
console('
# Mark values outside the possible range as missing.
# This records a decision; it does not invent a correction.
deliveries <- deliveries |>
  mutate(
    weight_tonnes = if_else(
      weight_tonnes <= 0,
      NA,
      weight_tonnes
    ),
    moisture_pct = if_else(
      moisture_pct > 100,
      NA,
      moisture_pct
    )
  )
')
```

An extreme value can still be valid. A canola yield of 95 bu/ac is unusual, but it may belong to a very good field. Check it against the source and, if uncertainty remains, compare the analysis with and without it.

#### Inconsistent Categories {#sec-categories}

Several spellings of the same category produce separate groups. Count each categorical column before using it in `group_by()`:

```{r}
#| eval: true
#| echo: false
console('
# Count the crop names as they arrived
deliveries |>
  count(crop, sort = TRUE)
')
```

Capitalization and spaces can be standardized with string functions:

```{r}
#| eval: true
#| echo: false
console('
# Remove extra spaces and use consistent capitalization
deliveries <- deliveries |>
  mutate(crop = str_squish(crop),
         crop = str_to_title(crop))

deliveries |>
  count(crop, sort = TRUE)
')
```

Synonyms require an explicit decision. This file contains none after spacing and capitalization are standardized. When documentation says that two labels mean the same thing, put that mapping in the script; for example:

```{r}
#| eval: true
#| echo: false
console('
# Reconcile known synonyms while leaving other values unchanged
deliveries <- deliveries |>
  mutate(crop = case_when(
    crop == "Canola/Rapeseed" ~ "Canola",
    crop == "Soybean" ~ "Soybeans",
    TRUE ~ crop
  ))

deliveries |>
  count(crop, sort = TRUE)
')
```

The output is unchanged because those two labels are illustrative and do not occur in this file. A mapping should change data only when the documented source labels are present.

#### Mixed Units {#sec-mixed-units}

Most moisture readings in the file are percentages near 8 to 15. Three values are below 1, which is not a plausible moisture percentage here; the intended construction of this teaching file confirms that they are proportions such as `0.099` rather than `9.9`. The values are numeric, so they produce no parsing error.

```{r}
#| eval: true
#| echo: false
console('
# Find moisture readings recorded as decimal proportions
sum(deliveries$moisture_pct < 1, na.rm = TRUE)

deliveries |>
  filter(moisture_pct < 1) |>
  select(ticket_id, moisture_pct)

# Convert decimal proportions to percentages
deliveries <- deliveries |>
  mutate(
    moisture_pct = if_else(
      moisture_pct < 1,
      moisture_pct * 100,
      moisture_pct
    )
  )
')
```

This rule depends on subject knowledge: a grain moisture value below 1 percent is not plausible here. If the file contains a unit column, use that column in `case_when()` instead of inferring the unit from the size of the value.

#### Missing Values {#sec-missing-values}

R represents a missing value as `NA`. Most calculations involving `NA` return `NA`. The canola trial data has three fields with no rainfall reading:

```{r}
#| eval: true
#| echo: false
console('
# Read a file containing three missing rainfall values
trial_data <- read_csv(
  "practice/data/canola_trial.csv",
  show_col_types = FALSE
)

mean(trial_data$rainfall_mm)
mean(trial_data$rainfall_mm, na.rm = TRUE)
')
```

The first call returns `NA` because three rainfall values are unknown. The second calculates the mean from the observed values. The argument `na.rm = TRUE` makes that decision visible. Many Excel functions skip blank cells automatically, so the missing values need to be counted separately there.

##### Missing values in disguise {.unnumbered}

A blank cell, `"NA"`, `"N/A"`, `"-"`, `"."` and `-999` may all mean that no value was recorded. The `na` argument in the first `read_csv()` call told R which text codes to treat as missing.

Count missing values in every column before deciding what to do with them:

```{r}
#| eval: true
#| echo: false
console('
# Count missing values in each column
deliveries |>
  summarise(
    across(everything(), \\(x) sum(is.na(x)))
  )
')
```

A numeric code such as `-999` is more dangerous. R reads it as a number and includes it in calculations. `summary()` exposes it as an implausible minimum unless it is converted to `NA` when the file is read.

##### Working with missing values {.unnumbered}

`is.na()` identifies missing values and `drop_na()` removes rows that contain them:

```{r}
#| eval: true
#| echo: false
console('
# Count missing weights
sum(is.na(deliveries$weight_tonnes))

# Count rows retained after dropping any missing value
deliveries |>
  drop_na() |>
  nrow()

# Count rows retained when only analysis variables matter
deliveries |>
  drop_na(weight_tonnes, moisture_pct) |>
  nrow()

# Calculate a mean from the observed moisture readings
mean(deliveries$moisture_pct, na.rm = TRUE)
')
```

The impossible-value decision above created missing values: there are now two missing weights and three missing moisture readings. The mean excludes the flagged moisture value of 250 rather than allowing it to distort the result. Dropping every incomplete row can remove far more data than dropping rows missing the variables used in a particular calculation.

##### When is dropping safe? {.unnumbered}

The answer depends on whether the chance a value is missing depends on the value itself. If a rain gauge failed at arbitrary times, the observed days may still represent the season, although there are fewer of them. If low-yielding farms were less likely to report, the remaining records will overstate average yield. In the deliveries file, omitting the three missing moisture readings would be less concerning if records were left blank for administrative reasons than if unusually wet loads were more likely to go unrecorded. Record how many values were missing, which observations were excluded and why the exclusion seemed reasonable.

#### Dates Stored as Text {#sec-dates-text}

`glimpse()` reported `delivery_date` as `<chr>` because most rows use `2025-09-07` while two use day/month/year. Text dates sort alphabetically and cannot be used in date arithmetic.

Parse the expected format and inspect the rows that fail. `lubridate::ymd()` warns when a value does not follow year-month-day order:

```{r}
#| eval: true
#| echo: false
console('
# Parse dates written as year-month-day
deliveries_with_dates <- deliveries |>
  mutate(date_parsed = lubridate::ymd(delivery_date))

# Display dates written in a different format
deliveries_with_dates |>
  filter(is.na(date_parsed)) |>
  select(ticket_id, delivery_date)
')
```

The value `03/11/2025` is ambiguous on its own. The other exceptional value, `15/10/2025`, can only be day/month/year, and all deliveries occur from September through November, so the teaching file's intended convention is day/month/year for both exceptional rows. Parse each stated format and combine the results:

```{r}
#| eval: true
#| echo: false
console('
deliveries <- deliveries |>
  mutate(
    delivery_date = coalesce(
      lubridate::ymd(delivery_date, quiet = TRUE),
      lubridate::dmy(delivery_date, quiet = TRUE)
    )
  )

deliveries |>
  select(ticket_id, delivery_date) |>
  filter(ticket_id %in% c("T0004", "T0047"))

glimpse(deliveries$delivery_date)
')
```

The choice of day/month/year comes from the file's construction and context, not from the ambiguous value alone. The final check is that the column is `<date>`, not `<chr>` (@sec-csv-files).

Once the checks are complete, write a derived file rather than editing the raw input:

```r
write_csv(
  deliveries,
  "data/processed/grain_deliveries_clean.csv"
)
```

The raw file remains unchanged, the script records every decision, and rerunning it regenerates the clean file—the same principle as **Refresh** in Power Query.

### A Checklist {#sec-cleaning-checklist .unnumbered}

Use these checks on a new file:

1. Use `glimpse()` to check the dimensions, column names and types.
2. Write down what one row represents and whether the table is tidy.
3. Use `summary()` to inspect the ranges of numeric variables.
4. Use `count()` on categorical variables and look for spelling and spacing differences.
5. Check for exact duplicates and repeated identifiers.
6. Count `NA`s in every column and look for other missing-value codes.
7. Check that units are consistent and dates were read as dates.
8. Record every correction, exclusion and assumption in the script or Power Query steps.
