# Graphing in R {#sec-graphing-r}

```{r}
#| include: false
#| eval: true
source("_console.R")
```

The Excel chapter built a bar chart of barley acreage, a pie chart, and line charts of acreage over time. This chapter makes the same charts with `ggplot2`. Excel starts with a selected range and a chart menu. In R, we describe the chart in code: the data, the variables assigned to visual properties, and the marks to draw. The `ggplot2` package turns that description into a figure.

The data are the same two files the Excel workbooks used. [`sask_barley_2025.csv`](practice/data/sask_barley_2025.csv) has one row per barley variety with its 2025 acres and average yield. [`sask_variety_yields.csv`](practice/data/sask_variety_yields.csv) has Saskatchewan Crop Insurance Corporation records by risk zone, crop, variety and year, 2021 to 2025 [@scic_smp]; suppressed observations have missing acres and yield.

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
suppressPackageStartupMessages(library(tidyverse))

# One row per barley variety: 2025 acres and average yield
barley_2025 <- read_csv(
  "practice/data/sask_barley_2025.csv",
  show_col_types = FALSE
)

# Risk-zone records for all crops, varieties and years
variety_yields <- read_csv(
  "practice/data/sask_variety_yields.csv",
  show_col_types = FALSE
)

glimpse(barley_2025)
')
```

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Build a plot with `ggplot()`, `aes()` and a geom.
2. Reproduce the Excel bar, pie and line charts in code.
3. Filter and summarise the data that feeds a chart, instead of editing it on a sheet.
4. Order bars with `fct_reorder()` and group small categories with `fct_lump_n()`.
5. Change a plot's title, labels, colours and theme.
6. Plot several series by mapping a variable to colour.
7. Save a finished plot with `ggsave()`.
:::

## Bar charts {#sec-r-bar-example}

In Excel we selected the variety and acres columns and inserted a chart. The R equivalent names the data and the mapping, then adds a geom:

```{r}
#| label: fig-r-bar-first
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.4
#| fig-cap: "The first attempt: every barley variety."
#| fig-alt: "A vertical bar chart with 32 overlapping variety names on the horizontal axis."
barley_2025 |>
  ggplot(aes(x = Variety, y = Acres)) +
  geom_col()
```

`ggplot()` names the data, `aes()` maps `Variety` to horizontal position and `Acres` to bar height, and `geom_col()` draws the bars. The result has the same problems as the first Excel attempt: 32 varieties crowd the axis and the names are unreadable.

In Excel we fixed this by filtering the worksheet and re-sorting the rows, because an Excel chart draws whatever block of cells it is pointed at, in sheet order. In R the chart is drawn from the data frame you pass in, so the fixes are `filter()` and `fct_reorder()` in the pipe:

```{r}
#| label: fig-r-bar-filtered
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.0
#| fig-cap: "Varieties over 10,000 acres, largest at the top."
#| fig-alt: "A horizontal bar chart of 14 varieties ordered by acres, Synergy AAC longest at about 393,000."
barley_2025 |>
  filter(Acres > 10000) |>                          # varieties over 10,000 acres
  ggplot(aes(x = Acres,
             y = fct_reorder(Variety, Acres))) +    # order bars by acres
  geom_col()
```

Two things happened at once. Putting `Acres` on the x axis and `Variety` on the y axis gives horizontal bars, which is `Change Chart Type > Bar Chart` in Excel; there is no separate chart type. And `fct_reorder(Variety, Acres)` sorts the bars by their value, which in Excel required sorting the worksheet rows. Alphabetical order is the default; ordering by value is one function call, so there is no reason to settle for a chart that is hard to read.

The titles, labels and colours that took several Excel menus are arguments to `labs()` and the geom:

```{r}
#| label: fig-r-bar-final
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.0
#| fig-cap: "The finished bar chart."
#| fig-alt: "The same 14 bars in green with acreage labels, titled Acres of barley varieties in Saskatchewan, 2025."
bar_plot <- barley_2025 |>
  filter(Acres > 10000) |>
  ggplot(aes(x = Acres,
             y = fct_reorder(Variety, Acres))) +
  geom_col(fill = "#3F6F52", width = 0.7) +          # green bars, less gap
  scale_x_continuous(labels = scales::comma) +       # 400,000 not 4e+05
  labs(title = "Acres of barley varieties in Saskatchewan, 2025",
       x = "Acres", y = NULL) +
  theme_minimal(base_size = 13)

bar_plot
```

`fill` sets every bar to one colour, because the bars all represent the same thing; mapping a variable to `fill` inside `aes()` is for when colour should carry information, as in the line chart below. `theme_minimal(base_size = 13)` removes the grey background and raises the font size -- the same complaint about small default fonts applies in both programs.

## Pie charts {#sec-r-pie-example}

In Excel we built the pie from a five-row block: the four largest varieties plus an `Other` row computed with `SUM`. The R version of that grouping is `fct_lump_n()`, which keeps the largest `n` categories, weighted here by acres, and lumps everything else into `Other`:

```{r}
#| label: fig-r-pie
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.4
#| fig-height: 3.4
#| fig-cap: "2025 barley acreage: the four largest varieties and everything else."
#| fig-alt: "A pie chart with five slices; Other and Synergy AAC are the two largest at roughly 400,000 acres each."
barley_2025 |>
  mutate(Variety = fct_lump_n(Variety, 4, w = Acres)) |>   # top 4 + Other
  group_by(Variety) |>
  summarise(Acres = sum(Acres)) |>
  ggplot(aes(x = "", y = Acres, fill = Variety)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +                               # wrap the bar into a circle
  labs(title = "2025 barley acreage by variety", x = NULL, y = NULL) +
  theme_void(base_size = 13)
```

There is no `geom_pie()`. A pie in ggplot2 is a single stacked bar wrapped into a circle with `coord_polar()`, which is a fair description of what a pie chart is. The `Other` slice comes to 412,760 acres, slightly more than Synergy itself -- the same total the Excel `SUM` produced, and a useful reminder that the 28 lumped varieties jointly matter as much as the largest one.

## Line charts {#sec-r-line-example}

The line charts use the risk-zone file, which has one row per zone-variety-year. A provincial line needs one value per year, so the pipe filters to the variety and sums acres within each year before plotting:

```{r}
#| label: fig-r-line-synergy
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.6
#| fig-cap: "Acres planted with Synergy barley."
#| fig-alt: "A line from 408,000 acres in 2021 dipping to 317,000 in 2022 and recovering to about 393,000 by 2025."
variety_yields |>
  filter(Crop == "Barley",
         Variety == "SYNERGY AAC") |>
  group_by(Year) |>
  summarise(acres = sum(Acres, na.rm = TRUE)) |>   # provincial total per year
  ggplot(aes(x = Year, y = acres)) +
  geom_line(colour = "#3F6F52", linewidth = 1) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(title = "Acres planted with Synergy AAC barley",
       x = "Year", y = "Total acres") +
  theme_minimal(base_size = 13)
```

In Excel the first attempt drew `Year` as a second data series, and fixing it meant a trip through **Select Data**. That mistake cannot happen here, because `aes(x = Year, y = acres)` states which variable is which axis.

To plot two varieties, Excel needed a column per variety and a new series added by hand. In R the data stays long: keep both varieties in the filter and map `Variety` to colour, and one line is drawn per variety with a legend:

```{r}
#| label: fig-r-line-two
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.6
#| fig-cap: "Synergy has held its acreage; Copeland has lost most of its own."
#| fig-alt: "Two lines: Copeland falls from 475,000 acres in 2021 to 126,000 in 2025 while Synergy stays near 400,000."
variety_yields |>
  filter(Crop == "Barley",
         Variety %in% c("SYNERGY AAC", "COPELAND CDC")) |>
  group_by(Variety, Year) |>
  summarise(acres = sum(Acres, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = Year, y = acres, colour = Variety)) +   # one line per variety
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(title = "Copeland's acres have gone somewhere -- much of it to Synergy's malting class",
       x = "Year", y = "Total acres", colour = NULL) +
  theme_minimal(base_size = 13)
```

The Excel chapter's answer to "what about all the varieties?" was a PivotChart, with the small varieties filtered out or grouped. The R answer is the same `fct_lump_n()` used for the pie: summarise acres by variety and year, keep the seven biggest varieties, and lump the rest into `Other`:

```{r}
#| label: fig-r-line-many
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.6
#| fig-height: 3.8
#| fig-cap: "Acres of the main barley varieties, 2021 to 2025."
#| fig-alt: "Eight lines showing Copeland declining, Synergy and Austenson roughly steady, and an Other line near 300,000 acres."
variety_yields |>
  filter(Crop == "Barley") |>
  group_by(Variety, Year) |>
  summarise(acres = sum(Acres, na.rm = TRUE), .groups = "drop") |>
  mutate(Variety = fct_lump_n(Variety, 7, w = acres)) |>   # 7 biggest + Other
  group_by(Variety, Year) |>
  summarise(acres = sum(acres), .groups = "drop") |>
  ggplot(aes(x = Year, y = acres, colour = Variety)) +
  geom_line(linewidth = 0.9) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(title = "Acres of barley varieties in Saskatchewan",
       x = "Year", y = "Total acres", colour = NULL) +
  theme_minimal(base_size = 13)
```

Eight lines is about the limit for one chart, and with lines this distinct the legend works; @sec-graphing-principles covers direct labels and other ways to keep a many-line chart readable.

## Save the finished chart {#sec-saving-plots}

Assigning a plot to a name makes the object passed to `ggsave()` explicit. Width and height are in inches unless another unit is supplied.

```{r}
#| eval: true
#| echo: false
console('
# Create the output folder if needed
dir.create("output", showWarnings = FALSE)

# Save the finished bar chart
ggsave(
  "output/barley_variety_bar.png",
  plot = bar_plot,
  width = 6.4,
  height = 4.0,
  dpi = 300
)

# Confirm that the file was written
file.info("output/barley_variety_bar.png")["size"]
')
```

PNG works for slides and web pages. PDF preserves vector lines and text for publication. Stating the plot object, dimensions and resolution in code makes the export reproducible.

::: {.callout-tip collapse="true"}
## Other resources -- ggplot2

- **R for Data Science (2nd ed.)** -- [Chapter 1, "Data visualization"](https://r4ds.hadley.nz/data-visualize) builds up the ggplot2 grammar with worked examples.
- **ggplot2 documentation** -- the [function reference](https://ggplot2.tidyverse.org/reference/) with examples for every geom, scale and theme.
- **R Graphics Cookbook** -- [recipes](https://r-graphics.org/) organized by the change you want to make to a chart.
:::
