# Merging Data {#sec-merging-data}

```{r}
#| include: false
#| eval: true
source("_console.R")
suppressMessages({library(tidyverse); library(janitor)})
```

Data arrives in pieces: yields in one file, prices in another, weather in a third.  Answering
a real question usually means putting two or more of the pieces together.  This chapter
covers joins: how they work, how they fail, and how to check them.

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Join two tables on a shared key with `left_join()` and friends.
2. Say what happens to rows that do not match, and check for them.
3. Explain why a join can return more rows than it started with.
4. Harmonize units and category names so that two real datasets can be merged.
5. Combine three tables with different keys and different units of observation.
:::

## Joining Two Tables {#sec-joins}

`XLOOKUP` (@sec-lookup) pulled a price across from a second table in Excel.  The R equivalent
is a **join**:

```r
library(tidyverse)

yields <- read_csv("data/field_yields.csv")
prices <- read_csv("data/crop_prices.csv")

yields |>
  left_join(prices, by = "crop")
```

The `by` argument names the **key**, the column both tables share, which decides which row
matches which.  Everything about joins comes back to the key.

### The four joins you need {.unnumbered}

They differ in which rows survive.

- **`left_join(a, b)`** keeps every row of `a` and fills in `NA` where `b` has no match. This
  is usually the one you want: your data stays intact and the extra columns come along.
- **`inner_join(a, b)`** keeps only rows that matched in both, and drops the rest without
  saying so.
- **`full_join(a, b)`** keeps everything from both, with `NA` wherever either side is missing.
- **`anti_join(a, b)`** keeps rows of `a` with no match in `b`. It returns no columns from `b`
  at all, which sounds useless until you need to find out what failed to match.

### Check what did not match {.unnumbered}

`left_join` does not complain when nothing matches.  It hands you a column of `NA` and lets you
carry on, so a join can fail completely without producing an error.

`anti_join` is how you look:

```r
yields |>
  anti_join(prices, by = "crop") |>
  distinct(crop)
```

Run that on real Saskatchewan data and the problem shows up straight away.  The variety trial
data calls it `"Canola/Rapeseed"` and the price list says `"Canola"`.  The trial says
`"Soybean"`, the prices say `"Soybeans"`.  Neither file is wrong.  They were written by
different people, and a join matches on exact text, so every canola row comes back with a
missing price.

Reconciling the categories is part of cleaning (@sec-cleaning, the next chapter).  Checking
before you assume the join worked is the habit worth building.

### The row-count trap {.unnumbered}

A join can return more rows than it started with.

This happens when the key is not unique in the second table.  If `prices` has one row per crop,
each yield row matches one price and the row count does not change.  If `prices` has one row
per crop *per year* and you join on crop alone, every yield row matches several price rows and
R returns all the combinations.  Sixty fields become two hundred and forty, and every summary
computed afterwards is wrong in a way that still looks plausible.

Check the count:

```r
nrow(yields)                                    # before
joined <- yields |> left_join(prices, by = "crop")
nrow(joined)                                    # after -- should be the same
```

If it grew, the key is incomplete.  Join on everything that identifies a row:

```r
yields |> left_join(prices, by = c("crop", "year"))
```

This is the composite key from @sec-lookup, where RM and year had to be glued together with `&`
to identify a row.  R lets you name several key columns instead of building one.

## A Worked Example: Canadian and US Canola Yields {#sec-canada-us-merge}

The Canadian and US canola data from @sec-getting-data make a realistic merge, because the
two tables agree on almost nothing: different column names, different units, different
geography labels.  The work before the join is most of the work.

Suppose the Canadian table has national average yield in kilograms per hectare, and the US
table reports pounds per acre.  Neither can be compared until both are in the same unit.
Prepare each table separately: keep the columns you need, convert to tonnes per hectare, and
give the yield column a name that says whose it is:

```r
canada <- canada_raw |>
  clean_names() |>
  filter(geo == "Canada") |>
  mutate(canada_t_ha = value / 1000) |>          # kg/ha -> t/ha
  select(year = ref_date, canada_t_ha)

us <- us_raw |>
  clean_names() |>
  mutate(us_t_ha = value * 0.00112) |>           # lb/ac -> t/ha
  select(year, us_t_ha)
```

Now the join is the easy part.  The key is `year`:

```r
both <- canada |>
  left_join(us, by = "year")

both
```

```
   year canada_t_ha us_t_ha
   <dbl>      <dbl>   <dbl>
 1  2015       1.85    1.66
 2  2016       2.06    1.71
 3  2017       2.28    1.87
 ...
```

One table, one row per year, both countries' yields in the same unit.  Run the checks anyway:
`nrow(both)` should equal `nrow(canada)`, and `anti_join(canada, us, by = "year")` shows any
years one source has and the other does not -- a data release usually lags a year on one side
or the other, and it is better to find that now than in a chart.

## A Worked Example: Three Tables {#sec-three-table-merge}

Most questions need more than two tables, and the tables are usually different kinds of
thing.  Here are three, all real, and all available to download:

- [`rm_yields_2015_2024.csv`](practice/data/rm_yields_2015_2024.csv): the Saskatchewan
  RM yields from @sec-direct-download, cut down to 2015 to 2024 and six crops.  One row per
  RM per year, yields in bushels per acre.
- [`rm_lookup.csv`](practice/data/rm_lookup.csv): one row per RM, with its name, census
  division and 2021 population from Statistics Canada, and the nearest of sixteen Environment
  Canada weather stations, with the distance to it in kilometres.
- [`station_precip.csv`](practice/data/station_precip.csv): one row per weather station per
  year, with total May to August precipitation in millimetres, computed from Environment
  Canada's daily records, and a count of days with no reading.

The question is whether canola yields track growing-season rain.  Answering it needs a yield,
a rain total, and a way to connect the two, and each of those is in a different file.  The
lookup table is the bridge: it is the only file that knows which station goes with which RM.

Read all three and look at them:

```{r}
#| eval: true
#| echo: false
console('
yields <- read_csv("data/rm_yields_2015_2024.csv") |> clean_names()
rms    <- read_csv("data/rm_lookup.csv") |> clean_names()
rain   <- read_csv("data/station_precip.csv") |> clean_names()

glimpse(yields)
glimpse(rms)
glimpse(rain)
')
```

Before joining anything, write down what identifies a row in each table.  In `yields` it is
`rm` and `year` together.  In `rms` it is `rm_number` alone.  In `rain` it is `station` and
`year` together.  Those are the keys, and the key column does not have the same name on both
sides, which `by` handles with `c("left_name" = "right_name")`.

### First join: yields to the lookup {.unnumbered}

Check what will not match, then join, then check the row count:

```{r}
#| eval: true
#| echo: false
console('
yields |>
  anti_join(rms, by = c("rm" = "rm_number")) |>
  distinct(rm)

with_station <- yields |>
  left_join(rms, by = c("rm" = "rm_number"))

nrow(yields)
nrow(with_station)
')
```

Every RM in the yield data has a row in the lookup, and the row count is unchanged.  That is
what a many-to-one join should look like: many yield rows for each RM, one lookup row per RM,
so each yield row picks up exactly one name and one station.

### Second join: add the rain {.unnumbered}

The key now has two parts.  The station name in `with_station` is called `weather_station`,
the year is called `year` on both sides, and both are needed to identify one rain total:

```{r}
#| eval: true
#| echo: false
console('
with_station |>
  anti_join(rain, by = c("weather_station" = "station", "year")) |>
  distinct(weather_station, year)

merged <- with_station |>
  left_join(rain, by = c("weather_station" = "station", "year"))

nrow(merged)
')
```

Here is what the row-count trap looks like when you leave `year` out of the key:

```{r}
#| eval: true
#| echo: false
console('
with_station |>
  left_join(rain, by = c("weather_station" = "station")) |>
  nrow()
')
```

Each RM-year matched all ten years of its station's rain, so 2,950 rows became 29,500.
Recent versions of dplyr print a warning about an unexpected many-to-many relationship when
this happens.  Read it when it appears; it is telling you the key is incomplete.

### What you can now do {.unnumbered}

One table, one row per RM-year, with the yield and the season's rain side by side:

```{r}
#| eval: true
#| echo: false
console('
merged |>
  select(year, rm, rm_name, canola, weather_station, precip_may_aug_mm)

merged |>
  filter(!is.na(canola)) |>
  group_by(year) |>
  summarise(canola_bu_ac = mean(canola),
            rain_mm = mean(precip_may_aug_mm))
')
```

The wet 2016 season has the highest canola yield in the table and the dry 2017 one of the
lowest, which is the pattern you would expect.  2021 is the exception: rainfall was ordinary
and yields collapsed, because that summer's problem was heat more than drought.  Rain explains
some of the variation in yields and not all of it.

Two limits of this particular merge belong in the README.  The nearest station is sometimes
far away: RM 1, Argyle, is matched to Kipling, 145 km off, because that is the closest of
sixteen stations.  And a few station-years are missing three weeks of daily readings, which
is what the `days_missing` column is for.
