# Data Manipulation {#sec-dplyr-verbs}

```{r}
#| include: false
#| eval: true
source("_console.R")
# No library() calls here: the first console() chunk runs library(tidyverse)
# itself, so its transcript shows the full attach message a student sees.
trial_data <- readr::read_csv("practice/data/canola_trial.csv", show_col_types = FALSE)
```

With a dataset in R and inspected, we can start doing things to it. Most data analysis
involves repeating a surprisingly small number of operations: keep certain rows, keep
certain columns, create or change a column, sort the rows, calculate summary statistics,
and do those calculations separately for different groups.

The tidyverse package `dplyr` gives us a function for each of these jobs (you will often see
them called the dplyr "verbs"). None of the operations are new -- each is the R form of
something from the Excel modules:

| Question                                | `dplyr` function | Excel counterpart        |
|-----------------------------------------|------------------|--------------------------|
| Which rows do I want?                   | `filter()`       | filtering                |
| Which columns do I want?                | `select()`       | keeping certain columns  |
| What new column do I want?              | `mutate()`       | a formula in a new column|
| In what order do I want the rows?       | `arrange()`      | sorting                  |
| What summary do I want?                 | `summarise()`    | summary formulas         |
| That summary separately for groups?     | `group_by()`     | a PivotTable             |

Once you understand these operations, most everyday data manipulation in R is a matter of
combining them.

We will continue with the canola trial data from the last chapter, in the same project folder.
Since this chapter does a new job, it gets a new script: `code/2_data_manipulation.R`.
Start it with a header, then load the tidyverse and read the data in:

```{r}
#| eval: true
#| echo: false
console('
# ---
# Title: Manipulating the canola trial data
# Author: Your Name
# Date: 2026-10-06
# Description:
#   Filters, transforms, and summarizes the canola trial data
#   using the dplyr functions.
# ---

# Load packages
library(tidyverse)

# Read in the data
trial_data <- read_csv("data/canola_trial.csv")
')
```

## Data Manipulation Functions {#sec-verbs-basic}

::: {.video-callout}
::: {.callout-note collapse="true" title="Video walkthrough of this section"}
{{< video https://youtu.be/kUNgyoJBnTc >}}
:::
:::

We will learn four functions in this section.  Two affect the rows:

- `filter()` chooses rows, and
- `arrange()` changes their order.

Two affect the columns:

- `select()` chooses columns, and
- `mutate()` creates or changes them.

Each function takes the data frame as its first argument, followed by what to do with it.

### `filter()`: Keep Certain Rows

`filter()` keeps the rows that satisfy a condition.  It's two arguments are the data frame and the condition.  For example, to keep only the fields with yields above 50 bushels per acre:

```{r}
#| eval: true
#| echo: false
console('
filter(trial_data, yield_bu_acre > 50)
')
```

Notice that `filter()` displays the high-yielding fields but does not remove the other fields from `trial_data` -- check `nrow(trial_data)` and all 120 rows are still there. Every function in this chapter works the same way: it takes a data frame as input and returns a *new* data frame as output.  Running the above code will only display that output, R does not save it. To keep a result, assign it to an object:

```{r}
#| eval: true
#| echo: false
console('
high_yield <- filter(trial_data, yield_bu_acre > 50)
')
```

(An assignment prints nothing -- the result went into `high_yield` instead of the console.)

Now there are two objects: `trial_data`, the original 120 fields, and `high_yield`, the filtered subset. That is a useful feature -- you can experiment without changing the original data. Throughout the chapter, an example without `<-` is just displaying a result.

Text variables work too:

```{r}
#| eval: true
#| echo: false
console('
filter(trial_data, variety == "InVigor")
')
```

Note the **two** equals signs: `==` asks whether two things are equal, so `variety == "InVigor"` asks "is `variety` equal to `"InVigor"`?". Also note the quotation marks: `"InVigor"` with quotes is text; `InVigor` without quotes would be read as the name of an object, which probably doesn't exist and would cause an error.

You can combine conditions with `&` (and -- both must be true), `|` (or -- at least one must be true), and `!` (not):

```{r}
#| eval: true
#| echo: false
console('
##Filter for yields above 50 with the variety InVigor
filter(trial_data, yield_bu_acre > 50 & variety == "InVigor")

##Filter for fertilizer below 80 or above 160
filter(trial_data, fertilizer_kg_ha < 80 | fertilizer_kg_ha > 160)

##Filter for yields above 60 with the variety not equal to InVigor
filter(trial_data, yield_bu_acre > 60 & !variety == "InVigor")
')
```

When you want to match several possible values of the same variable, `%in%` is easier to read:

```{r}
#| eval: true
#| echo: false
console('

##The following two filters are identical

filter(trial_data, variety == "InVigor" | variety == "DEKALB")

filter(trial_data, variety %in% c("InVigor", "DEKALB"))
')
```


### `arrange()`: Sort Rows

The function `arrange()` changes the order of the rows.  It also takes a minimum of two arguments: the data frame and the variable to sort by.  The code `arrange(trial_data, yield_bu_acre)` sorts from lowest yield to highest.  If you want to arrange from highest to lowest then use the `desc()` (descending order) function. 

```{r}
#| eval: true
#| echo: false
console('

##Sort data by yield (lowest to highest)
arrange(trial_data, yield_bu_acre)

##Sort data by yield (highest to lowest)
arrange(trial_data, desc(yield_bu_acre))
')
```

You can sort by more than one variable -- `arrange(trial_data, variety, desc(yield_bu_acre))` sorts by variety and then, within each variety, from highest yield to lowest. Like the other functions, `arrange()` does not permanently rearrange `trial_data` unless you save the result.

```{r}
#| eval: true
#| echo: false
console('

## Sort data by variety first, then yield (highest to lowest)
arrange(trial_data, variety, desc(yield_bu_acre))
')
```

### `select()`: Keep Certain Columns

`select()` chooses variables rather than observations.  It's arguments are the data frame and the names of the variables to keep.  For example, to keep only the field ID, variety, and yield:

```{r}
#| eval: true
#| echo: false
console('
select(trial_data, field_id, variety, yield_bu_acre)
')
```

The result has the same rows but only those three columns -- especially useful with datasets containing dozens or hundreds of variables. You can also drop a column by putting `-` in front of its name, or keep a range of adjacent columns:

```{r}
#| eval: true
#| echo: false
console('

## Select  everything except rainfall
select(trial_data, -rainfall_mm)                 

# Select columns from field_id through rainfall_mm
select(trial_data, field_id:rainfall_mm)         
')
```

A relative of `select()` is `rename()`, which changes a column's name without dropping anything: `rename(trial_data, yield = yield_bu_acre)`.

### `mutate()`: Create or Change Columns

Potentially the most useful function is `mutate()`, which creates a new variable or changes an existing one. Suppose we want yield in tonnes per hectare rather than bushels per acre. We can accomplish this by running the code:

```{r}
#| eval: true
#| echo: false
console('
## Create a new variable which is yield in tons per ha using the conversion factor of 0.056.  Save as trial data
trial_data <- mutate(trial_data, yield_t_ha = yield_bu_acre * 0.0560)

## Print trial data to see our new column
head(trial_data)
')
```

Note that we have saved the resulting data frame back into `trial_data`. In R we can always overwrite an object.  The result contains all the original columns plus a new `yield_t_ha`. Read the expression as: create `yield_t_ha` and set it equal to `yield_bu_acre` × 0.0560. 

You can create several variables in one call:

```{r}
#| eval: true
#| echo: false
console('
## Create new rainfall and fertilizer variables
trial_data <- mutate(trial_data,
       rainfall_cm = rainfall_mm / 10,
       kg_n_per_t = fertilizer_kg_ha / yield_t_ha)

## Print trial data to see our new columns
print(trial_data)
')
```

In the second variable the hectares cancel -- fertilizer and yield are both per hectare -- leaving kilograms of nitrogen per tonne of grain.


## The Pipe `|>` {#sec-pipe}

::: {.video-callout}
::: {.callout-note collapse="true" title="Video walkthrough of this section"}
{{< video https://youtu.be/m5c5gwby2tY >}}
:::
:::

Many questions that we want to resolve require more than one function. Suppose we want to know: among the InVigor fields, what are the field IDs and yields, ordered from highest to lowest? We know every operation we need -- filter to InVigor, select the two columns, arrange by descending yield. We can combine these using the **pipe**, which is written as `|>`.  The pipe takes whatever is on its *left* and feeds it as the first argument to the function on its *right*.  Essentially, think of the pipe as a command that says "then".  For example, suppose we want to

- filter the data to InVigor, THEN
- select the columns field_id and yield_bu_acre, THEN
- arrange by yield (highest to lowest).

We can write this in R code as:

```{r}
#| eval: true
#| echo: false
console('
trial_data |>                         ##Take our trial data THEN  
  filter(variety == "InVigor") |>     ##Filter it to InVigor THEN
  select(field_id, yield_bu_acre) |>  ##Select field_id and yield_bu_acre columns THEN
  arrange(desc(yield_bu_acre))        ##Sort by yield (highest to lowest)
')
```

One thing to note is that we have split this code across multiple lines.  We could have written it all one one line as:
```{r}
#| eval: true
#| echo: false
console('
trial_data |> filter(variety == "InVigor") |> select(field_id, yield_bu_acre) |> arrange(desc(yield_bu_acre))
')
```

This becomes difficult to read. R knows that when a line ends with the pipe `|>`, the command is not finished, so it continues reading the next line. This lets us write a long sequence of operations across several lines rather than cramming everything onto one line.

This is not something special about the pipe. More generally, R will keep reading whenever it can tell that a command is incomplete. For example, we can also split a function across multiple lines:

```{r}
#| eval: true
#| echo: false
console('
##These two commands are identical
filter(trial_data,
        yield_bu_acre > 50)

filter(trial_data, yield_bu_acre > 50)
  
')
```
In the first command, the opening parenthesis `(` tells R that the command will not be complete until there is a closing parenthesis `)`. 

To keep a result, assign the whole pipeline to an object:

```{r}
#| eval: true
#| echo: false
console('
invigor_yields <- trial_data |>
  filter(variety == "InVigor") |>
  select(field_id, yield_bu_acre) |>
  arrange(desc(yield_bu_acre))

head(invigor_yields)
')
```

## `summarise()` and `group_by()` {#sec-summarise-group}

::: {.video-callout}
::: {.callout-note collapse="true" title="Video walkthrough of this section and Writing Output"}
{{< video https://youtu.be/a01D6EyHtvA >}}
:::
:::

The four functions so far rearrange or add information while keeping observations as rows. Sometimes we want to *collapse* the observations into summary statistics. That is the job of `summarise()`.

### `summarise()`: Many Rows Become One

Suppose we want to get the mean of yields in the data.  We can use the summarise function and inside define a column that is the mean of yields.  The result is a new data frame with a single row and a single column, `mean_yield`:
```{r}
#| eval: true
#| echo: false
console('
trial_data |>
  summarise(mean_yield = mean(yield_bu_acre, na.rm = TRUE))
')
```

We can extend this to get several summary statistics at the same time:

```{r}
#| eval: true
#| echo: false
console('
trial_data |>
  summarise(
    mean_yield = mean(yield_bu_acre, na.rm = TRUE),
    sd_yield = sd(yield_bu_acre, na.rm = TRUE),
    n_fields = n()
  )
')
```


The names on the left (`mean_yield`, `sd_yield`, `n_fields`) become column names in the
result; the expressions on the right calculate the values. `n()` counts the rows.   Note again that we have split the arguments of summarise across multiple lines.  We could also write:


```{r}
#| eval: true
#| echo: false
console('
trial_data |>
  summarise(mean_yield = mean(yield_bu_acre, na.rm = TRUE), sd_yield = sd(yield_bu_acre, na.rm = TRUE), n_fields = n())
')
```


The contrast with `mutate()` matters. `mutate()` keeps the observations and adds
information; `summarise()` collapses the observations into summaries. So
`trial_data |> mutate(mean_yield = mean(yield_bu_acre))` still has 120 rows, with the
overall mean repeated on every row, while the `summarise()` version returns one row.

Note that we could have used potentially simpler commands from "base R" to get the mean and standard deviation above:

```{r}
#| eval: true
#| echo: false
console('
##Get mean and st dev of yields using base R

mean(trial_data$yield_bu_acre)
sd(trial_data$yield_bu_acre)
')
```

The power of summarise comes from being able to combine it with other commands via the pipe.  For example, we can combine it with `filter()` to get the mean of only the InVigor fields:

```{r}
#| eval: true
#| echo: false
console('
trial_data |>
  filter(variety == "InVigor") |>
  summarise(
    mean_yield = mean(yield_bu_acre, na.rm = TRUE)
    )
')
```

Or we may want to get the means for each variety -- not just InVigor.  In this case we can use `group_by()`

### `group_by()`: Calculate Separately for Groups
`group_by()` doesn't change the data -- it tells subsequent operations to apply *per group*:

```{r}
#| eval: true
#| echo: false
console('
trial_data |>
  group_by(variety) |>
  summarise(
    mean_yield = mean(yield_bu_acre, na.rm = TRUE),
    sd_yield = sd(yield_bu_acre, na.rm = TRUE),
    n_fields = n()
  )
')
```

Take `trial_data`, *then* divide the observations by variety, *then* calculate the
statistics separately for each variety. The result has one row per variety. This is the R equivalent of a PivotTable.

The central pattern -- `data |> group_by(group_variable) |> summarise(...)` -- appears
throughout the rest of the book.

## Writing Output {#sec-writing-output}

Results so far have existed only inside R. To save one as a file that can be opened in Excel
or sent to someone else, pass the object and a filename to `write_csv`. With our project
layout, results go in `output/`:

```{r}
#| eval: true
#| echo: false
console('
## Summarise mean yields by variety
variety_summary <- trial_data |>
  group_by(variety) |>
  summarise(mean_yield = mean(yield_bu_acre, na.rm = TRUE),
            n_fields = n())

## Write the summary table to the output folder
write_csv(variety_summary, "output/variety_summary.csv")
')
```

Notice that `write_csv` prints nothing to the console -- the result is the file itself,
which appears in `output/`. `read_csv` brings data into R; `write_csv` sends a data frame
out to a file. Because the output is created by the script, you can always recreate it by
running the script again.

## Combining It All {#sec-dplyr-combining}

We now have enough tools to answer a realistic question: for each variety, among fields
that received at least 100 kg/ha of nitrogen, what was the mean yield in tonnes per
hectare, and how many fields were observed? Show the highest-yielding variety first.

```{r}
#| eval: true
#| echo: false
console('
variety_summary_100 <- trial_data |>
  filter(fertilizer_kg_ha >= 100) |>
  mutate(yield_t_ha = yield_bu_acre * 0.0560) |>
  group_by(variety) |>
  summarise(mean_yield_t_ha = mean(yield_t_ha, na.rm = TRUE),
            n_fields = n()) |>
  arrange(desc(mean_yield_t_ha))
variety_summary_100
')
```

Do not read it as one giant command; read it one step at a time:

1. **Filter** to the fields that received at least 100 kg/ha.
2. **Create** the variable the question needs: metric yield.
3. **Group** the observations by variety.
4. **Summarise** each group down to one row: its mean yield and its number of fields.
5. **Sort** from highest mean yield to lowest.

Each line does one simple thing, and that is the reason for pipes: a good pipeline makes the
logic of an analysis visible. Saving the table is one more line:
`write_csv(variety_summary_100, "output/variety_summary_100.csv")`.


## Debugging {#sec-debugging}

Your code will break. That is normal -- everyone's does -- and finding the problem is a skill of its own. It has two parts: reading what R tells you, and narrowing down where things went wrong.

### Reading error messages {.unnumbered}

R's error messages name the problem more precisely than they first appear to. Here are the three you will meet most often, each with its usual cause.

A misspelled function name (or a package you forgot to load):

```{r}
#| eval: true
#| echo: false
console('
quantle(trial_data$yield_bu_acre, 0.5)
')
```

R is exact about names -- `quantle` is not `quantile`, and no amount of context will make R guess. The same message appears when the function is spelled correctly but its package is not loaded; then the fix is `library(tidyverse)`.

A misspelled object name, or an object that was never created:

```{r}
#| eval: true
#| echo: false
console('
mean(trial_dta$yield_bu_acre)
')
```

Check the Variables pane: is the object there, spelled exactly that way? If it is missing entirely, the line that creates it has not been run -- which often means the script was not run from the top.

Using `=` where `==` belongs:

```{r}
#| eval: true
#| echo: false
console('
filter(trial_data, variety = "InVigor")
')
```

This one diagnoses itself -- the message even suggests the fix.

### The wrong answer with no error {.unnumbered}

The harder case is code that runs and returns something wrong. R is exact about the *values* in the data too:

```{r}
#| eval: true
#| echo: false
console('
filter(trial_data, variety == "invigor")
')
```

No error -- just zero rows, because no variety is spelled `"invigor"` with a small i. A comparison matches text exactly, capitals included.

For a pipeline that produces a wrong-looking result, do not stare at the whole thing and guess. Run it one step at a time: run the first step alone, check the result, add the second step, check again, and keep going. The first point where the data stops looking the way you expect is where the problem is.

::: {.callout-tip collapse="true"}
## Other resources — dplyr

- **R for Data Science (2nd ed.)** -- [Chapter 3, "Data transformation"](https://r4ds.hadley.nz/data-transform.html): the core dplyr functions, the pipe, and grouped summaries.
- **dplyr documentation** -- the [dplyr reference site](https://dplyr.tidyverse.org/), including the one-page [cheatsheet](https://rstudio.github.io/cheatsheets/html/data-transformation.html).
- **Video** -- Equitable Equations, [*The five main dplyr verbs*](https://www.youtube.com/watch?v=b2ODCVAsXlw) -- filter, select, mutate, arrange, and summarise in one sitting.
:::
