# Graphing in R {#sec-graphing-r}

```{r}
#| include: false
#| eval: true
source("_console.R")
```

Excel starts with a selected range and a chart menu. In R, we describe the chart in code: the data, the variables assigned to visual properties, and the marks to draw. The `ggplot2` package turns that description into a figure.

The examples use [`field_yields.csv`](practice/data/field_yields.csv), with one row for each of sixty hypothetical canola fields.

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the field data
suppressPackageStartupMessages(library(tidyverse))

yields <- read_csv(
  "practice/data/field_yields.csv",
  show_col_types = FALSE
)
glimpse(yields)
')
```

The file is simulated for teaching, so its patterns are not facts about canola. Three fields have no recorded yield.

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Build a plot with `ggplot()`, `aes()` and a geom.
2. Choose a geom for a comparison, trend, relationship or distribution.
3. Distinguish mapping an aesthetic to a variable from setting it to a constant.
4. Add titles, axis labels and a theme.
5. Divide a plot into small multiples with facets.
6. save a plot with `ggsave()`.
:::

## The Grammar of Graphics {#sec-grammar}

`ggplot2` follows the **grammar of graphics**. A plot is assembled from a few parts:

- **Data:** the data frame to plot.
- **Aesthetic mappings:** columns assigned to position, colour, size or shape.
- **Geometries:** points, lines, bars, boxes and other marks.
- **Scales:** the rules connecting data values to axis positions or colours.
- **Facets:** separate panels for groups.

Some geoms also calculate a statistic. A histogram bins values before drawing bars, and a smooth fits a line or curve before drawing it.

## A First Plot {#sec-first-ggplot}

This scatter plot places field size on the horizontal axis and yield on the vertical axis:

```{r}
#| label: fig-r-first-scatter
#| eval: true
#| echo: true
#| message: false
#| fig-width: 5.4
#| fig-height: 3.2
#| fig-cap: "Field size and canola yield for sixty hypothetical fields."
ggplot(yields, aes(x = acres, y = yield_bu_acre)) +
  geom_point()
```

Read the code in three parts:

- `ggplot(yields, ...)` names the data frame.
- `aes(x = acres, y = yield_bu_acre)` maps two columns to position.
- `geom_point()` draws one point for each row with both values present.

Layers are joined with `+`. A pipe passes a data object into a function; the `+` adds a layer to a ggplot object.

The warning says that three rows were removed because their yields are missing. It is information, not an error. Once you have verified that the count is expected, `na.rm = TRUE` suppresses the message. Excel drops blank values from a chart without reporting how many it omitted.

## Common Geoms {#sec-geoms}

### `geom_point()`: scatter plots {.unnumbered}

Use points for the relationship between two numeric variables. The `alpha` argument makes overlapping points partly transparent.

```{r}
#| label: fig-r-alpha
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.4
#| fig-height: 3.2
#| fig-cap: "Transparency makes overlapping observations easier to see."
ggplot(yields, aes(x = acres, y = yield_bu_acre)) +
  geom_point(alpha = 0.55, na.rm = TRUE)
```

### `geom_line()`: change over time {.unnumbered}

A line chart needs one value for each point along the horizontal axis. The field file has only two years, which is not enough to show a meaningful trend, so use the ten-year RM-yield series from Module 3:

```{r}
#| label: fig-r-line
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.4
#| fig-height: 3.0
#| fig-cap: "Mean published RM canola yield, 2015–2024."
annual_mean <- read_csv(
  "practice/data/rm_yields_2015_2024.csv",
  show_col_types = FALSE
) |>
  group_by(Year) |>
  summarise(mean_yield = mean(Canola, na.rm = TRUE)) |>
  rename(year = Year)

ggplot(annual_mean, aes(x = year, y = mean_yield)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 2015:2024)
```

The summary creates one row per year. `geom_line()` connects those rows in year order, and `scale_x_continuous()` states the integer-year tick marks. Excel's Line chart usually treats years as evenly spaced category labels instead.

### `geom_col()` and `geom_bar()`: bars {.unnumbered}

Use `geom_col()` when the bar heights have already been calculated.

```{r}
#| label: fig-r-col
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.4
#| fig-height: 3.0
#| fig-cap: "Mean field yield by region."
region_means <- yields |>
  group_by(region) |>
  summarise(mean_yield = mean(yield_bu_acre, na.rm = TRUE))

ggplot(region_means, aes(x = region, y = mean_yield)) +
  geom_col()
```

A bar of means hides how many fields are behind it and how much they vary. The box plot below shows the spread and is usually more informative for comparing these regions. In Excel, sort the summary table to reorder the bars.

Use `geom_bar()` when the desired value is the number of rows. It counts the observations in each category:

```{r}
#| label: fig-r-bar
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.0
#| fig-height: 2.8
#| fig-cap: "Number of fields in each region."
ggplot(yields, aes(x = region)) +
  geom_bar()
```

Excel has no counting chart: use `COUNTIF` or a PivotTable to create the counts, then chart that summary.

### `geom_histogram()` and `geom_boxplot()`: distributions {.unnumbered}

A histogram shows the shape of one numeric variable. State the number of bins or the bin width rather than accepting the default without looking at it.

```{r}
#| label: fig-r-hist
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.2
#| fig-height: 3.0
#| fig-cap: "Distribution of recorded field yields."
ggplot(yields, aes(x = yield_bu_acre)) +
  geom_histogram(bins = 12, na.rm = TRUE)
```

With only 57 recorded yields, the apparent shape changes noticeably with 5, 12 or 30 bins; none is a uniquely true display. In Excel use **Insert > Statistical Chart > Histogram**, then **Format Axis** to set Bin width or Number of bins.

A box plot makes distributions compact enough to compare across categories.

```{r}
#| label: fig-r-box
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.2
#| fig-height: 3.0
#| fig-cap: "Field-yield distributions by region."
ggplot(yields, aes(x = region, y = yield_bu_acre)) +
  geom_boxplot(na.rm = TRUE)
```

Excel's **Box and Whisker** chart is in the same Statistical Chart menu, but it expects each group's values in a separate column.

### `geom_smooth()`: fitted patterns {.unnumbered}

`geom_smooth(method = "lm")` adds a fitted straight line and an interval around it:

```{r}
#| label: fig-r-smooth
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.4
#| fig-height: 3.2
#| fig-cap: "Field size and yield with a fitted straight line."
ggplot(yields, aes(x = acres, y = yield_bu_acre)) +
  geom_point(alpha = 0.45, na.rm = TRUE) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    na.rm = TRUE
  )
```

Here the line is nearly flat and the 95% confidence band is wide: these synthetic fields show no clear relationship between field size and yield. A fitted line is drawn whether or not a clear relationship exists, so read the points and band as well as the line. In Excel use **Chart Design > Add Chart Element > Trendline > Linear**; Excel can display the equation and R-squared but does not add this confidence band. Module 7 explains what the line estimates.

## Mapping and Setting Aesthetics {#sec-mapping-vs-setting}

Put an aesthetic inside `aes()` when it should depend on a variable. Put it outside `aes()` when every mark should receive the same value.

```{r}
#| label: fig-r-map-set
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 4.7
#| fig-height: 3.0
#| fig-subcap:
#|   - "A. Colour mapped to region"
#|   - "B. One colour set for all points"
#| layout-ncol: 2
# Map colour to a column in the data
ggplot(yields, aes(x = acres, y = yield_bu_acre, colour = region)) +
  geom_point(na.rm = TRUE)

# Set one colour for every point
ggplot(yields, aes(x = acres, y = yield_bu_acre)) +
  geom_point(colour = "steelblue", na.rm = TRUE)
```

Putting `colour = "steelblue"` inside `aes()` would ask ggplot to map the literal category `"steelblue"` to a colour scale, producing an unnecessary legend.

In Excel, colouring points by a category requires one series per group: restructure the data with `FILTER()` or a PivotTable, then use **Select Data > Add** for each series. The one-line `aes(colour = region)` is a practical advantage of ggplot2.

## Labels and Themes {#sec-ggplot-polish}

`labs()` supplies the text and `theme_minimal()` changes the non-data elements of the plot:

```{r}
#| label: fig-r-polished
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.0
#| fig-height: 3.4
#| fig-cap: "A labelled scatter plot with region mapped to colour."
ggplot(yields, aes(x = acres, y = yield_bu_acre, colour = region)) +
  geom_point(alpha = 0.65, na.rm = TRUE) +
  labs(
    title = "Field size and yield, by region",
    subtitle = "Sixty hypothetical Saskatchewan fields, 2024--2025",
    x = "Field size (acres)",
    y = "Yield (bu/acre)",
    colour = "Region",
    caption = "Source: synthetic AREC 261 data"
  ) +
  theme_minimal()
```

Other built-in themes include `theme_classic()` and `theme_bw()`. A theme changes the surrounding presentation; it does not change the data or mappings.

## Faceting {#sec-faceting}

Faceting divides one plot into panels. Each panel below uses the same histogram definition for a different region:

```{r}
#| label: fig-r-facet
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.2
#| fig-height: 3.2
#| fig-cap: "One yield histogram for each region."
ggplot(yields, aes(x = yield_bu_acre)) +
  geom_histogram(bins = 8, na.rm = TRUE) +
  facet_wrap(~ region) +
  labs(x = "Yield (bu/acre)", y = "Fields")
```

`facet_grid(region ~ variety)` creates a two-dimensional grid. Use it only when the resulting number of panels remains readable.

Excel has no direct faceting feature. Use a PivotChart with a slicer to view one group at a time, or build one chart per group and set identical axis bounds manually.

## Saving Plots {#sec-saving-plots}

Assigning a plot to a name before `ggsave()` makes the saved object explicit and lets you reuse it. Create the destination folder first:

```{r}
#| eval: true
#| echo: false
console('
# Build and save the regional box plot
dir.create("output", showWarnings = FALSE)

region_plot <- ggplot(
  yields,
  aes(x = region, y = yield_bu_acre)
) +
  geom_boxplot(na.rm = TRUE) +
  labs(x = NULL, y = "Yield (bu/acre)")

ggsave(
  "output/region_boxplot.png",
  plot = region_plot,
  width = 6,
  height = 4,
  dpi = 300
)
')
```

Width and height are in inches unless another unit is supplied. PNG works for slides and web pages; PDF preserves vector lines and text for publication. SVG output requires the **svglite** package. Unlike Excel's **Save as Picture**, `ggsave()` lets you state the output dimensions and resolution in code.

::: {.callout-tip collapse="true"}
## Other resources -- ggplot2

- **R for Data Science (2nd ed.)** -- [Chapter 1, "Data visualization"](https://r4ds.hadley.nz/data-visualize) introduces `ggplot()`, aesthetic mappings, geoms and facets.
- **ggplot2 documentation** -- the [function reference](https://ggplot2.tidyverse.org/reference/) lists the arguments and examples for each geom.
- **RStudio cheatsheet** -- [Data visualization with ggplot2](https://rstudio.github.io/cheatsheets/html/data-visualization.html).
:::
