# Graphing in R {#sec-graphing-r}

```{r}
#| include: false
#| eval: true
source("_console.R")
```

The Excel chapter built a bar chart, scatter plot and histogram from the canola variety trial. This chapter builds the same three charts with `ggplot2`, then adds value labels, a fitted line, a direct field label and a distribution comparison by variety.

The examples use [`canola_trial.csv`](practice/data/canola_trial.csv). The data are simulated for teaching and should not be interpreted as results from a real variety trial.

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
suppressPackageStartupMessages(library(tidyverse))

# Read the canola variety trial
canola <- read_csv(
  "practice/data/canola_trial.csv",
  show_col_types = FALSE
)

# Calculate one mean and count per variety
variety_means <- canola |>
  group_by(variety) |>
  summarise(
    mean_yield = mean(yield_bu_acre),
    fields = n(),
    .groups = "drop"
  ) |>
  arrange(desc(mean_yield))

# Inspect the table that will drive the bars
variety_means
')
```

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Build a plot with `ggplot()`, `aes()` and a geom.
2. Reproduce the Excel bar chart, scatter plot and histogram.
3. Change a plot's title, axis labels, limits and tick marks.
4. Distinguish mapping an aesthetic to a variable from setting it to a constant.
5. Add value labels, a fitted line and a direct label.
6. Compare field-level distributions and save a finished plot.
:::

## The parts of a ggplot {#sec-grammar}

A ggplot is assembled from a data frame, aesthetic mappings and one or more layers. `ggplot(variety_means, aes(x = variety, y = mean_yield))` names the data and maps columns to horizontal and vertical position. `geom_col()` draws the bars.

Layers are joined with `+`. A pipe passes a data frame into a function; the `+` adds another layer or scale to the plot object.

## Build the bar chart {#sec-r-bar-example}

The Excel chart used a three-row summary calculated with `AVERAGEIF`. The `variety_means` data frame performs the same calculation in R.

```{r}
#| label: fig-r-variety-bars-basic
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.6
#| fig-height: 3.2
#| fig-cap: "Mean simulated field yield by canola variety."
#| fig-alt: "Three green bars show InVigor at 54.5, Clearfield at 53.5 and DEKALB at 52.2 bushels per acre."
# Put the highest mean first
variety_means |>
  mutate(
    variety = fct_reorder(
      variety,
      mean_yield,
      .desc = TRUE
    )
  ) |>
  # Map variety and mean yield to position
  ggplot(aes(x = variety, y = mean_yield)) +
  # Draw values that were already calculated
  geom_col(fill = "#3F6F52")
```

`geom_col()` uses the numeric values supplied to `y`. `geom_bar()` would count rows instead, which answers a different question. The `fill` is outside `aes()` because every bar receives the same colour.

InVigor has the highest mean at 54.5 bu/ac, followed by Clearfield at 53.5 and DEKALB at 52.2. The differences are small, and a bar of means does not show the variation among fields within each variety.

## Change the title and axes {#sec-r-title-axes}

`labs()` supplies the title and axis text. `scale_y_continuous()` controls the displayed range and tick marks. These are separate parts of the plot, so changing the scale does not change the data or mean calculations.

```{r}
#| label: fig-r-variety-bars-polished
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.8
#| fig-height: 3.5
#| fig-cap: "A finished bar chart with a title, labelled axis and values above the bars."
#| fig-alt: "A labelled bar chart shows InVigor at 54.5, Clearfield at 53.5 and DEKALB at 52.2 bushels per acre on an axis from zero to sixty."
# Store the finished plot so it can be saved later
bar_plot <- variety_means |>
  mutate(
    variety = fct_reorder(
      variety,
      mean_yield,
      .desc = TRUE
    )
  ) |>
  ggplot(aes(x = variety, y = mean_yield)) +
  # Set one colour for the whole series
  geom_col(
    width = 0.7,
    fill = "#3F6F52"
  ) +
  # Map formatted values to text labels
  geom_text(
    aes(label = sprintf("%.1f", mean_yield)),
    vjust = -0.35,
    size = 3.4
  ) +
  # Match the Excel bounds and tick spacing
  scale_y_continuous(
    limits = c(0, 60),
    breaks = seq(0, 60, by = 10),
    expand = expansion(mult = c(0, 0.04))
  ) +
  # Supply the title and axis text
  labs(
    title = "Average canola yield by variety",
    x = NULL,
    y = "Mean yield (bu/ac)"
  ) +
  # Use a restrained theme
  theme_minimal(base_size = 11) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor = element_blank()
  )

# Print the stored plot
bar_plot
```

`limits = c(0, 60)` gives the bars a zero baseline. `breaks = seq(0, 60, by = 10)` places the labelled ticks. `expand` removes unused space below zero and leaves a little room for the value labels.

The category names already explain the horizontal position, so `x = NULL` removes an unnecessary axis title. The unit remains on the vertical axis. `geom_text()` is useful here because there are only three values; the same treatment would make a chart with dozens of bars harder to read.

## Map or set an aesthetic {#sec-r-mapping-vs-setting}

Put an aesthetic inside `aes()` when it should depend on a column. Put it outside `aes()` when every mark should receive the same value.

```{r}
#| label: fig-r-bar-map-set
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.6
#| fig-height: 3.0
#| fig-subcap:
#|   - "A. Fill mapped to variety"
#|   - "B. One fill set for all bars"
#| fig-alt: "The left chart uses three colours and a redundant variety legend; the right chart uses one green colour and no legend."
#| layout-ncol: 2
# Map fill to the variety column
ggplot(
  variety_means,
  aes(x = variety, y = mean_yield, fill = variety)
) +
  geom_col() +
  scale_y_continuous(limits = c(0, 60)) +
  labs(x = NULL, y = "Mean yield (bu/ac)") +
  theme_minimal(base_size = 10)

# Set one fill for every bar
ggplot(
  variety_means,
  aes(x = variety, y = mean_yield)
) +
  geom_col(fill = "#3F6F52") +
  scale_y_continuous(limits = c(0, 60)) +
  labs(x = NULL, y = "Mean yield (bu/ac)") +
  theme_minimal(base_size = 10)
```

The mapped version creates a legend because colour now represents a column. That legend repeats the category labels already printed below the bars. The set version is closer to the Excel chart.

## Compare rainfall and yield {#sec-r-scatter-example}

A scatter plot assigns one numeric column to each position. Three fields have missing rainfall, so 117 points are drawn.

```{r}
#| eval: true
#| echo: false
console('
# Select the field with the highest yield
highest_field <- canola |>
  slice_max(
    yield_bu_acre,
    n = 1,
    with_ties = FALSE
  )

# Inspect the row that will receive a label
highest_field |>
  select(
    field_id,
    rainfall_mm,
    yield_bu_acre,
    variety
  )
')
```

```{r}
#| label: fig-r-rainfall-scatter
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.9
#| fig-height: 3.5
#| fig-cap: "Canola yield and growing-season rainfall with a fitted straight line."
#| fig-alt: "A scatter plot of 117 fields slopes upward; field C112 is labelled near 70 bushels per acre."
# Map rainfall and yield to position
ggplot(
  canola,
  aes(x = rainfall_mm, y = yield_bu_acre)
) +
  # Draw one point for each complete row
  geom_point(
    colour = "#2878A5",
    alpha = 0.6,
    na.rm = TRUE
  ) +
  # Add a fitted straight line and its interval
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    colour = "#2F5940",
    na.rm = TRUE
  ) +
  # Label only the highest-yield field
  geom_text(
    data = highest_field,
    aes(label = field_id),
    nudge_x = 8,
    nudge_y = 1,
    size = 3.2
  ) +
  # Set readable displayed ranges
  scale_x_continuous(
    limits = c(120, 335),
    breaks = seq(120, 320, by = 40)
  ) +
  scale_y_continuous(
    limits = c(35, 75),
    breaks = seq(40, 70, by = 10)
  ) +
  # Supply title, labels and units
  labs(
    title = "Canola yield by growing-season rainfall",
    x = "Growing-season rainfall (mm)",
    y = "Yield (bu/ac)"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title.position = "plot")
```

The correlation is about 0.65, which agrees with the upward slope. The fitted line summarizes the pattern but does not replace the field-level points. `geom_text()` labels C112 directly without creating a legend.

## Show the yield distribution {#sec-r-histogram-example}

`geom_histogram()` performs the binning shown in the Excel formula table. The settings below match the Excel boundaries: bins are 4 bu/ac wide and begin at 38.

```{r}
#| label: fig-r-canola-histogram
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.8
#| fig-height: 3.3
#| fig-cap: "Distribution of 120 simulated canola yields using 4 bu/ac bins."
#| fig-alt: "A histogram of 120 yields has its tallest bar between 50 and 54 bushels per acre."
# Map field yield to the horizontal axis
ggplot(canola, aes(x = yield_bu_acre)) +
  # Match the Excel width and boundary
  geom_histogram(
    binwidth = 4,
    boundary = 38,
    closed = "left",
    fill = "#D8A940",
    colour = "white"
  ) +
  # Show the bin boundaries
  scale_x_continuous(
    breaks = seq(38, 74, by = 4)
  ) +
  # Supply title, labels and units
  labs(
    title = "Canola trial yields (4 bu/ac bins)",
    x = "Yield (bu/ac)",
    y = "Fields"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor = element_blank()
  )
```

The nine bin counts add to 120. With a small dataset, a different bin width can change the visible shape, so the width is part of the chart definition rather than a cosmetic setting.

## Show what the means hide {#sec-r-variety-distributions}

The bar chart compares three means but hides the spread of the field yields. A box plot and jittered points show the observations behind each bar.

```{r}
#| label: fig-r-variety-box-jitter
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.8
#| fig-height: 3.4
#| fig-cap: "Field-yield distributions by canola variety."
#| fig-alt: "Three box plots with jittered field observations show substantial and similar spread within InVigor, Clearfield and DEKALB."
# Map variety and field yield to position
ggplot(
  canola,
  aes(x = variety, y = yield_bu_acre)
) +
  # Summarize each distribution
  geom_boxplot(
    width = 0.55,
    fill = "#E4EEE7",
    colour = "#2F5940",
    outlier.shape = NA
  ) +
  # Show the individual fields
  geom_jitter(
    width = 0.12,
    colour = "#2878A5",
    alpha = 0.55,
    size = 1.5
  ) +
  labs(
    title = "Yield variation within each canola variety",
    x = NULL,
    y = "Yield (bu/ac)"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title.position = "plot")
```

The within-variety ranges are much larger than the gaps among the three means. Use a bar chart when the intended comparison is the averages; use a box plot or field-level display when spread and consistency matter.

## Save the finished chart {#sec-saving-plots}

Assigning a plot to a name makes the object passed to `ggsave()` explicit. Width and height are in inches unless another unit is supplied.

```{r}
#| eval: true
#| echo: false
console('
# Create the output folder if needed
dir.create("output", showWarnings = FALSE)

# Save the finished bar chart
ggsave(
  "output/canola_variety_bar.png",
  plot = bar_plot,
  width = 6,
  height = 4,
  dpi = 300
)

# Confirm that the file was written
file.info(
  "output/canola_variety_bar.png"
)["size"]
')
```

PNG works for slides and web pages. PDF preserves vector lines and text for publication. Stating the plot object, dimensions and resolution in code makes the export reproducible.

::: {.callout-tip collapse="true"}
## Other resources -- ggplot2

- [ggplot2 documentation](https://ggplot2.tidyverse.org/)
- [Data Visualization chapter in R for Data Science](https://r4ds.hadley.nz/data-visualize)
:::
