# Merging Data {#sec-merging-data}

```{r}
#| include: false
#| eval: true
source("_console.R")
```

The data you need to answer one question often arrive in several tables. For example, suppose we wanted to understand the relationship between RM yields and precipitation. The yield data would come in one table, and the precipitation data would likely come in another. To answer our question, we would need to **merge** or **join** these tables.

Throughout this section, we will work with three datasets:

- [`rm_yields_2015_2024.csv`](practice/data/rm_yields_2015_2024.csv), with one row per RM per year. It contains 2,950 rows for ten years and six crops. Yields are measured in bushels per acre.
- [`station_precip.csv`](practice/data/station_precip.csv), with one row per weather station per year. It contains May to August precipitation for sixteen stations and a count of days with no reading.
- [`rm_lookup.csv`](practice/data/rm_lookup.csv), with one row per RM. It contains the RM name, census division, 2021 population and the closest of the sixteen weather stations.

We will do two merges:

1. Merge the RM yields with the RM lookup. This will give us a dataset containing the closest weather station for each yield observation.
2. Merge the dataset from step 1 with the weather station data. This will give us the growing-season precipitation for each observation.

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Identify the key that connects two tables.
2. Distinguish one-to-one, many-to-one and many-to-many relationships.
3. Merge tables in Excel with `XLOOKUP`, including a lookup that uses two columns.
4. Merge tables in R with `left_join()` and related functions.
5. Check for unmatched rows and unexpected changes in the row count.
:::


### Keys {.unnumbered}

A key is a column, or set of columns, used to identify and match rows across datasets. When we merge, we say that we merge on the key. The key must be present in both datasets.


In the three tables above, the keys for the two merges are:

1. RM yields to RM lookup: RM.
2. The resulting dataset to weather-station data: weather station and year.

### One-to-one, many-to-one and many-to-many {.unnumbered}

There are three types of merges:


![Three kinds of merge. Each line connects a grades row to the row it matches.](images/join_types.svg){#fig-join-types}

- **One-to-one (1:1).** Each key value appears once in both tables. Grades and attendance both have one row per student-class, so the merged table has the same number of rows as either input.
- **Many-to-one (n:1).** Values in the key column may repeat in one dataset but are unique in the other. In this example, students take many classes, so each student appears multiple times in the table on the left. Each class has a single professor, however, so each class appears only once in the table on the right.
- **Many-to-many (n:n).** Values in the key column repeat in both datasets. In the figure, students take multiple classes (the table on the left), and each class can have multiple teaching assistants (the table on the right).

Before joining, decide which relationship you expect. Then compare the row counts before and after.

Both of our merges will be many-to-one:

1. The RM yields dataset contains many observations for the same RM, but the RM lookup contains only one row for each RM.
2. The merged dataset from step 1 contains many observations with the same closest weather station and year, but the weather-station dataset contains only one observation for each station and year.

## Merging in Excel {#sec-merge-excel}

### A simple lookup {#sec-lookup .unnumbered}

Data often arrive split across tables. You might have yields by field in one table and crop prices in another but need both pieces of information for your analysis. A lookup lets us use information in one table to bring a corresponding value into another.

`XLOOKUP` lets us look up a value. It takes three main arguments: the value to find, the range to search and the range containing the value to return. For example:

```
=XLOOKUP(B11, $A$5:$A$7, $B$5:$B$7)
```

Excel searches `$A$5:$A$7` for the crop named in `B11`, finds its position and returns the value from the same position in `$B$5:$B$7`. Note the `$` signs: the crop being looked up changes from row to row, but the price table stays in place, so its ranges are locked.

If there are multiple matches—for example, if Canola appears twice in the price list—`XLOOKUP` returns the value associated with the first match. The value you look up should therefore identify a unique observation in the lookup table.

The workbook below has two tables: a short crop-price list at the top and the field data underneath. Column E uses `XLOOKUP` to find the price for each field's crop and bring it across, with the formula shown in column F.

::: {.workbook-callout}
::: {.excel-embed}
<iframe width="402" height="700" frameborder="0" scrolling="no" src="https://usaskca1-my.sharepoint.com/personal/pjs998_usask_ca/_layouts/15/Doc.aspx?sourcedoc=%7B1dc1b36e-393e-462b-b395-b6bd1fe4eae0%7D&action=embedview&wdAllowInteractivity=False&wdDownloadButton=True&wdInConfigurator=True"></iframe>
:::

[Download this workbook](textbook_examples/mod01_lookup.xlsx) · [blank version to fill in yourself](textbook_examples/mod01_lookup_practice.xlsx) · [Open online](https://usaskca1-my.sharepoint.com/personal/pjs998_usask_ca/_layouts/15/Doc.aspx?sourcedoc=%7B1dc1b36e-393e-462b-b395-b6bd1fe4eae0%7D&action=default) (File → Save a Copy to view formulas and edit)
:::

::: {.callout-tip collapse="true"}
## Other resources—lookup functions (VLOOKUP, XLOOKUP, INDEX/MATCH)

- **Excel for Dummies** -- *Microsoft 365 Excel for Dummies*, the chapter on lookup and reference functions.
- **Microsoft Support** -- [VLOOKUP](https://support.microsoft.com/en-us/office/vlookup-function-0bbc8083-26fe-4963-8ab8-93a18ad188a1), [XLOOKUP](https://support.microsoft.com/en-us/office/xlookup-function-b7fd680e-6d10-43e6-84f9-8818e00d0651), [INDEX](https://support.microsoft.com/en-us/office/index-function-a5dcf0dd-996d-40a4-a822-b56b061328bd), [MATCH](https://support.microsoft.com/en-us/office/match-function-e8dffd45-c762-47d6-bf89-533f4a37673a).
- **Video (XLOOKUP)** -- Leila Gharani, [*How an Excel pro uses XLOOKUP*](https://www.youtube.com/watch?v=xnLvEhXWSas) (XLOOKUP vs VLOOKUP and INDEX/MATCH).
- **Video (VLOOKUP for beginners)** -- Kevin Stratvert, [*VLOOKUP in Excel -- step-by-step tutorial*](https://www.youtube.com/watch?v=xIynD1gFOLo).
- **Video (INDEX/MATCH)** -- Leila Gharani, [*The definitive guide to INDEX and MATCH*](https://www.youtube.com/watch?v=lCKglBM2jyk).
:::

### Looking up values with two keys {.unnumbered}

Sometimes one column is not enough to identify a row uniquely. Suppose each row represents one rural municipality in one year. RM 1 appears once for every year in the dataset, so looking up RM 1 alone will return the first match. To identify a particular observation, we need both the RM and the year.

One way to do this in Excel is to combine the two values into a single key and use that key for the lookup:

```
=XLOOKUP(1 & "|" & 2023, B2:B10650 & "|" & A2:A10650, E2:E10650)
```

Excel effectively creates identifiers such as `1|2023`, `1|2024` and `2|2023`, then looks for the one we requested. The `|` is a separator between the two pieces of information. Without a separator, different combinations can accidentally produce the same identifier. For example, 1 and 2023 would become `12023`, but so would 12 and 023.

A lookup needs enough information to identify the observation you want uniquely. Sometimes one variable is enough; sometimes you need a combination of variables.

The **Merge** sheet in the workbook below contains thirty yield rows and the two lookup columns beside them. The `FORMULATEXT` column shows the precipitation formula, and the checks below the data count `#N/A` results.

The first lookup matches the RM number and returns its station. The second lookup needs station and year, so it joins the two values on each side with `& "|" &` to make a composite key. The shaded column deliberately leaves out year. It returns the first rain value for that station in the rain sheet—2015 here—for every year, without producing an error. These formulas require Microsoft 365's `XLOOKUP`, which University of Saskatchewan students can use through the university.

::: {.workbook-callout}
<!-- Public view-only link expires 2026-11-20 under the University of Saskatchewan 90-day policy. -->

::: {.excel-embed}
<iframe width="402" height="700" frameborder="0" scrolling="no" src="https://usaskca1-my.sharepoint.com/personal/pjs998_usask_ca/_layouts/15/Doc.aspx?sourcedoc=%7BE4286D57-D90E-4E60-9E9A-C493B5099E8E%7D&action=embedview&wdAllowInteractivity=False&wdDownloadButton=True&wdInConfigurator=True&ActiveCell=%27Merge%27!A1"></iframe>
:::

[Download this workbook](textbook_examples/mod03_merge_excel.xlsx) · [Open online](https://usaskca1-my.sharepoint.com/:x:/g/personal/pjs998_usask_ca/IQBXbSjkDtlgTp6axJO1CZ6OAa9jBZn7UhVyZRjoYXeXy3Q?e=AcDC7T) (File → Save a Copy to view formulas and edit)
:::

An unmatched `XLOOKUP` returns `#N/A`, which is easy to count. An incomplete key can be harder to find because the formula returns a value. Before a lookup, confirm that the key on the lookup side is unique: `XLOOKUP` silently returns the first duplicate. Check a few rows against the source tables, especially where the year or another part of the key changes. Excel lookups ignore case but not stray spaces, and the number `1` does not match the text `"1"`; if an entire column returns `#N/A`, check the key types and use `TRIM()` to remove extra spaces.

For a merge that must be repeated, Power Query records the source, key columns and join type. Import each table with **Data > Get Data > From Text/CSV > Transform Data**. In the editor choose **Home > Merge Queries**, select the key columns in the same order on both sides using Ctrl-click (Command-click on a Mac), and keep **Left Outer** as the join kind. The dialog reports how many rows match. After clicking **OK**, expand the new Table column with the double-arrow icon in its header. Power Query also offers Inner, Full Outer and Left Anti joins, corresponding to the R joins below.

## Merging in R {#sec-merge-r}

Read the three files and inspect their columns:

```{r}
#| eval: true
#| echo: false
console('
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

# Read the three tables and clean their column names
yields <- read_csv(
  "practice/data/rm_yields_2015_2024.csv",
  show_col_types = FALSE
) |> clean_names()
rms <- read_csv(
  "practice/data/rm_lookup.csv",
  show_col_types = FALSE
) |> clean_names()
rain <- read_csv(
  "practice/data/station_precip.csv",
  show_col_types = FALSE
) |> clean_names()

glimpse(yields)
glimpse(rms)
glimpse(rain)
')
```

`left_join()` takes two data frames and a `by` argument that identifies the key. If a key has different names in the two tables, write `join_by(left_name == right_name)`; a column with the same name on both sides is written once. The older `by = c("left_name" = "right_name")` form still appears in other people's code.

### Four joins {.unnumbered}

The join functions differ in which rows they keep:

- **`left_join(a, b)`** keeps every row of `a`. New columns contain `NA` where `b` has no match.
- **`inner_join(a, b)`** keeps only rows that matched in both tables.
- **`full_join(a, b)`** keeps every row from both tables, with `NA` where either side has no match.
- **`anti_join(a, b)`** returns the rows of `a` that have no match in `b`. It is useful for finding failed matches before a join.

We will use `left_join()` because the yield rows are the observations we want to keep.

Before joining, check that the columns expected to identify one row are actually unique:

```{r}
#| eval: true
#| echo: false
console('
rms |>
  count(rm_number) |>
  filter(n > 1)

rain |>
  count(station, year) |>
  filter(n > 1)
')
```

Both checks return empty tibbles. A key is only a key if each combination occurs no more than once on the side where uniqueness is expected.

### First join: yields and the RM lookup {.unnumbered}

The RM column is called `rm` in the yields table and `rm_number` in the lookup. Check for unmatched RMs, perform the join and compare the row counts:

```{r}
#| eval: true
#| echo: false
console('
# List RM numbers that have no lookup row
yields |>
  anti_join(rms, by = join_by(rm == rm_number)) |>
  distinct(rm)

# Add the RM information and nearest station
with_station <- yields |>
  left_join(rms,
            by = join_by(rm == rm_number),
            relationship = "many-to-one")

nrow(yields)
nrow(with_station)
')
```

The `anti_join()` returns an empty tibble, so every RM has a match. `yields` has 2,950 rows and so does `with_station`, which is what a many-to-one join should give. The `relationship = "many-to-one"` argument records what we expect and makes `left_join()` stop if the lookup contains more than one row for an RM.

In R, text keys are sensitive to both case and spaces. For example, `"canola"`, `"Canola"` and `"Canola "` are three different keys. These differences have to be reconciled before the join (@sec-categories).

### A failed match on purpose {.unnumbered}

The complete rain table matches every station-year. To see how the joins differ when data are missing, temporarily remove 2021:

```{r}
#| eval: true
#| echo: false
console('
rain_short <- rain |>
  filter(year != 2021)

# Which yield rows no longer have rain?
with_station |>
  anti_join(
    rain_short,
    by = join_by(weather_station == station, year)
  ) |>
  count(weather_station)

# A left join keeps all 2,950 yield rows,
# including 295 rows with missing rain.
with_station |>
  left_join(
    rain_short,
    by = join_by(weather_station == station, year)
  ) |>
  filter(is.na(precip_may_aug_mm)) |>
  nrow()

# An inner join drops those 295 rows.
with_station |>
  inner_join(
    rain_short,
    by = join_by(weather_station == station, year)
  ) |>
  nrow()

# A full join also keeps a right-side row
# that has no yield-side match.
rain_extra <- rain_short |>
  add_row(
    station = "Course example",
    year = 2025,
    precip_may_aug_mm = 200,
    days_missing = 0
  )

with_station |>
  full_join(
    rain_extra,
    by = join_by(weather_station == station, year)
  ) |>
  filter(weather_station == "Course example") |>
  select(year, weather_station, rm, canola)
')
```

`anti_join()` identifies the failed keys, `left_join()` makes the missing matches visible as `NA`, and `inner_join()` silently removes them. The final `full_join()` also retains the deliberately added rain row that has no yield-side match, filling its yield columns with `NA`.

### Second join: add the rain {.unnumbered}

The second key has two parts. The station column is called `weather_station` on the left and `station` on the right; `year` has the same name in both tables.

```{r}
#| eval: true
#| echo: false
console('
# Find station-years with no precipitation row
with_station |>
  anti_join(rain,
            by = join_by(weather_station == station, year)) |>
  distinct(weather_station, year)

# Add precipitation for the matching station and year
merged <- with_station |>
  left_join(rain,
            by = join_by(weather_station == station, year),
            relationship = "many-to-one")

nrow(merged)
')
```

Again there are no unmatched keys and the row count remains 2,950.

### The row-count trap {.unnumbered}

Now leave `year` out of the key:

```{r}
#| eval: true
#| echo: false
console('
# This key is incomplete: every row matches ten years of rain
with_station |>
  left_join(rain,
            by = join_by(weather_station == station)) |>
  nrow()

# State the expected relationship to turn
# the warning into an error.
try(
  with_station |>
    left_join(
      rain,
      by = join_by(weather_station == station),
      relationship = "many-to-one"
    )
)
')
```

Each yield row matches all ten years for its station, so 2,950 rows become 29,500. By default, `dplyr` warns about the unexpected many-to-many relationship. When `relationship = "many-to-one"` states what was intended, the same mistake becomes an error and the analysis stops. Excel's station-only `XLOOKUP` makes the mistake differently: it returns only the first matching year and gives no warning.

## What the Merge Gives You {#sec-merge-result}

The merged table has one row per RM-year, with yield and precipitation beside one another:

```{r}
#| eval: true
#| echo: false
console('
# Display the columns used in the analysis
merged |>
  select(year, rm, rm_name, canola,
         weather_station, precip_may_aug_mm)

# Calculate annual mean yield and precipitation
merged |>
  filter(!is.na(canola)) |>
  summarise(
    canola_bu_ac = mean(canola),
    rain_mm = mean(precip_may_aug_mm),
    max_days_missing = max(days_missing),
    .by = year
  )

# Count yield values excluded from those means
merged |>
  summarise(missing_canola = sum(is.na(canola)))

# Describe association, not causation
merged |>
  summarise(
    yield_rain_correlation = cor(
      canola,
      precip_may_aug_mm,
      use = "complete.obs"
    )
  )
')
```

Across RM-years, wetter observations tend to have higher canola yield, but the relationship is loose: the correlation is about 0.41. The annual table also resists a simple rainfall story. Both 2021 and 2024 have low average yields despite ordinary or above-average rainfall. These data cannot say why. Testing whether heat contributed would require temperature data; explaining the remaining differences would require other variables as well.

The annual means exclude 46 RM-years with no published canola yield. They also average rainfall over RM rows, so a station serving more RMs receives more weight than a station serving fewer RMs. The merge has further limits that should be recorded in the project's README. The median RM is 62 km from its assigned station, one quarter are more than 85 km away, and the maximum is 161 km, so station rainfall is a coarse proxy for rainfall across an RM. Some station-years also have many missing daily observations—up to 25 days in this file—and their seasonal totals sum only the available days, which biases those totals downward.
