# Graphing in R {#sec-graphing-r}

```{r}
#| include: false
#| eval: true
source("_console.R")
```

Excel starts with a selected range and a chart menu. In R, we describe the chart in code: the data, the variables assigned to visual properties, and the marks to draw. The `ggplot2` package turns that description into a figure.

The examples use [`field_yields.csv`](practice/data/field_yields.csv), with one row for each of sixty hypothetical canola fields.

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse and read the field data
library(tidyverse)

yields <- read_csv("data/field_yields.csv")
glimpse(yields)
')
```

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Build a plot with `ggplot()`, `aes()` and a geom.
2. Choose a geom for a comparison, trend, relationship or distribution.
3. Distinguish mapping an aesthetic to a variable from setting it to a constant.
4. Add titles, axis labels and a theme.
5. Divide a plot into small multiples with facets.
6. save a plot with `ggsave()`.
:::

## The Grammar of Graphics {#sec-grammar}

`ggplot2` follows the **grammar of graphics**. A plot is assembled from a few parts:

- **Data:** the data frame to plot.
- **Aesthetic mappings:** columns assigned to position, colour, size or shape.
- **Geometries:** points, lines, bars, boxes and other marks.
- **Scales:** the rules connecting data values to axis positions or colours.
- **Facets:** separate panels for groups.

Some geoms also calculate a statistic. A histogram bins values before drawing bars, and a smooth fits a line or curve before drawing it.

## A First Plot {#sec-first-ggplot}

This scatter plot places field size on the horizontal axis and yield on the vertical axis:

```{r}
#| label: fig-r-first-scatter
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.4
#| fig-height: 3.2
#| fig-cap: "Field size and canola yield for sixty hypothetical fields."
ggplot(yields, aes(x = acres, y = yield_bu_acre)) +
  geom_point(na.rm = TRUE)
```

Read the code in three parts:

- `ggplot(yields, ...)` names the data frame.
- `aes(x = acres, y = yield_bu_acre)` maps two columns to position.
- `geom_point()` draws one point for each row with both values present.

Layers are joined with `+`. A pipe passes a data object into a function; the `+` adds a layer to a ggplot object.

## Common Geoms {#sec-geoms}

### `geom_point()`: scatter plots {.unnumbered}

Use points for the relationship between two numeric variables. The `alpha` argument makes overlapping points partly transparent.

```{r}
#| label: fig-r-alpha
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.4
#| fig-height: 3.2
#| fig-cap: "Transparency makes overlapping observations easier to see."
ggplot(yields, aes(x = acres, y = yield_bu_acre)) +
  geom_point(alpha = 0.55, na.rm = TRUE)
```

### `geom_line()`: change over time {.unnumbered}

A line chart needs one value for each point along the horizontal axis. Our file has many fields in each year, so calculate the annual means first:

```{r}
#| label: fig-r-line
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.4
#| fig-height: 3.0
#| fig-cap: "Mean field yield in 2024 and 2025."
annual_mean <- yields |>
  group_by(year) |>
  summarise(mean_yield = mean(yield_bu_acre, na.rm = TRUE))

ggplot(annual_mean, aes(x = year, y = mean_yield)) +
  geom_line() +
  geom_point()
```

The summary creates one row per year. `geom_line()` then connects those two rows in year order.

### `geom_col()` and `geom_bar()`: bars {.unnumbered}

Use `geom_col()` when the bar heights have already been calculated.

```{r}
#| label: fig-r-col
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.4
#| fig-height: 3.0
#| fig-cap: "Mean field yield by region."
region_means <- yields |>
  group_by(region) |>
  summarise(mean_yield = mean(yield_bu_acre, na.rm = TRUE))

ggplot(region_means, aes(x = region, y = mean_yield)) +
  geom_col()
```

Use `geom_bar()` when the desired value is the number of rows. It counts the observations in each category:

```{r}
#| label: fig-r-bar
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.0
#| fig-height: 2.8
#| fig-cap: "Number of fields in each region."
ggplot(yields, aes(x = region)) +
  geom_bar()
```

### `geom_histogram()` and `geom_boxplot()`: distributions {.unnumbered}

A histogram shows the shape of one numeric variable. State the number of bins or the bin width rather than accepting the default without looking at it.

```{r}
#| label: fig-r-hist
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.2
#| fig-height: 3.0
#| fig-cap: "Distribution of recorded field yields."
ggplot(yields, aes(x = yield_bu_acre)) +
  geom_histogram(bins = 12, na.rm = TRUE)
```

A box plot makes distributions compact enough to compare across categories.

```{r}
#| label: fig-r-box
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.2
#| fig-height: 3.0
#| fig-cap: "Field-yield distributions by region."
ggplot(yields, aes(x = region, y = yield_bu_acre)) +
  geom_boxplot(na.rm = TRUE)
```

### `geom_smooth()`: fitted patterns {.unnumbered}

`geom_smooth(method = "lm")` adds a fitted straight line and an interval around it:

```{r}
#| label: fig-r-smooth
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.4
#| fig-height: 3.2
#| fig-cap: "Field size and yield with a fitted straight line."
ggplot(yields, aes(x = acres, y = yield_bu_acre)) +
  geom_point(alpha = 0.45, na.rm = TRUE) +
  geom_smooth(method = "lm", na.rm = TRUE)
```

The line summarizes the relationship present in these observations. Module 7 explains what the fitted line estimates and how to interpret it.

## Mapping and Setting Aesthetics {#sec-mapping-vs-setting}

Put an aesthetic inside `aes()` when it should depend on a variable. Put it outside `aes()` when every mark should receive the same value.

```{r}
#| label: fig-r-map-set
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 4.7
#| fig-height: 3.0
#| fig-subcap:
#|   - "A. Colour mapped to region"
#|   - "B. One colour set for all points"
#| layout-ncol: 2
# Map colour to a column in the data
ggplot(yields, aes(x = acres, y = yield_bu_acre, colour = region)) +
  geom_point(na.rm = TRUE)

# Set one colour for every point
ggplot(yields, aes(x = acres, y = yield_bu_acre)) +
  geom_point(colour = "steelblue", na.rm = TRUE)
```

Putting `colour = "steelblue"` inside `aes()` would ask ggplot to map the literal category `"steelblue"` to a colour scale, producing an unnecessary legend.

## Labels and Themes {#sec-ggplot-polish}

`labs()` supplies the text and `theme_minimal()` changes the non-data elements of the plot:

```{r}
#| label: fig-r-polished
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.0
#| fig-height: 3.4
#| fig-cap: "A labelled scatter plot with region mapped to colour."
ggplot(yields, aes(x = acres, y = yield_bu_acre, colour = region)) +
  geom_point(alpha = 0.65, na.rm = TRUE) +
  labs(
    title = "Canola yield by field size",
    subtitle = "Sixty hypothetical Saskatchewan fields, 2024--2025",
    x = "Field size (acres)",
    y = "Yield (bu/acre)",
    colour = "Region",
    caption = "Source: synthetic AREC 261 data"
  ) +
  theme_minimal()
```

Other built-in themes include `theme_classic()` and `theme_bw()`. A theme changes the surrounding presentation; it does not change the data or mappings.

## Faceting {#sec-faceting}

Faceting divides one plot into panels. Each panel below uses the same histogram definition for a different region:

```{r}
#| label: fig-r-facet
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.2
#| fig-height: 3.2
#| fig-cap: "One yield histogram for each region."
ggplot(yields, aes(x = yield_bu_acre)) +
  geom_histogram(bins = 8, na.rm = TRUE) +
  facet_wrap(~ region) +
  labs(x = "Yield (bu/acre)", y = "Fields")
```

`facet_grid(region ~ variety)` creates a two-dimensional grid. Use it only when the resulting number of panels remains readable.

## Saving Plots {#sec-saving-plots}

Save a ggplot object before passing it to `ggsave()`:

```{r}
#| eval: true
#| echo: false
console('
# Build and save the regional box plot
region_plot <- ggplot(
  yields,
  aes(x = region, y = yield_bu_acre)
) +
  geom_boxplot(na.rm = TRUE) +
  labs(x = NULL, y = "Yield (bu/acre)")

ggsave(
  "output/region_boxplot.png",
  plot = region_plot,
  width = 6,
  height = 4,
  dpi = 300
)
')
```

Width and height are in inches unless another unit is supplied. PNG works for slides and web pages; PDF or SVG preserves vector lines and text for publication.

::: {.callout-tip collapse="true"}
## Other resources -- ggplot2

- **R for Data Science (2nd ed.)** -- [Chapter 1, "Data visualization"](https://r4ds.hadley.nz/data-visualize) introduces `ggplot()`, aesthetic mappings, geoms and facets.
- **ggplot2 documentation** -- the [function reference](https://ggplot2.tidyverse.org/reference/) lists the arguments and examples for each geom.
- **RStudio cheatsheet** -- [Data visualization with ggplot2](https://rstudio.github.io/cheatsheets/html/data-visualization.html).
:::
