# Graphing with AI {#sec-graphing-ai}

```{r}
#| include: false
#| eval: true
source("_console.R")
```

Graphing is a good place to begin using AI assistance. `ggplot2` is a great package, but details such as annotations, direct labels and axis breaks require arguments that are easy to forget and somewhat difficult to implement -- for example, getting text in just the right place can take trial and error. An assistant can draft that syntax quickly.

It is also fairly easy to verify the work an AI assistant does on a chart.  If you make the raw chart yourself, then you only need to verify that the aesthetics of the graph have been updated correctly, which can be seen visually as opposed to through the code.

These exercises use R because ggplot2 code is text that an assistant can draft and you can inspect. Copilot inside Excel is a separately licensed product and is not unlocked by GitHub Copilot. A browser assistant can suggest Excel steps if you provide headers and made-up rows, but you must carry out and verify the clicks yourself.

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Ask an AI assistant for `ggplot2` code using the actual structure of a dataset.
2. Check the data supplied to an AI-produced chart.
3. Check the marks, labels, groups and scales drawn by the chart.
4. Recognize when repeated prompting is no longer improving the result.
:::

## Set up the readme file {.unnumbered}
Our first thing to do is to establish a readme file that the assistant can refer to.  This should establish the structure of the project.  We don't need to be too detailed about the data we are using, because the AI itself can continue to build this readme file.  

The common file format is a .md file.  You can write this as essentially a plain text file, denoting sections with hashtags.  For example, my first README.md is below  

```markdown
# AREC 261 project

I am a student in AREC 261 Agricultural Data Analytics.  In this class, we
will work with several agricultural datasets to learn concepts about data
analysis, data visualization, and data communication.

## Tools

We will be using R for data analysis.

When you write R code ensure that you are using the tidyverse package and
that your code is well-commented so that I can learn.

## Datasets

We will continually re-use datasets.  So when we first encounter a dataset,
update this document with a summary of the data set and what data it
includes.
```

That's it.  Save it as `README.md` at the top of your project folder, beside `data/`, `code/` and `output/`.

Now we can get the assistant to read our variety data and summarize it for us.  With the readme in place, the request can be short, because the standing instructions are already written down:

::: {.ai-prompt}
Read data/sask_variety_yields.csv.  Summarize its structure -- the columns, what one row represents, the years covered, and anything I should know about missing values -- and add that summary to the Datasets section of README.md.
:::

The assistant reads the file, writes a summary along these lines into the README, and every later conversation that reads it starts already knowing the dataset:

```markdown
### sask_variety_yields.csv

Saskatchewan crop insurance records with one row per risk zone, crop,
variety and year, 2021-2025.  Columns: Risk_Zone, Crop, Variety, Year,
Acres, Yield.  13,625 rows; 4,186 have no recorded yield, mostly because
the variety was not grown in that zone and year.
```

Before trusting the summary, check it the same way you would check any other AI output.  The row and missing-value counts it wrote down are exactly what `glimpse()` and a `summarise()` by year report, which is where the next section starts.

## A Chart Does Not Check Its Data {#sec-visual-limits}

A chart can show whether the drawing is wrong. It cannot show whether the wrong rows were supplied to it.

A clean chart of 2024 is still wrong when the question asked about 2025. Code that removes every row with missing acreage may also create a convincing figure while changing the population being described.

Check both sides of the plot:

1. Before plotting, inspect the number of rows, groups, missing values and numerical ranges.
2. After plotting, inspect the marks, labels, axes and panels.

The synthetic teaching file contains one row for a variety in a risk zone and year. Its rectangular layout includes combinations with no acreage and yield, often because that variety is absent from that zone-year. Consequently, comparisons across years also compare changing sets of varieties. The file has 13,625 rows, including 4,186 with no recorded yield. Inspect the counts before drawing the figure:

```{r}
#| eval: true
#| echo: false
console('
suppressPackageStartupMessages(library(tidyverse))

d <- read_csv(
  "practice/data/sask_variety_yields.csv",
  show_col_types = FALSE
)

glimpse(d)

d |>
  summarise(
    rows = n(),
    missing_yield = sum(is.na(Yield)),
    .by = Year
  )
')
```

```{r}
#| label: fig-missing
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| fig-width: 5.0
#| fig-height: 2.3
#| fig-cap: "Rows with and without a recorded yield, by year."
source("R/ai_failures.R")
fig_missing_by_year()
```

About half the 2021 rows have no yield, while almost all 2025 rows have one. A trend therefore reflects turnover in the varieties represented as well as changes in yield. In Excel, a PivotTable with Year in Rows and both Count of Variety and Count of Yield in Values provides the same check.

## The Exercises {#sec-ai-graph-exercises}

All five exercises use [`sask_variety_yields.csv`](practice/data/sask_variety_yields.csv). Its columns are `Risk_Zone`, `Crop`, `Variety`, `Year`, `Acres` and `Yield`.

Work in Positron with an assistant connected. For each exercise, describe the result, read the proposed code, run it and complete the checks. When the draft is wrong, tell the assistant what is wrong rather than asking it to produce the same chart again.

### 1. Annotate one point {.unnumbered}

Plot the unweighted mean yield by year for `Crop == "Wheat - Hard Red Spring"`. Add a note with an arrow pointing to 2021 and identify it as a drought year; that label comes from outside weather knowledge, while the file itself shows only a low-yield year.

**Checks:** Does the arrow point to 2021? Is the note inside the plotting area? Which `Crop` value did the code filter? How many recorded yields contribute to each annual point? Is the mean weighted or unweighted, and is that what you intended?

### 2. Replace a legend with direct labels {.unnumbered}

Plot average yield by year for `Wheat - Hard Red Spring`, `Canola/Rapeseed`, `Barley` and `Oats`. Put each crop name beside the right end of its line and remove the legend.

**Checks:** Are all four labels present and beside the correct line? Has a label been clipped at the panel edge? Can the series still be distinguished when printed in grey?

### 3. Shade a range behind a line {.unnumbered}

Plot average `Wheat - Hard Red Spring` yield by year and add a `geom_ribbon()` band showing the range across risk-zone means. Summarize first by zone-year and then by year; draw the ribbon before the line so it sits behind it.

**Checks:** Does the band contain the mean line in every year? Is it drawn behind the line? Do the axis limits include it? How many recorded yields sit behind each zone-year mean?

### 4. Build small multiples {.unnumbered}

Find the six crops with the greatest total acreage. Make one panel per crop showing average yield over time.

**Checks:** Print the acreage ranking used by the code. Were missing acres excluded or treated as zero? Are any selected categories sub-classes of another, such as Lentils and Lentils - Red? Are there exactly six panels? Decide whether the default shared scale or `scales = "free_y"` better answers the question.

### 5. Rebuild a chart from Module 4 {.unnumbered}

Choose a figure from @sec-chart-types and reproduce it using only the description in the text. Do not open `R/chart_types.R` until the new version is finished. Then compare the two scripts and figures.

**Checks:** Compare the pre-plot summary, aesthetic mappings and geoms; ignore differences that are only theme or colour. Does each figure make the same comparison? Is either version clearer at the size used in the book?

## A Useful Prompt Structure {#sec-ai-graph-prompt}

A graphing request should state four things:

- the data frame and relevant columns;
- the rows or groups to include;
- the calculation required before plotting;
- the chart form and the specific design change needed.

For example:

::: {.ai-prompt}
The data frame d has columns Crop, Year, Risk_Zone, Acres and Yield. Keep recorded rows for Wheat - Hard Red Spring, Canola/Rapeseed, Barley and Oats. Calculate unweighted mean Yield by Year and Crop, then plot one line per crop. Put each crop name at the right end of its line and remove the legend. Show the summary table before the plot code.
:::

Asking for the summary table is useful. It separates the calculation from the drawing and gives you something numerical to inspect. For a workbook prompt, paste the header and several made-up rows and state which fields are categories—`Risk_Zone` is an identifier, not a quantity.

## When the Assistant Gets Stuck {#sec-ai-stuck}

An assistant may repeat the same failed fix after each error report. Repeating the same symptom can return the same approach.

Try one of these instead:

- **Change the description.** Restate the desired result rather than repeating the symptom.
- **Ask for an explanation.** Understanding the error may reveal a one-line correction.
- **Reduce the problem.** Remove extra layers until the basic plot works, then add them back one at a time.
- **Start a fresh chat.** A conversation built around a mistaken approach can keep returning to it.

When those fail, use the package documentation or a worked example. The assistant is one route to the syntax, not the authority on it.

::: {.callout-tip collapse="true"}
## Other resources -- ggplot2

- **ggplot2: Elegant Graphics for Data Analysis**, [ggplot2-book.org](https://ggplot2-book.org/) -- an explanation of the grammar and its components.
- **R Graph Gallery**, [r-graph-gallery.com](https://r-graph-gallery.com/) -- examples organized by the kind of chart being built.
- **ggplot2 reference**, [ggplot2.tidyverse.org/reference](https://ggplot2.tidyverse.org/reference/) -- function arguments and current syntax.
:::
