# Graphing in R {#sec-graphing-r}

```{r}
#| include: false
#| eval: true
source("_console.R")
```

The Excel chapter built four charts from worksheet ranges. In R we describe the same charts in code: the data, the variables assigned to visual properties and the marks to draw. The `ggplot2` package turns that description into a figure.

The examples use [`field_yields.csv`](practice/data/field_yields.csv) and [`rm_yields_2015_2024.csv`](practice/data/rm_yields_2015_2024.csv), the same files used in the Excel workbook. The field data are simulated; the RM yield series comes from the Saskatchewan Agriculture provincial dashboard.

```{r}
#| eval: true
#| echo: false
console('
# Load the tidyverse
suppressPackageStartupMessages(library(tidyverse))

# Read the simulated field data
yields <- read_csv(
  "practice/data/field_yields.csv",
  show_col_types = FALSE
)

# Read the published RM yield data
rm_yields <- read_csv(
  "practice/data/rm_yields_2015_2024.csv",
  show_col_types = FALSE
)

# Calculate one regional mean per row
region_means <- yields |>
  group_by(region) |>
  summarise(
    mean_yield = mean(yield_bu_acre, na.rm = TRUE),
    fields = sum(!is.na(yield_bu_acre)),
    .groups = "drop"
  )

# Calculate one annual mean per row
annual_mean <- rm_yields |>
  group_by(Year) |>
  summarise(
    mean_yield = mean(Canola, na.rm = TRUE),
    .groups = "drop"
  ) |>
  rename(year = Year)

# Inspect both chart-driving summaries
region_means
annual_mean
')
```

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Build a plot with `ggplot()`, `aes()` and a geom.
2. Reproduce a bar, line, scatter and histogram chart from the Excel examples.
3. Distinguish mapping an aesthetic to a variable from setting it to a constant.
4. Add titles, axis labels, direct labels and annotations.
5. Add a fitted pattern and divide a plot into facets.
6. Save a plot with stated dimensions and resolution.
:::

## The grammar of graphics {#sec-grammar}

`ggplot2` follows the **grammar of graphics**. A plot is assembled from a few parts:

- **Data:** the data frame to plot.
- **Aesthetic mappings:** columns assigned to position, colour, size or shape.
- **Geometries:** points, lines, bars, boxes and other marks.
- **Scales:** the rules connecting data values to axis positions or colours.
- **Facets:** separate panels for groups.

Some geoms also calculate a statistic. A histogram bins values before drawing bars, and a smooth fits a line or curve before drawing it.

Layers are joined with `+`. A pipe passes a data object into a function; the `+` adds another layer to a ggplot object.

## Compare regional means with bars {#sec-r-bar-example}

The Excel chart used a three-row summary calculated with `AVERAGEIF`. The `region_means` data frame performs the same job in R. `geom_col()` draws values that have already been calculated.

```{r}
#| label: fig-r-region-bars
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.5
#| fig-height: 3.2
#| fig-cap: "Mean simulated field yield by region."
#| fig-alt: "Three green bars show Central at 47.6, North at 47.3 and South at 42.7 bushels per acre."
# Put the highest mean first
region_means |>
  mutate(region = fct_reorder(region, mean_yield, .desc = TRUE)) |>
  ggplot(aes(x = region, y = mean_yield)) +
  # Draw the calculated means as bars
  geom_col(width = 0.7, fill = "#3F6F52") +
  # Keep a zero baseline for bar length
  scale_y_continuous(
    limits = c(0, 55),
    expand = expansion(mult = c(0, 0.04))
  ) +
  # State the variables and unit
  labs(
    x = NULL,
    y = "Mean yield (bu/ac)"
  ) +
  # Remove unnecessary decoration
  theme_minimal(base_size = 11)
```

Central and North differ by 0.3 bu/ac. The bar chart shows the difference, but it does not show the variation among the nineteen recorded fields behind each mean.

## Show change through time with a line {#sec-r-line-example}

The annual summary contains one mean for each year from 2015 to 2024. `geom_line()` connects those values in chronological order and `geom_point()` shows the observed annual means.

```{r}
#| label: fig-r-annual-line
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.0
#| fig-height: 3.2
#| fig-cap: "Mean published RM canola yield, 2015--2024."
#| fig-alt: "A line chart of ten annual means reaches its lowest point, 21.9 bushels per acre, in 2021."
# Start with one row per year
ggplot(annual_mean, aes(x = year, y = mean_yield)) +
  # Connect the years in time order
  geom_line(colour = "#2878A5", linewidth = 0.7) +
  # Mark each annual mean
  geom_point(colour = "#2878A5", size = 2) +
  # Show each integer year
  scale_x_continuous(breaks = 2015:2024) +
  # Use the same displayed range as the Excel example
  scale_y_continuous(limits = c(15, 45)) +
  # State the unit
  labs(
    x = NULL,
    y = "Mean yield (bu/ac)"
  ) +
  # Use a restrained theme
  theme_minimal(base_size = 11)
```

The lowest annual mean is 21.9 bu/ac in 2021. As in Excel, blank RM yields are omitted from each annual mean rather than treated as zero.

## Compare two numeric variables with points {#sec-r-scatter-example}

A scatter plot assigns one numeric column to each position. Rows missing either value do not produce a point.

```{r}
#| label: fig-r-field-scatter
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.7
#| fig-height: 3.3
#| fig-cap: "Field size and yield for sixty simulated fields."
#| fig-alt: "Fifty-seven points form a broad cloud with little upward or downward slope."
# Map field size and yield to position
ggplot(yields, aes(x = acres, y = yield_bu_acre)) +
  # Draw one point for each complete row
  geom_point(
    colour = "#2878A5",
    alpha = 0.65,
    na.rm = TRUE
  ) +
  # State both units
  labs(
    x = "Field size (acres)",
    y = "Yield (bu/ac)"
  ) +
  # Keep both horizontal and vertical guides
  theme_minimal(base_size = 11)
```

Three rows have missing yield, so the chart contains 57 points. The correlation is about -0.07, which agrees with the nearly flat cloud.

## Show a distribution with a histogram {#sec-r-histogram-example}

`geom_histogram()` performs the binning that the Excel workbook displayed in a formula-driven table. The settings below match the Excel boundaries: bins are 3 bu/ac wide and begin at 30.

```{r}
#| label: fig-r-yield-histogram
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.8
#| fig-height: 3.2
#| fig-cap: "Distribution of recorded simulated field yields using 3 bu/ac bins."
#| fig-alt: "A histogram of 57 yields has its tallest bin from 48 to 51 bushels per acre."
# Map the recorded yields to the horizontal axis
ggplot(yields, aes(x = yield_bu_acre)) +
  # Match the Excel bin width and boundary
  geom_histogram(
    binwidth = 3,
    boundary = 30,
    closed = "left",
    fill = "#D8A940",
    colour = "white",
    na.rm = TRUE
  ) +
  # Show the bin boundaries
  scale_x_continuous(breaks = seq(30, 63, by = 3)) +
  # State the unit and count
  labs(
    x = "Yield (bu/ac)",
    y = "Fields"
  ) +
  # Remove unnecessary decoration
  theme_minimal(base_size = 11)
```

With only 57 recorded yields, the apparent shape changes noticeably when the bin width changes. The bin definition is therefore part of the analysis rather than a formatting detail.

## Mapping and setting aesthetics {#sec-mapping-vs-setting}

Put an aesthetic inside `aes()` when it should depend on a column. Put it outside `aes()` when every mark should receive the same value.

```{r}
#| label: fig-r-map-set
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.6
#| fig-height: 3.0
#| fig-subcap:
#|   - "A. Colour mapped to region"
#|   - "B. One colour set for every point"
#| fig-alt: "The left scatter plot uses three colours for regions; the right uses one blue colour for all fields."
#| layout-ncol: 2
# Map colour to the region column
ggplot(
  yields,
  aes(x = acres, y = yield_bu_acre, colour = region)
) +
  geom_point(alpha = 0.65, na.rm = TRUE) +
  labs(x = "Field size (acres)", y = "Yield (bu/ac)") +
  theme_minimal(base_size = 10)

# Set one colour for all points
ggplot(yields, aes(x = acres, y = yield_bu_acre)) +
  geom_point(
    colour = "#2878A5",
    alpha = 0.65,
    na.rm = TRUE
  ) +
  labs(x = "Field size (acres)", y = "Yield (bu/ac)") +
  theme_minimal(base_size = 10)
```

Putting `colour = "#2878A5"` inside `aes()` would ask ggplot to treat the colour code as a data value and would create an unnecessary legend. In Excel, colouring points by region requires a separate series for each region.

## Add a fitted pattern and direct label {#sec-r-direct-labels}

Text can identify a small number of observations without forcing the reader to search a legend or table. The highest recorded field yield belongs to field F044.

```{r}
#| eval: true
#| echo: false
console('
# Find the field with the highest recorded yield
highest_field <- yields |>
  filter(!is.na(yield_bu_acre)) |>
  slice_max(yield_bu_acre, n = 1, with_ties = FALSE)

# Inspect the row that will receive a label
highest_field |>
  select(field_id, acres, yield_bu_acre)
')
```

```{r}
#| label: fig-r-labelled-smooth
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.8
#| fig-height: 3.4
#| fig-cap: "Field size and yield with a fitted line and the highest-yield field labelled."
#| fig-alt: "A nearly flat fitted line crosses a scatter plot; field F044 is labelled near 61 bushels per acre."
# Draw the complete field records
ggplot(yields, aes(x = acres, y = yield_bu_acre)) +
  geom_point(alpha = 0.45, na.rm = TRUE) +
  # Fit a straight line with its confidence band
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    colour = "#2878A5",
    na.rm = TRUE
  ) +
  # Label only the selected field
  geom_text(
    data = highest_field,
    aes(label = field_id),
    nudge_x = 12,
    nudge_y = 0.5,
    hjust = 0,
    size = 3.2
  ) +
  # Leave room for the direct label
  scale_x_continuous(limits = c(0, 520)) +
  labs(
    x = "Field size (acres)",
    y = "Yield (bu/ac)"
  ) +
  theme_minimal(base_size = 11)
```

The fitted line is nearly flat and its interval is wide. `geom_smooth()` will draw a line whether or not the relationship is useful, so the points and uncertainty remain part of the interpretation.

## Put explanatory text inside a chart {#sec-r-annotations}

`geom_text()` attaches labels to rows in a data frame. `annotate()` places text at coordinates supplied in the plot code. The line chart below uses both.

```{r}
#| eval: true
#| echo: false
console('
# Select the lowest annual mean for a data-driven label
lowest_year <- annual_mean |>
  slice_min(mean_yield, n = 1, with_ties = FALSE) |>
  mutate(label = paste0(year, ": ", round(mean_yield, 1), " bu/ac"))

# Inspect the selected label row
lowest_year
')
```

```{r}
#| label: fig-r-annotated-line
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.0
#| fig-height: 3.4
#| fig-cap: "Annual mean canola yield with a direct label and a short note inside the plotting area."
#| fig-alt: "The 2021 low point is labelled 2021: 21.9 bu/ac, and a note states that blank RM yields were omitted."
# Draw the annual series
ggplot(annual_mean, aes(x = year, y = mean_yield)) +
  geom_line(colour = "#2878A5", linewidth = 0.7) +
  geom_point(colour = "#2878A5", size = 2) +
  # Attach the value to the selected row
  geom_text(
    data = lowest_year,
    aes(label = label),
    nudge_y = -1.6,
    size = 3.2
  ) +
  # Add a note that is not stored in the data
  annotate(
    "text",
    x = 2015,
    y = 44,
    label = "Blank RM yields omitted",
    hjust = 0,
    size = 3.2,
    colour = "#5F6E66"
  ) +
  scale_x_continuous(breaks = 2015:2024) +
  scale_y_continuous(limits = c(15, 45)) +
  labs(x = NULL, y = "Mean yield (bu/ac)") +
  theme_minimal(base_size = 11)
```

Use text for a small number of points or one necessary qualification. Labelling every point would make this ten-year series harder to read.

## Divide a comparison into facets {#sec-faceting}

Faceting applies the same chart definition to subsets of the data. Each panel below uses the same 3 bu/ac bins and axis limits for a different region.

```{r}
#| label: fig-r-facet
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7.0
#| fig-height: 3.2
#| fig-cap: "Recorded field-yield distributions shown separately by region."
#| fig-alt: "Three small histograms use common bins and scales for Central, North and South."
# Define one histogram
ggplot(yields, aes(x = yield_bu_acre)) +
  geom_histogram(
    binwidth = 3,
    boundary = 30,
    closed = "left",
    fill = "#D8A940",
    colour = "white",
    na.rm = TRUE
  ) +
  # Repeat it for each region
  facet_wrap(~ region, nrow = 1) +
  scale_x_continuous(breaks = seq(30, 60, by = 10)) +
  labs(x = "Yield (bu/ac)", y = "Fields") +
  theme_minimal(base_size = 10)
```

Excel has no direct faceting feature. Separate charts can make the same comparison, but their bin definitions and axis bounds must be kept identical.

## Save a plot {#sec-saving-plots}

Assigning a plot to a name before `ggsave()` makes the saved object explicit and lets it be reused. Create the destination folder first.

```{r}
#| eval: true
#| echo: false
console('
# Create the output folder if needed
dir.create("output", showWarnings = FALSE)

# Store the finished plot in an object
regional_plot <- region_means |>
  mutate(region = fct_reorder(region, mean_yield, .desc = TRUE)) |>
  ggplot(aes(x = region, y = mean_yield)) +
  geom_col(width = 0.7, fill = "#3F6F52") +
  scale_y_continuous(
    limits = c(0, 55),
    expand = expansion(mult = c(0, 0.04))
  ) +
  labs(x = NULL, y = "Mean yield (bu/ac)") +
  theme_minimal(base_size = 11)

# Save the stated object at a reproducible size
ggsave(
  "output/region_yield_bar.png",
  plot = regional_plot,
  width = 6,
  height = 4,
  dpi = 300
)
')
```

Width and height are in inches unless another unit is supplied. PNG works for slides and web pages; PDF preserves vector lines and text for publication. SVG output requires the `svglite` package.

::: {.callout-tip collapse="true"}
## Other resources -- ggplot2

- **R for Data Science (2nd ed.)** -- [Chapter 1, "Data visualization"](https://r4ds.hadley.nz/data-visualize) introduces `ggplot()`, aesthetic mappings, geoms and facets.
- **ggplot2 documentation** -- the [function reference](https://ggplot2.tidyverse.org/reference/) lists the arguments and examples for each geom.
- **RStudio cheatsheet** -- [Data visualization with ggplot2](https://rstudio.github.io/cheatsheets/html/data-visualization.html).
:::
