# Cleaning Data {#sec-cleaning}

```{r}
#| include: false
#| eval: true
source("_console.R")
```

A file that opens without an error may still contain duplicate rows, inconsistent names, mixed units and missing values recorded as ordinary text. Cleaning starts by finding these problems. The decisions that follow depend on what the values mean and where the file came from.

The examples use [`grain_deliveries_messy.csv`](practice/data/grain_deliveries_messy.csv), a small file of grain delivery tickets. Read it without letting `read_csv()` trim spaces so that we can see the spaces in some crop names:

```{r}
#| eval: true
#| echo: false
console('
# Load the packages used in this chapter
library(tidyverse)
library(janitor)

# Read the file and identify common missing-value codes
deliveries <- read_csv(
  "data/grain_deliveries_messy.csv",
  na = c("", "NA", "N/A", "-", "."),
  trim_ws = FALSE
)

glimpse(deliveries)
')
```

The output already shows two problems. `weight_tonnes` was read as text because one value contains a unit, and `delivery_date` is text because the file mixes date formats.

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Explain what tidy data means and reshape a dataset with `pivot_longer()` and `pivot_wider()`.
2. Clean column names with `rename()` and `janitor::clean_names()`.
3. Find exact and conflicting duplicates.
4. Find impossible values, inconsistent categories and mixed units.
5. Recognize common codes for missing values and decide when observations can be dropped.
6. Identify dates that were imported as text.
7. Reshape and clean a table in Excel with Power Query.
:::

## Tidy Data {#sec-tidy-data}

The tidyverse is organized around **tidy data** [@r4ds]. A dataset is tidy when:

1. Every variable has its own column.
2. Every observation has its own row.
3. Every value has its own cell.

Suppose a table records yield for three crops in two regions. In a tidy version the columns are `region`, `crop` and `yield`, and each row is one region-crop combination. In a wide version the columns are `region`, `Canola`, `Wheat` and `Barley`. The crop names are values of the variable `crop`, but they have been stored as column names.

The wide table may be easier to read. The long, tidy table is usually easier to analyze because `crop` is a column that can be filtered, grouped and plotted. This is the wide-versus-long distinction from @sec-wide-long.

### Recognizing untidy data {.unnumbered}

Common signs include:

- **Values stored in column names.** Columns named `2024`, `2025` and `2026` usually contain values of a variable called `year`.
- **Several variables stored in one column.** A value such as `"South_InVigor"` combines a region and a variety.
- **One observation spread over several rows.** Price in one row and quantity in another row for the same market and year prevents either row from describing the observation on its own.

When you open a table, write down what one row represents and what each column represents. This usually makes the shape problem visible.

::: {.callout-tip collapse="true"}
## Other resources -- tidy data

- **R for Data Science (2nd ed.)** -- [Chapter 5, "Data tidying"](https://r4ds.hadley.nz/data-tidy) develops the tidy-data idea and shows how to reshape datasets between long and wide forms.
- **Video** -- Riffomonas Project, [*Reshaping data to be long or wide with pivot_longer and pivot_wider*](https://www.youtube.com/watch?v=ZVddiwrbrr4).
:::

## Tidying and Cleaning in Excel {#sec-cleaning-excel}

Excel offers three approaches. Formulas can clean a file once while leaving the raw columns visible. Power Query can reshape the table and records each cleaning step. Find & Replace changes cells directly and leaves little evidence of what was changed.

### Reshaping with Power Query {#sec-unpivot}

The RM yields file (@sec-direct-download) is wide, with one column for each crop. Power Query calls the wide-to-long operation **Unpivot**. Import the CSV with **Data > Get Data > From Text/CSV**, choose **Transform Data**, select `Year` and `RM`, and then choose **Transform > Unpivot Columns > Unpivot Other Columns**. Rename the two new columns `crop` and `yield_bu_ac` before loading the result.

The original 2,950 RM-year rows become 17,700 RM-year-crop rows. The reverse operation is **Pivot Column** in Power Query or a PivotTable on a worksheet.

### Cleaning with formulas {.unnumbered}

The workbook below keeps the raw delivery data in columns A through F and places cleaning formulas beside it. The formulas standardize crop names, rescale moisture values entered as decimals, count repeated ticket numbers and flag impossible weights. `FORMULATEXT` cells show the formulas, and summary checks appear below the table.

::: {.workbook-callout}
<!-- Public view-only link expires 2026-11-20 under the University of Saskatchewan 90-day policy. -->

::: {.excel-embed}
<iframe width="402" height="700" frameborder="0" scrolling="no" src="https://usaskca1-my.sharepoint.com/personal/pjs998_usask_ca/_layouts/15/Doc.aspx?sourcedoc=%7B6F136BC8-96DB-499F-AE4B-A04CFDF86DED%7D&action=embedview&wdAllowInteractivity=False&wdDownloadButton=True&wdInConfigurator=True&ActiveCell=%27Cleaning%27!A1"></iframe>
:::

[Download this workbook](textbook_examples/mod03_cleaning_excel.xlsx) · [Open online](https://usaskca1-my.sharepoint.com/:x:/g/personal/pjs998_usask_ca/IQDIaxNv25afSa5LoEz9-G3tAXoBxYJcoWjP4fMOie4JVrQ?e=PsWLYM) (File → Save a Copy to view formulas and edit)
:::

The filter dropdown is useful when checking categories because it lists each distinct value. `MIN`, `MAX`, `COUNTBLANK` and a count of repeated identifiers cover many of the numerical checks. Keep the raw columns beside the cleaned ones so that a changed value can be traced back.

### Cleaning with Power Query {.unnumbered}

Power Query has recorded steps for the same work:

- **Remove Top Rows** and **Use First Row as Headers** deal with files that begin with a title or notes.
- The type icon in a column header sets a column to Text, Whole Number, Decimal or Date. A value that cannot be converted becomes an error that can be inspected.
- **Transform > Format > Trim** removes extra spaces, and **Capitalize Each Word** standardizes capitalization.
- **Replace Values** handles synonyms and missing-value codes. Power Query stores a missing value as `null`.
- **Remove Duplicates** drops exact duplicates. To find a repeated identifier with different values, use **Group By** to count rows for each identifier and filter to counts above one.
- **Split Column > By Delimiter** separates values that were stored together.
- **Fill > Down** copies a category into blank cells below it when a spreadsheet records the label only once for a group of rows.

Each operation appears in the **Applied Steps** list. Clicking a step shows the table at that point. When a new version of the source file arrives, **Refresh** runs the steps again.

## Tidying and Cleaning in R {#sec-cleaning-r}

### Reshaping with `pivot_longer()` and `pivot_wider()` {#sec-reshaping}

`pivot_longer()` does the same job as Unpivot. It takes several columns, puts their names in one column and puts their values in another. Here is the RM yields table:

```{r}
#| eval: true
#| echo: false
console('
# Read the wide table and clean its column names
yields_wide <- read_csv("data/rm_yields_2015_2024.csv") |>
  clean_names()

yields_wide

# Put crop names and crop yields into columns
yields_long <- yields_wide |>
  pivot_longer(
    cols = spring_wheat:peas,
    names_to = "crop",
    values_to = "yield_bu_ac"
  )

yields_long
nrow(yields_long)
')
```

The column selection `spring_wheat:peas` means every column from `spring_wheat` through `peas`. The result has one row per RM-year-crop and six times as many rows as the wide table.

`pivot_wider()` goes in the other direction. `names_from` supplies the new column names and `values_from` supplies the values placed under them:

```{r}
#| eval: true
#| echo: false
console('
# Return the crop column to six separate crop columns
yields_long |>
  pivot_wider(
    names_from = crop,
    values_from = yield_bu_ac
  )
')
```

Analysis often uses the long form, while a report may use the wide form.

### Awkward Column Names {#sec-clean-names}

Column names such as `Yield (bu/ac)`, `REGION_NAME` and `Field ID Number` are cumbersome in code. This book uses **snake case**: lower-case words separated by underscores, such as `yield_bu_ac`, `region_name` and `field_id`.

Use `rename()` when only a few names need changing. The new name goes on the left and the old name on the right:

```{r}
#| eval: true
#| echo: false
console('
# Make a small table with awkward names
awkward_names <- tibble(
  `Weight (t)` = c(42.1, 37.8),
  CROP_NAME = c("Canola", "Wheat")
)

awkward_names |>
  rename(
    weight_tonnes = `Weight (t)`,
    crop = CROP_NAME
  )
')
```

`janitor::clean_names()` converts every name in a data frame to snake case. The package is installed once with `install.packages("janitor")`, then loaded in each new R session with `library(janitor)`.

```{r}
#| eval: true
#| echo: false
console('
# Clean every column name in one call
awkward_names |>
  clean_names() |>
  names()

deliveries <- deliveries |> clean_names()
')
```

### Text in a Numeric Column {#sec-text-numbers}

The first `glimpse()` reported `weight_tonnes` as `<chr>`, or character text. One row contains `"38.2 t"`, so `read_csv()` could not treat the whole column as numeric. Find the values containing something other than a number, decimal point or minus sign:

```{r}
#| eval: true
#| echo: false
console('
# Find weights containing a unit or other text
deliveries |>
  filter(str_detect(weight_tonnes, "[^0-9.-]")) |>
  select(ticket_id, weight_tonnes)
')
```

The file documentation says every weight is in tonnes, so the suffix can be removed. `parse_number()` keeps the numeric part:

```{r}
#| eval: true
#| echo: false
console('
# Convert the weight column after checking its units
deliveries <- deliveries |>
  mutate(weight_tonnes = parse_number(weight_tonnes))

glimpse(deliveries)
')
```

If the suffix indicated a different unit, removing it would give the wrong answer. Convert the value to the common unit before discarding the label.

### Duplicates {#sec-duplicates}

An **exact duplicate** repeats the entire row. `duplicated()` finds the later copy, while `distinct()` keeps one copy of every unique row:

```{r}
#| eval: true
#| echo: false
console('
# Show exact duplicates
deliveries |>
  filter(duplicated(deliveries))

nrow(deliveries)

# Keep one copy of each unique row
deliveries <- deliveries |> distinct()
nrow(deliveries)
')
```

A **conflicting duplicate** repeats an identifier but not the whole row. Ticket `T0030` appears with two different weights, so `distinct()` keeps both. Count the identifier to find it:

```{r}
#| eval: true
#| echo: false
console('
# Find ticket numbers that still occur more than once
deliveries |>
  count(ticket_id) |>
  filter(n > 1)

# Inspect every version of the repeated ticket
deliveries |>
  filter(ticket_id == "T0030")
')
```

The data alone does not say which row is correct. Check the source record. If that is impossible, report how the conflict was handled.

### Impossible Values {#sec-impossible-values}

Every numeric variable has a possible range. A delivery weight cannot be negative and a moisture percentage cannot be 250. `summary()` is a quick way to inspect the minimum and maximum:

```{r}
#| eval: true
#| echo: false
console('
# Inspect the ranges of the numeric measurements
deliveries |>
  select(weight_tonnes, moisture_pct) |>
  summary()

# Display values outside their possible ranges
deliveries |>
  filter(weight_tonnes <= 0 | moisture_pct > 100) |>
  select(ticket_id, weight_tonnes, moisture_pct)
')
```

A negative weight may be a sign error or a reversed transaction. A moisture value of 250 may have lost its decimal point. Those are possible explanations, not facts. Correct a value only when the source supports the correction.

An extreme value can still be valid. A canola yield of 95 bu/ac is unusual, but it may belong to a very good field. Check it against the source and, if uncertainty remains, compare the analysis with and without it.

### Inconsistent Categories {#sec-categories}

Several spellings of the same category produce separate groups. Count each categorical column before using it in `group_by()`:

```{r}
#| eval: true
#| echo: false
console('
# Count the crop names as they arrived
deliveries |>
  count(crop, sort = TRUE)
')
```

Capitalization and spaces can be standardized with string functions:

```{r}
#| eval: true
#| echo: false
console('
# Remove extra spaces and use consistent capitalization
deliveries <- deliveries |>
  mutate(crop = str_squish(crop),
         crop = str_to_title(crop))

deliveries |>
  count(crop, sort = TRUE)
')
```

Synonyms require an explicit decision. Put that decision in the script:

```{r}
#| eval: true
#| echo: false
console('
# Reconcile known synonyms while leaving other values unchanged
deliveries <- deliveries |>
  mutate(crop = case_when(
    crop == "Canola/Rapeseed" ~ "Canola",
    crop == "Soybean" ~ "Soybeans",
    TRUE ~ crop
  ))
')
```

### Mixed Units {#sec-mixed-units}

Most moisture readings in the file are percentages near 8 to 15. Three were entered as decimal proportions, such as `0.099` rather than `9.9`. The values are numeric, so they produce no parsing error.

```{r}
#| eval: true
#| echo: false
console('
# Find moisture readings recorded as decimal proportions
deliveries |>
  filter(moisture_pct < 1) |>
  select(ticket_id, moisture_pct)

# Convert decimal proportions to percentages
deliveries <- deliveries |>
  mutate(
    moisture_pct = if_else(
      moisture_pct < 1,
      moisture_pct * 100,
      moisture_pct
    )
  )
')
```

This rule depends on subject knowledge: a grain moisture value below 1 percent is not plausible here. If the file contains a unit column, use that column in `case_when()` instead of inferring the unit from the size of the value.

### Missing Values {#sec-missing-values}

R represents a missing value as `NA`. Most calculations involving `NA` return `NA`. The canola trial data has three fields with no rainfall reading:

```{r}
#| eval: true
#| echo: false
console('
# Read a file containing three missing rainfall values
trial_data <- read_csv("data/canola_trial.csv")

mean(trial_data$rainfall_mm)
mean(trial_data$rainfall_mm, na.rm = TRUE)
')
```

The first call returns `NA` because three rainfall values are unknown. The second calculates the mean from the observed values. The argument `na.rm = TRUE` makes that decision visible. Many Excel functions skip blank cells automatically, so the missing values need to be counted separately there.

#### Missing values in disguise {.unnumbered}

A blank cell, `"NA"`, `"N/A"`, `"-"`, `"."` and `-999` may all mean that no value was recorded. The `na` argument in the first `read_csv()` call told R which text codes to treat as missing.

Count missing values in every column before deciding what to do with them:

```{r}
#| eval: true
#| echo: false
console('
# Count missing values in each column
deliveries |>
  summarise(across(everything(), ~ sum(is.na(.))))
')
```

A numeric code such as `-999` is more dangerous. R reads it as a number and includes it in calculations. `summary()` exposes it as an implausible minimum unless it is converted to `NA` when the file is read.

#### Working with missing values {.unnumbered}

`is.na()` identifies missing values and `drop_na()` removes rows that contain them:

```{r}
#| eval: true
#| echo: false
console('
# Count missing weights
sum(is.na(deliveries$weight_tonnes))

# Remove rows missing any variable
deliveries |> drop_na()

# Remove rows missing selected variables
deliveries |> drop_na(weight_tonnes, moisture_pct)

# Calculate a mean from the observed moisture readings
mean(deliveries$moisture_pct, na.rm = TRUE)
')
```

Dropping every incomplete row can remove far more data than dropping rows missing the variables used in a particular calculation.

#### When is dropping safe? {.unnumbered}

The answer depends on why the values are missing. If a rain gauge failed at arbitrary times, the observed days may still represent the season, although there are fewer of them. If low-yielding farms were less likely to report, the remaining records will overstate average yield. Record how many values were missing, which observations were excluded and why the exclusion seemed reasonable.

### Dates Stored as Text {#sec-dates-text}

`glimpse()` reported `delivery_date` as `<chr>` because most rows use `2025-09-07` while two use day/month/year. Text dates sort alphabetically and cannot be used in date arithmetic.

Parse the expected format and inspect the rows that fail:

```{r}
#| eval: true
#| echo: false
console('
# Parse dates written as year-month-day
deliveries_with_dates <- deliveries |>
  mutate(date_parsed = as.Date(
    delivery_date,
    format = "%Y-%m-%d"
  ))

# Display dates written in a different format
deliveries_with_dates |>
  filter(is.na(date_parsed)) |>
  select(ticket_id, delivery_date)
')
```

Those rows need a second stated format or a date parser that accepts several formats. For this course, the main check is in `glimpse()`: a date column should be `<date>`, not `<chr>` (@sec-col-types).

## A Checklist {#sec-cleaning-checklist}

Use these checks on a new file:

1. Use `glimpse()` to check the dimensions, column names and types.
2. Write down what one row represents and whether the table is tidy.
3. Use `summary()` to inspect the ranges of numeric variables.
4. Use `count()` on categorical variables and look for spelling and spacing differences.
5. Check for exact duplicates and repeated identifiers.
6. Count `NA`s in every column and look for other missing-value codes.
7. Check that units are consistent and dates were read as dates.
8. Record every correction, exclusion and assumption in the script or Power Query steps.
