# Merging Data {#sec-merging-data}

```{r}
#| include: false
#| eval: true
source("_console.R")
```

Data for one question often arrives in several tables. The Saskatchewan yield data from @sec-direct-download identifies rural municipalities and years. The weather data identifies stations and years. A third table says which weather station is closest to each RM.

A **merge**, also called a **join**, matches rows that refer to the same thing and puts their columns together. In this chapter we will merge the three tables so that every RM-year has its crop yields and May to August precipitation in one row.

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Identify the key that connects two tables.
2. Distinguish one-to-one, many-to-one and many-to-many relationships.
3. Merge tables in Excel with `XLOOKUP`, including a lookup that uses two columns.
4. Merge tables in R with `left_join()` and related functions.
5. Check for unmatched rows and unexpected changes in the row count.
:::

## The Three Tables {#sec-three-tables}

The files are:

- [`rm_yields_2015_2024.csv`](practice/data/rm_yields_2015_2024.csv), with one row per RM per year. It contains 2,950 rows for ten years and six crops. Yields are measured in bushels per acre.
- [`station_precip.csv`](practice/data/station_precip.csv), with one row per weather station per year. It contains May to August precipitation for sixteen stations and a count of days with no reading.
- [`rm_lookup.csv`](practice/data/rm_lookup.csv), with one row per RM. It contains the RM name, census division, 2021 population and the closest of the sixteen weather stations.

The yields table and rain table have no column in common. The lookup table connects them. We first attach a weather station to each yield row, then attach the rain recorded at that station in that year.

### Keys {.unnumbered}

A **key** is a column, or a set of columns, used to identify and match rows. The key follows from what one row represents:

- A row in the yields table is one RM in one year, so its key is `RM` and `Year` together.
- A row in the lookup table is one RM, so its key is `RM_Number`.
- A row in the rain table is one station in one year, so its key is `Station` and `Year` together.

The first merge uses RM number. The second uses station and year. Station alone is not enough because each station appears once in every year.

### One-to-one, many-to-one and many-to-many {.unnumbered}

The number of rows sharing a key determines what a join returns. @fig-join-types uses a table of student grades to show the three cases.

![Three kinds of merge. Each line connects a grades row to the row it matches. Attendance is one-to-one: each student-class appears once on both sides. Professors is many-to-one: several grades rows point to the same professor. Teaching assistants is many-to-many: each class has two TAs, so each grades row matches twice and the result has eight rows instead of four.](images/join_types.svg){#fig-join-types}

- **One-to-one (1:1).** Each key appears once in both tables. Grades and attendance both have one row per student-class, so the merged table has the same number of rows as either input.
- **Many-to-one (n:1).** The key repeats in one table and is unique in the other. Several grade rows can share one professor. In our data, each RM has ten yield rows but only one lookup row. A many-to-one join should have the same number of rows as the table on the many side.
- **Many-to-many (n:n).** The key repeats in both tables. Each row on one side matches every row with the same key on the other. If a class has two TAs, each grade row appears twice. If we join yields to rain by station but leave out year, each yield row matches ten years of rain.

Before joining, decide which relationship you expect. Then compare the row counts before and after.

## Merging in Excel {#sec-merge-excel}

In Excel, `XLOOKUP` can bring a column from one table into another (@sec-lookup). The **Merge** sheet in the workbook below contains thirty yield rows and the two lookup columns beside them. The `FORMULATEXT` column shows the precipitation formula, and the checks below the data count `#N/A` results.

The first lookup matches the RM number and returns its station. The second lookup needs station and year, so it joins the two values on each side with `& "|" &` to make a composite key. The shaded column deliberately leaves out year. It returns the first rain value for the station in every year, without producing an error.

::: {.workbook-callout}
<!-- Public view-only link expires 2026-11-20 under the University of Saskatchewan 90-day policy. -->

::: {.excel-embed}
<iframe width="402" height="700" frameborder="0" scrolling="no" src="https://usaskca1-my.sharepoint.com/personal/pjs998_usask_ca/_layouts/15/Doc.aspx?sourcedoc=%7BE4286D57-D90E-4E60-9E9A-C493B5099E8E%7D&action=embedview&wdAllowInteractivity=False&wdDownloadButton=True&wdInConfigurator=True&ActiveCell=%27Merge%27!A1"></iframe>
:::

[Download this workbook](textbook_examples/mod03_merge_excel.xlsx) · [Open online](https://usaskca1-my.sharepoint.com/:x:/g/personal/pjs998_usask_ca/IQBXbSjkDtlgTp6axJO1CZ6OAa9jBZn7UhVyZRjoYXeXy3Q?e=AcDC7T) (File → Save a Copy to view formulas and edit)
:::

An unmatched `XLOOKUP` returns `#N/A`, which is easy to count. An incomplete key can be harder to find because the formula returns a value. Check a few rows against the source tables, especially rows where the year or another part of the key changes.

For a merge that must be repeated, Power Query records the source, the key columns and the join type. Use **Data > Get Data** to load the tables, then **Merge Queries** to select the matching columns. A two-column key is selected in the same order in both tables.

## Merging in R {#sec-merge-r}

Read the three files and inspect their columns:

```{r}
#| eval: true
#| echo: false
console('
# Load the packages used in this chapter
library(tidyverse)
library(janitor)

# Read the three tables and clean their column names
yields <- read_csv("data/rm_yields_2015_2024.csv") |> clean_names()
rms <- read_csv("data/rm_lookup.csv") |> clean_names()
rain <- read_csv("data/station_precip.csv") |> clean_names()

glimpse(yields)
glimpse(rms)
glimpse(rain)
')
```

`left_join()` takes two data frames and a `by` argument that identifies the key. If the key has different names in the two tables, write `by = c("left_name" = "right_name")`.

### Four joins {.unnumbered}

The join functions differ in which rows they keep:

- **`left_join(a, b)`** keeps every row of `a`. New columns contain `NA` where `b` has no match.
- **`inner_join(a, b)`** keeps only rows that matched in both tables.
- **`full_join(a, b)`** keeps every row from both tables, with `NA` where either side has no match.
- **`anti_join(a, b)`** returns the rows of `a` that have no match in `b`. It is useful for finding failed matches before a join.

We will use `left_join()` because the yield rows are the observations we want to keep.

### First join: yields and the RM lookup {.unnumbered}

The RM column is called `rm` in the yields table and `rm_number` in the lookup. Check for unmatched RMs, perform the join and compare the row counts:

```{r}
#| eval: true
#| echo: false
console('
# List RM numbers that have no lookup row
yields |>
  anti_join(rms, by = c("rm" = "rm_number")) |>
  distinct(rm)

# Add the RM information and nearest station
with_station <- yields |>
  left_join(rms,
            by = c("rm" = "rm_number"),
            relationship = "many-to-one")

nrow(yields)
nrow(with_station)
')
```

The `anti_join()` returns no rows, so every RM has a match. Both tables have 2,950 rows after the join. The `relationship = "many-to-one"` argument records what we expect and makes `left_join()` stop if the lookup contains more than one row for an RM.

Text keys must match exactly. For example, `"Canola/Rapeseed"` does not match `"Canola"`, and `"Soybean"` does not match `"Soybeans"`. These differences have to be reconciled before the join (@sec-categories).

### Second join: add the rain {.unnumbered}

The second key has two parts. The station column is called `weather_station` on the left and `station` on the right; `year` has the same name in both tables.

```{r}
#| eval: true
#| echo: false
console('
# Find station-years with no precipitation row
with_station |>
  anti_join(rain,
            by = c("weather_station" = "station", "year")) |>
  distinct(weather_station, year)

# Add precipitation for the matching station and year
merged <- with_station |>
  left_join(rain,
            by = c("weather_station" = "station", "year"),
            relationship = "many-to-one")

nrow(merged)
')
```

Again there are no unmatched keys and the row count remains 2,950.

### The row-count trap {.unnumbered}

Now leave `year` out of the key:

```{r}
#| eval: true
#| echo: false
console('
# This key is incomplete: every row matches ten years of rain
with_station |>
  left_join(rain,
            by = c("weather_station" = "station")) |>
  nrow()
')
```

Each yield row matches all ten years for its station, so 2,950 rows become 29,500. `dplyr` warns about the unexpected many-to-many relationship. Excel's station-only `XLOOKUP` makes the same mistake differently: it returns only the first matching year and gives no warning.

## What the Merge Gives You {#sec-merge-result}

The merged table has one row per RM-year, with yield and precipitation beside one another:

```{r}
#| eval: true
#| echo: false
console('
# Display the columns used in the analysis
merged |>
  select(year, rm, rm_name, canola,
         weather_station, precip_may_aug_mm)

# Calculate annual mean yield and precipitation
merged |>
  filter(!is.na(canola)) |>
  group_by(year) |>
  summarise(canola_bu_ac = mean(canola),
            rain_mm = mean(precip_may_aug_mm))
')
```

Canola yield is highest in wet 2016 and among its lowest values in dry 2017. The pattern is not exact. In 2021 rainfall was ordinary but yields fell sharply, because heat also mattered.

The merge also has limits that should be recorded in the project's README. Some RMs are far from the nearest station in this small set. RM 1, Argyle, is matched to Kipling, 145 km away. Some station-years are missing several weeks of daily readings, which is why the rain table includes `days_missing`.
