# Graphing in R {#sec-graphing-r}

```{r}
#| include: false
#| eval: true
source("_console.R")
# Chunks in this chapter run with practice/ as the working directory, so the
# student-facing paths (data/..., output/...) resolve at render time.
knitr::opts_knit$set(root.dir = "practice")
```

The Excel chapter built a bar chart of barley acreage, a pie chart, and line charts of acreage over time. This chapter makes the same charts with `ggplot2`. Excel starts with a selected range and a chart menu. In R, we describe the chart in code: the data, the variables assigned to visual properties, and the marks to draw. The `ggplot2` package turns that description into a figure.

The data are the same two files the Excel workbooks used. [`sask_barley_2025.csv`](practice/data/sask_barley_2025.csv) has one row per barley variety with its 2025 acres and average yield. [`sask_variety_yields.csv`](practice/data/sask_variety_yields.csv) has Saskatchewan Crop Insurance Corporation records by risk zone, crop, variety and year, 2021 to 2025 [@scic_smp]; suppressed observations have missing acres and yield.

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Build a plot with `ggplot()`, `aes()` and a geom.
2. Reproduce the Excel bar, pie and line charts in code.
3. Filter and summarise the data that feeds a chart, instead of editing it on a sheet.
4. Order bars with `fct_reorder()` and group small categories into an "Other" category.
5. Change a plot's title, labels, colours and theme.
6. Plot several series by mapping a variable to colour.
7. Save a finished plot with `ggsave()`.
:::

## Introduction to ggplot2 {#sec-ggplot-intro}

`ggplot2` loads with the tidyverse, and every chart it draws is built from the same three parts:

- the **data**: a data frame with the data to be plotted;
- the **mapping**: `aes()`, which assigns variables to visual properties -- horizontal position, vertical position, colour;
- a **geom**: the mark to draw -- `geom_col()` for bars, `geom_line()` for lines, `geom_point()` for points.

The template is always:

```r
data |>
  ggplot(aes(x = ..., y = ...)) +
  geom_something() + ...
```

We can then add other options to the graph with `+`.  For example, we can change the formatting, font size, or axis labels.  

If we want an analogy to Excel, we can think of `data` as the spreadsheet containing our data, `aes()` as choosing which columns to plot on the x- and y-axes, and the `geom` as choosing the chart type. Other formatting options are then added using the + operator.

## Bar charts {#sec-r-bar-example}

Let's again start with a simple example using the barley variety data for 2025.  When we read the file in, we also convert the column names to lowercase with `clean_names()` (@sec-clean-names), so they are easier to type.  We can then plot the acres of all the varieties with the following code:

```{r}
#| label: fig-r-bar-first
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.4
#| fig-cap: "The first attempt: every barley variety."
#| fig-alt: "A vertical bar chart with 32 overlapping variety names on the horizontal axis."
# Load the tidyverse and janitor
library(tidyverse)
library(janitor)

# One row per barley variety: 2025 acres and average yield
barley_2025 <- read_csv(
  "data/sask_barley_2025.csv",
  show_col_types = FALSE
) |>
  clean_names()          # column names to lowercase

## Show the first few rows of data
head(barley_2025)

## Create a bar chart
barley_2025 |>
  ggplot(aes(x = variety, y = acres)) +
  geom_col()
```

Here we can see our formula in action: `barley_2025` is the data, `aes()` puts `variety` on the x-axis and `acres` on the y-axis, and then we use the function `geom_col()` to specify that we want a bar chart -- that is a bar for each row.  

This gives us a graph that looks a lot like the terrible graph that we first made in Excel.  

As we did in Excel, let's filter to only varieties with more than 10,000 acres.  We can do this with the `filter()` function that we learned earlier.  We can also flip this to a horizontal bar chart by putting `variety` on the y-axis (vertical axis) and `acres` on the x-axis (horizontal axis).

```{r}
#| label: fig-r-bar-filtered
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.0
#| fig-cap: "Varieties over 10,000 acres, largest at the top."
#| fig-alt: "A horizontal bar chart of 14 varieties ordered by acres, Synergy AAC longest at about 393,000."
barley_2025 |>
  filter(acres > 10000) |>       # varieties over 10,000 acres
  ggplot(aes(x = acres,
             y = variety)) +
  geom_col()
```

First, let's note what we did in the first two lines.  Recall that the pipe `|>` is like saying "then".  So we are saying to R, take `barley_2025`, then filter it to only include rows where `acres` is greater than 10,000.  Then assign acres to the x-axis and variety to the y-axis.  Then draw the bars with `geom_col()`, which infers that the bars should run along the numeric axis.

Great!  Now we can start adding additional options.  First, we want to order the bars by their value, not alphabetically.  We can do this with the `fct_reorder()` function: `fct_reorder(variety, acres)` orders the varieties by their acres, which in Excel required sorting the worksheet rows.


```{r}
#| label: fig-r-bar-ordered
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.0
#| fig-cap: "The same varieties, ordered by acres."
#| fig-alt: "A horizontal bar chart of 14 varieties ordered by acres, Synergy AAC longest at about 393,000."
barley_2025 |>
  filter(acres > 10000) |>
  ggplot(aes(x = acres,
             y = fct_reorder(variety, acres))) +    # order bars by acres
  geom_col()
```

Finally we can add all of our additional options to the chart -- the colour of the bars, a title and axis labels, comma formatting on the axis, and a theme with a larger font -- with additional functions and with arguments inside some of the functions.

```{r}
#| label: fig-r-bar-final
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.0
#| fig-cap: "The finished bar chart."
#| fig-alt: "The same 14 bars in green with acreage labels, titled Acres of barley varieties in Saskatchewan, 2025."
bar_plot <- barley_2025 |>
  filter(acres > 10000) |>
  ggplot(aes(x = acres,
             y = fct_reorder(variety, acres))) +
  geom_col(fill = "DarkGreen") +                     # dark green bars
  scale_x_continuous(labels = scales::comma) +       # 400,000 not 4e+05
  labs(title = "Acres of barley varieties in Saskatchewan, 2025",
       x = "Acres", y = NULL) +
  theme_classic(base_size = 13)

bar_plot
```

Let's read this code one line at a time:
1. Take the data `barley_2025` *then*
2. Filter to only include rows where `acres` is greater than 10,000 *then*
3. Assign `acres` to the x-axis and `variety` to the y-axis, ordering the bars by `acres` *then*
4. Draw bars with `geom_col()` and colour them dark green *then*
5. Change the scale on the x axis so that labels are in comma format (400,000 instead of 4e+05) *then*
6. Add a title and axis labels *then*
7. Change the theme to `theme_classic()` and increase the base font size to 13.

All of this is a bit overwhelming -- even for me who has been using R for years.  Whereas in Excel you can kind of find the options through a user interface, with R you need to know the functions that you can add on to ggplot and what their arguments are.  To be honest, I rarely remember the exact function names or arguments. I used to have to look them up -- now I get AI to do that for me -- in the next module we will start graphing with AI.  For now, it is only important that you understand the very basic graphing commands:

1. `ggplot()` to start the graph, and `aes()` to assign variables to axes and other visual properties.
2. `geom_col()` to make a bar chart, `geom_line()` to make a line chart, and `geom_point()` to make a scatter plot.  If you don't supply any arguments to these functions, defaults will be used.  But you can change things like the colour and size of the bars, lines and points inside them, as we did with `fill = "DarkGreen"`.
3. `labs()` to add a title and axis labels.
4. The `theme_*()` functions, such as `theme_minimal()` and `theme_classic()`, to control the overall look of the chart, including the font size.


## Pie charts {#sec-r-pie-example}
Truth be told, ggplot is not great at producing pie charts.  In part, this is because data analysts don't much like pie charts, so perhaps ggplot developers do not want to encourage their use.  But we can still make a pie chart with ggplot by adding `+coord_polar(theta = "y")`.  This is like saying take the bar chart that we had created *then* wrapped into a circle along the y axis. 

When we do exactly this we get the same kind of terrible pie chart as we had in Excel.  Note that I removed some of the lines of code from the bar chart code above that don't apply to the pie chart -- the `fct_reorder()` function, because we don't want to order the slices, the colour of the bars, and the scale.

```{r}
#| label: fig-r-pie-first
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.0
#| fig-height: 3.8
#| fig-cap: "The first attempt: a pie with fourteen slices."
#| fig-alt: "A pie chart with fourteen similarly coloured slices and a long legend of variety names."
barley_2025 |>
  filter(acres > 10000) |>
  ggplot(aes(x = "", y = acres, fill = variety)) +
  geom_col() +
  labs(title = "Acres of barley varieties in Saskatchewan, 2025") +
  theme_void(base_size = 13) + 
  coord_polar(theta = "y") 

```

Again, let's break down this code:
1. Take the data `barley_2025` *then*
2. Filter to only include rows where `acres` is greater than 10,000 *then*
3. Assign `acres` to the y-axis and `variety` to the fill colour, and assign an empty string to the x-axis (this is a bit of a hack since the x-axis does not exist on our pie chart) *then*
4. Draw bars with `geom_col()` *then*
5. Add a title with `labs()` *then*
6. Use `theme_void()` to remove the axes and gridlines *then*
7. Wrap the bars into a circle with `coord_polar(theta = "y")`.

To make this graph interpretable, we likely want to group the small varieties into an "Other" category, as we did in Excel.  We can do this through the following code:

```{r}
#| eval: true
#| echo: false
console('
# Group the small varieties into "Other", then total the acres
pie_data <- barley_2025 |>
  mutate(variety = if_else(acres > 100000, variety, "Other")) |>
  group_by(variety) |>
  summarise(acres = sum(acres))

pie_data
')
```

Again, let's break down this code:
1. Set `pie_data` equal to `barley_2025` *then*
2. Use `mutate()` to create a new column `variety` (this overwrites the existing column) that assigns the value `"Other"` to any row where `acres` is less than or equal to 100,000, and keeps the original `variety` value otherwise *then*
3. Group the data by the new `variety` column *then*
4. Summarise the data by calculating the total `acres` for each group

Now we can see that pie_data is made up of four rows with acreage for the four largest varieties, and one row with acreage for all of the other varieties combined.    

When we make a pie chart with the new data we get:
```{r}
#| label: fig-r-pie
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.4
#| fig-height: 3.4
#| fig-cap: "2025 barley acreage: the four largest varieties and everything else."
#| fig-alt: "A pie chart with five slices; Other and Synergy AAC are the two largest at roughly 400,000 acres each."
pie_data |>
  ggplot(aes(x = "", y = acres, fill = variety)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  labs(title = "2025 barley acreage by variety",
       x = NULL, y = NULL, fill = NULL) +
  theme_void(base_size = 13)
```

Now this pie looks a bit nicer than our previous one.  There are ways of adding labels to the slices and including the percentages in each slice, but we will leave that for another time.  

## Line charts {#sec-r-line-example}
Line charts are (fortunately) much easier to make in ggplot than pie charts.  Let's start by making the same line chart we had in Excel that plots Synergy's acreage from 2021 to 2025.  

We need to start by reading in the risk-zone data, which has one row per zone-variety-year.  We again convert its column names to lowercase on the way in.  To keep things simple, we start with a single risk zone:

```{r}
#| label: fig-r-line-synergy
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.6
#| fig-cap: "Acres planted with Synergy barley in risk zone 1."
#| fig-alt: "A line rising unevenly from about 4,300 acres in 2021 to about 9,200 in 2025."
# Risk-zone records for all crops, varieties and years
variety_yields <- read_csv(
  "data/sask_variety_yields.csv",
  show_col_types = FALSE
) |>
  clean_names()          # column names to lowercase

variety_yields |>
  filter(variety == "SYNERGY AAC", risk_zone == 1) |>
  ggplot(aes(x = year, y = acres)) +
  geom_line()
```


Breaking this down:

1. Read the risk-zone file and convert its column names to lowercase *then*
2. Filter to the rows where `variety` is Synergy **and** `risk_zone` is 1 -- two conditions in one `filter()` call keep only the rows that satisfy both *then*
3. Assign `year` to the x-axis and `acres` to the y-axis *then*
4. Draw a line through the points, in year order, with `geom_line()`.

This works, but returns a quite ugly chart.  Once again, we can add the same formatting options as we added to the bar chart to make it more readable.  

```{r}
#| label: fig-r-line-formatted
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.6
#| fig-cap: "The same line with a title, labelled axes and a zero baseline."
#| fig-alt: "The formatted line chart of Synergy acres in risk zone 1, rising from about 4,300 to about 9,200 acres."
variety_yields |>
  filter(variety == "SYNERGY AAC", risk_zone == 1) |>
  ggplot(aes(x = year, y = acres)) +
  geom_line(colour = "DarkGreen", linewidth = 1) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(title = "Acres planted with Synergy AAC barley, risk zone 1",
       x = "Year", y = "Acres") +
  theme_classic(base_size = 13)
```


Now let's add Copeland to the chart.  This is actually quite easy -- we just adjust the filter to allow both varieties, and then map the `variety` variable to colour inside `aes()`.  This will draw one line per variety, and add a legend automatically.

```{r}
#| label: fig-r-line-two
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.6
#| fig-cap: "Synergy and Copeland in risk zone 1."
#| fig-alt: "Two lines crossing: Copeland falls from about 21,000 acres in 2021 to 7,900 in 2025 while Synergy rises from 4,300 to 9,200."
variety_yields |>
  filter(variety %in% c("SYNERGY AAC", "COPELAND CDC"),
         risk_zone == 1) |>
  ggplot(aes(x = year, y = acres, colour = variety)) +   # one line per variety
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(title = "Acres planted with Synergy and Copeland in risk zone 1",
       x = "Year", y = "Acres", colour = NULL) +
  theme_classic(base_size = 13) 
```

Breaking down what changed:

1. `%in%` keeps a row when `variety` matches either name in the list, where `==` could match only one *then*
2. `colour = variety` is now *inside* `aes()`.  So what is in `aes()` now reads -- assign year to the x-axis, acres to the y-axis, and give each variety its own colour (i.e., its own line) *then*
3. `colour = NULL` inside `labs()` removes the legend's title, since the variety names speak for themselves.

Now we want to add all the varieties to our plot and to sum acreage across risk zones.  Our first challenge is to do what the PivotTable effectively did in Excel: sum the acres for each variety and year across risk zones.  

```{r}
#| eval: true
#| echo: false
console('
# Total acres for each variety and year
barley_by_year <- variety_yields |>
  filter(crop == "Barley") |>
  group_by(variety, year) |>
  summarise(acres = sum(acres, na.rm = TRUE), .groups = "drop")

barley_by_year
')
```



This gives us one row per variety per year, but a chart with every variety would still have too many lines.  As we did for the pie chart, we want to group the small varieties into an "Other" category.  This time a variety's size is its total acres across all five years, so we need to calculate that total before we can rename anything.  

```{r}
#| eval: true
#| echo: false
console('
##First get the total acreage of each variety across all risk zones and years
barley_variety_yields <- variety_yields |>
  filter(crop == "Barley") |>
  group_by(variety) |>
  mutate(total_acres = sum(acres, na.rm = TRUE)) |>
  ungroup()

## For varieties with less than 150,000 acres across all years, rename them to "Other"
barley_variety_yields <- barley_variety_yields |>
    mutate(variety = if_else(total_acres<150000, "Other", variety))

# Total acres for each variety and year
barley_by_year <- barley_variety_yields |>
  group_by(variety, year) |>
  summarise(acres = sum(acres, na.rm = TRUE)) |>
  ungroup()
')
```

Reading this code line by line:

1. Take the barley records, group them by `variety`, and use `mutate()` to create a `total_acres` column holding each variety's acres summed across every zone and year.  Where `summarise()` would collapse the groups down to one row each, `mutate()` after a `group_by()` keeps every row and attaches the group total to each of them.  `ungroup()` then removes the grouping so later steps work on the whole table *then*
2. Use `if_else()` to rename every variety whose `total_acres` is under 150,000 -- the same threshold we used in the PivotChart filter -- to `"Other"` *then*
3. Re-total the acres by variety and year, so `"Other"` is a single line rather than fifty separate ones.



Then the plotting step is the same line chart as before, with `variety` mapped to colour:

```{r}
#| label: fig-r-line-many
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.6
#| fig-height: 3.8
#| fig-cap: "Acres of the main barley varieties, 2021 to 2025."
#| fig-alt: "Nine lines showing Copeland declining, Synergy and Austenson roughly steady, and an Other line near 300,000 acres."
barley_by_year |>
  ggplot(aes(x = year, y = acres, colour = variety)) +
  geom_line(linewidth = 0.9) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(title = "Acres of barley varieties in Saskatchewan",
       x = "Year", y = "Total acres", colour = NULL) +
  theme_classic(base_size = 13)
```


## Save the finished chart {#sec-saving-plots}

In our code to date we have just displayed the plots in Positron.  We can also save these plots as an object and then save them to a file.  The first step is to assign the plot to a name.  This means the object is saved in memory and can be displayed again later.  

We can then save that object to a file with `ggsave()`.  The first argument is the file name, and the second is the plot object.  



```{r}
#| eval: true
#| echo: false
console('
## Create plot (as above) and save it as the object variety_line_chart
variety_line_chart <- barley_by_year |>
  ggplot(aes(x = year, y = acres, colour = variety)) +
  geom_line(linewidth = 0.9) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(title = "Acres of barley varieties in Saskatchewan",
       x = "Year", y = "Total acres", colour = NULL) +
  theme_classic(base_size = 13)

## Visualize variety line chart by printing it in the console
variety_line_chart

# Save the finished line chart
ggsave(
  "output/variety_line_chart.png",
  plot = variety_line_chart,
  width = 6.4,
  height = 4.0,
  dpi = 300
)

')
```

I have chosen to save this as a PNG object by specifying the file type as `.png`. PNG works for slides and web pages. PDF preserves vector lines and text for publication. We can also alter the width, height, and quality of the image (in dpi or dots per inch).    

::: {.callout-tip collapse="true"}
## Other resources -- ggplot2

- **R for Data Science (2nd ed.)** -- [Chapter 1, "Data visualization"](https://r4ds.hadley.nz/data-visualize) builds up the ggplot2 grammar with worked examples.
- **ggplot2 documentation** -- the [function reference](https://ggplot2.tidyverse.org/reference/) with examples for every geom, scale and theme.
- **R Graphics Cookbook** -- [recipes](https://r-graphics.org/) organized by the change you want to make to a chart.
:::
