# Data Manipulation {#sec-dplyr-verbs}

```{r}
#| include: false
#| eval: true
source("_console.R")
suppressMessages({library(dplyr); library(readr); library(tidyr)})
trial_data <- readr::read_csv("practice/data/canola_trial.csv", show_col_types = FALSE)
```

With a dataset in R and inspected, we can start doing things to it. Most data analysis
involves repeating a surprisingly small number of operations: keep certain rows, keep
certain columns, create or change a column, sort the rows, calculate summary statistics,
and do those calculations separately for different groups.

The tidyverse package `dplyr` gives us a function for each of these jobs (you will often see
them called the dplyr "verbs"). None of the operations are new -- each is the R form of
something from the Excel modules:

| Question                                | `dplyr` function | Excel counterpart        |
|-----------------------------------------|------------------|--------------------------|
| Which rows do I want?                   | `filter()`       | filtering                |
| Which columns do I want?                | `select()`       | keeping certain columns  |
| What new column do I want?              | `mutate()`       | a formula in a new column|
| In what order do I want the rows?       | `arrange()`      | sorting                  |
| What summary do I want?                 | `summarise()`    | summary formulas         |
| That summary separately for groups?     | `group_by()`     | a PivotTable             |

Once you understand these operations, most everyday data manipulation in R is a matter of
combining them.

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Use `filter()`, `select()`, `mutate()`, and `arrange()` to manipulate a data frame.
2. Chain several operations together using the pipe, `|>`.
3. Use `summarise()` to calculate statistics from a data frame.
4. Use `group_by()` with `summarise()` to calculate statistics separately for groups.
5. Read a longer pipeline from top to bottom and explain what each step does.
6. Save a result to a CSV file.
:::

## The Data {.unnumbered}

We continue with the canola trial data from the last chapter, in the same project folder.
Since this chapter does a new job, it gets a new script: `code/2_data_manipulation.R`.
Start it with a header, then load the tidyverse and read the data in:

```r
# ---
# Title: Manipulating the canola trial data
# Author: Your Name
# Date: 2026-10-06
# Description:
#   Filters, transforms, and summarizes the canola trial data
#   using the dplyr functions.
# ---

# Load packages
library(tidyverse)

# Read in the data
trial_data <- read_csv("data/canola_trial.csv")
```

## Data Manipulation Functions {#sec-verbs-basic}

We will learn four functions in this section.  Two affect the rows:

- `filter()` chooses rows, and
- `arrange()` changes their order.

Two affect the columns:

- `select()` chooses columns, and
- `mutate()` creates or changes them.

Each function takes the data frame as its first argument, followed by what to do with it.

### `filter()`: Keep Certain Rows

`filter()` keeps the rows that satisfy a condition:

```{r}
#| eval: true
#| echo: false
console('
filter(trial_data, yield_bu_acre > 50)
')
```

Read this as: from `trial_data`, keep the rows where `yield_bu_acre` is greater than 50.

Notice that `filter()` displays the high-yielding fields but does not remove the other fields from `trial_data` -- check `nrow(trial_data)` and all 120 rows are still there. Every function in this chapter works the same way: it takes a data frame as input and returns a *new* data frame as output, and if we only display that output, R does not save it. To keep a result, assign it to an object:

```{r}
#| eval: true
#| echo: false
console('
high_yield <- filter(trial_data, yield_bu_acre > 50)
')
```

(An assignment prints nothing -- the result went into `high_yield` instead of the console.)

Now there are two objects: `trial_data`, the original 120 fields, and `high_yield`, the filtered subset. That is a useful feature -- you can experiment without changing the original data. Throughout the chapter, an example without `<-` is just displaying a result.

Text variables work too:

```{r}
#| eval: true
#| echo: false
console('
filter(trial_data, variety == "InVigor")
')
```

Note the **two** equals signs: `==` asks whether two things are equal, so `variety == "InVigor"` asks "is `variety` equal to `"InVigor"`?". A single `=` has a different job in R -- naming arguments, as in `mean(x, na.rm = TRUE)` -- so writing `variety = "InVigor"` inside `filter()` will not do what you want. Also note the quotation marks: `"InVigor"` with quotes is text; `InVigor` without quotes would be read as the name of an object, which probably doesn't exist and would cause an error.

You can combine conditions with `&` (and -- both must be true), `|` (or -- at least one must be true), and `!` (not):

```{r}
#| eval: true
#| echo: false
console('
##Filter for yields above 50 with the variety InVigor
filter(trial_data, yield_bu_acre > 50 & variety == "InVigor")

##Filter for fertilizer below 80 or above 160
filter(trial_data, fertilizer_kg_ha < 80 | fertilizer_kg_ha > 160)

##Filter for yields above 60 with the variety not equal to InVigor
filter(trial_data, yield_bu_acre > 60 & !variety == "InVigor")
')
```

When you want to match several possible values of the same variable, `%in%` is easier to read:

```{r}
#| eval: true
#| echo: false
console('

##The following two filters are identical

filter(trial_data, variety == "InVigor" | variety == "DEKALB")

filter(trial_data, variety %in% c("InVigor", "DEKALB"))
')
```


### `arrange()`: Sort Rows

`arrange()` changes the order of the rows. `arrange(trial_data, yield_bu_acre)` sorts from lowest yield to highest.  If you want to arrange from highest to lowest then use the `desc()` (descending order) function. 

```{r}
#| eval: true
#| echo: false
console('

##Sort data by yield (lowest to highest)
arrange(trial_data, yield_bu_acre)

##Sort data by yield (highest to lowest)
arrange(trial_data, desc(yield_bu_acre))
')
```

You can sort by more than one variable -- `arrange(trial_data, variety, desc(yield_bu_acre))` sorts by variety and then, within each variety, from highest yield to lowest. Like the other functions, `arrange()` does not permanently rearrange `trial_data` unless you save the result.

```{r}
#| eval: true
#| echo: false
console('

## Sort data by variety first, then yield (highest to lowest)
arrange(trial_data, variety, desc(yield_bu_acre))
')
```

### `select()`: Keep Certain Columns

`select()` chooses variables rather than observations:

```{r}
#| eval: true
#| echo: false
console('
select(trial_data, field_id, variety, yield_bu_acre)
')
```

The result has the same rows but only those three columns -- especially useful with datasets containing dozens or hundreds of variables. You can also drop a column by putting `-` in front of its name, or keep a range of adjacent columns:

```{r}
#| eval: true
#| echo: false
console('

## Select  everything except rainfall
select(trial_data, -rainfall_mm)                 

# Select columns from field_id through rainfall_mm
select(trial_data, field_id:rainfall_mm)         
')
```

A relative of `select()` is `rename()`, which changes a column's name without dropping anything: `rename(trial_data, yield = yield_bu_acre)`.

Keep the pair straight: `filter()` works on rows, `select()` on columns.

### `mutate()`: Create or Change Columns

`mutate()` creates a new variable or changes an existing one. Suppose we want yield in
tonnes per hectare rather than bushels per acre:

```{r}
#| eval: true
#| echo: false
console('
mutate(trial_data, yield_t_ha = yield_bu_acre * 0.0560)
')
```

The result contains all the original columns plus a new `yield_t_ha`. Read the expression
as: create `yield_t_ha` and set it equal to `yield_bu_acre` × 0.0560. This is the R version
of adding a formula column in Excel.

As always, the display above is not saved anywhere. To keep the new column, assign the result -- and here it is common to overwrite the original object with the version that has the extra column:

```{r}
#| eval: true
#| echo: false
console('
trial_data <- mutate(trial_data, yield_t_ha = yield_bu_acre * 0.0560)

names(trial_data)
')
```

`trial_data` now carries `yield_t_ha` for the rest of the session.

You can create several variables in one call, and later expressions can use variables
created earlier in the same `mutate()`:

```{r}
#| eval: true
#| echo: false
console('
mutate(trial_data,
       yield_t_ha = yield_bu_acre * 0.0560,
       kg_n_per_bu = fertilizer_kg_ha / yield_bu_acre)
')
```


## The Pipe `|>` {#sec-pipe}

Real questions usually take more than one function. Suppose we want to know: among the
InVigor fields, what are the field IDs and yields, ordered from highest to lowest? We know
every operation we need -- filter to InVigor, select the two columns, arrange by descending
yield. Without a pipe, each step's output becomes the next step's first argument, and the
calls nest:

```{r}
#| eval: true
#| echo: false
console('
arrange(select(filter(trial_data, variety == "InVigor"), field_id, yield_bu_acre),
        desc(yield_bu_acre))
')
```

That works, but you have to start reading in the middle and work outward. The **pipe**,
`|>`, lets us write the analysis in the order we think about it. The pipe takes whatever is
on its *left* and feeds it as the first argument to the function on its *right*, so
`trial_data |> filter(variety == "InVigor")` is exactly
`filter(trial_data, variety == "InVigor")`. Read `|>` as the word "then":

```{r}
#| eval: true
#| echo: false
console('
trial_data |>
  filter(variety == "InVigor") |>
  select(field_id, yield_bu_acre) |>
  arrange(desc(yield_bu_acre))
')
```

Take `trial_data`, *then* keep the InVigor fields, *then* keep two columns, *then* sort.
Once we use the pipe, we no longer repeat the data-frame name in every function, and a
multi-step analysis reads top to bottom. From here on the examples use the pipe. (Older code
uses `%>%` for the same idea; `|>` is the modern built-in version.)

To keep a result, assign the whole pipeline to an object:

```{r}
#| eval: true
#| echo: false
console('
invigor_yields <- trial_data |>
  filter(variety == "InVigor") |>
  select(field_id, yield_bu_acre) |>
  arrange(desc(yield_bu_acre))
')
```

### Debugging a pipeline {.unnumbered}

Pipelines also give you a way to debug. When a pipeline produces a result that looks wrong,
do not stare at the whole thing and guess. Run it one step at a time: run the first step
alone, check the result, add the second step, check again, and keep going. The first point
where the data stops looking the way you expect is where the problem is. This works because
every step does one thing -- and the same habit is how you check a pipeline an AI assistant
wrote for you.

## `summarise()` and `group_by()` {#sec-summarise-group}

The four functions so far rearrange or add information while keeping observations as rows.
Sometimes we want to *collapse* the observations into summary statistics. That is the job of
`summarise()`.

### `summarise()`: Many Rows Become One

```{r}
#| eval: true
#| echo: false
console('
trial_data |>
  summarise(
    mean_yield = mean(yield_bu_acre, na.rm = TRUE),
    sd_yield = sd(yield_bu_acre, na.rm = TRUE),
    n_fields = n()
  )
')
```

The names on the left (`mean_yield`, `sd_yield`, `n_fields`) become column names in the
result; the expressions on the right calculate the values. `n()` counts the rows.

The contrast with `mutate()` matters. `mutate()` keeps the observations and adds
information; `summarise()` collapses the observations into summaries. So
`trial_data |> mutate(mean_yield = mean(yield_bu_acre))` still has 120 rows, with the
overall mean repeated on every row, while the `summarise()` version returns one row.

### `group_by()`: Calculate Separately for Groups

The overall mean may not be what we want. Suppose we want the mean yield for each variety.
`group_by()` doesn't change the data -- it tells subsequent operations to apply *per group*:

```{r}
#| eval: true
#| echo: false
console('
trial_data |>
  group_by(variety) |>
  summarise(
    mean_yield = mean(yield_bu_acre, na.rm = TRUE),
    sd_yield = sd(yield_bu_acre, na.rm = TRUE),
    n_fields = n()
  )
')
```

Take `trial_data`, *then* divide the observations by variety, *then* calculate the
statistics separately for each variety. The result has one row per variety. This is the R
equivalent of a PivotTable.

Grouping is state that can affect later operations. After a grouped `summarise()`, ask for
an ordinary, ungrouped data frame back with `.groups = "drop"` when you plan to keep
manipulating the result -- the capstone example below does this. (There is also an
`ungroup()` function, which appears later in this chapter in a situation where it genuinely
matters.)

The central pattern -- `data |> group_by(group_variable) |> summarise(...)` -- appears
throughout the rest of the book.

## Unit Conversion: A Worked Example {#sec-unit-conversion}

Canadian ag data often comes in mixed units. Canola yields are traditionally reported in bushels per acre (bu/ac) in Canada, but metric tonnes per hectare (t/ha) in global datasets. The conversion for canola is approximately:

$$
\text{yield (t/ha)} = \text{yield (bu/ac)} \times 0.0560
$$

This factor comes from two facts: a bushel of canola weighs about 22.68 kg (canola's standard 50-lb bushel), and an acre is about 0.4047 ha. So one bu/ac is $22.68 \div 0.4047 \div 1000 \approx 0.0560$ t/ha. (The exact factor depends on the crop's bushel weight -- a wheat or corn bushel is 60 or 56 lb and gives a different number -- so in a real analysis you would look up the right bushel weight for your crop and cite your source.)

The transformation is exactly what `mutate()` is for:

```{r}
#| eval: true
#| echo: false
console('
trial_converted <- trial_data |>
  mutate(yield_t_ha = yield_bu_acre * 0.0560)

trial_converted |>
  select(field_id, yield_bu_acre, yield_t_ha)
')
```

The original `yield_bu_acre` stays in the data alongside the new `yield_t_ha`, which makes
the transformation easy to check.

### An extension: different rules for different rows {.unnumbered}

Sometimes a new variable depends on a condition. Suppose we want to class each field's
nitrogen rate as low, medium, or high. `case_when` handles it -- read this one to
understand it rather than to memorize it:

```{r}
#| eval: true
#| echo: false
console('
trial_data |>
  mutate(n_rate = case_when(
    fertilizer_kg_ha < 100 ~ "low",
    fertilizer_kg_ha < 150 ~ "medium",
    TRUE                   ~ "high"
  )) |>
  select(field_id, fertilizer_kg_ha, n_rate)
')
```

Each line reads "if this condition is true, use this value", and the first matching
condition wins: a field at 120 kg/ha fails the first test and passes the second, so it is
"medium". The `TRUE ~ ...` line is the catch-all -- anything that matched neither condition
is "high". `case_when` is the R analogue of nested `IF`s in Excel, and it comes back when we
clean data in @sec-cleaning. The point for now is that `mutate()` can create variables using
rules, not just arithmetic.

## Writing Output {#sec-writing-output}

Results so far have existed only inside R. To save one as a file that can be opened in Excel
or sent to someone else, pass the object and a filename to `write_csv`. With our project
layout, results go in `output/`:

```r
variety_summary <- trial_data |>
  group_by(variety) |>
  summarise(mean_yield = mean(yield_bu_acre, na.rm = TRUE),
            n_fields = n())

write_csv(variety_summary, "output/variety_summary.csv")
```

`read_csv` brings data into R; `write_csv` sends a data frame out to a file. Because the
output is created by the script, you can always recreate it by running the script again.

## Combining It All {#sec-dplyr-combining}

We now have enough tools to answer a realistic question: for each variety and
nitrogen-rate class, what was the mean yield in t/ha, and how many fields were observed?
Show the highest-yielding class first within each variety.

```{r}
#| eval: true
#| echo: false
console('
variety_n_summary <- trial_data |>
  mutate(yield_t_ha = yield_bu_acre * 0.0560,
         n_rate = case_when(
           fertilizer_kg_ha < 100 ~ "low",
           fertilizer_kg_ha < 150 ~ "medium",
           TRUE                   ~ "high")) |>
  group_by(variety, n_rate) |>
  summarise(mean_yield_t_ha = mean(yield_t_ha, na.rm = TRUE),
            n_fields = n(),
            .groups = "drop") |>
  arrange(variety, desc(mean_yield_t_ha))
variety_n_summary
')
```

Do not read it as one giant command; read it one step at a time:

1. **Create** the variables the question needs: metric yield, and the nitrogen-rate class.
2. **Group** the observations into variety and nitrogen-rate combinations.
3. **Summarise** each group down to one row: its mean yield and its number of fields. With
   two grouping variables, `.groups = "drop"` tells `summarise` to hand back an ordinary,
   ungrouped data frame.
4. **Sort** by variety and, within each variety, from highest mean yield to lowest.

Each line does one simple thing, and that is the reason for pipes: a good pipeline makes the
logic of an analysis visible. Saving the table is one more line:
`write_csv(variety_n_summary, "output/variety_n_summary.csv")`.

### A slightly harder example {.unnumbered}

Which fields yielded above the average for their own variety? This needs one new idea:
`mutate()` can operate within groups.

```{r}
#| eval: true
#| echo: false
console('
above_variety_average <- trial_data |>
  group_by(variety) |>
  mutate(variety_mean = mean(yield_bu_acre)) |>
  ungroup() |>
  filter(yield_bu_acre > variety_mean) |>
  select(field_id, variety, yield_bu_acre, variety_mean) |>
  arrange(variety, desc(yield_bu_acre))
above_variety_average
')
```

The important step is `group_by(variety)` followed by `mutate()`. Because the data is
grouped by variety, `mean()` is calculated separately within each variety -- and unlike
`summarise()`, `mutate()` keeps every field and adds its variety's mean as a new column on
each one. This is also where `ungroup()` earns its place: we are finished calculating within
varieties, so we remove the grouping and the later steps treat the result as an ordinary
data frame.

You do not need to produce this second pipeline from memory. You should be able to read it
from top to bottom and explain what each step is doing. As analyses become more complicated
-- and especially once AI helps generate the code -- that is the skill that matters: looking
at a sequence of operations and deciding whether it actually answers the question you
intended to ask.

::: {.callout-tip collapse="true"}
## Other resources — dplyr

- **R for Data Science (2nd ed.)** -- [Chapter 3, "Data transformation"](https://r4ds.hadley.nz/data-transform.html): the core dplyr functions, the pipe, and grouped summaries.
- **dplyr documentation** -- the [dplyr reference site](https://dplyr.tidyverse.org/), including the one-page [cheatsheet](https://rstudio.github.io/cheatsheets/html/data-transformation.html).
- **Video** -- Equitable Equations, [*The five main dplyr verbs*](https://www.youtube.com/watch?v=b2ODCVAsXlw) -- filter, select, mutate, arrange, and summarise in one sitting.
:::
