# Thinking Spatially {#sec-spatial-data}

```{r}
#| include: false
#| eval: true
source("_console.R")
# Chunks in this chapter run with practice/ as the working directory, so the
# student-facing paths (data/...) resolve at render time.
knitr::opts_knit$set(root.dir = "practice")
suppressMessages({library(tidyverse); library(sf)})
```

Almost everything in agriculture happens somewhere: a yield belongs to a field, a rainfall reading to a gauge, an insurance premium to a risk zone. Spatial data is data that keeps the *where* attached to the numbers, so that location itself can be part of the analysis.

Spatial data comes in two shapes.

**Vector** data describes things with boundaries or positions: points (a weather station, an elevator), lines (a road, a river), and polygons (a field, a rural municipality, a risk zone). A vector dataset is a table like any other, except that each row carries a geometry column describing its shape. The RM yields we have used all term are one join away from being spatial: Saskatchewan publishes the RM boundaries as polygons, and the RM number links a row of yields to a shape on the map.

Here is that idea in practice. Reading the RM boundaries with the **sf** package gives a data frame with one extra column:

```{r}
#| eval: true
#| echo: false
console('
# Load the package for spatial data
library(sf)

# Read the RM boundary file
rms <- st_read("data/sask_rms.geojson", quiet = TRUE)

# An sf object is a data frame with a geometry column
head(rms)
')
```

Each row is an RM; the `geometry` column holds its polygon. Drawing the table draws the map:

```{r}
#| label: fig-rm-boundaries
#| eval: true
#| echo: true
#| message: false
#| fig-width: 5.5
#| fig-height: 6.2
#| fig-cap: "Vector data: the 296 rural municipalities of Saskatchewan. Each polygon is one row of the table; the shaded polygon is the row for RM 344, Corman Park."
#| fig-alt: "A map of southern Saskatchewan divided into small rural municipality polygons, with one polygon near the centre shaded green."
# Draw every row of the table as its polygon
ggplot(rms) +
  geom_sf(fill = "grey97", colour = "grey40", linewidth = 0.1) +
  # Shade a single row: RM 344, Corman Park
  geom_sf(data = filter(rms, RMNO == "344"), fill = "#4A7C59") +
  theme_void()
```

**Raster** data divides the land into a grid of equal cells and stores one value per cell, like a photograph where every pixel is a measurement. Satellite imagery is raster data, and so is anything derived from it: a 30-metre grid where each cell holds an estimated crop type, a soil-moisture surface, an elevation model.

Here is a small piece of one: a 50-by-50 window of the 2023 Annual Crop Inventory, covering about two square kilometres of central Saskatchewan. Each cell is 30 metres on a side and holds a single value, the crop class:

```{r}
#| label: fig-raster-field
#| eval: true
#| echo: true
#| message: false
#| fig-width: 6.4
#| fig-height: 4.6
#| fig-cap: "Raster data: 30-metre cells from the 2023 Annual Crop Inventory. The white lines show the cells; fields, a road allowance and a wetland emerge from nothing but a grid of single values."
#| fig-alt: "A square grid of small coloured cells forming recognizable field shapes: blocks of canola and spring wheat, patches of trees and pasture, and a thin line of developed land."
# One row per 30-metre cell: its position and its crop class
field <- read_csv("data/aci_field_raster.csv", show_col_types = FALSE)

# The class table pairs each value with a crop name and a map colour
classes <- read_csv("data/aci_crop_classes.csv", show_col_types = FALSE)

# Attach the names and colours to the cells
field_classed <- field |>
  left_join(classes, by = "value")

# The named palette: each crop paired with its official colour
palette <- field_classed |>
  distinct(crop, colour) |>
  deframe()

# Draw every cell as a tile
ggplot(field_classed, aes(x = x, y = y, fill = crop)) +
  geom_tile(colour = "white", linewidth = 0.15) +
  scale_fill_manual(values = palette) +
  coord_fixed() +
  theme_void() +
  labs(fill = NULL)
```

Zoom in on either figure and the difference between the two shapes is plain: the vector map is exact boundaries, the raster is a grid pretending to be boundaries. Both are tables underneath, which is why the rest of the module can treat them with tools you already have.

## Coordinates and reference systems

A location is usually written as a longitude and a latitude. The two conventions disagree on the order: maps, GPS units and Google write latitude first, while R and most spatial file formats want longitude first, because longitude is the x coordinate and latitude is the y. @fig-coord-swap shows the same two numbers entered in both orders.

```{r}
#| label: fig-coord-swap
#| eval: true
#| echo: true
#| message: false
#| fig-width: 6.4
#| fig-height: 3.4
#| fig-cap: "The coordinates of a field near Saskatoon, entered longitude-first (green) and latitude-first (red). The red point is in the Southern Ocean."
#| fig-alt: "A world map with a green point on Saskatchewan and a red point in the ocean south of Africa."
# A world outline to plot against
world <- map_data("world")

ggplot() +
  # Country outlines in light grey
  geom_polygon(data = world, aes(long, lat, group = group), fill = "grey90") +
  # The field, longitude first
  annotate("point", x = -106.55, y = 52.19, colour = "#4A7C59", size = 3) +
  # The same numbers, latitude first
  annotate("point", x = 52.19, y = -106.55, colour = "firebrick", size = 3) +
  coord_fixed() +
  theme_void()
```

Degrees are also not distances. A degree of latitude is about 111 km everywhere, but a degree of longitude shrinks toward the pole -- at Saskatchewan's latitude it is about 68 km. Subtracting two longitudes therefore does not give a distance, an area computed in degrees means nothing, and a map drawn as if degrees were equal in both directions comes out the wrong shape. @fig-degrees-squash draws the RM centres that way, and the province is stretched wide. `geom_sf()` drew @fig-rm-boundaries correctly because it knew the coordinates were degrees.

```{r}
#| label: fig-degrees-squash
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.2
#| fig-cap: "The centre of each RM, plotted as if a degree of longitude equalled a degree of latitude. Compare the shape with @fig-rm-boundaries."
#| fig-alt: "A scatterplot of RM centre points forming a version of southern Saskatchewan that is noticeably too wide and flat."
# One point per RM: the centre of its polygon
rm_centres <- rms |>
  st_centroid() |>       # centre point of each polygon
  st_coordinates() |>    # the longitude and latitude of each centre
  as_tibble()

# Degrees plotted as if equal in both directions
ggplot(rm_centres, aes(X, Y)) +
  geom_point(size = 0.4, colour = "grey40") +
  coord_fixed() +
  theme_void()
```

A **coordinate reference system** (CRS) is the statement of what a file's coordinates mean, and there are two kinds. A geographic CRS records positions in degrees on the globe; the one GPS uses is called WGS 84, or EPSG:4326. A projected CRS flattens a region onto a plane and records positions in metres; most Saskatchewan agency data uses UTM Zone 13. The EPSG number is how software identifies a CRS, and those two -- 4326 for degrees, UTM Zone 13 for metres -- cover nearly everything in this module.

Every spatial file carries its CRS. Check it rather than guessing: Saskatchewan is wide enough that its western edge falls in UTM Zone 12, so even the usual choice for the province is sometimes wrong. The working rule is to put everything in one CRS before measuring distance or area or combining two layers, and to use a projected CRS when measuring. The functions that check and convert a CRS, `st_crs()` and `st_transform()`, are in @sec-spatial-r.

Canadian files sometimes say NAD 83 rather than WGS 84. The two differ by less than a metre here, which is less than a combine header, and this course ignores the difference.

## Resolution, scale and extent

**Resolution** is the size of the smallest thing the data can tell apart. For a raster it is the cell: the Annual Crop Inventory's cells are 30 metres, so a quarter section is about 27 cells across, and anything smaller than a cell -- a slough's edge, a yard site -- is rounded into whichever class dominates the cell. Vector data has a resolution too, in how finely its boundaries are drawn; the RM file this chapter uses has been simplified to within a few hundred metres, which is fine for a provincial map and too coarse for measuring a boundary line. Data also has a resolution in time: a yield monitor logs a point every second, the crop inventory gives one value per year.

**Extent** is how much ground the data covers: one field for a machinery file, the province for the RM yields, all of Canada for the crop inventory.

The two pull against each other. Fine cells over a large extent make enormous files, so most datasets choose one or the other, and a table's place on that trade-off tells you what questions it can answer.

| Dataset | Resolution | Extent |
|---|---|---|
| Yield monitor | a point every second, a few metres apart | one field |
| Annual Crop Inventory | 30-metre cells, once a year | Canada |
| RM crop yields | one value per RM per year | Saskatchewan |
| StatCan field crops | one value per province per year | Canada |

Matching the data to the question is mostly a matter of direction. Fine data can always be aggregated up to answer a coarse question -- yield-monitor points average up to a field mean, ACI cells count up to an RM's canola area. Going the other way is not possible: an RM yield says nothing about which quarters in the RM did well, and a provincial average hides the differences between RMs.

@fig-resolution shows what aggregation costs. It is the land from @fig-raster-field again, coarsened so that every 10-by-10 block of cells becomes one 300-metre cell holding the block's most common class.

```{r}
#| label: fig-resolution
#| eval: true
#| echo: true
#| message: false
#| fig-width: 5.6
#| fig-height: 4.2
#| fig-cap: "The land from @fig-raster-field in 300-metre cells. The large fields survive the coarsening; the wetland, the road allowance and the smaller bluffs are gone."
#| fig-alt: "A five-by-five grid of large coloured cells showing only the dominant crops of the same area, with the small features no longer visible."
# Coarsen the window: each 10-by-10 block of cells becomes one 300 m cell
coarse <- field_classed |>
  mutate(block_x = ceiling(x / 10),
         block_y = ceiling(y / 10)) |>       # assign each cell to a block
  count(block_x, block_y, crop, colour) |>   # cells of each class per block
  arrange(desc(n)) |>                        # most common class first
  distinct(block_x, block_y, .keep_all = TRUE)  # keep it, one per block

# Draw the coarsened cells
ggplot(coarse, aes(block_x, block_y, fill = crop)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = palette) +
  coord_fixed() +
  theme_void() +
  labs(fill = NULL)
```

## Where agricultural spatial data comes from

- **[AAFC Annual Crop Inventory](https://open.canada.ca/data/en/dataset/ba2645d5-4458-414d-b196-6303ac06c1c9).** A raster classification of every 30-metre cell of Canadian farmland, produced from satellite imagery every year since 2009: this cell was canola, that one spring wheat, that one was too wet to seed. Accuracy for the major crops is roughly 85 to 90 percent.
- **[Saskatchewan RM boundaries](https://geohub.saskatchewan.ca/).** The polygons for every rural municipality, which is what you need to put the RM yield data on a map.
- **[Canadian Drought Monitor](https://agriculture.canada.ca/en/agricultural-production/weather/canadian-drought-monitor).** Monthly drought-condition polygons for the whole country.
- **Soils.** The [Saskatchewan Soil Information System](https://sksis.usask.ca/) maps soil classes and properties across the province.

## The Annual Crop Inventory

The Annual Crop Inventory is the dataset this module works with. It answers a question no survey can: not "how much canola did this RM grow" but "what grew in *this exact field*, each year, going back to 2009."

::: {.callout-tip title="What grew on your land?"}
[This page](practice/field_viewer/index.html) puts the Annual Crop Inventory on a map. Find a field you know -- a home quarter, a field you drive past -- and click it to see the classified crop for every year. Field edges and small features misclassify more often than field centres, which is worth remembering whenever you use classified data.
:::

The viewer works by sending a small request to an AAFC server every time you click, and the server answers with the pixel value at that point. That is an API doing its job. In @sec-rgee-lab you make the same kind of request yourself, from R, through Google Earth Engine -- and get back data you can tidy, join and plot with the tools you already have.
