# Thinking Spatially {#sec-spatial-data}

```{r}
#| include: false
#| eval: true
source("_console.R")
# Chunks in this chapter run with practice/ as the working directory, so the
# student-facing paths (data/...) resolve at render time.
knitr::opts_knit$set(root.dir = "practice")
suppressMessages({library(tidyverse); library(sf)})
```

Almost everything in agriculture happens somewhere: a yield belongs to a field, a rainfall reading to a gauge, an insurance premium to a risk zone. Spatial data is data that keeps the *where* attached to the numbers, so that location itself can be part of the analysis.

Spatial data comes in two shapes.

**Vector** data describes things with boundaries or positions: points (a weather station, an elevator), lines (a road, a river), and polygons (a field, a rural municipality, a risk zone). A vector dataset is a table like any other, except that each row carries a geometry column describing its shape. The RM yields we have used all term are one join away from being spatial: Saskatchewan publishes the RM boundaries as polygons, and the RM number links a row of yields to a shape on the map.

Here is that idea in practice. Reading the RM boundaries with the **sf** package gives a data frame with one extra column:

```{r}
#| eval: true
#| echo: false
console('
# Load the package for spatial data
library(sf)

# Read the RM boundary file
rms <- st_read("data/sask_rms.geojson", quiet = TRUE)

# An sf object is a data frame with a geometry column
head(rms)
')
```

Each row is an RM; the `geometry` column holds its polygon. Drawing the table draws the map:

```{r}
#| label: fig-rm-boundaries
#| eval: true
#| echo: true
#| message: false
#| fig-width: 5.5
#| fig-height: 6.2
#| fig-cap: "Vector data: the 297 rural municipalities of Saskatchewan. Each polygon is one row of the table; the shaded polygon is the row for RM 344, Corman Park."
#| fig-alt: "A map of southern Saskatchewan divided into small rural municipality polygons, with one polygon near the centre shaded green."
# Draw every row of the table as its polygon
ggplot(rms) +
  geom_sf(fill = "grey97", colour = "grey40", linewidth = 0.1) +
  # Shade a single row: RM 344, Corman Park
  geom_sf(data = filter(rms, RMNO == "344"), fill = "#4A7C59") +
  theme_void()
```

**Raster** data divides the land into a grid of equal cells and stores one value per cell, like a photograph where every pixel is a measurement. Satellite imagery is raster data, and so is anything derived from it: a 30-metre grid where each cell holds an estimated crop type, a soil-moisture surface, an elevation model.

Here is a small piece of one: a 50-by-50 window of the 2023 Annual Crop Inventory, covering about two square kilometres of central Saskatchewan. Each cell is 30 metres on a side and holds a single value, the crop class:

```{r}
#| label: fig-raster-field
#| eval: true
#| echo: true
#| message: false
#| fig-width: 6.4
#| fig-height: 4.6
#| fig-cap: "Raster data: 30-metre cells from the 2023 Annual Crop Inventory. The white lines show the cells; fields, a road allowance and a wetland emerge from nothing but a grid of single values."
#| fig-alt: "A square grid of small coloured cells forming recognizable field shapes: blocks of canola and spring wheat, patches of trees and pasture, and a thin line of developed land."
# One row per 30-metre cell: its position and its crop class
field <- read_csv("data/aci_field_raster.csv", show_col_types = FALSE)

# The class table pairs each value with a crop name and a map colour
classes <- read_csv("data/aci_crop_classes.csv", show_col_types = FALSE)

# Attach the names and colours to the cells
field_classed <- field |>
  left_join(classes, by = "value")

# The named palette: each crop paired with its official colour
palette <- field_classed |>
  distinct(crop, colour) |>
  deframe()

# Draw every cell as a tile
ggplot(field_classed, aes(x = x, y = y, fill = crop)) +
  geom_tile(colour = "white", linewidth = 0.15) +
  scale_fill_manual(values = palette) +
  coord_fixed() +
  theme_void() +
  labs(fill = NULL)
```

Zoom in on either figure and the difference between the two shapes is plain: the vector map is exact boundaries, the raster is a grid pretending to be boundaries. Both are tables underneath, which is why the rest of the module can treat them with tools you already have.

## Coordinates and reference systems

::: {.callout-note}
This section is a draft outline.
:::

- Latitude and longitude
- Coordinate reference systems
- Geographic versus projected coordinates

## Resolution, scale and extent

::: {.callout-note}
This section is a draft outline.
:::

- Spatial resolution, scale, and extent
- Agricultural examples at field, RM, provincial, and national scales

## Where agricultural spatial data comes from

- **[AAFC Annual Crop Inventory](https://open.canada.ca/data/en/dataset/ba2645d5-4458-414d-b196-6303ac06c1c9).** A raster classification of every 30-metre cell of Canadian farmland, produced from satellite imagery every year since 2009: this cell was canola, that one spring wheat, that one was too wet to seed. Accuracy for the major crops is roughly 85 to 90 percent.
- **[Saskatchewan RM boundaries](https://geohub.saskatchewan.ca/).** The polygons for every rural municipality, which is what you need to put the RM yield data on a map.
- **[Canadian Drought Monitor](https://agriculture.canada.ca/en/agricultural-production/weather/canadian-drought-monitor).** Monthly drought-condition polygons for the whole country.
- **Soils.** The [Saskatchewan Soil Information System](https://sksis.usask.ca/) maps soil classes and properties across the province.

## The Annual Crop Inventory

The Annual Crop Inventory is the dataset this module works with. It answers a question no survey can: not "how much canola did this RM grow" but "what grew in *this exact field*, each year, going back to 2009."

::: {.callout-tip title="What grew on your land?"}
[This page](practice/field_viewer/index.html) puts the Annual Crop Inventory on a map. Find a field you know -- a home quarter, a field you drive past -- and click it to see the classified crop for every year. Field edges and small features misclassify more often than field centres, which is worth remembering whenever you use classified data.
:::

The viewer works by sending a small request to an AAFC server every time you click, and the server answers with the pixel value at that point. That is an API doing its job. In @sec-rgee-lab you make the same kind of request yourself, from R, through Google Earth Engine -- and get back data you can tidy, join and plot with the tools you already have.
