# Graphing in R {#sec-graphing-r}

```{r}
#| include: false
#| eval: true
source("_console.R")
# Chunks in this chapter run with practice/ as the working directory, so the
# student-facing paths (data/..., output/...) resolve at render time.
knitr::opts_knit$set(root.dir = "practice")
```

The Excel chapter built a bar chart of barley acreage, a pie chart, and line charts of acreage over time. This chapter makes the same charts with `ggplot2`. Excel starts with a selected range and a chart menu. In R, we describe the chart in code: the data, the variables assigned to visual properties, and the marks to draw. The `ggplot2` package turns that description into a figure.

The data are the same two files the Excel workbooks used. [`sask_barley_2025.csv`](practice/data/sask_barley_2025.csv) has one row per barley variety with its 2025 acres and average yield. [`sask_variety_yields.csv`](practice/data/sask_variety_yields.csv) has Saskatchewan Crop Insurance Corporation records by risk zone, crop, variety and year, 2021 to 2025 [@scic_smp]; suppressed observations have missing acres and yield.

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Build a plot with `ggplot()`, `aes()` and a geom.
2. Reproduce the Excel bar, pie and line charts in code.
3. Filter and summarise the data that feeds a chart, instead of editing it on a sheet.
4. Order bars with `fct_reorder()` and group small categories into an "Other" category.
5. Change a plot's title, labels, colours and theme.
6. Plot several series by mapping a variable to colour.
7. Save a finished plot with `ggsave()`.
:::

## Introduction to ggplot2 {#sec-ggplot-intro}

`ggplot2` loads with the tidyverse, and every chart it draws is built from the same three parts:

- the **data**: a data frame with the data to be plotted;
- the **mapping**: `aes()`, which assigns variables to visual properties -- horizontal position, vertical position, colour;
- a **geom**: the mark to draw -- `geom_col()` for bars, `geom_line()` for lines, `geom_point()` for points.

The template is always:

```r
data |>
  ggplot(aes(x = ..., y = ...)) +
  geom_something() + ...
```

We can then add other options to the graph with `+`.  For example, we can change the formatting, font size, or axis labels.  

If we want an analogy to Excel, we can think of `data` as the spreadsheet containing our data, `aes()` as choosing which columns to plot on the x- and y-axes, and the `geom` as choosing the chart type. Other formatting options are then added using the + operator.

## Bar charts {#sec-r-bar-example}

Let's again start with a simple example using the barley variety data for 2025.  When we read the file in, we also convert the column names to lowercase with `clean_names()` (@sec-clean-names), so they are easier to type.  We can then plot the acres of all the varieties with the following code:

```{r}
#| label: fig-r-bar-first
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.4
#| fig-cap: "The first attempt: every barley variety."
#| fig-alt: "A vertical bar chart with 32 overlapping variety names on the horizontal axis."
# Load the tidyverse and janitor
library(tidyverse)
library(janitor)

# One row per barley variety: 2025 acres and average yield
barley_2025 <- read_csv(
  "data/sask_barley_2025.csv",
  show_col_types = FALSE
) |>
  clean_names()          # column names to lowercase

## Show the first few rows of data
head(barley_2025)

## Create a bar chart
barley_2025 |>
  ggplot(aes(x = variety, y = acres)) +
  geom_col()
```

Here we can see our formula in action: `barley_2025` is the data, `aes()` puts `variety` on the x-axis and `acres` on the y-axis, and then we use the function `geom_col()` to specify that we want a bar chart -- that is a bar for each row.  

This gives us a graph that looks a lot like the terrible graph that we first made in Excel.  

As we did in Excel, let's filter to only varieties with more than 10,000 acres.  We can do this with the `filter()` function that we learned earlier.  We can also flip this to a horizontal bar chart by putting `variety` on the y-axis (vertical axis) and `acres` on the x-axis (horizontal axis).

```{r}
#| label: fig-r-bar-filtered
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.0
#| fig-cap: "Varieties over 10,000 acres, largest at the top."
#| fig-alt: "A horizontal bar chart of 14 varieties ordered by acres, Synergy AAC longest at about 393,000."
barley_2025 |>
  filter(acres > 10000) |>       # varieties over 10,000 acres
  ggplot(aes(x = acres,
             y = variety)) +
  geom_col()
```

First, let's note what we did in the first two lines.  Recall that the pipe `|>` is like saying "then".  So we are saying to R, take `barley_2025`, then filter it to only include rows where `acres` is greater than 10,000.  Then assign acres to the x-axis and variety to the y-axis.  Then draw the bars with `geom_col()`, which infers that the bars should run along the numeric axis.

Great!  Now we can start adding additional options.  First, we want to order the bars by their value, not alphabetically.  We can do this with the `fct_reorder()` function: `fct_reorder(variety, acres)` orders the varieties by their acres, which in Excel required sorting the worksheet rows.


```{r}
#| label: fig-r-bar-ordered
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.0
#| fig-cap: "The same varieties, ordered by acres."
#| fig-alt: "A horizontal bar chart of 14 varieties ordered by acres, Synergy AAC longest at about 393,000."
barley_2025 |>
  filter(acres > 10000) |>
  ggplot(aes(x = acres,
             y = fct_reorder(variety, acres))) +    # order bars by acres
  geom_col()
```

Finally we can add all of our additional options to the chart -- the colour of the bars, a title and axis labels, comma formatting on the axis, and a theme with a larger font -- with additional functions and with arguments inside some of the functions.

```{r}
#| label: fig-r-bar-final
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 4.0
#| fig-cap: "The finished bar chart."
#| fig-alt: "The same 14 bars in green with acreage labels, titled Acres of barley varieties in Saskatchewan, 2025."
bar_plot <- barley_2025 |>
  filter(acres > 10000) |>
  ggplot(aes(x = acres,
             y = fct_reorder(variety, acres))) +
  geom_col(fill = "DarkGreen") +                     # dark green bars
  scale_x_continuous(labels = scales::comma) +       # 400,000 not 4e+05
  labs(title = "Acres of barley varieties in Saskatchewan, 2025",
       x = "Acres", y = NULL) +
  theme_classic(base_size = 13)

bar_plot
```

Let's read this code one line at a time:
1. Take the data `barley_2025` *then*
2. Filter to only include rows where `acres` is greater than 10,000 *then*
3. Assign `acres` to the x-axis and `variety` to the y-axis, ordering the bars by `acres` *then*
4. Draw bars with `geom_col()` and colour them dark green *then*
5. Change the scale on the x axis so that labels are in comma format (400,000 instead of 4e+05) *then*
6. Add a title and axis labels *then*
7. Change the theme to `theme_classic()` and increase the base font size to 13.

All of this is a bit overwhelming -- even for me who has been using R for years.  Whereas in Excel you can kind of find the options through a user interface, with R you need to know the functions that you can add on to ggplot and what their arguments are.  To be honest, I rarely remember the exact function names or arguments. I used to have to look them up -- now I get AI to do that for me -- in the next module we will start graphing with AI.  For now, it is only important that you understand the very basic graphing commands:

1. `ggplot()` to start the graph, and `aes()` to assign variables to axes and other visual properties.
2. `geom_col()` to make a bar chart, `geom_line()` to make a line chart, and `geom_point()` to make a scatter plot.  If you don't supply any arguments to these functions, defaults will be used.  But you can change things like the colour and size of the bars, lines and points inside them, as we did with `fill = "DarkGreen"`.
3. `labs()` to add a title and axis labels.
4. The `theme_*()` functions, such as `theme_minimal()` and `theme_classic()`, to control the overall look of the chart, including the font size.


## Pie charts {#sec-r-pie-example}

In Excel we built the pie from a five-row block: the four largest varieties plus an `Other` row computed with `SUM`. We can build the same five-row table with functions we already know. First the data step: `if_else()` renames every variety under 100,000 acres to `"Other"`, and `group_by()` with `summarise()` adds the small varieties up -- the same job the `SUM` did in Excel.

```{r}
#| eval: true
#| echo: false
console('
# Group the small varieties into "Other", then total the acres
pie_data <- barley_2025 |>
  mutate(variety = if_else(acres > 100000, variety, "Other")) |>
  group_by(variety) |>
  summarise(acres = sum(acres))

pie_data
')
```

Five rows, exactly like the block the Excel pie was built from. Now the plotting step:

```{r}
#| label: fig-r-pie
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 5.4
#| fig-height: 3.4
#| fig-cap: "2025 barley acreage: the four largest varieties and everything else."
#| fig-alt: "A pie chart with five slices; Other and Synergy AAC are the two largest at roughly 400,000 acres each."
pie_data |>
  ggplot(aes(x = "", y = acres, fill = variety)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +      # wrap the bar into a circle
  labs(title = "2025 barley acreage by variety",
       x = NULL, y = NULL, fill = NULL) +
  theme_void(base_size = 13)
```

There is no `geom_pie()`. A pie in ggplot2 is a single stacked bar wrapped into a circle with `coord_polar()`, which is a fair description of what a pie chart is. The `Other` slice comes to 412,760 acres, slightly more than Synergy itself -- the same total the Excel `SUM` produced, and a useful reminder that the 28 grouped varieties jointly matter as much as the largest one.

## Line charts {#sec-r-line-example}

The line charts use the risk-zone file, which has one row per zone-variety-year; we again convert its column names to lowercase on the way in. A provincial line needs one value per year, so the pipe filters to the variety and sums acres within each year before plotting:

```{r}
#| label: fig-r-line-synergy
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.6
#| fig-cap: "Acres planted with Synergy barley."
#| fig-alt: "A line from 408,000 acres in 2021 dipping to 317,000 in 2022 and recovering to about 393,000 by 2025."
# Risk-zone records for all crops, varieties and years
variety_yields <- read_csv(
  "data/sask_variety_yields.csv",
  show_col_types = FALSE
) |>
  clean_names()          # column names to lowercase

variety_yields |>
  filter(crop == "Barley",
         variety == "SYNERGY AAC") |>
  group_by(year) |>
  summarise(acres = sum(acres, na.rm = TRUE)) |>   # provincial total per year
  ggplot(aes(x = year, y = acres)) +
  geom_line(colour = "#3F6F52", linewidth = 1) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(title = "Acres planted with Synergy AAC barley",
       x = "Year", y = "Total acres") +
  theme_minimal(base_size = 13)
```

In Excel the first attempt drew `year` as a second data series, and fixing it meant a trip through **Select Data**. That mistake cannot happen here, because `aes(x = year, y = acres)` states which variable is which axis.

To plot two varieties, Excel needed a column per variety and a new series added by hand. In R the data stays long: keep both varieties in the filter and map `variety` to colour, and one line is drawn per variety with a legend:

```{r}
#| label: fig-r-line-two
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.4
#| fig-height: 3.6
#| fig-cap: "Synergy has held its acreage; Copeland has lost most of its own."
#| fig-alt: "Two lines: Copeland falls from 475,000 acres in 2021 to 126,000 in 2025 while Synergy stays near 400,000."
variety_yields |>
  filter(crop == "Barley",
         variety %in% c("SYNERGY AAC", "COPELAND CDC")) |>
  group_by(variety, year) |>
  summarise(acres = sum(acres, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = year, y = acres, colour = variety)) +   # one line per variety
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(title = "Copeland's acreage has fallen by three-quarters; Synergy's has held",
       x = "Year", y = "Total acres", colour = NULL) +
  theme_minimal(base_size = 13)
```

The Excel chapter's answer to "what about all the varieties?" was a PivotChart, with the small varieties filtered out or grouped. We can do the same grouping here, in a data step and a plotting step. The data step builds the variety-year totals, works out which varieties have more than 150,000 acres over the five years -- the same threshold we used in the PivotChart filter -- and renames the rest to `"Other"`:

```{r}
#| eval: true
#| echo: false
console('
# Total acres for each variety and year
barley_by_year <- variety_yields |>
  filter(crop == "Barley") |>
  group_by(variety, year) |>
  summarise(acres = sum(acres, na.rm = TRUE), .groups = "drop")

# Which varieties have more than 150,000 acres over the five years?
big_varieties <- barley_by_year |>
  group_by(variety) |>
  summarise(total_acres = sum(acres)) |>
  filter(total_acres > 150000)

big_varieties

# Group everything else into "Other" and re-total
main_varieties <- barley_by_year |>
  mutate(variety = if_else(variety %in% big_varieties$variety,
                           variety, "Other")) |>
  group_by(variety, year) |>
  summarise(acres = sum(acres), .groups = "drop")
')
```

Then the plotting step is the same line chart as before, with `variety` mapped to colour:

```{r}
#| label: fig-r-line-many
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-width: 6.6
#| fig-height: 3.8
#| fig-cap: "Acres of the main barley varieties, 2021 to 2025."
#| fig-alt: "Nine lines showing Copeland declining, Synergy and Austenson roughly steady, and an Other line near 300,000 acres."
main_varieties |>
  ggplot(aes(x = year, y = acres, colour = variety)) +
  geom_line(linewidth = 0.9) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(title = "Acres of barley varieties in Saskatchewan",
       x = "Year", y = "Total acres", colour = NULL) +
  theme_minimal(base_size = 13)
```

Nine lines is about the limit for one chart, and with lines this distinct the legend works; @sec-graphing-principles covers direct labels and other ways to keep a many-line chart readable.

## Save the finished chart {#sec-saving-plots}

Assigning a plot to a name makes the object passed to `ggsave()` explicit. Width and height are in inches unless another unit is supplied.

```{r}
#| eval: true
#| echo: false
console('
# Create the output folder if needed
dir.create("output", showWarnings = FALSE)

# Save the finished bar chart
ggsave(
  "output/barley_variety_bar.png",
  plot = bar_plot,
  width = 6.4,
  height = 4.0,
  dpi = 300
)

# Confirm that the file was written
file.info("output/barley_variety_bar.png")["size"]
')
```

PNG works for slides and web pages. PDF preserves vector lines and text for publication. Stating the plot object, dimensions and resolution in code makes the export reproducible.

::: {.callout-tip collapse="true"}
## Other resources -- ggplot2

- **R for Data Science (2nd ed.)** -- [Chapter 1, "Data visualization"](https://r4ds.hadley.nz/data-visualize) builds up the ggplot2 grammar with worked examples.
- **ggplot2 documentation** -- the [function reference](https://ggplot2.tidyverse.org/reference/) with examples for every geom, scale and theme.
- **R Graphics Cookbook** -- [recipes](https://r-graphics.org/) organized by the change you want to make to a chart.
:::
