# The `dplyr` Verbs {#sec-dplyr-verbs}

```{r}
#| include: false
#| eval: true
source("_console.R")
suppressMessages({library(dplyr); library(readr); library(tidyr)})
yields <- readr::read_csv("practice/data/field_yields.csv", show_col_types = FALSE)
```

The tidyverse package `dplyr` provides a small set of **verbs** for data transformation
that are powerful enough to express most of what you will ever need.  Most data manipulation
is a *sequence of simple operations*, and with the right small vocabulary the complex
transformations become readable.

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Use the core `dplyr` verbs (`filter`, `select`, `mutate`, `arrange`) to manipulate a data frame.
2. Chain operations together with the pipe (`|>`).
3. Summarise data overall and by group with `summarise` and `group_by`.
4. Use `case_when` for multi-condition transformations, such as unit conversion.
5. Combine several verbs into a single readable pipeline that answers a real question.
:::

## The Data {.unnumbered}

The examples use the same [`field_yields.csv`](practice/data/field_yields.csv) dataset as the
last chapter -- 60 canola fields with columns `field_id`, `region`, `variety`, `acres`,
`yield_bu_acre`, `year`, `seeding_date`, and `units`. Start your script with:

```r
library(tidyverse)

yields <- read_csv("data/field_yields.csv")
```

**One crucial thing to understand before we start.** Every `dplyr` verb *returns a new
data frame* -- it does **not** change the original. So this:

```{r}
#| eval: true
#| echo: false
console('
filter(yields, yield_bu_acre > 50)
')
```

computes the filtered result and *prints it*, but `yields` itself is unchanged -- the next
time you use `yields`, all 60 rows are still there. If you want to **keep** a result, you
must assign it to an object:

```r
big_fields <- filter(yields, yield_bu_acre > 50)   # save the result
```

That is deliberate: your original data stays intact while you experiment.  You can see it --
the filter above returned 14 rows, and `yields` still has all 60:

```{r}
#| eval: true
#| echo: false
console('
nrow(yields)
')
```

Throughout this chapter, when an example just shows an operation without `<-`, assume we are
looking at the result on screen; add `name <- ` in front when you want to save it.

## The Four Row-and-Column Verbs {#sec-verbs-basic}

Each verb is a function whose first argument is the data frame, followed by what to do
with it.

### `filter()`: Keep Certain Rows {.unnumbered}

`filter()` keeps rows where a condition is true.

```r
filter(yields, yield_bu_acre > 50)
```

This keeps only rows where yield exceeds 50. You can combine conditions with `&` (and), `|` (or), and `!` (not):

```r
filter(yields, yield_bu_acre > 50 & region == "South")
filter(yields, region == "South" | region == "Central")
filter(yields, !is.na(yield_bu_acre))  # drop rows with missing yield
```

A shortcut for "in this set of values":

```r
filter(yields, region %in% c("South", "Central"))
```

Important distinction: use **`==`** (two equals signs) to *test* whether two things are equal. A single `=` means something different in R -- it is used for assigning values and naming arguments, not for testing equality -- so writing `region = "South"` inside `filter()` will not do what you want. Always use `==` for comparisons. And note `"South"` -- with quotes -- is a **string** (a piece of text); `South` -- without quotes -- would be read as the name of an object, which probably doesn't exist and would cause an error.

### `select()`: Keep Certain Columns {.unnumbered}

`select()` keeps (or drops) columns.

```{r}
#| eval: true
#| echo: false
console('
select(yields, field_id, region, yield_bu_acre)
')
```

Useful when you have a wide dataset and want to focus on a few columns.

### `mutate()`: Add or Modify Columns {.unnumbered}

`mutate()` creates new columns or modifies existing ones.

```{r}
#| eval: true
#| echo: false
console('
mutate(yields, yield_t_ha = yield_bu_acre * 0.0560)
')
```

The new column is computed from the existing ones. You can create multiple columns in one call, and later columns can use earlier ones:

```r
mutate(yields,
       yield_t_ha = yield_bu_acre * 0.0560,
       above_average = yield_bu_acre > mean(yield_bu_acre, na.rm = TRUE),
       total_bu = yield_bu_acre * acres)
```

### `arrange()`: Sort Rows {.unnumbered}

```{r}
#| eval: true
#| echo: false
console('
arrange(yields, desc(yield_bu_acre))
')
```

`arrange(yields, yield_bu_acre)` sorts the other way, and you can give it more than one
column -- `arrange(yields, region, desc(yield_bu_acre))` sorts by region, then by yield
within each region.

## The Pipe `|>` {#sec-pipe}

Real questions usually take more than one verb. Suppose you want the field ID and yield of
the South-region fields, sorted from highest to lowest. Written the way we have so far, each
step's output becomes the next step's first argument, and the calls nest:

```r
arrange(select(filter(yields, region == "South"), field_id, yield_bu_acre),
        desc(yield_bu_acre))
```

That works, but you have to read it inside-out. The **pipe**, `|>`, fixes it. The pipe takes
whatever is on its *left* and feeds it as the first argument to the function on its *right*.
So `yields |> filter(region == "South")` is exactly `filter(yields, region == "South")`.
Read `|>` as the word **"then."** The nested call above becomes a top-to-bottom sequence:

```{r}
#| eval: true
#| echo: false
console('
yields |>
  filter(region == "South") |>
  select(field_id, yield_bu_acre) |>
  arrange(desc(yield_bu_acre))
')
```

Take `yields`, *then* keep the South fields, *then* keep two columns, *then* sort. From here
on the examples use the pipe. (Older code uses `%>%` for the same idea; `|>` is the modern
built-in version.)

## `summarise()` and `group_by()` {#sec-summarise-group}

### `summarise()`: Collapse to Summary {.unnumbered}

`summarise()` computes summary statistics, reducing many rows to one.

```{r}
#| eval: true
#| echo: false
console('
yields |>
  summarise(
    mean_yield = mean(yield_bu_acre, na.rm = TRUE),
    sd_yield = sd(yield_bu_acre, na.rm = TRUE),
    n = n()
  )
')
```

The result is a data frame with one row containing the requested summaries. (`n()` counts
the rows.)

### `group_by()`: Do the Above by Group {.unnumbered}

This is where things get powerful. `group_by()` doesn't change the data -- but it tells subsequent operations to apply *per group*.

```{r}
#| eval: true
#| echo: false
console('
yields |>
  group_by(region) |>
  summarise(
    mean_yield = mean(yield_bu_acre, na.rm = TRUE),
    sd_yield = sd(yield_bu_acre, na.rm = TRUE),
    n = n()
  )
')
```

The result is a data frame with one row per region, containing the summary statistics for that region. This is the R equivalent of a PivotTable. You can group by multiple columns:

```r
yields |>
  group_by(region, variety) |>
  summarise(mean_yield = mean(yield_bu_acre, na.rm = TRUE))
```

After you are done summarising, it is good practice to `ungroup()` -- otherwise the grouping persists and can surprise you later:

```r
yields |>
  group_by(region) |>
  summarise(mean_yield = mean(yield_bu_acre, na.rm = TRUE)) |>
  ungroup()
```

## Unit Conversion: A Worked Example {#sec-unit-conversion}

Canadian ag data often comes in mixed units. Canola yields are traditionally reported in bushels per acre (bu/ac) in Canada, but metric tonnes per hectare (t/ha) in global datasets. The conversion for canola is approximately:

$$
\text{yield (t/ha)} = \text{yield (bu/ac)} \times 0.0560
$$

This factor comes from two facts: a bushel of canola weighs about 22.68 kg (canola's standard 50-lb bushel), and an acre is about 0.4047 ha. So one bu/ac is $22.68 \div 0.4047 \div 1000 \approx 0.0560$ t/ha. (The exact factor depends on the crop's bushel weight -- a wheat or corn bushel is 60 or 56 lb and gives a different number -- so in a real analysis you would look up the right bushel weight for your crop and cite your source.)

```r
yields_converted <- yields |>
  mutate(yield_t_ha = yield_bu_acre * 0.0560)
```

Now suppose your dataset is really messy -- some rows are in bu/ac and some are already in t/ha, and there is a `units` column telling you which. You can use `case_when`:

```r
yields_converted <- yields |>
  mutate(
    yield_t_ha = case_when(
      units == "bu/ac" ~ yield_bu_acre * 0.0560,   # convert the bu/ac rows
      units == "t/ha"  ~ yield_bu_acre,             # already metric: leave as-is
      TRUE             ~ NA_real_                    # anything else becomes missing
    )
  )
```

(Here we assume the value lives in a `yield_bu_acre` column and a separate `units` column records the unit of each row. `NA_real_` is just R's way of writing a missing *numeric* value -- the numeric version of `NA`.)

`case_when` is the R analogue of nested `IF`s in Excel. Each line is a condition and a value. The first matching condition wins. The `TRUE ~ ...` line is the catch-all "else." Use `case_when` whenever you have multi-way conditions; it is much cleaner than nested `if_else`.

## Writing Output {#sec-writing-output}

To save a result, pass the object and a filename to `write_csv`. With the project layout
from @sec-project-folder, results go in `output/`:

```r
region_summary <- yields |>
  group_by(region) |>
  summarise(mean_yield = mean(yield_bu_acre, na.rm = TRUE))

write_csv(region_summary, "output/region_summary.csv")
```

## Combining It All {#sec-dplyr-combining}

Let's do something realistic. Suppose we want to answer: "For each region, what was the mean yield (in t/ha) of fields larger than 100 acres, in 2025, for the top three varieties by total production?"

```{r}
#| eval: true
#| echo: false
console('
yields |>
  filter(year == 2025, acres > 100) |>
  mutate(yield_t_ha = yield_bu_acre * 0.0560,
         total_tonnes = yield_t_ha * acres * 0.4047) |>
  group_by(variety) |>
  mutate(variety_total = sum(total_tonnes, na.rm = TRUE)) |>
  ungroup() |>
  filter(dense_rank(desc(variety_total)) <= 3) |>
  group_by(region, variety) |>
  summarise(mean_yield_t_ha = mean(yield_t_ha, na.rm = TRUE),
            n_fields = n(),
            .groups = "drop") |>
  arrange(region, desc(mean_yield_t_ha))
')
```

Read it top to bottom and notice how each step does one thing:

1. **Filter** to 2025 and fields over 100 acres.
2. **Compute** metric yield and total production for each field.
3. For each variety, compute its total production across all its fields. Here we use `group_by(variety)` followed by `mutate` (not `summarise`) on purpose: `summarise` would collapse each variety down to a single row, but `mutate` keeps *all* the rows and just adds the group total as a new column on each one. (After a grouped operation, call `ungroup()` to remove the grouping so later steps behave normally.)
4. **Keep the top three varieties** by total production. `desc(variety_total)` sorts largest-first; `dense_rank(...)` assigns rank 1 to the biggest, 2 to the next, and so on; and `<= 3` keeps ranks 1, 2, and 3.
5. **Regroup** by region and variety and `summarise` down to one row per group, computing the mean yield and the number of fields. The `.groups = "drop"` argument just tells `summarise` to return an ordinary, ungrouped data frame afterward (without it, R prints a chatty message about how it left the grouping).
6. **Sort** the final table.

Do not worry about writing something this long yet -- the point is that each line is one small, readable step. Writing it as a pipeline makes the logic visible. Compare with how you would do this in Excel -- probably three PivotTables, a manual filter, and a lot of copy-pasting between sheets.

::: {.callout-tip collapse="true"}
## Other resources — dplyr

- **R for Data Science (2nd ed.)** -- [Chapter 3, "Data transformation"](https://r4ds.hadley.nz/data-transform.html): the verbs, the pipe, and grouped summaries.
- **dplyr documentation** -- the [dplyr reference site](https://dplyr.tidyverse.org/), including the one-page [cheatsheet](https://rstudio.github.io/cheatsheets/html/data-transformation.html).
- **Video** -- Equitable Equations, [*The five main dplyr verbs*](https://www.youtube.com/watch?v=b2ODCVAsXlw) -- filter, select, mutate, arrange, and summarise in one sitting.
:::

## Test Bank Sample {#sec-module2c-test}

1. (Verbs.) What does each of `filter`, `select`, `mutate`, and `arrange` do? Which work on rows and which on columns?
2. (Assignment.) `filter(yields, region == "South")` runs without error but `yields` still has all its rows afterward. Explain why.
3. (Comparison.) What is the difference between `=` and `==` inside `filter()`?
4. (Pipe.) Rewrite `arrange(filter(yields, acres > 100), desc(yield_bu_acre))` using the pipe.
5. (dplyr.) Write a pipeline that filters `yields` to fields in the South region, computes yield in t/ha, and sorts descending by yield.
6. (group_by.) Write a pipeline that computes the mean and standard deviation of yield for each variety, for fields larger than 50 acres.
7. (case_when.) A dataset has a `yield` column and a `units` column. Write code to create a `yield_t_ha` column that converts as needed.
8. (mutate vs summarise.) After `group_by(variety)`, what is the difference between `mutate(total = sum(acres))` and `summarise(total = sum(acres))`?
9. (Output.) Write the line that saves a data frame called `region_summary` to your project's `output/` folder.

## Practice Exercises {#sec-module2c-practice}

Use the [`field_yields.csv`](practice/data/field_yields.csv) dataset for these.

1. Reproduce a table of summary statistics (mean, median, and standard deviation of yield) using `summarise`.
2. Filter to one region and one year, and find the variety with the highest mean yield.
3. Create a new column that flags whether each field is "above average" in yield for its region.
4. Add a `yield_t_ha` column converting `yield_bu_acre` to tonnes per hectare (× 0.0560), then sort the fields from highest to lowest metric yield.
5. Rewrite your Module 1 worked example (canola yields) as an R script using dplyr verbs. Compare the result with what you got in Excel.
6. Break your script intentionally (misspell a column name) and read the error message. Can you fix it?
