# GitHub Copilot {#sec-copilot}

```{r}
#| include: false
#| eval: true
source("_console.R")
```

The AI codign assistant we will use for this course is GitHub Copilot.  This assistant works well within the Positron environment.  It can function at all three levels of AI assistance -- as a chat bot, an inline editor, and an agent.  Most importantly, it is free for students and faculty.  

Note that GitHub Copilot and Microsoft 365 Copilot are separate products. GitHub Copilot supplies coding help in Positron and is the product used here. Copilot in Excel requires separate Microsoft licensing; GitHub Education does not unlock it.

## Learning Objectives {.unnumbered}

::: {.objectives}
By the end of this chapter you should be able to:

1. Connect a GitHub Copilot account to Positron.
2. Use inline completions and Posit Assistant chat for different tasks.
3. Give the assistant enough information about a dataset to avoid guessing.
4. Decide whether to accept, revise or reject suggested code.
5. Check AI-assisted code before it becomes part of an analysis.
:::

## Getting Access {#sec-copilot-signup}

A GitHub Pro plan is free for post-secondary students and faculty.  This provides you with the equivalent of \$15 of monthly credits, which should be more than enough for our class.  To apply for the student benefit:

1. Create a personal account at [github.com](https://github.com) if you do not have one.
2. Open your [GitHub Education benefits](https://github.com/settings/education/benefits) and start a student application.
3. Add and verify your `@usask.ca` address if GitHub asks for an academic email.
4. Supply current proof of enrolment, such as a student card with an enrolment date, a class schedule, a transcript or an enrolment letter.
5. After verification, return to the benefits page and activate the Copilot benefit. Education approval and Copilot activation are separate steps.

Apply early as verification can take several days.  

## Connecting Copilot to Positron {#sec-copilot-positron}

Recent versions of Positron include AI support directly. You do not need to install the GitHub Copilot extension from the Extensions pane.

1. Update Positron before beginning.
2. Open **View > Command Palette**, or use `Cmd+Shift+P` on macOS or `Ctrl+Shift+P` on Windows.
3. Run **Authentication: Configure Language Model Providers**.
4. Select **GitHub Copilot** and sign in through the browser window that opens.
5. Return to Positron and open Posit Assistant chat. Copilot should appear as an available provider.

Copilot's grey inline completions become available after the same sign-in. If chat or completions do not appear, restart Positron. If the account was connected in an older Positron release, sign out of GitHub from the Accounts menu and sign in again.

The command **AI: Create Diagnostic Report** lists the active providers and AI settings without exposing the content of your prompts. It is the quickest check when Positron appears connected but produces no suggestions.

These command names were checked in August 2026. If a later Positron release uses different wording, search the Command Palette for “language model” or “AI” and consult the configuration page linked below.

## Two Ways to Use It {#sec-copilot-modes}

### Inline completions {.unnumbered}

Inline completions appear as grey text while you type. Press `Tab` to accept a suggestion, `Esc` to dismiss it or keep typing to replace it.

A comment can provide a short instruction:

```r
# Read the yield data and count rows by crop
```

A plausible completion might invent both the file and column name:

```r
yields <- read_csv("yields.csv")
count(yields, crop)
```

Neither name was supplied by the comment. Grey text is a suggestion, not evidence that the assistant found the correct file.

Inline completion is useful for predictable syntax, repeated arguments and the next line of a pattern already present in the script. It is a poor place to ask for an explanation because the result arrives as code rather than a discussion.

### Posit Assistant chat {.unnumbered}

Chat is useful for questions: why an error occurred, what a line does, how to reorganize a plot or what checks belong after a join. Select the relevant code or file before asking so the assistant receives the right context.

Posit Assistant is the default chat interface in Positron. GitHub Copilot can supply its language model after you connect the provider. This distinction matters when following documentation: Copilot is the account and model provider; Posit Assistant is the chat interface you see in Positron.

## Show the Shape of the Data {#sec-copilot-context}

Copilot cannot infer a file's columns from its filename. If it has not seen `sask_variety_yields.csv`, code using an assumed column name is a guess.

Run `glimpse()` and inspect the categories rather than typing them from memory:

```{r}
#| eval: true
#| echo: false
console('
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
})

d <- read_csv(
  "practice/data/sask_variety_yields.csv",
  show_col_types = FALSE
) |>
  clean_names()

glimpse(d)
d |> distinct(crop)
')
```

Paste that output, then make the requested missing-value decision explicit:

```r
# d has columns: risk_zone, crop, variety, year, acres, yield
# The complete crop list from distinct(crop) is pasted above.
# Drop rows missing acres or yield and report how many were dropped.
#
# Compute mean 2025 yield by crop, weighted by acres
```

This supplies the exact names, category spellings, missing-value decision and intended calculation. A reasonable suggestion is:

```{r}
#| eval: true
#| echo: false
console('
d_2025 <- d |>
  filter(
    year == 2025,
    !is.na(yield),
    !is.na(acres),
    acres > 0
  )

d_2025 |>
  summarise(
    mean_yield = weighted.mean(yield, acres),
    rows = n(),
    .by = crop
  )
')
```

Context also has a privacy cost. A file included in an AI request leaves your machine. Public course data is suitable; confidential farm, employee, client or research data may not be. Follow the data owner's rules and the restrictions in @sec-ai-privacy.

## Accepting and Rejecting {#sec-copilot-accept}

Accept code when you can explain what it does and can name the check that would reveal an error.

- **Usually safe to accept after reading:** syntax you recognize, plotting boilerplate and repetitive argument lists.
- **Read and test carefully:** joins, filters, grouped summaries, missing-value handling and any code that changes the number of rows.
- **Stop and investigate:** unfamiliar functions or packages, an unexpectedly long solution or code whose result you cannot predict.
- **For Excel, stop at `#NAME?`:** the function may require a newer Microsoft 365 build. Ask for a `SUMIFS` or `INDEX`/`MATCH` alternative when compatibility matters.

Rejecting a suggestion costs almost nothing. Accepting code you do not understand leaves you responsible for a calculation you cannot defend or repair.

## Checking the Work {#sec-copilot-verify}

Use the same checks after every AI-assisted change:

1. Read every line and identify which data it uses.
2. Count rows before and after a filter, join or grouped operation.
3. Test the code on a small case whose answer is known.
4. Recalculate one result another way.
5. Inspect the final table or chart against the question that was asked.

The assistant can help write a check. It cannot replace the independent evidence supplied by that check.

For the weighted-mean example, count the rows and independently recalculate one crop from its numerator and denominator:

```{r}
#| eval: true
#| echo: false
console('
nrow(d)
nrow(d_2025)

d_2025 |>
  filter(crop == "Barley") |>
  summarise(
    checked_mean = sum(acres * yield) / sum(acres)
  )
')
```

In Excel, press F2 or use **Formulas > Evaluate Formula** to see which cells a formula reads, compare the result with a PivotTable, copy the formula down to test relative and absolute references, compare `COUNT` with `COUNTA` to find numbers stored as text, and require exact matching in every lookup.

::: {.callout-tip collapse="true"}
## Other resources -- Copilot and Positron

- **GitHub Docs**, [Access Copilot as a student](https://docs.github.com/en/copilot/how-tos/copilot-on-github/set-up-copilot/enable-copilot/set-up-for-students) -- current eligibility and activation instructions.
- **GitHub Docs**, [Apply to GitHub Education](https://docs.github.com/en/education/about-github-education/github-education-for-students/apply-to-github-education-as-a-student) -- accepted documents and the application process.
- **Positron**, [Configure AI features](https://positron.posit.co/ai-configuration.html) -- providers, completions, chat and diagnostic settings.
:::
